### 2017-11-08 outline.Rmd
outline

---
title: Outline of plyranges a grammar of genomic data manipulation
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
bibliography: references.bib
link-citations: true

---

Overview of structure, references and layout for the plyranges package paper.

# Abstract

The Bioconductor project has created many powerful abstractions for
reasoning about genomics data, such as the _Ranges_ data structures 
for representing genomic intervals. By recognising that these data structures
follow 'tidy' data principles we have created a grammar of genomic data 
manipulation that defines verbs for performing actions on and between genomic 
interval data. This grammar simplifies performing common genomic data analysis 
tasks via method chaining, type consistency and results in creating human 
readable pipelines. We have implemented this grammar as an Bioconductor/R 
package  plyranges.

# Introduction

Putting the work into context.

## What do we mean by genomic data?

  - How does that relate to 'tidy' data?
  - Why tidy data is a useful abstraction.
  - Why _Ranges_ follow this pattern.

Key references: @Wickham2014-jc, @Lawrence2013-wg

## How can we analyse genomic data?

  - the Ranges data structures
  - other abstractions for analysing genomic data
    - the query language approaches
    - the bedtools approach
    - relational algebras

Key references: @Quinlan2010-gc, @Dale2011-js, @Kaitoua2017-pw, @Ovaska2013-gd, @Mungall2014-dg @Kozanitis2016-bm

## What does plyranges contribute?

`plyranges` API provides

- a consistent grammar for analysing genomic data.
- fluent interface enables readable pipelines 
- reproducibility via code clarity, declaraitve programming,  and return consistency

Key references:

# Design and Implementation

Discussion of the API and its key features

- every action on a Range is a verb (functional composition, method chaining)
- fluent (human readable, code describes what to do rather than 'how')
- every function returns a class familiar to the user
- an expressive algebra for performing arithmetic (anchoring functions)
- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)
- recasting overlapping/nearest etc as joins

# Results

A case study showing plyranges makes life easier...
Perhaps could use long reads data from Matt and his post-doc Charity (interesting
use case of looking for intron-exon junctions)

# Conclusion (future directions)

- availability and maintenance of package
- extending to SummarisedExperiment class
- natural fit with ggbio2

# References

### 2017-11-22 paper.Rmd
plos format added

--- 2017-11-08 outline.Rmd
@@
 ---
-title: Outline of plyranges a grammar of genomic data manipulation
-output:
-  bookdown::pdf_document2:
-    toc: false
-    number_sections: false
+title: "plyranges: a grammar of graphics for genomics data" 
+author:
+  - name: Stuart Lee
+    email: stuart.lee1@monash.edu
+    affiliation: Monash University
+  - name: Michael Lawrence
+    email: michael.lawrence@gene.com
+    affiliation: Genentech
+  - name: Di Cook
+    email: di.cook@monash.edu
+    affiliation: Monash University
+address:
+  - code: Monash University
+    address: Department of Econometrics and Business Statistics, Clayton, Victoria, Australia
+  - code: Genentech
+    address: Bioinformatics and Computational Biology, Genentech, Inc., South San Francisco, California, United States of America 
+abstract: |
+    The Bioconductor project has created many powerful abstractions for reasoning about genomics data, such as the _Ranges_ data structures for representing genomic intervals.
+By recognising that these data structures follow 'tidy' data principles we have created a grammar of genomic data manipulation that defines verbs for performing actions on and between genomic interval data.
+This grammar simplifies performing common genomic data analysis tasks via method chaining, type consistency and results in creating human readable pipelines.
+We have implemented this grammar as an Bioconductor/R package called plyranges.
+
 bibliography: references.bib
-link-citations: true
-
+output: rticles::plos_article
+csl: plos.csl
 ---
 
-Overview of structure, references and layout for the plyranges package paper.
-
-# Abstract
-
-The Bioconductor project has created many powerful abstractions for
-reasoning about genomics data, such as the _Ranges_ data structures 
-for representing genomic intervals. By recognising that these data structures
-follow 'tidy' data principles we have created a grammar of genomic data 
-manipulation that defines verbs for performing actions on and between genomic 
-interval data. This grammar simplifies performing common genomic data analysis 
-tasks via method chaining, type consistency and results in creating human 
-readable pipelines. We have implemented this grammar as an Bioconductor/R 
-package  plyranges.
-
 # Introduction
-
-Putting the work into context.
-
 ## What do we mean by genomic data?
@@
 
-Key references:
-
-# Design and Implementation
-
-Discussion of the API and its key features
-
-- every action on a Range is a verb (functional composition, method chaining)
-- fluent (human readable, code describes what to do rather than 'how')
-- every function returns a class familiar to the user
-- an expressive algebra for performing arithmetic (anchoring functions)
-- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)
-- recasting overlapping/nearest etc as joins
-
-# Results
-
-A case study showing plyranges makes life easier...
-Perhaps could use long reads data from Matt and his post-doc Charity (interesting
-use case of looking for intron-exon junctions)
-
-# Conclusion (future directions)
-
-- availability and maintenance of package
-- extending to SummarisedExperiment class
-- natural fit with ggbio2
-
-# References
+# References {#references .unnumbered}

### 2017-11-27 paper.Rmd
first pass at introduction

@@
 # Introduction
-## What do we mean by genomic data?
 
-  - How does that relate to 'tidy' data?
-  - Why tidy data is a useful abstraction.
-  - Why _Ranges_ follow this pattern.
+Genomic data may be naturally represented as sets of pairs of integers 
+corresponding to the start and end points of sequences. Further information
+such as strandedness and chromosome name may be added to these sets to 
+provide biological context. Because of the flexibility of this representation 
+supplemental information such as measurements obtained from an experimental 
+assay or annotations from a genome database can be joined to their relevant
+sequences.
+In the Bioconductor/R packages `IRanges` and `GenomicRanges` these representations have been implemented as a suite of data structures called _Ranges_ (@Lawrence2013-wg).
+These data structures cover many common data types encountered in bioinformatics analyses - a gene can be represented with its coordinates, along with its identifier and the identifiers of its exons; or
+an RNA-seq experiment may be represented as sets of genes with a matching count column.
 
-Key references: @Wickham2014-jc, @Lawrence2013-wg
+The Bioconductor infrastructure for computing with genomic ranges are highly 
+effecient and powerful, however the application programming interface (API) for performing analysis tasks with _Ranges_ is complex due to its large number of
+methods and classes. It also makes resulting scripts written 
+difficult for a non-programmer to read and reason about. An alternative approach 
+would be to implement a domain specific language (DSL) as a fluent interface 
+built on top _Ranges_. The goal of fluent interface is to enable users to write 
+human-readable code via method chaining and consistent function returns. Fluent 
+interfaces fit naturally in the context of Bioinformatics workflows because they 
+enable writing succinct pipelines.
 
-## How can we analyse genomic data?
+Several other DSLs have been implemented to reason about genomics data. Broadly, 
+these are either implemented as query languages or as command line tools 
+embedded in the unix environment ^[other ideas to mention GROK @Ovaska2013-gd]. An example of the former is the Genome Query Language (GQL) and its distributed 
+implementation GenAp which use an SQL-like syntax for fast retrieval of 
+information from genomic databases and BAM files (@Kozanitis2014-va, @Kozanitis2016-bm). 
+Another  example is the Genometric Query Language (GMQL) which implements a relational algebra for combining big genomic datasets (@Kaitoua2017-pw).  The command line application 
+BEDtools develops an extensive algebra for performing arithmetic between two or 
+more sets of genomic regions  (@Quinlan2010-gc). It also has a python interface 
+which simplifies construcing scripts for performing analyses based on BEDTools  (@Dale2011-js). ^[probably should also mention something about their cons, and more detaiil]
 
-  - the Ranges data structures
-  - other abstractions for analysing genomic data
-    - the query language approaches
-    - the bedtools approach
-    - relational algebras
+The abstraction provided by the _Ranges_ data structures aligns with the concept
+of tidy data developed by @Wickham2014-jc. The tidy data pattern is useful because it allows us to see how the data relates to the design of an experiment
+and the variables measured.
+Consequently, it makes both the modelling and manipulation of data more systematic.
+The _Ranges_ data structure follows this abstraction: it is a rectanglular table corresponding to a single biological context.
+Each row contains a single observation and each column a variable about that observation.
 
-Key references: @Quinlan2010-gc, @Dale2011-js, @Kaitoua2017-pw, @Ovaska2013-gd, @Mungall2014-dg @Kozanitis2016-bm
-
-## What does plyranges contribute?
-
-`plyranges` API provides
-
-- a consistent grammar for analysing genomic data.
-- fluent interface enables readable pipelines 
-- reproducibility via code clarity, declaraitve programming,  and return consistency
+The tidy data abstraction has motivated the development of `plyranges`
+a grammar of genomic data manipulation based on the _Ranges_ data structures. It
+implements and extends the grammar defined by the R package `dplyr` (@Wickham2017-dplyr).
+The grammar provides a consistent way of interacting with and analysing genomic data via methods for constructing, grouping, mutating, filtering, and summarising _Ranges_ and an algebra for reasoning about actions on _Ranges_ and relationships between _Ranges_.
 

### 2017-11-27 paper.Rmd
adding in outline for second paragraph

@@
 
+# Design and Implementation
+
+Discussion of the API and its key features
+
+- every action on a Range is a verb (functional composition, method chaining)
+- fluent (human readable, code describes what to do rather than 'how')
+- every function returns a class familiar to the user
+- an expressive algebra for performing arithmetic (anchoring functions)
+- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)
+- recasting overlapping/nearest etc as joins
+
 # References {#references .unnumbered}

### 2017-11-27 paper.Rmd
typo in title; track pdf

@@
 ---
-title: "plyranges: a grammar of graphics for genomics data" 
+title: "plyranges: a grammar for manipulating genomics data" 
 author:
@@
 sequences.
-In the Bioconductor/R packages `IRanges` and `GenomicRanges` these representations have been implemented as a suite of data structures called _Ranges_ (@Lawrence2013-wg).
+In the Bioconductor/R packages `IRanges` and `GenomicRanges` these representations have been implemented as a suite of data structures called _Ranges_ @Lawrence2013-wg.
 These data structures cover many common data types encountered in bioinformatics analyses - a gene can be represented with its coordinates, along with its identifier and the identifiers of its exons; or
@@
 implementation GenAp which use an SQL-like syntax for fast retrieval of 
-information from genomic databases and BAM files (@Kozanitis2014-va, @Kozanitis2016-bm). 
-Another  example is the Genometric Query Language (GMQL) which implements a relational algebra for combining big genomic datasets (@Kaitoua2017-pw).  The command line application 
+information from genomic databases and BAM files @Kozanitis2014-va; @Kozanitis2016-bm. 
+Another  example is the Genometric Query Language (GMQL) which implements a relational algebra for combining big genomic datasets @Kaitoua2017-pw.  The command line application 
 BEDtools develops an extensive algebra for performing arithmetic between two or 
-more sets of genomic regions  (@Quinlan2010-gc). It also has a python interface 
-which simplifies construcing scripts for performing analyses based on BEDTools  (@Dale2011-js). ^[probably should also mention something about their cons, and more detaiil]
+more sets of genomic regions  @Quinlan2010-gc. It also has a python interface 
+which simplifies construcing scripts for performing analyses based on BEDTools  @Dale2011-js. ^[probably should also mention something about their cons, and more detaiil]
 
 The abstraction provided by the _Ranges_ data structures aligns with the concept
-of tidy data developed by @Wickham2014-jc. The tidy data pattern is useful because it allows us to see how the data relates to the design of an experiment
+of tidy data @Wickham2014-jc. The tidy data pattern is useful because it allows us to see how the data relates to the design of an experiment
 and the variables measured.
@@
 a grammar of genomic data manipulation based on the _Ranges_ data structures. It
-implements and extends the grammar defined by the R package `dplyr` (@Wickham2017-dplyr).
+implements and extends the grammar defined by the R package `dplyr` @Wickham2017-dplyr.
 The grammar provides a consistent way of interacting with and analysing genomic data via methods for constructing, grouping, mutating, filtering, and summarising _Ranges_ and an algebra for reasoning about actions on _Ranges_ and relationships between _Ranges_.

### 2018-02-11 paper.Rmd
initial draft of design and implemenation, needs more examples

@@
 
-Discussion of the API and its key features
+The _plyranges_ API implements a domain specific language using the existing
+_IRanges_ and _GenomicRanges_ packages in Bioconductor as a backend. Consequently,
+our API still has the speed and effeciency of the aforementioned packages but
+with a more coherent interface. The API also
+extends the grammar elements in _dplyr_ for performing genomic specific
+manipulations such as finding overlapping regions or nearest neighbour regions
+between many _Ranges_. The _plyranges_ API is specifically designed to enable
+fast interactive analysis of _Ranges_ objects but can also be used for scripting 
+genomic data workflows.
 
-- every action on a Range is a verb (functional composition, method chaining)
-- fluent (human readable, code describes what to do rather than 'how')
-- every function returns a class familiar to the user
-- an expressive algebra for performing arithmetic (anchoring functions)
-- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)
+We have desgined the API to be fluent. Every function call corresponds to an
+action on a _Ranges_ object (they are named verbs) and where possible
+functions have few arguments. Each verb is constructed
+to enable a tab completion based workflow. Both of these aspects reduce the 
+cognitive load on a new user  since most manipulations can be performed with a 
+vocabulary of several verbs, rather than having to memorise functions with many 
+arguments that are nouns (as is required in the existing Bioconductor packages). 
+This is also has the advantage of allowing users to write human readable code
+because verbs describe what the code is doing rather than how its doing it.
+
+Workflows can be composed by chaining verbs together via the pipe operator,`%>%` 
+(exported from the R package _magrittr_). This is possible because every 
+function call is endomorphic: when the input is _Ranges_ object the output
+will also be a _Ranges_ object. One advantage of this static typing is that it 
+does not  require any additional learning of classes beyond _Ranges_ and 
+the _DataFrame_ classes. This is similar to the bedtools API, where the
+output is usually a BED file. However, it strongly deviates from the design of 
+the _Ranges_ Bioconductor packages, where many methods return a new class
+upon return. The Bioconductor design enables effecient computing as users
+are exposed to low-level features of its API which _plyranges_ trys to abstract
+away. Method chaining via the pipe operator can also be difficult to debug,
+as there multiple points of failure.
+
+## Working with Ranges
+
+The _plyranges_ API exports the six core verbs from the _dplyr_ package
+and modifys them for use with simple _Ranges_ and _Genomic Ranges_. The
+verb `mutate()` takes a Ranges object and a set of name-value pairs
+and generates a new Ranges object that with modified or new metadata
+columns or modified core components (start, end, width, seqnames, strand). 
+The `filter()` function takes a Ranges object and logical expressions and
+restricts Ranges object to where the logical expression evaluates to true.
+The `summarise()` function takes a Ranges object and a set of name-value pairs
+and aggregates the Ranges according to functions evaluated in the name-value pairs.
+As `summarise()` is an aggregation it may break the structure of the of a Ranges
+object, hence it returns a DataFrame object. The `select()` function determines
+which metadata columns are returned and the order they are returned in. The 
+`arrange()` function sorts a Ranges object by named variables.
+The `group_by()` function creates an implicit grouping of Ranges object according
+to variables in the Ranges object. This modifies the actions of `mutate()`,
+`summarise()` and `filter()` (and also the set operations, `reduce_ranges()`
+and `disjoin_ranges()`), so they are performed on each partition created by the
+grouping.
+
+The API has an expressive algebra for performing arithmetic on Ranges via
+the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
+modifies the width of a Ranges object, while `stretch` extends the start and
+end of a Ranges object. These can be chained with the anchoring functions
+`anchor_start()`, `anchor_end()`, `anchor_center()`, `anchor_3p()` or 
+`anchor_5p()`, which fix the coordinates of a Ranges object in place. 
+Moreover, the `shift_` and `flank_` family of functions can be used to shift
+all coordinates in a Ranges object or generate flanking regions from a Ranges 
+object to the left, right, upstream or downstream of the input. Unlike, the
+Bioconductor API, _plyranges_ makes it expclit via function calls whether to
+take into account the strand information of a _Ranges_ object.
+
 - recasting overlapping/nearest etc as joins
 
+- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)
+
+- note on the NSE
+
+# Results
+
+# Availablilty and Future Work
+
+The _plyranges_ package is avaiable on the Bioconductor project website 
+[https://bioconductor.org](https://bioconductor.org) or can be accessed
+via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
+We aim to continue developing the `plyranges` package and extend it for use
+with more complex data structures outside of genomics such as the 
+_SummarizedExperiment_ class. As the `plyranges` interface encourages tidy data 
+practices it also integrates well with the grammar of graphics, we aim to use
+it for the visualisation of genomic ranges.
+
 # References {#references .unnumbered}

### 2018-02-12 paper.Rmd
added section on overlaps

@@
 columns or modified core components (start, end, width, seqnames, strand).
+The use of `mutate()` means that a user no longer needs knowledge of the 
+accessors of the Ranges object, as they can modify them in place.
 The `filter()` function takes a Ranges object and logical expressions and
@@
 
+A common operation to perform between two _Ranges_ objects is to find overlaps
+or nearest neighbours. The _plyranges_ API recasts these operations as 'joins'
+or 'pairing' operations. For overlaps, there are three join operations: 
+`join_overlap_intersect()`, `join_overlap_inner()` and `join_overlap_left()`
+which are shown in figure (). These operations consider any overlap between
+two input ranges and return any corresponding metadata from both Ranges objects
+as metadata. The intersect join takes the intersect of the start and end coordinates
+of overlapping intervals of the query and subject Ranges 
+(for GenomicRanges it also accounts for sequence name), when there is a overlap
+the metadata corresponding to the query and subject Ranges are returned. 
+Similarly, inner join takes the start and end coordinates of the query Ranges
+that overlap the subject Ranges and returns metadata of the overlapping query
+and subject Ranges. Finally, the left join performs a left outer join between
+the query and subject Ranges, it returns all genomic intervals from the query
+ranges, and returns missing values in metadata columns when there is no overlap.
+A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
+`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
+be modified with the `within` suffix (which changes the type of overlap 
+from 'any' to 'within') or the `directed` suffix (which takes into account
+the strand of a GenomicRanges object.).
+
+For nearest neighbours, the `plyranges` API provides `join_nearest()`, 
+`join_precede()`, and `join_follow()` functions. These functions are similar
+to the overlapping functions, in that they return the query ranges that are
+nearest (or precede or follow) the subject ranges and add metadata from
+the subject ranges when the query is a nearest neighbour of the subject. Like
+the overlap joins, these functions can modified with suffixes to find nearest
+neighbours that are left, right, upstream or downstream of the subject.
+
+and returns
+
 - recasting overlapping/nearest etc as joins

### 2018-02-12 paper.Rmd
higher level headings

@@
 
+### Core verbs
+
 The _plyranges_ API exports the six core verbs from the _dplyr_ package
@@
 
+### Arithimetic
+
 The API has an expressive algebra for performing arithmetic on Ranges via
@@
 take into account the strand information of a _Ranges_ object.
+
+### Overlapping Ranges
 
@@
 
-and returns
+The pairing operations, `pair_overlap()`, `pair_nearest()`, `pair_follow()`, 
+and `pair_precede()` are similar to the join operation but instead of returning
+a Ranges, they pair up the subject and query Ranges objects into a DataFrame,
+alongside their metadata columns. This data structure is similar to the _Pairs_
+data structure in the _S4Vectors_ package or the BED-PE file format. 
+<!--- should also create convience methods for grouping by pairs and pairing up, 
+unpairing, would also be useful for alignments too--->
 
-- recasting overlapping/nearest etc as joins
-
-- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)
-
-- note on the NSE
+### Non-standard evaluation
 

### 2018-02-12 paper.Rmd
outline of design and implementation

@@
 
-### Core verbs
+### Construction and Import/Output
+
+The _plyranges_ API provides the methods `as_granges()` and `as_iranges()`
+for constructing Ranges from tabular data structures, such as data.frames in
+base R. These methods use non-standard evaluation so columns in a data.frame
+can be modified before a Ranges object is created.
+
+The package also provides a consistent framework for reading and writing 
+files from and to common genomic data formats, using the `rtracklayer` package
+as a backend. The methods are implemented in the `read_/write_` family of 
+functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
+narrowPeaks, GFF/GTF, WIG and BigWig files.
+
+### Manipulation of Ranges
 
@@
 
-### Arithimetic
+### Arithimetic on Ranges
 
@@
 We aim to continue developing the `plyranges` package and extend it for use
-with more complex data structures outside of genomics such as the 
-_SummarizedExperiment_ class. As the `plyranges` interface encourages tidy data 
-practices it also integrates well with the grammar of graphics, we aim to use
-it for the visualisation of genomic ranges.
+with more complex data structures  such as the 
+_SummarizedExperiment_ class, which can be used for analysing transcriptomic
+and variant data. As the `plyranges` interface encourages tidy data 
+practices it integrates well with the principles of the grammar of graphics, 
+we aim to use it for the visualisation of genomic ranges.
 

### 2018-02-12 paper.tex
outline of design and implementation

% Template for PLoS
% Version 3.4 January 2017
\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
% \bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below
\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

\usepackage{forarray}
\usepackage{xstring}
\newcommand{\getIndex}[2]{
  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
}

\setcounter{secnumdepth}{0}

\newcommand{\getAff}[1]{
  \getIndex{#1}{Monash University,Genentech}
}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{plyranges: a grammar for manipulating genomics data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
Stuart Lee\textsuperscript{\getAff{Monash University}},
Michael Lawrence\textsuperscript{\getAff{Genentech}},
Di Cook\textsuperscript{\getAff{Monash University}}\\
\bigskip
\textbf{\getAff{Monash University}}Department of Econometrics and Business Statistics, Clayton, Victoria,
Australia\\
\textbf{\getAff{Genentech}}Bioinformatics and Computational Biology, Genentech, Inc., South San
Francisco, California, United States of America\\
\bigskip
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The Bioconductor project has created many powerful abstractions for
reasoning about genomics data, such as the \emph{Ranges} data structures
for representing genomic intervals. By recognising that these data
structures follow `tidy' data principles we have created a grammar of
genomic data manipulation that defines verbs for performing actions on
and between genomic interval data. This grammar simplifies performing
common genomic data analysis tasks via method chaining, type consistency
and results in creating human readable pipelines. We have implemented
this grammar as an Bioconductor/R package called plyranges.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Genomic data may be naturally represented as sets of pairs of integers
corresponding to the start and end points of sequences. Further
information such as strandedness and chromosome name may be added to
these sets to provide biological context. Because of the flexibility of
this representation supplemental information such as measurements
obtained from an experimental assay or annotations from a genome
database can be joined to their relevant sequences. In the
Bioconductor/R packages \texttt{IRanges} and \texttt{GenomicRanges}
these representations have been implemented as a suite of data
structures called \emph{Ranges} {[}1{]}. These data structures cover
many common data types encountered in bioinformatics analyses - a gene
can be represented with its coordinates, along with its identifier and
the identifiers of its exons; or an RNA-seq experiment may be
represented as sets of genes with a matching count column.

The Bioconductor infrastructure for computing with genomic ranges are
highly effecient and powerful, however the application programming
interface (API) for performing analysis tasks with \emph{Ranges} is
complex due to its large number of methods and classes. It also makes
resulting scripts written difficult for a non-programmer to read and
reason about. An alternative approach would be to implement a domain
specific language (DSL) as a fluent interface built on top
\emph{Ranges}. The goal of fluent interface is to enable users to write
human-readable code via method chaining and consistent function returns.
Fluent interfaces fit naturally in the context of Bioinformatics
workflows because they enable writing succinct pipelines.

Several other DSLs have been implemented to reason about genomics data.
Broadly, these are either implemented as query languages or as command
line tools embedded in the unix environment.\footnote{other ideas to
  mention GROK {[}2{]}} An example of the former is the Genome Query
Language (GQL) and its distributed implementation GenAp which use an
SQL-like syntax for fast retrieval of information from genomic databases
and BAM files {[}3{]}; {[}4{]}. Another example is the Genometric Query
Language (GMQL) which implements a relational algebra for combining big
genomic datasets {[}5{]}. The command line application BEDtools develops
an extensive algebra for performing arithmetic between two or more sets
of genomic regions {[}6{]}. It also has a python interface which
simplifies construcing scripts for performing analyses based on BEDTools
{[}7{]}.\footnote{probably should also mention something about their
  cons, and more detaiil}

The abstraction provided by the \emph{Ranges} data structures aligns
with the concept of tidy data {[}8{]}. The tidy data pattern is useful
because it allows us to see how the data relates to the design of an
experiment and the variables measured. Consequently, it makes both the
modelling and manipulation of data more systematic. The \emph{Ranges}
data structure follows this abstraction: it is a rectanglular table
corresponding to a single biological context. Each row contains a single
observation and each column a variable about that observation.

The tidy data abstraction has motivated the development of
\texttt{plyranges} a grammar of genomic data manipulation based on the
\emph{Ranges} data structures. It implements and extends the grammar
defined by the R package \texttt{dplyr} {[}9{]}. The grammar provides a
consistent way of interacting with and analysing genomic data via
methods for constructing, grouping, mutating, filtering, and summarising
\emph{Ranges} and an algebra for reasoning about actions on
\emph{Ranges} and relationships between \emph{Ranges}.

\hypertarget{design-and-implementation}{%
\section{Design and Implementation}\label{design-and-implementation}}

The \emph{plyranges} API implements a domain specific language using the
existing \emph{IRanges} and \emph{GenomicRanges} packages in
Bioconductor as a backend. Consequently, our API still has the speed and
effeciency of the aforementioned packages but with a more coherent
interface. The API also extends the grammar elements in \emph{dplyr} for
performing genomic specific manipulations such as finding overlapping
regions or nearest neighbour regions between many \emph{Ranges}. The
\emph{plyranges} API is specifically designed to enable fast interactive
analysis of \emph{Ranges} objects but can also be used for scripting
genomic data workflows.

We have desgined the API to be fluent. Every function call corresponds
to an action on a \emph{Ranges} object (they are named verbs) and where
possible functions have few arguments. Each verb is constructed to
enable a tab completion based workflow. Both of these aspects reduce the
cognitive load on a new user since most manipulations can be performed
with a vocabulary of several verbs, rather than having to memorise
functions with many arguments that are nouns (as is required in the
existing Bioconductor packages). This is also has the advantage of
allowing users to write human readable code because verbs describe what
the code is doing rather than how its doing it.

Workflows can be composed by chaining verbs together via the pipe
operator,\texttt{\%\textgreater{}\%} (exported from the R package
\emph{magrittr}). This is possible because every function call is
endomorphic: when the input is \emph{Ranges} object the output will also
be a \emph{Ranges} object. One advantage of this static typing is that
it does not require any additional learning of classes beyond
\emph{Ranges} and the \emph{DataFrame} classes. This is similar to the
bedtools API, where the output is usually a BED file. However, it
strongly deviates from the design of the \emph{Ranges} Bioconductor
packages, where many methods return a new class upon return. The
Bioconductor design enables effecient computing as users are exposed to
low-level features of its API which \emph{plyranges} trys to abstract
away. Method chaining via the pipe operator can also be difficult to
debug, as there multiple points of failure.

\hypertarget{working-with-ranges}{%
\subsection{Working with Ranges}\label{working-with-ranges}}

\hypertarget{construction-and-importoutput}{%
\subsubsection{Construction and
Import/Output}\label{construction-and-importoutput}}

The \emph{plyranges} API provides the methods \texttt{as\_granges()} and
\texttt{as\_iranges()} for constructing Ranges from tabular data
structures, such as data.frames in base R. These methods use
non-standard evaluation so columns in a data.frame can be modified
before a Ranges object is created.

The package also provides a consistent framework for reading and writing
files from and to common genomic data formats, using the
\texttt{rtracklayer} package as a backend. The methods are implemented
in the \texttt{read\_/write\_} family of functions, currently
\emph{plyranges} can read and write BAM, BED, BEDPE, narrowPeaks,
GFF/GTF, WIG and BigWig files.

\hypertarget{manipulation-of-ranges}{%
\subsubsection{Manipulation of Ranges}\label{manipulation-of-ranges}}

The \emph{plyranges} API exports the six core verbs from the
\emph{dplyr} package and modifys them for use with simple \emph{Ranges}
and \emph{Genomic Ranges}. The verb \texttt{mutate()} takes a Ranges
object and a set of name-value pairs and generates a new Ranges object
that with modified or new metadata columns or modified core components
(start, end, width, seqnames, strand). The use of \texttt{mutate()}
means that a user no longer needs knowledge of the accessors of the
Ranges object, as they can modify them in place. The \texttt{filter()}
function takes a Ranges object and logical expressions and restricts
Ranges object to where the logical expression evaluates to true. The
\texttt{summarise()} function takes a Ranges object and a set of
name-value pairs and aggregates the Ranges according to functions
evaluated in the name-value pairs. As \texttt{summarise()} is an
aggregation it may break the structure of the of a Ranges object, hence
it returns a DataFrame object. The \texttt{select()} function determines
which metadata columns are returned and the order they are returned in.
The \texttt{arrange()} function sorts a Ranges object by named
variables. The \texttt{group\_by()} function creates an implicit
grouping of Ranges object according to variables in the Ranges object.
This modifies the actions of \texttt{mutate()}, \texttt{summarise()} and
\texttt{filter()} (and also the set operations,
\texttt{reduce\_ranges()} and \texttt{disjoin\_ranges()}), so they are
performed on each partition created by the grouping.

\hypertarget{arithimetic-on-ranges}{%
\subsubsection{Arithimetic on Ranges}\label{arithimetic-on-ranges}}

The API has an expressive algebra for performing arithmetic on Ranges
via the verbs \texttt{set\_width()} and \texttt{stretch()}. As the names
suggest \texttt{set\_width()} modifies the width of a Ranges object,
while \texttt{stretch} extends the start and end of a Ranges object.
These can be chained with the anchoring functions
\texttt{anchor\_start()}, \texttt{anchor\_end()},
\texttt{anchor\_center()}, \texttt{anchor\_3p()} or
\texttt{anchor\_5p()}, which fix the coordinates of a Ranges object in
place. Moreover, the \texttt{shift\_} and \texttt{flank\_} family of
functions can be used to shift all coordinates in a Ranges object or
generate flanking regions from a Ranges object to the left, right,
upstream or downstream of the input. Unlike, the Bioconductor API,
\emph{plyranges} makes it expclit via function calls whether to take
into account the strand information of a \emph{Ranges} object.

\hypertarget{overlapping-ranges}{%
\subsubsection{Overlapping Ranges}\label{overlapping-ranges}}

A common operation to perform between two \emph{Ranges} objects is to
find overlaps or nearest neighbours. The \emph{plyranges} API recasts
these operations as `joins' or `pairing' operations. For overlaps, there
are three join operations: \texttt{join\_overlap\_intersect()},
\texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
are shown in figure (). These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges
objects as metadata. The intersect join takes the intersect of the start
and end coordinates of overlapping intervals of the query and subject
Ranges (for GenomicRanges it also accounts for sequence name), when
there is a overlap the metadata corresponding to the query and subject
Ranges are returned. Similarly, inner join takes the start and end
coordinates of the query Ranges that overlap the subject Ranges and
returns metadata of the overlapping query and subject Ranges. Finally,
the left join performs a left outer join between the query and subject
Ranges, it returns all genomic intervals from the query ranges, and
returns missing values in metadata columns when there is no overlap. A
user may also restrict or group by overlaps with the
\texttt{filter\_by\_overlaps()}, \texttt{filter\_by\_non\_overlaps()}
and \texttt{group\_by\_overlaps()}. All overlap methods can be modified
with the \texttt{within} suffix (which changes the type of overlap from
`any' to `within') or the \texttt{directed} suffix (which takes into
account the strand of a GenomicRanges object.).

For nearest neighbours, the \texttt{plyranges} API provides
\texttt{join\_nearest()}, \texttt{join\_precede()}, and
\texttt{join\_follow()} functions. These functions are similar to the
overlapping functions, in that they return the query ranges that are
nearest (or precede or follow) the subject ranges and add metadata from
the subject ranges when the query is a nearest neighbour of the subject.
Like the overlap joins, these functions can modified with suffixes to
find nearest neighbours that are left, right, upstream or downstream of
the subject.

The pairing operations, \texttt{pair\_overlap()},
\texttt{pair\_nearest()}, \texttt{pair\_follow()}, and
\texttt{pair\_precede()} are similar to the join operation but instead
of returning a Ranges, they pair up the subject and query Ranges objects
into a DataFrame, alongside their metadata columns. This data structure
is similar to the \emph{Pairs} data structure in the \emph{S4Vectors}
package or the BED-PE file format.

\hypertarget{non-standard-evaluation}{%
\subsubsection{Non-standard evaluation}\label{non-standard-evaluation}}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{availablilty-and-future-work}{%
\section{Availablilty and Future
Work}\label{availablilty-and-future-work}}

The \emph{plyranges} package is avaiable on the Bioconductor project
website \url{https://bioconductor.org} or can be accessed via Github
\url{https://github.com/sa-lee/plyranges}. We aim to continue developing
the \texttt{plyranges} package and extend it for use with more complex
data structures such as the \emph{SummarizedExperiment} class, which can
be used for analysing transcriptomic and variant data. As the
\texttt{plyranges} interface encourages tidy data practices it
integrates well with the principles of the grammar of graphics, we aim
to use it for the visualisation of genomic ranges.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Lawrence2013-wg}{}%
1. Lawrence M, Huber W, Pagès H, Aboyoun P, Carlson M, Gentleman R, et
al. Software for computing and annotating genomic ranges. PLoS Comput
Biol. 2013;9.
doi:\href{https://doi.org/10.1371/journal.pcbi.1003118}{10.1371/journal.pcbi.1003118}

\leavevmode\hypertarget{ref-Ovaska2013-gd}{}%
2. Ovaska K, Lyly L, Sahu B, Jänne OA, Hautaniemi S. Genomic region
operation kit for flexible processing of deep sequencing data. IEEE/ACM
Trans Comput Biol Bioinform. 2013;10: 200--206.
doi:\href{https://doi.org/10.1109/TCBB.2012.170}{10.1109/TCBB.2012.170}

\leavevmode\hypertarget{ref-Kozanitis2014-va}{}%
3. Kozanitis C, Heiberg A, Varghese G, Bafna V. Using genome query
language to uncover genetic variation. Bioinformatics. 2014;30: 1--8.
doi:\href{https://doi.org/10.1093/bioinformatics/btt250}{10.1093/bioinformatics/btt250}

\leavevmode\hypertarget{ref-Kozanitis2016-bm}{}%
4. Kozanitis C, Patterson DA. GenAp: A distributed SQL interface for
genomic data. BMC Bioinformatics. 2016;17: 63.
doi:\href{https://doi.org/10.1186/s12859-016-0904-1}{10.1186/s12859-016-0904-1}

\leavevmode\hypertarget{ref-Kaitoua2017-pw}{}%
5. Kaitoua A, Pinoli P, Bertoni M, Ceri S. Framework for supporting
genomic operations. IEEE Trans Comput. 2017;66: 443--457.
doi:\href{https://doi.org/10.1109/TC.2016.2603980}{10.1109/TC.2016.2603980}

\leavevmode\hypertarget{ref-Quinlan2010-gc}{}%
6. Quinlan AR, Hall IM. BEDTools: A flexible suite of utilities for
comparing genomic features. Bioinformatics. 2010;26: 841--842.
doi:\href{https://doi.org/10.1093/bioinformatics/btq033}{10.1093/bioinformatics/btq033}

\leavevmode\hypertarget{ref-Dale2011-js}{}%
7. Dale RK, Pedersen BS, Quinlan AR. Pybedtools: A flexible python
library for manipulating genomic datasets and annotations.
Bioinformatics. 2011;27: 3423--3424.
doi:\href{https://doi.org/10.1093/bioinformatics/btr539}{10.1093/bioinformatics/btr539}

\leavevmode\hypertarget{ref-Wickham2014-jc}{}%
8. Wickham H. Tidy data. Journal of Statistical Software, Articles.
2014;59: 1--23.
doi:\href{https://doi.org/10.18637/jss.v059.i10}{10.18637/jss.v059.i10}

\leavevmode\hypertarget{ref-Wickham2017-dplyr}{}%
9. Wickham H, Francois R, Henry L, Müller K. Dplyr: A grammar of data
manipulation {[}Internet{]}. 2017. Available:
\url{https://CRAN.R-project.org/package=dplyr}

\nolinenumbers

\end{document}

### 2018-02-12 README.Rmd
move outline into README

--- 2017-11-08 outline.Rmd
@@
 ---
-title: Outline of plyranges a grammar of genomic data manipulation
-output:
-  bookdown::pdf_document2:
-    toc: false
-    number_sections: false
+output: github_document
 bibliography: references.bib
 link-citations: true
+---
 
----
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>"
+)
+```
+# plyranges-paper
 

### 2018-02-12 paper.Rmd
some analysis interspersed in the text

@@
 ---
-
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>", 
+                      fig.width = 6, fig.height = 4)
+library(ggbio)
+library(ggplot2)
+library(plyranges)
+```
 # Introduction
@@
 
+In order to provide a compatible API with _dplyr_, _plyranges_ makes use
+of non-standard evaluation in R via the _rlang_ package. 
+<!--- need to mention something about how this could be detrimental --->
+
 ## Working with Ranges
@@
 base R. These methods use non-standard evaluation so columns in a data.frame
-can be modified before a Ranges object is created.
+can be modified before a Ranges object is created. The API also has
+convienence methods for annotating or extracting annotations from Genomic Ranges
+objects with the `set_genome_info()` and `get_genome_info()` functions.
 
-The package also provides a consistent framework for reading and writing 
+There is a consistent framework for reading and writing 
 files from and to common genomic data formats, using the `rtracklayer` package
@@
 narrowPeaks, GFF/GTF, WIG and BigWig files.
+
+As an example, we use the Bioconductor package AnnotationHub to search for 
+BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project.
+We choose to focus on assays for primary T CD8+ memory cells from peripheral 
+blood. We can then read the BigWig file  corresponding to the H3K27Me3 
+methylation mark over chromosome 10. First we gather the BigWig file
+and extract its annotation information and filter it to chromosome 10. 
+Then we read the BigWig file only extracting scores if they overlap chromosome
+10. We then add annotation information to ensure we know we are working
+with the hg19 genome build.
+
+```{r load-bw, cache = TRUE}
+# E048 corresponds to CD8+ T cells from peripheral blood
+library(plyranges)
+library(AnnotationHub)
+q <-query(AnnotationHub(), c("EpigenomeRoadMap", "E048"))
+
+bw_file <- q[["AH33458"]]
+
+chr10_ranges <- bw_file %>% 
+  get_genome_info() %>%
+  filter(seqnames == "chr10")
+
+chr10_scores <- bw_file %>%
+  read_bigwig(overlap_ranges = chr10_ranges) %>%
+  set_genome_info(genome = "hg19")
+chr10_scores
+```
 
@@
 As `summarise()` is an aggregation it may break the structure of the of a Ranges
-object, hence it returns a DataFrame object. The `select()` function determines
+object, hence it returns a DataFrame object.
+The `select()` function determines
 which metadata columns are returned and the order they are returned in. The 
@@
 
+Returning to the Ranges object containing normalised coverage scores for
+the methylation data, we can filter to find the coordinates of the peak 
+containing maximum coverage score. We can then find 
+a 5000 nt region centered around the maximum position.
+
+```{r max-score-region}
+chr10_max_score_region <- chr10_scores %>%
+  filter(score == max(score)) %>% 
+  anchor_center() %>%
+  set_width(5000)
+```
+
 ### Overlapping Ranges
@@
 ranges, and returns missing values in metadata columns when there is no overlap.
+
+The overlap innner join could be used to restrict the chromosome 10
+normalised coverage scores that are within the 5000nt region that contains
+the max peak on chromosome 10 (visualised in figure ()).
+
+```{r peak_region, fig.align = "center", fig.cap = "Visualisation of peak coverage score accross 5000nt region of chromosome 10"}
+peak_region <- chr10_scores %>%
+  join_overlap_inner(., 
+                     chr10_max_score_region %>% 
+                       select(-score))
+ggbio::autoplot(peak_region, 
+                aes(y = score), 
+                geom = "blank") + 
+  geom_area(aes(x = start))
+```
+
 A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
@@
 
-### Non-standard evaluation
-
 # Results

### 2018-02-12 paper.tex
some analysis interspersed in the text

@@
 \newcommand{\ipsum}{{\bf IPSUM}}
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 
@@
 
+In order to provide a compatible API with \emph{dplyr}, \emph{plyranges}
+makes use of non-standard evaluation in R via the \emph{rlang} package.
+
 \hypertarget{working-with-ranges}{%
@@
 non-standard evaluation so columns in a data.frame can be modified
-before a Ranges object is created.
+before a Ranges object is created. The API also has convienence methods
+for annotating or extracting annotations from Genomic Ranges objects
+with the \texttt{set\_genome\_info()} and \texttt{get\_genome\_info()}
+functions.
 
-The package also provides a consistent framework for reading and writing
-files from and to common genomic data formats, using the
-\texttt{rtracklayer} package as a backend. The methods are implemented
-in the \texttt{read\_/write\_} family of functions, currently
-\emph{plyranges} can read and write BAM, BED, BEDPE, narrowPeaks,
-GFF/GTF, WIG and BigWig files.
+There is a consistent framework for reading and writing files from and
+to common genomic data formats, using the \texttt{rtracklayer} package
+as a backend. The methods are implemented in the \texttt{read\_/write\_}
+family of functions, currently \emph{plyranges} can read and write BAM,
+BED, BEDPE, narrowPeaks, GFF/GTF, WIG and BigWig files.
+
+As an example, we use the Bioconductor package AnnotationHub to search
+for BigWig files from ChIP-Seq experiments from the Human Epigenome
+Roadmap project. We choose to focus on assays for primary T CD8+ memory
+cells from peripheral blood. We can then read the BigWig file
+corresponding to the H3K27Me3 methylation mark over chromosome 10. First
+we gather the BigWig file and extract its annotation information and
+filter it to chromosome 10. Then we read the BigWig file only extracting
+scores if they overlap chromosome 10. We then add annotation information
+to ensure we know we are working with the hg19 genome build.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# E048 corresponds to CD8+ T cells from peripheral blood}
+\KeywordTok{library}\NormalTok{(plyranges)}
+\KeywordTok{library}\NormalTok{(AnnotationHub)}
+\NormalTok{q <-}\KeywordTok{query}\NormalTok{(}\KeywordTok{AnnotationHub}\NormalTok{(), }\KeywordTok{c}\NormalTok{(}\StringTok{"EpigenomeRoadMap"}\NormalTok{, }\StringTok{"E048"}\NormalTok{))}
+
+\NormalTok{bw_file <-}\StringTok{ }\NormalTok{q[[}\StringTok{"AH33458"}\NormalTok{]] }
+
+\NormalTok{chr10_ranges <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{get_genome_info}\NormalTok{() }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{filter}\NormalTok{(seqnames }\OperatorTok{==}\StringTok{ "chr10"}\NormalTok{)}
+
+\NormalTok{chr10_scores <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{read_bigwig}\NormalTok{(}\DataTypeTok{overlap_ranges =}\NormalTok{ chr10_ranges) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{set_genome_info}\NormalTok{(}\DataTypeTok{genome =} \StringTok{"hg19"}\NormalTok{)}
+\NormalTok{chr10_scores}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#> GRanges object with 5789841 ranges and 1 metadata column:
+#>             seqnames                 ranges strand |              score
+#>                <Rle>              <IRanges>  <Rle> |          <numeric>
+#>         [1]    chr10         [    1, 60602]      * | 0.0422799997031689
+#>         [2]    chr10         [60603, 60781]      * |  0.163240000605583
+#>         [3]    chr10         [60782, 60816]      * |  0.372139990329742
+#>         [4]    chr10         [60817, 60995]      * |  0.163240000605583
+#>         [5]    chr10         [60996, 61625]      * | 0.0422799997031689
+#>         ...      ...                    ...    ... .                ...
+#>   [5789837]    chr10 [135524723, 135524734]      * |  0.144319996237755
+#>   [5789838]    chr10 [135524735, 135524775]      * |  0.250230014324188
+#>   [5789839]    chr10 [135524776, 135524784]      * |  0.427789986133575
+#>   [5789840]    chr10 [135524785, 135524806]      * |  0.730019986629486
+#>   [5789841]    chr10 [135524807, 135524837]      * |   1.03103005886078
+#>   -------
+#>   seqinfo: 25 sequences from hg19 genome
+\end{verbatim}
 
@@
 
+Returning to the Ranges object containing normalised coverage scores for
+the methylation data, we can filter to find the coordinates of the peak
+containing maximum coverage score. We can then find a 5000 nt region
+centered around the maximum position.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{chr10_max_score_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{filter}\NormalTok{(score }\OperatorTok{==}\StringTok{ }\KeywordTok{max}\NormalTok{(score)) }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{anchor_center}\NormalTok{() }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{set_width}\NormalTok{(}\DecValTok{5000}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
 \hypertarget{overlapping-ranges}{%
@@
 Ranges, it returns all genomic intervals from the query ranges, and
-returns missing values in metadata columns when there is no overlap. A
-user may also restrict or group by overlaps with the
+returns missing values in metadata columns when there is no overlap.
+
+The overlap innner join could be used to restrict the chromosome 10
+normalised coverage scores that are within the 5000nt region that
+contains the max peak on chromosome 10 (visualised in figure ()).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{peak_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(., }
+\NormalTok{                     chr10_max_score_region }\OperatorTok{%>%}\StringTok{ }
+\StringTok{                       }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{score))}
+\NormalTok{ggbio}\OperatorTok{::}\KeywordTok{autoplot}\NormalTok{(peak_region, }
+                \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ score), }
+                \DataTypeTok{geom =} \StringTok{"blank"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
+\StringTok{  }\KeywordTok{geom_area}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ start))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/peak_region-1}
+
+}
+
+\caption{Visualisation of peak coverage score accross 5000nt region of chromosome 10}\label{fig:peak_region}
+\end{figure}
+
+A user may also restrict or group by overlaps with the
 \texttt{filter\_by\_overlaps()}, \texttt{filter\_by\_non\_overlaps()}
@@
 package or the BED-PE file format.
-
-\hypertarget{non-standard-evaluation}{%
-\subsubsection{Non-standard evaluation}\label{non-standard-evaluation}}
 

### 2018-02-13 paper.Rmd
design section mostly complete; now onto results

@@
 bibliography: references.bib
-output: rticles::plos_article
+output: 
+  rticles::plos_article:
 csl: plos.csl
@@
 knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>", 
-                      fig.width = 6, fig.height = 4)
+                      fig.width = 5, fig.height = 3, fig.align = "center")
+options(digits = 3)
 library(ggbio)
@@
 assay or annotations from a genome database can be joined to their relevant
-sequences.
-In the Bioconductor/R packages `IRanges` and `GenomicRanges` these representations have been implemented as a suite of data structures called _Ranges_ @Lawrence2013-wg.
-These data structures cover many common data types encountered in bioinformatics analyses - a gene can be represented with its coordinates, along with its identifier and the identifiers of its exons; or
-an RNA-seq experiment may be represented as sets of genes with a matching count column.
+sequences. In the Bioconductor/R packages `IRanges` and `GenomicRanges` these 
+representations have been implemented as a suite of data structures called
+_Ranges_ @Lawrence2013-wg. These data structures cover many common data types 
+encountered in bioinformatics analyses - a gene can be represented with its 
+coordinates, along with its identifier and the identifiers of its exons; or
+an RNA-seq experiment may be represented as sets of genes with a matching count 
+column.
 
 The Bioconductor infrastructure for computing with genomic ranges are highly 
-effecient and powerful, however the application programming interface (API) for performing analysis tasks with _Ranges_ is complex due to its large number of
+effecient and powerful, however the application programming interface (API) for 
+performing analysis tasks with _Ranges_ is complex due to its large number of
 methods and classes. It also makes resulting scripts written 
@@
 these are either implemented as query languages or as command line tools 
-embedded in the unix environment ^[other ideas to mention GROK @Ovaska2013-gd]. An example of the former is the Genome Query Language (GQL) and its distributed 
+embedded in the unix environment ^[other ideas to mention GROK @Ovaska2013-gd]. 
+An example of the former is the Genome Query Language (GQL) and its distributed 
 implementation GenAp which use an SQL-like syntax for fast retrieval of 
 information from genomic databases and BAM files @Kozanitis2014-va; @Kozanitis2016-bm. 
-Another  example is the Genometric Query Language (GMQL) which implements a relational algebra for combining big genomic datasets @Kaitoua2017-pw.  The command line application 
+Another  example is the Genometric Query Language (GMQL) which implements a 
+relational algebra for combining big genomic datasets @Kaitoua2017-pw.  
+The command line application 
 BEDtools develops an extensive algebra for performing arithmetic between two or 
 more sets of genomic regions  @Quinlan2010-gc. It also has a python interface 
-which simplifies construcing scripts for performing analyses based on BEDTools  @Dale2011-js. ^[probably should also mention something about their cons, and more detaiil]
+which simplifies constructing scripts for performing analyses based on BEDTools 
+@Dale2011-js. 
+<!--- probably should also mention something about their cons, and more detail --->
 
 The abstraction provided by the _Ranges_ data structures aligns with the concept
-of tidy data @Wickham2014-jc. The tidy data pattern is useful because it allows us to see how the data relates to the design of an experiment
-and the variables measured.
-Consequently, it makes both the modelling and manipulation of data more systematic.
-The _Ranges_ data structure follows this abstraction: it is a rectanglular table corresponding to a single biological context.
-Each row contains a single observation and each column a variable about that observation.
+of tidy data @Wickham2014-jc. The tidy data pattern is useful because it allows 
+us to see how the data relates to the design of an experiment
+and the variables measured. Consequently, it makes both the modelling and 
+manipulation of data more systematic. The _Ranges_ data structure follows 
+this abstraction: it is a rectanglular table corresponding to a single biological 
+context. Each row contains a single observation and each column a variable about
+that observation.
 
@@
 implements and extends the grammar defined by the R package `dplyr` @Wickham2017-dplyr. 
-The grammar provides a consistent way of interacting with and analysing genomic data via methods for constructing, grouping, mutating, filtering, and summarising _Ranges_ and an algebra for reasoning about actions on _Ranges_ and relationships between _Ranges_.
+The grammar provides a consistent way of interacting with and analysing genomic 
+data via methods for constructing, grouping, mutating, filtering, and 
+summarising _Ranges_ and an algebra for reasoning about actions on _Ranges_
+and relationships between _Ranges_.
 
@@
 
-In order to provide a compatible API with _dplyr_, _plyranges_ makes use
-of non-standard evaluation in R via the _rlang_ package. 
-<!--- need to mention something about how this could be detrimental --->
+In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
+use of non-standard evaluation in R via the _rlang_ package. Simply, this
+means that compuations are performed and evaluated in the context of the 
+Ranges objects, which emphasises the interactive nature of the API. This has
+the disadvantage that programming with _plyranges_ becomes more difficult 
+because a user needs to capture expressions inside function calls and then
+unquote them.
+<!--- need to make this clearer? --->
 
@@
 The _plyranges_ API provides the methods `as_granges()` and `as_iranges()`
-for constructing Ranges from tabular data structures, such as data.frames in
-base R. These methods use non-standard evaluation so columns in a data.frame
-can be modified before a Ranges object is created. The API also has
+for constructing Ranges from tabular data structures, such as the _data.frame_ in
+base R. These methods use non-standard evaluation so columns in a _data.frame_
+can be modified before a _Ranges_ object is created. The API also has
 convienence methods for annotating or extracting annotations from Genomic Ranges
@@
 We choose to focus on assays for primary T CD8+ memory cells from peripheral 
-blood. We can then read the BigWig file  corresponding to the H3K27Me3 
+blood. We can then read the BigWig file  corresponding to the H3 lysine 27 
+ trimethylation (H3K27Me3) 
 methylation mark over chromosome 10. First we gather the BigWig file
 and extract its annotation information and filter it to chromosome 10.
-Then we read the BigWig file only extracting scores if they overlap chromosome
-10. We then add annotation information to ensure we know we are working
-with the hg19 genome build.
 
@@
   filter(seqnames == "chr10")
+```
 
+Then we read the BigWig file only extracting scores if they overlap chromosome
+10. We then add annotation information to ensure we know we are working
+with the hg19 genome build.
+
+```{r}
 chr10_scores <- bw_file %>%
@@
 to variables in the Ranges object. This modifies the actions of `mutate()`,
-`summarise()` and `filter()` (and also the set operations, `reduce_ranges()`
-and `disjoin_ranges()`), so they are performed on each partition created by the
-grouping.
+`summarise()` and `filter()`, so they are performed on each partition created 
+by the grouping. The `group_by()` operation acts as a replacement for 
+the _GenomicRangesList_ class in the original _GenomicRanges_ API.
+
+The _plyranges_ API introduces additional summary verbs, `reduce_ranges()`
+and `disjoin_ranges()`that return Ranges objects after being returned. The
+`reduce_ranges()` operation merges overlapping and neighbour ranges, while
+`disjoin_ranges()` expands ranges by taking the union of end points.
+
+The `reduce_ranges()` operation could be used to find coverage peaks 
+accross chromosome 10. We can manually set a threshold to restrict genomic
+regions to have a coverage score of greater than 8, and
+then merge those regions and compute the maximum coverage over all the regions
+that were reduced.
+
+```{r}
+all_peaks <- chr10_scores %>% 
+  filter(score > 8) %>% 
+  reduce_ranges(score = max(score))
+```
+
+The combination of these verbs together encapsulate most operations that
+can be performed in the original _IRanges_ and _GenomicRanges_ packages
+without the user be exposed to new classes. In those packages to perform
+the above operation requires users to understand what an run length encoding
+object and views object means and perform additional manipulations to return
+the result back to a _Ranges_.
 
@@
 containing maximum coverage score. We can then find 
-a 5000 nt region centered around the maximum position.
+a 5000 nt region centered around the maximum position by anchoring and 
+modifying the the width.
 
@@
 `join_overlap_intersect()`, `join_overlap_inner()` and `join_overlap_left()`
-which are shown in figure (). These operations consider any overlap between
+which are shown in figure (\ref{fig:olap}).
+
+```{r olap, echo = FALSE, fig.cap="The three overlap joins: the query and subject ranges are coloured by their metadata.
+When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.", fig.width = 5, fig.height = 7}
+a <- GRanges(seqnames = "chr1",
+               strand = c("+", "-"),
+               ranges = IRanges(c(1, 9), c(7, 10)),
+               key.a = letters[1:2])
+
+b <- GRanges(seqnames = "chr1",
+               strand = c("-", "+"),
+               ranges = IRanges(c(2, 6), c(4, 8)),
+               key.b = LETTERS[1:2])
+
+c <- join_overlap_intersect(a,b)
+d <- join_overlap_inner(a, b)
+e <- join_overlap_left(a, b)
+
+p1 <- autoplot(a, aes(fill = key.a)) + 
+  scale_fill_brewer(type = "qual", palette = "Set2") + 
+  guides(fill = FALSE)
+p2 <- autoplot(b, aes(fill = key.b))  + 
+  scale_fill_brewer(type = "qual", palette = "Set1") +
+  guides(fill = FALSE)
+p3 <- autoplot(c, aes(fill = key.a, colour = key.b)) + 
+  scale_fill_brewer(type = "qual", palette = "Set2") +
+  scale_color_brewer(type = "qual", palette = "Set1") +
+  guides(fill = FALSE, color = FALSE) 
+p4 <- autoplot(d, aes(fill = key.a, colour = key.b)) + 
+  scale_fill_brewer(type = "qual", palette = "Set2") +
+  scale_color_brewer(type = "qual", palette = "Set1") +
+  guides(fill = FALSE, colour = FALSE)
+p5 <- autoplot(e, aes(fill = key.a, colour = key.b))  +
+  scale_fill_brewer(type = "qual", palette = "Set2") +
+  scale_color_brewer(type = "qual", palette = "Set1") +
+  guides(fill = FALSE, colour = FALSE)
+
+tracks(query = p1, subject = p2, 
+       `intersect` = p3, 
+       `inner` = p4, 
+       `left` = p5) 
+```
+
+These operations consider any overlap between
 two input ranges and return any corresponding metadata from both Ranges objects
@@
 normalised coverage scores that are within the 5000nt region that contains
-the max peak on chromosome 10 (visualised in figure ()).
+the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).
 
-```{r peak_region, fig.align = "center", fig.cap = "Visualisation of peak coverage score accross 5000nt region of chromosome 10"}
+```{r peak_region}
 peak_region <- chr10_scores %>%
@@
                        select(-score))
+
+```
+
+```{r peak-viz, echo = FALSE, fig.cap = "Visualisation of normalised coverage scores accross a 5000nt region of chromosome 10 from H3K27Me3 ChIP-Seq assay from the Human Epigenome Roadmap project."}
 ggbio::autoplot(peak_region, 
@@
 
+- Redo an analysis in https://www.nature.com/articles/nature14248
+
+in particular section: Trait-associated variants enrich in tissue-specific marks
+
 # Availablilty and Future Work
@@
 practices it integrates well with the principles of the grammar of graphics, 
-we aim to use it for the visualisation of genomic ranges.
+we aim to use it for the visualisation of multimodal biological datasets.
 

### 2018-02-13 paper.tex
design section mostly complete; now onto results

@@
 Language (GMQL) which implements a relational algebra for combining big
-genomic datasets {[}5{]}. The command line application BEDtools develops
-an extensive algebra for performing arithmetic between two or more sets
-of genomic regions {[}6{]}. It also has a python interface which
-simplifies construcing scripts for performing analyses based on BEDTools
-{[}7{]}.\footnote{probably should also mention something about their
-  cons, and more detaiil}
+genomic datasets {[}5{]}.\\
+The command line application BEDtools develops an extensive algebra for
+performing arithmetic between two or more sets of genomic regions
+{[}6{]}. It also has a python interface which simplifies constructing
+scripts for performing analyses based on BEDTools {[}7{]}.
 
@@
 In order to provide a compatible API with \emph{dplyr}, \emph{plyranges}
-makes use of non-standard evaluation in R via the \emph{rlang} package.
+makes extensive use of non-standard evaluation in R via the \emph{rlang}
+package. Simply, this means that compuations are performed and evaluated
+in the context of the Ranges objects, which emphasises the interactive
+nature of the API. This has the disadvantage that programming with
+\emph{plyranges} becomes more difficult because a user needs to capture
+expressions inside function calls and then unquote them.
 
@@
 \texttt{as\_iranges()} for constructing Ranges from tabular data
-structures, such as data.frames in base R. These methods use
-non-standard evaluation so columns in a data.frame can be modified
-before a Ranges object is created. The API also has convienence methods
-for annotating or extracting annotations from Genomic Ranges objects
-with the \texttt{set\_genome\_info()} and \texttt{get\_genome\_info()}
-functions.
+structures, such as the \emph{data.frame} in base R. These methods use
+non-standard evaluation so columns in a \emph{data.frame} can be
+modified before a \emph{Ranges} object is created. The API also has
+convienence methods for annotating or extracting annotations from
+Genomic Ranges objects with the \texttt{set\_genome\_info()} and
+\texttt{get\_genome\_info()} functions.
 
@@
 cells from peripheral blood. We can then read the BigWig file
-corresponding to the H3K27Me3 methylation mark over chromosome 10. First
-we gather the BigWig file and extract its annotation information and
-filter it to chromosome 10. Then we read the BigWig file only extracting
-scores if they overlap chromosome 10. We then add annotation information
-to ensure we know we are working with the hg19 genome build.
+corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation
+mark over chromosome 10. First we gather the BigWig file and extract its
+annotation information and filter it to chromosome 10.
 
@@
 \StringTok{  }\KeywordTok{filter}\NormalTok{(seqnames }\OperatorTok{==}\StringTok{ "chr10"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
+Then we read the BigWig file only extracting scores if they overlap
+chromosome 10. We then add annotation information to ensure we know we
+are working with the hg19 genome build.
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{chr10_scores <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}
@@
 This modifies the actions of \texttt{mutate()}, \texttt{summarise()} and
-\texttt{filter()} (and also the set operations,
-\texttt{reduce\_ranges()} and \texttt{disjoin\_ranges()}), so they are
-performed on each partition created by the grouping.
+\texttt{filter()}, so they are performed on each partition created by
+the grouping. The \texttt{group\_by()} operation acts as a replacement
+for the \emph{GenomicRangesList} class in the original
+\emph{GenomicRanges} API.
+
+The \emph{plyranges} API introduces additional summary verbs,
+\texttt{reduce\_ranges()} and \texttt{disjoin\_ranges()}that return
+Ranges objects after being returned. The \texttt{reduce\_ranges()}
+operation merges overlapping and neighbour ranges, while
+\texttt{disjoin\_ranges()} expands ranges by taking the union of end
+points.
+
+The \texttt{reduce\_ranges()} operation could be used to find coverage
+peaks accross chromosome 10. We can manually set a threshold to restrict
+genomic regions to have a coverage score of greater than 8, and then
+merge those regions and compute the maximum coverage over all the
+regions that were reduced.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{all_peaks <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{filter}\NormalTok{(score }\OperatorTok{>}\StringTok{ }\DecValTok{8}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{reduce_ranges}\NormalTok{(}\DataTypeTok{score =} \KeywordTok{max}\NormalTok{(score))}
+\end{Highlighting}
+\end{Shaded}
+
+The combination of these verbs together encapsulate most operations that
+can be performed in the original \emph{IRanges} and \emph{GenomicRanges}
+packages without the user be exposed to new classes. In those packages
+to perform the above operation requires users to understand what an run
+length encoding object and views object means and perform additional
+manipulations to return the result back to a \emph{Ranges}.
 
@@
 containing maximum coverage score. We can then find a 5000 nt region
-centered around the maximum position.
+centered around the maximum position by anchoring and modifying the the
+width.
 
@@
 \texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
-are shown in figure (). These operations consider any overlap between
-two input ranges and return any corresponding metadata from both Ranges
-objects as metadata. The intersect join takes the intersect of the start
-and end coordinates of overlapping intervals of the query and subject
-Ranges (for GenomicRanges it also accounts for sequence name), when
-there is a overlap the metadata corresponding to the query and subject
-Ranges are returned. Similarly, inner join takes the start and end
-coordinates of the query Ranges that overlap the subject Ranges and
-returns metadata of the overlapping query and subject Ranges. Finally,
-the left join performs a left outer join between the query and subject
-Ranges, it returns all genomic intervals from the query ranges, and
-returns missing values in metadata columns when there is no overlap.
+are shown in figure (\ref{fig:olap}).
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/olap-1}
+
+}
+
+\caption{The three overlap joins: the query and subject ranges are coloured by their metadata.
+When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.}\label{fig:olap}
+\end{figure}
+
+These operations consider any overlap between two input ranges and
+return any corresponding metadata from both Ranges objects as metadata.
+The intersect join takes the intersect of the start and end coordinates
+of overlapping intervals of the query and subject Ranges (for
+GenomicRanges it also accounts for sequence name), when there is a
+overlap the metadata corresponding to the query and subject Ranges are
+returned. Similarly, inner join takes the start and end coordinates of
+the query Ranges that overlap the subject Ranges and returns metadata of
+the overlapping query and subject Ranges. Finally, the left join
+performs a left outer join between the query and subject Ranges, it
+returns all genomic intervals from the query ranges, and returns missing
+values in metadata columns when there is no overlap.
 
@@
 normalised coverage scores that are within the 5000nt region that
-contains the max peak on chromosome 10 (visualised in figure ()).
+contains the max peak on chromosome 10 (visualised in figure
+\ref{fig:peak-viz}).
 
@@
 \StringTok{                       }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{score))}
-\NormalTok{ggbio}\OperatorTok{::}\KeywordTok{autoplot}\NormalTok{(peak_region, }
-                \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ score), }
-                \DataTypeTok{geom =} \StringTok{"blank"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
-\StringTok{  }\KeywordTok{geom_area}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ start))}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/peak_region-1}
+{\centering \includegraphics{paper_files/figure-latex/peak-viz-1}
 
@@
 
-\caption{Visualisation of peak coverage score accross 5000nt region of chromosome 10}\label{fig:peak_region}
+\caption{Visualisation of normalised coverage scores accross a 5000nt region of chromosome 10 from H3K27Me3 ChIP-Seq assay from the Human Epigenome Roadmap project.}\label{fig:peak-viz}
 \end{figure}
@@
 
+\begin{itemize}
+\tightlist
+\item
+  Redo an analysis in https://www.nature.com/articles/nature14248
+\end{itemize}
+
 \hypertarget{availablilty-and-future-work}{%
@@
 integrates well with the principles of the grammar of graphics, we aim
-to use it for the visualisation of genomic ranges.
+to use it for the visualisation of multimodal biological datasets.
 

### 2018-02-25 paper.Rmd
not happy with that idea ---
updates

@@
 By recognising that these data structures follow 'tidy' data principles we have created a grammar of genomic data manipulation that defines verbs for performing actions on and between genomic interval data.
-This grammar simplifies performing common genomic data analysis tasks via method chaining, type consistency and results in creating human readable pipelines.
+This grammar simplifies performing common genomic data analysis tasks via a consistent interface to existing Bioconductor infrastructure which results in creating human readable analysis pipelines.
 We have implemented this grammar as an Bioconductor/R package called plyranges.
@@
                       fig.width = 5, fig.height = 3, fig.align = "center")
-options(digits = 3)
+options(digits = 2)
 library(ggbio)
@@
 library(plyranges)
+
+library(AnnotationHub)
+q <- AnnotationHub()
+bw_file <- q[["AH33458"]] 
 ```
@@
 representations have been implemented as a suite of data structures called
-_Ranges_ @Lawrence2013-wg. These data structures cover many common data types 
-encountered in bioinformatics analyses - a gene can be represented with its 
+_Ranges_ @Lawrence2013-wg; @Huber2015-ei. These data structures cover 
+many common data types  encountered in bioinformatics analyses. For
+example, a gene can be represented with its 
 coordinates, along with its identifier and the identifiers of its exons; or
@@
 The Bioconductor infrastructure for computing with genomic ranges are highly 
-effecient and powerful, however the application programming interface (API) for 
+efficient and powerful, however the application programming interface (API) for 
 performing analysis tasks with _Ranges_ is complex due to its large number of
@@
 would be to implement a domain specific language (DSL) as a fluent interface 
-built on top _Ranges_. The goal of fluent interface is to enable users to write 
-human-readable code via method chaining and consistent function returns. Fluent 
-interfaces fit naturally in the context of Bioinformatics workflows because they 
+built on top _Ranges_ [@Fowler2010-zd]. The goal of a fluent interface is to 
+enable users to write  human-readable code via method chaining and consistent function returns. Fluent  interfaces fit naturally in the context of bioinformatics workflows because they 
 enable writing succinct pipelines.
@@
 these are either implemented as query languages or as command line tools 
-embedded in the unix environment ^[other ideas to mention GROK @Ovaska2013-gd]. 
-An example of the former is the Genome Query Language (GQL) and its distributed 
+embedded in the Unix environment.  An example of the former is the 
+Genome Query Language (GQL) and its distributed 
 implementation GenAp which use an SQL-like syntax for fast retrieval of 
@@
 The abstraction provided by the _Ranges_ data structures aligns with the concept
-of tidy data @Wickham2014-jc. The tidy data pattern is useful because it allows 
+of tidy data @Wickham2014-jc. A tidy dataset is defined to be to a tabular
+data structure that has observations as rows and columns as variables, and each
+tidy dataset represents measurements from a single observational unit.
+The tidy data pattern is useful because it allows 
 us to see how the data relates to the design of an experiment
@@
 manipulation of data more systematic. The _Ranges_ data structure follows 
-this abstraction: it is a rectanglular table corresponding to a single biological 
-context. Each row contains a single observation and each column a variable about
+this abstraction: it is a rectangular table corresponding to a single biological 
+context. Each row contains a single observation and each column is a variable about
 that observation.
 
-The tidy data abstraction has motivated the development of `plyranges`
-a grammar of genomic data manipulation based on the _Ranges_ data structures. It
-implements and extends the grammar defined by the R package `dplyr` @Wickham2017-dplyr. 
-The grammar provides a consistent way of interacting with and analysing genomic 
-data via methods for constructing, grouping, mutating, filtering, and 
-summarising _Ranges_ and an algebra for reasoning about actions on _Ranges_
-and relationships between _Ranges_.
+The _plyranges_ API implements a domain specific language using the existing
+_IRanges_ and _GenomicRanges_ packages in Bioconductor as a back-end. Consequently,
+our API still has the speed and efficiency of the aforementioned packages but
+with a more coherent interface. The API also
+extends the grammar elements in _dplyr_ [@Wickham2017-dplyr] 
+for performing genomic specific
+manipulations such as finding overlapping regions or nearest neighbour regions
+between many _Ranges_ . The _plyranges_ API is specifically designed to enable
+fast interactive analysis of _Ranges_ objects but can also be used for scripting 
+genomic data workflows.
 
@@
 
-The _plyranges_ API implements a domain specific language using the existing
-_IRanges_ and _GenomicRanges_ packages in Bioconductor as a backend. Consequently,
-our API still has the speed and effeciency of the aforementioned packages but
-with a more coherent interface. The API also
-extends the grammar elements in _dplyr_ for performing genomic specific
-manipulations such as finding overlapping regions or nearest neighbour regions
-between many _Ranges_. The _plyranges_ API is specifically designed to enable
-fast interactive analysis of _Ranges_ objects but can also be used for scripting 
-genomic data workflows.
-
-We have desgined the API to be fluent. Every function call corresponds to an
+We have designed the API to be fluent. Every function call corresponds to an
 action on a _Ranges_ object (they are named verbs) and where possible
@@
 
-Workflows can be composed by chaining verbs together via the pipe operator,`%>%` 
-(exported from the R package _magrittr_). This is possible because every 
-function call is endomorphic: when the input is _Ranges_ object the output
-will also be a _Ranges_ object. One advantage of this static typing is that it 
+Workflows can be composed by chaining verbs together via the forward pipe operator,`%>%` 
+(exported from the R package _magrittr_ @R-magrittr). This is possible because every 
+function call is an endomorphism: when the input is _Ranges_ object the output
+will also be a _Ranges_ object. One advantage of endomorphism is that it 
 does not  require any additional learning of classes beyond _Ranges_ and 
-the _DataFrame_ classes. This is similar to the bedtools API, where the
-output is usually a BED file. However, it strongly deviates from the design of 
+the _DataFrame_ classes. This strongly deviates from the design of 
 the _Ranges_ Bioconductor packages, where many methods return a new class
-upon return. The Bioconductor design enables effecient computing as users
-are exposed to low-level features of its API which _plyranges_ trys to abstract
+upon return. The Bioconductor design enables efficient computing as users
+are exposed to low-level features of its API which _plyranges_ abstracts
 away. Method chaining via the pipe operator can also be difficult to debug,
@@
 In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
-use of non-standard evaluation in R via the _rlang_ package. Simply, this
-means that compuations are performed and evaluated in the context of the 
-Ranges objects, which emphasises the interactive nature of the API. This has
+use of non-standard evaluation in R via the _rlang_ package @R-rlang. Simply, this
+means that computations are performed and evaluated in the context of the 
+_Ranges_ objects, which emphasises the interactive nature of the API. This has
 the disadvantage that programming with _plyranges_ becomes more difficult 
@@
 
-## Working with Ranges
-
-### Construction and Import/Output
-
-The _plyranges_ API provides the methods `as_granges()` and `as_iranges()`
-for constructing Ranges from tabular data structures, such as the _data.frame_ in
-base R. These methods use non-standard evaluation so columns in a _data.frame_
-can be modified before a _Ranges_ object is created. The API also has
-convienence methods for annotating or extracting annotations from Genomic Ranges
-objects with the `set_genome_info()` and `get_genome_info()` functions.
-
-There is a consistent framework for reading and writing 
-files from and to common genomic data formats, using the `rtracklayer` package
-as a backend. The methods are implemented in the `read_/write_` family of 
-functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
-narrowPeaks, GFF/GTF, WIG and BigWig files.
-
-As an example, we use the Bioconductor package AnnotationHub to search for 
-BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project.
-We choose to focus on assays for primary T CD8+ memory cells from peripheral 
-blood. We can then read the BigWig file  corresponding to the H3 lysine 27 
- trimethylation (H3K27Me3) 
-methylation mark over chromosome 10. First we gather the BigWig file
-and extract its annotation information and filter it to chromosome 10.
-
-```{r load-bw, cache = TRUE}
-# E048 corresponds to CD8+ T cells from peripheral blood
-library(plyranges)
-library(AnnotationHub)
-q <-query(AnnotationHub(), c("EpigenomeRoadMap", "E048"))
-
-bw_file <- q[["AH33458"]]
-
-chr10_ranges <- bw_file %>% 
-  get_genome_info() %>%
-  filter(seqnames == "chr10")
-```
-
-Then we read the BigWig file only extracting scores if they overlap chromosome
-10. We then add annotation information to ensure we know we are working
-with the hg19 genome build.
-
-```{r}
-chr10_scores <- bw_file %>%
-  read_bigwig(overlap_ranges = chr10_ranges) %>%
-  set_genome_info(genome = "hg19")
-chr10_scores
-```
-
-### Manipulation of Ranges
+## Actions on Ranges
 
 The _plyranges_ API exports the six core verbs from the _dplyr_ package
-and modifys them for use with simple _Ranges_ and _Genomic Ranges_. The
+and modifies them for use with _Ranges_ objects. The
 verb `mutate()` takes a Ranges object and a set of name-value pairs
@@
 As `summarise()` is an aggregation it may break the structure of the of a Ranges
-object, hence it returns a DataFrame object.
+object, hence it returns a _DataFrame_ object.
 The `select()` function determines
@@
 
-The `reduce_ranges()` operation could be used to find coverage peaks 
-accross chromosome 10. We can manually set a threshold to restrict genomic
-regions to have a coverage score of greater than 8, and
-then merge those regions and compute the maximum coverage over all the regions
-that were reduced.
+## Arithimetic on Ranges
 
-```{r}
-all_peaks <- chr10_scores %>% 
-  filter(score > 8) %>% 
-  reduce_ranges(score = max(score))
-```
-
-The combination of these verbs together encapsulate most operations that
-can be performed in the original _IRanges_ and _GenomicRanges_ packages
-without the user be exposed to new classes. In those packages to perform
-the above operation requires users to understand what an run length encoding
-object and views object means and perform additional manipulations to return
-the result back to a _Ranges_.
-
-### Arithimetic on Ranges
-
-The API has an expressive algebra for performing arithmetic on Ranges via
+The _plyranges_ API has an expressive algebra for performing arithmetic on Ranges via
 the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
@@
 object to the left, right, upstream or downstream of the input. Unlike, the
-Bioconductor API, _plyranges_ makes it expclit via function calls whether to
+Bioconductor API, _plyranges_ makes it explicit via function calls whether to
 take into account the strand information of a _Ranges_ object.
 
-Returning to the Ranges object containing normalised coverage scores for
-the methylation data, we can filter to find the coordinates of the peak 
-containing maximum coverage score. We can then find 
-a 5000 nt region centered around the maximum position by anchoring and 
-modifying the the width.
-
-```{r max-score-region}
-chr10_max_score_region <- chr10_scores %>%
-  filter(score == max(score)) %>% 
-  anchor_center() %>%
-  set_width(5000)
-```
-
-### Overlapping Ranges
+## Overlapping Ranges
 
@@
 as metadata. The intersect join takes the intersect of the start and end coordinates
-of overlapping intervals of the query and subject Ranges 
-(for GenomicRanges it also accounts for sequence name), when there is a overlap
-the metadata corresponding to the query and subject Ranges are returned. 
-Similarly, inner join takes the start and end coordinates of the query Ranges
-that overlap the subject Ranges and returns metadata of the overlapping query
-and subject Ranges. Finally, the left join performs a left outer join between
-the query and subject Ranges, it returns all genomic intervals from the query
+of overlapping intervals of the query and subject _Ranges_ 
+(for _GenomicRanges_ it also accounts for sequence name), when there is a overlap
+the metadata corresponding to the query and subject _Ranges_ are returned. 
+Similarly, inner join takes the start and end coordinates of the query _Ranges_
+that overlap the subject _Ranges_ and returns metadata of the overlapping query
+and subject _Ranges_. Finally, the left join performs a left outer join between
+the query and subject _Ranges_, it returns all genomic intervals from the query
 ranges, and returns missing values in metadata columns when there is no overlap.
 
-The overlap innner join could be used to restrict the chromosome 10
+A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
+`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
+be modified with the `within` suffix (which changes the type of overlap 
+from 'any' to 'within') or the `directed` suffix (which takes into account
+the strand of a _GenomicRanges_ object.).
+
+For nearest neighbours, the _plyranges_ API provides `join_nearest()`, 
+`join_precede()`, and `join_follow()` functions. These functions are similar
+to the overlapping functions, in that they return the query ranges that are
+nearest (or precede or follow) the subject ranges and add metadata from
+the subject ranges when the query is a nearest neighbour of the subject. Like
+the overlap joins, these functions can modified with suffixes to find nearest
+neighbours that are left, right, upstream or downstream of the subject.
+
+The pairing operations, `pair_overlap()`, `pair_nearest()`, `pair_follow()`, 
+and `pair_precede()` are similar to the join operation but instead of returning
+a _Ranges_, they pair up the subject and query _Ranges_ objects into a _DataFrame_,
+alongside their metadata columns. This data structure is similar to the _Pairs_
+data structure in the _S4Vectors_ @R-S4vectors package or the BED-PE file format.
+
+The combination of verbs we have defined above encapsulate all operations that
+can be performed in the original _IRanges_ and _GenomicRanges_ packages
+without the user be exposed to new classes. In those packages to perform
+most operations requires users to learn many classes and perform 
+additional manipulations to return the results of their computation back to a
+_Ranges_ object.
+
+## Construction and Import/Output
+
+The methods `as_granges()` and `as_iranges()`
+for constructing _Ranges_ from tabular data structures, such as the _data.frame_ in
+base R. These methods use non-standard evaluation so columns in a _data.frame_
+can be modified before a _Ranges_ object is created. The API also has
+convenience methods for annotating or extracting annotations from _GRanges_
+objects with the `set_genome_info()` and `get_genome_info()` functions.
+
+There is a consistent framework for reading and writing 
+files from and to common genomic data formats, using the _rtracklayer_ package
+as a back-end @Lawrence2009-nt. The methods are implemented in the `read_/write_` family of 
+functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
+narrowPeaks, GFF/GTF, WIG and BigWig files.
+
+# Results
+
+As an example, we use the Bioconductor package _AnnotationHub_ @R-ahub to search for 
+BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project @Roadmap_Epigenomics_Consortium2015-pr.
+We choose to focus on assays for primary T CD8+ memory cells from peripheral 
+blood. We can then read the BigWig file  corresponding to the H3 lysine 27 
+ trimethylation (H3K27Me3) 
+methylation mark over chromosome 10.
+
+First, we gather the BigWig file
+and extract its annotation information and filter it to chromosome 10.
+
+```{r load-bw, cache = TRUE}
+library(plyranges)
+chr10_ranges <- bw_file %>% 
+  get_genome_info() %>%
+  filter(seqnames == "chr10")
+```
+
+Then we read the BigWig file only extracting scores if they overlap chromosome
+10. The annotation information from the file is automatically included (in this
+case the hg19 genome build).
+
+```{r}
+chr10_scores <- bw_file %>%
+  read_bigwig(overlap_ranges = chr10_ranges) %>%
+  set_genome_info(genome = "hg19")
+chr10_scores
+```
+
+The `reduce_ranges()` operation is used to find coverage peaks 
+across chromosome 10. We can manually set a threshold to restrict genomic
+regions to have a coverage score of greater than 8, and
+then merge nearby regions. The maximum coverage is computed over all the
+coverage scores in regions that were reduced.
+
+```{r}
+all_peaks <- chr10_scores %>% 
+  filter(score > 8) %>% 
+  reduce_ranges(score = max(score))
+```
+
+Returning to the _Ranges_ object containing normalised coverage scores for
+the methylation data, we can filter to find the coordinates of the peak 
+containing maximum coverage score. We can then find 
+a 5000 nt region centered around the maximum position by anchoring and 
+modifying the the width.
+
+```{r max-score-region}
+chr10_max_score_region <- chr10_scores %>%
+  filter(score == max(score)) %>% 
+  anchor_center() %>%
+  set_width(5000)
+```
+
+Finally, the overlap inner join could be used to restrict the chromosome 10
 normalised coverage scores that are within the 5000nt region that contains
@@
 peak_region <- chr10_scores %>%
-  join_overlap_inner(., 
-                     chr10_max_score_region %>% 
+  join_overlap_inner(chr10_max_score_region %>% 
                        select(-score))
@@
 
-A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
-`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
-be modified with the `within` suffix (which changes the type of overlap 
-from 'any' to 'within') or the `directed` suffix (which takes into account
-the strand of a GenomicRanges object.).
-
-For nearest neighbours, the `plyranges` API provides `join_nearest()`, 
-`join_precede()`, and `join_follow()` functions. These functions are similar
-to the overlapping functions, in that they return the query ranges that are
-nearest (or precede or follow) the subject ranges and add metadata from
-the subject ranges when the query is a nearest neighbour of the subject. Like
-the overlap joins, these functions can modified with suffixes to find nearest
-neighbours that are left, right, upstream or downstream of the subject.
-
-The pairing operations, `pair_overlap()`, `pair_nearest()`, `pair_follow()`, 
-and `pair_precede()` are similar to the join operation but instead of returning
-a Ranges, they pair up the subject and query Ranges objects into a DataFrame,
-alongside their metadata columns. This data structure is similar to the _Pairs_
-data structure in the _S4Vectors_ package or the BED-PE file format. 
-<!--- should also create convience methods for grouping by pairs and pairing up, 
-unpairing, would also be useful for alignments too--->
-
-# Results
-
-- Redo an analysis in https://www.nature.com/articles/nature14248
-
-in particular section: Trait-associated variants enrich in tissue-specific marks
-
 # Availablilty and Future Work
 
-The _plyranges_ package is avaiable on the Bioconductor project website 
+The _plyranges_ package is available on the Bioconductor project website 
 [https://bioconductor.org](https://bioconductor.org) or can be accessed
 via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
-We aim to continue developing the `plyranges` package and extend it for use
+We aim to continue developing the _plyranges_ package and extend it for use
 with more complex data structures  such as the 
 _SummarizedExperiment_ class, which can be used for analysing transcriptomic
-and variant data. As the `plyranges` interface encourages tidy data 
+and variant data. As the _plyranges_ interface encourages tidy data 
 practices it integrates well with the principles of the grammar of graphics, 

### 2018-02-25 paper.tex
updates

@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
 \newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
 \newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 
@@
 and between genomic interval data. This grammar simplifies performing
-common genomic data analysis tasks via method chaining, type consistency
-and results in creating human readable pipelines. We have implemented
-this grammar as an Bioconductor/R package called plyranges.
+common genomic data analysis tasks via a consistent interface to
+existing Bioconductor infrastructure which results in creating human
+readable analysis pipelines. We have implemented this grammar as an
+Bioconductor/R package called plyranges.
 
@@
 % Use "Eq" instead of "Equation" for equation citations.
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 these representations have been implemented as a suite of data
-structures called \emph{Ranges} {[}1{]}. These data structures cover
-many common data types encountered in bioinformatics analyses - a gene
-can be represented with its coordinates, along with its identifier and
-the identifiers of its exons; or an RNA-seq experiment may be
-represented as sets of genes with a matching count column.
+structures called \emph{Ranges} {[}1{]}; {[}2{]}. These data structures
+cover many common data types encountered in bioinformatics analyses. For
+example, a gene can be represented with its coordinates, along with its
+identifier and the identifiers of its exons; or an RNA-seq experiment
+may be represented as sets of genes with a matching count column.
 
 The Bioconductor infrastructure for computing with genomic ranges are
-highly effecient and powerful, however the application programming
+highly efficient and powerful, however the application programming
 interface (API) for performing analysis tasks with \emph{Ranges} is
@@
 reason about. An alternative approach would be to implement a domain
-specific language (DSL) as a fluent interface built on top
-\emph{Ranges}. The goal of fluent interface is to enable users to write
+specific language (DSL) as a fluent interface built on top \emph{Ranges}
+{[}3{]}. The goal of a fluent interface is to enable users to write
 human-readable code via method chaining and consistent function returns.
-Fluent interfaces fit naturally in the context of Bioinformatics
+Fluent interfaces fit naturally in the context of bioinformatics
 workflows because they enable writing succinct pipelines.
@@
 Broadly, these are either implemented as query languages or as command
-line tools embedded in the unix environment.\footnote{other ideas to
-  mention GROK {[}2{]}} An example of the former is the Genome Query
-Language (GQL) and its distributed implementation GenAp which use an
-SQL-like syntax for fast retrieval of information from genomic databases
-and BAM files {[}3{]}; {[}4{]}. Another example is the Genometric Query
-Language (GMQL) which implements a relational algebra for combining big
-genomic datasets {[}5{]}.\\
+line tools embedded in the Unix environment. An example of the former is
+the Genome Query Language (GQL) and its distributed implementation GenAp
+which use an SQL-like syntax for fast retrieval of information from
+genomic databases and BAM files {[}4{]}; {[}5{]}. Another example is the
+Genometric Query Language (GMQL) which implements a relational algebra
+for combining big genomic datasets {[}6{]}.\\
 The command line application BEDtools develops an extensive algebra for
 performing arithmetic between two or more sets of genomic regions
-{[}6{]}. It also has a python interface which simplifies constructing
-scripts for performing analyses based on BEDTools {[}7{]}.
+{[}7{]}. It also has a python interface which simplifies constructing
+scripts for performing analyses based on BEDTools {[}8{]}.
 
 The abstraction provided by the \emph{Ranges} data structures aligns
-with the concept of tidy data {[}8{]}. The tidy data pattern is useful
-because it allows us to see how the data relates to the design of an
-experiment and the variables measured. Consequently, it makes both the
-modelling and manipulation of data more systematic. The \emph{Ranges}
-data structure follows this abstraction: it is a rectanglular table
-corresponding to a single biological context. Each row contains a single
-observation and each column a variable about that observation.
-
-The tidy data abstraction has motivated the development of
-\texttt{plyranges} a grammar of genomic data manipulation based on the
-\emph{Ranges} data structures. It implements and extends the grammar
-defined by the R package \texttt{dplyr} {[}9{]}. The grammar provides a
-consistent way of interacting with and analysing genomic data via
-methods for constructing, grouping, mutating, filtering, and summarising
-\emph{Ranges} and an algebra for reasoning about actions on
-\emph{Ranges} and relationships between \emph{Ranges}.
-
-\hypertarget{design-and-implementation}{%
-\section{Design and Implementation}\label{design-and-implementation}}
+with the concept of tidy data {[}9{]}. A tidy dataset is defined to be
+to a tabular data structure that has observations as rows and columns as
+variables, and each tidy dataset represents measurements from a single
+observational unit. The tidy data pattern is useful because it allows us
+to see how the data relates to the design of an experiment and the
+variables measured. Consequently, it makes both the modelling and
+manipulation of data more systematic. The \emph{Ranges} data structure
+follows this abstraction: it is a rectangular table corresponding to a
+single biological context. Each row contains a single observation and
+each column is a variable about that observation.
 
@@
 existing \emph{IRanges} and \emph{GenomicRanges} packages in
-Bioconductor as a backend. Consequently, our API still has the speed and
-effeciency of the aforementioned packages but with a more coherent
-interface. The API also extends the grammar elements in \emph{dplyr} for
-performing genomic specific manipulations such as finding overlapping
-regions or nearest neighbour regions between many \emph{Ranges}. The
-\emph{plyranges} API is specifically designed to enable fast interactive
-analysis of \emph{Ranges} objects but can also be used for scripting
-genomic data workflows.
+Bioconductor as a back-end. Consequently, our API still has the speed
+and efficiency of the aforementioned packages but with a more coherent
+interface. The API also extends the grammar elements in \emph{dplyr}
+{[}10{]} for performing genomic specific manipulations such as finding
+overlapping regions or nearest neighbour regions between many
+\emph{Ranges} . The \emph{plyranges} API is specifically designed to
+enable fast interactive analysis of \emph{Ranges} objects but can also
+be used for scripting genomic data workflows.
 
-We have desgined the API to be fluent. Every function call corresponds
+\section{Design and Implementation}\label{design-and-implementation}
+
+We have designed the API to be fluent. Every function call corresponds
 to an action on a \emph{Ranges} object (they are named verbs) and where
@@
 
-Workflows can be composed by chaining verbs together via the pipe
-operator,\texttt{\%\textgreater{}\%} (exported from the R package
-\emph{magrittr}). This is possible because every function call is
-endomorphic: when the input is \emph{Ranges} object the output will also
-be a \emph{Ranges} object. One advantage of this static typing is that
-it does not require any additional learning of classes beyond
-\emph{Ranges} and the \emph{DataFrame} classes. This is similar to the
-bedtools API, where the output is usually a BED file. However, it
-strongly deviates from the design of the \emph{Ranges} Bioconductor
-packages, where many methods return a new class upon return. The
-Bioconductor design enables effecient computing as users are exposed to
-low-level features of its API which \emph{plyranges} trys to abstract
-away. Method chaining via the pipe operator can also be difficult to
-debug, as there multiple points of failure.
+Workflows can be composed by chaining verbs together via the forward
+pipe operator,\texttt{\%\textgreater{}\%} (exported from the R package
+\emph{magrittr} {[}11{]}). This is possible because every function call
+is an endomorphism: when the input is \emph{Ranges} object the output
+will also be a \emph{Ranges} object. One advantage of endomorphism is
+that it does not require any additional learning of classes beyond
+\emph{Ranges} and the \emph{DataFrame} classes. This strongly deviates
+from the design of the \emph{Ranges} Bioconductor packages, where many
+methods return a new class upon return. The Bioconductor design enables
+efficient computing as users are exposed to low-level features of its
+API which \emph{plyranges} abstracts away. Method chaining via the pipe
+operator can also be difficult to debug, as there multiple points of
+failure.
 
@@
 makes extensive use of non-standard evaluation in R via the \emph{rlang}
-package. Simply, this means that compuations are performed and evaluated
-in the context of the Ranges objects, which emphasises the interactive
-nature of the API. This has the disadvantage that programming with
-\emph{plyranges} becomes more difficult because a user needs to capture
-expressions inside function calls and then unquote them.
+package {[}12{]}. Simply, this means that computations are performed and
+evaluated in the context of the \emph{Ranges} objects, which emphasises
+the interactive nature of the API. This has the disadvantage that
+programming with \emph{plyranges} becomes more difficult because a user
+needs to capture expressions inside function calls and then unquote
+them.
 
-\hypertarget{working-with-ranges}{%
-\subsection{Working with Ranges}\label{working-with-ranges}}
+\subsection{Actions on Ranges}\label{actions-on-ranges}
 
-\hypertarget{construction-and-importoutput}{%
-\subsubsection{Construction and
-Import/Output}\label{construction-and-importoutput}}
+The \emph{plyranges} API exports the six core verbs from the
+\emph{dplyr} package and modifies them for use with \emph{Ranges}
+objects. The verb \texttt{mutate()} takes a Ranges object and a set of
+name-value pairs and generates a new Ranges object that with modified or
+new metadata columns or modified core components (start, end, width,
+seqnames, strand). The use of \texttt{mutate()} means that a user no
+longer needs knowledge of the accessors of the Ranges object, as they
+can modify them in place. The \texttt{filter()} function takes a Ranges
+object and logical expressions and restricts Ranges object to where the
+logical expression evaluates to true. The \texttt{summarise()} function
+takes a Ranges object and a set of name-value pairs and aggregates the
+Ranges according to functions evaluated in the name-value pairs. As
+\texttt{summarise()} is an aggregation it may break the structure of the
+of a Ranges object, hence it returns a \emph{DataFrame} object. The
+\texttt{select()} function determines which metadata columns are
+returned and the order they are returned in. The \texttt{arrange()}
+function sorts a Ranges object by named variables. The
+\texttt{group\_by()} function creates an implicit grouping of Ranges
+object according to variables in the Ranges object. This modifies the
+actions of \texttt{mutate()}, \texttt{summarise()} and
+\texttt{filter()}, so they are performed on each partition created by
+the grouping. The \texttt{group\_by()} operation acts as a replacement
+for the \emph{GenomicRangesList} class in the original
+\emph{GenomicRanges} API.
 
-The \emph{plyranges} API provides the methods \texttt{as\_granges()} and
-\texttt{as\_iranges()} for constructing Ranges from tabular data
-structures, such as the \emph{data.frame} in base R. These methods use
-non-standard evaluation so columns in a \emph{data.frame} can be
-modified before a \emph{Ranges} object is created. The API also has
-convienence methods for annotating or extracting annotations from
-Genomic Ranges objects with the \texttt{set\_genome\_info()} and
-\texttt{get\_genome\_info()} functions.
+The \emph{plyranges} API introduces additional summary verbs,
+\texttt{reduce\_ranges()} and \texttt{disjoin\_ranges()}that return
+Ranges objects after being returned. The \texttt{reduce\_ranges()}
+operation merges overlapping and neighbour ranges, while
+\texttt{disjoin\_ranges()} expands ranges by taking the union of end
+points.
+
+\subsection{Arithimetic on Ranges}\label{arithimetic-on-ranges}
+
+The \emph{plyranges} API has an expressive algebra for performing
+arithmetic on Ranges via the verbs \texttt{set\_width()} and
+\texttt{stretch()}. As the names suggest \texttt{set\_width()} modifies
+the width of a Ranges object, while \texttt{stretch} extends the start
+and end of a Ranges object. These can be chained with the anchoring
+functions \texttt{anchor\_start()}, \texttt{anchor\_end()},
+\texttt{anchor\_center()}, \texttt{anchor\_3p()} or
+\texttt{anchor\_5p()}, which fix the coordinates of a Ranges object in
+place. Moreover, the \texttt{shift\_} and \texttt{flank\_} family of
+functions can be used to shift all coordinates in a Ranges object or
+generate flanking regions from a Ranges object to the left, right,
+upstream or downstream of the input. Unlike, the Bioconductor API,
+\emph{plyranges} makes it explicit via function calls whether to take
+into account the strand information of a \emph{Ranges} object.
+
+\subsection{Overlapping Ranges}\label{overlapping-ranges}
+
+A common operation to perform between two \emph{Ranges} objects is to
+find overlaps or nearest neighbours. The \emph{plyranges} API recasts
+these operations as `joins' or `pairing' operations. For overlaps, there
+are three join operations: \texttt{join\_overlap\_intersect()},
+\texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
+are shown in figure (\ref{fig:olap}).
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/olap-1}
+
+}
+
+\caption{The three overlap joins: the query and subject ranges are coloured by their metadata.
+When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.}\label{fig:olap}
+\end{figure}
+
+These operations consider any overlap between two input ranges and
+return any corresponding metadata from both Ranges objects as metadata.
+The intersect join takes the intersect of the start and end coordinates
+of overlapping intervals of the query and subject \emph{Ranges} (for
+\emph{GenomicRanges} it also accounts for sequence name), when there is
+a overlap the metadata corresponding to the query and subject
+\emph{Ranges} are returned. Similarly, inner join takes the start and
+end coordinates of the query \emph{Ranges} that overlap the subject
+\emph{Ranges} and returns metadata of the overlapping query and subject
+\emph{Ranges}. Finally, the left join performs a left outer join between
+the query and subject \emph{Ranges}, it returns all genomic intervals
+from the query ranges, and returns missing values in metadata columns
+when there is no overlap.
+
+A user may also restrict or group by overlaps with the
+\texttt{filter\_by\_overlaps()}, \texttt{filter\_by\_non\_overlaps()}
+and \texttt{group\_by\_overlaps()}. All overlap methods can be modified
+with the \texttt{within} suffix (which changes the type of overlap from
+`any' to `within') or the \texttt{directed} suffix (which takes into
+account the strand of a \emph{GenomicRanges} object.).
+
+For nearest neighbours, the \emph{plyranges} API provides
+\texttt{join\_nearest()}, \texttt{join\_precede()}, and
+\texttt{join\_follow()} functions. These functions are similar to the
+overlapping functions, in that they return the query ranges that are
+nearest (or precede or follow) the subject ranges and add metadata from
+the subject ranges when the query is a nearest neighbour of the subject.
+Like the overlap joins, these functions can modified with suffixes to
+find nearest neighbours that are left, right, upstream or downstream of
+the subject.
+
+The pairing operations, \texttt{pair\_overlap()},
+\texttt{pair\_nearest()}, \texttt{pair\_follow()}, and
+\texttt{pair\_precede()} are similar to the join operation but instead
+of returning a \emph{Ranges}, they pair up the subject and query
+\emph{Ranges} objects into a \emph{DataFrame}, alongside their metadata
+columns. This data structure is similar to the \emph{Pairs} data
+structure in the \emph{S4Vectors} {[}13{]} package or the BED-PE file
+format.
+
+The combination of verbs we have defined above encapsulate all
+operations that can be performed in the original \emph{IRanges} and
+\emph{GenomicRanges} packages without the user be exposed to new
+classes. In those packages to perform most operations requires users to
+learn many classes and perform additional manipulations to return the
+results of their computation back to a \emph{Ranges} object.
+
+\subsection{Construction and
+Import/Output}\label{construction-and-importoutput}
+
+The methods \texttt{as\_granges()} and \texttt{as\_iranges()} for
+constructing \emph{Ranges} from tabular data structures, such as the
+\emph{data.frame} in base R. These methods use non-standard evaluation
+so columns in a \emph{data.frame} can be modified before a \emph{Ranges}
+object is created. The API also has convenience methods for annotating
+or extracting annotations from \emph{GRanges} objects with the
+\texttt{set\_genome\_info()} and \texttt{get\_genome\_info()} functions.
 
 There is a consistent framework for reading and writing files from and
-to common genomic data formats, using the \texttt{rtracklayer} package
-as a backend. The methods are implemented in the \texttt{read\_/write\_}
-family of functions, currently \emph{plyranges} can read and write BAM,
-BED, BEDPE, narrowPeaks, GFF/GTF, WIG and BigWig files.
+to common genomic data formats, using the \emph{rtracklayer} package as
+a back-end {[}14{]}. The methods are implemented in the
+\texttt{read\_/write\_} family of functions, currently \emph{plyranges}
+can read and write BAM, BED, BEDPE, narrowPeaks, GFF/GTF, WIG and BigWig
+files.
 
-As an example, we use the Bioconductor package AnnotationHub to search
-for BigWig files from ChIP-Seq experiments from the Human Epigenome
-Roadmap project. We choose to focus on assays for primary T CD8+ memory
-cells from peripheral blood. We can then read the BigWig file
-corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation
-mark over chromosome 10. First we gather the BigWig file and extract its
-annotation information and filter it to chromosome 10.
+\section{Results}\label{results}
+
+As an example, we use the Bioconductor package \emph{AnnotationHub}
+{[}15{]} to search for BigWig files from ChIP-Seq experiments from the
+Human Epigenome Roadmap project {[}16{]}. We choose to focus on assays
+for primary T CD8+ memory cells from peripheral blood. We can then read
+the BigWig file corresponding to the H3 lysine 27 trimethylation
+(H3K27Me3) methylation mark over chromosome 10.
+
+First, we gather the BigWig file and extract its annotation information
+and filter it to chromosome 10.
 
@@
 \begin{Highlighting}[]
-\CommentTok{# E048 corresponds to CD8+ T cells from peripheral blood}
 \KeywordTok{library}\NormalTok{(plyranges)}
-\KeywordTok{library}\NormalTok{(AnnotationHub)}
-\NormalTok{q <-}\KeywordTok{query}\NormalTok{(}\KeywordTok{AnnotationHub}\NormalTok{(), }\KeywordTok{c}\NormalTok{(}\StringTok{"EpigenomeRoadMap"}\NormalTok{, }\StringTok{"E048"}\NormalTok{))}
-
-\NormalTok{bw_file <-}\StringTok{ }\NormalTok{q[[}\StringTok{"AH33458"}\NormalTok{]] }
-
 \NormalTok{chr10_ranges <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}\StringTok{ }
@@
 Then we read the BigWig file only extracting scores if they overlap
-chromosome 10. We then add annotation information to ensure we know we
-are working with the hg19 genome build.
+chromosome 10. The annotation information from the file is automatically
+included (in this case the hg19 genome build).
 
@@
 
-\hypertarget{manipulation-of-ranges}{%
-\subsubsection{Manipulation of Ranges}\label{manipulation-of-ranges}}
-
-The \emph{plyranges} API exports the six core verbs from the
-\emph{dplyr} package and modifys them for use with simple \emph{Ranges}
-and \emph{Genomic Ranges}. The verb \texttt{mutate()} takes a Ranges
-object and a set of name-value pairs and generates a new Ranges object
-that with modified or new metadata columns or modified core components
-(start, end, width, seqnames, strand). The use of \texttt{mutate()}
-means that a user no longer needs knowledge of the accessors of the
-Ranges object, as they can modify them in place. The \texttt{filter()}
-function takes a Ranges object and logical expressions and restricts
-Ranges object to where the logical expression evaluates to true. The
-\texttt{summarise()} function takes a Ranges object and a set of
-name-value pairs and aggregates the Ranges according to functions
-evaluated in the name-value pairs. As \texttt{summarise()} is an
-aggregation it may break the structure of the of a Ranges object, hence
-it returns a DataFrame object. The \texttt{select()} function determines
-which metadata columns are returned and the order they are returned in.
-The \texttt{arrange()} function sorts a Ranges object by named
-variables. The \texttt{group\_by()} function creates an implicit
-grouping of Ranges object according to variables in the Ranges object.
-This modifies the actions of \texttt{mutate()}, \texttt{summarise()} and
-\texttt{filter()}, so they are performed on each partition created by
-the grouping. The \texttt{group\_by()} operation acts as a replacement
-for the \emph{GenomicRangesList} class in the original
-\emph{GenomicRanges} API.
-
-The \emph{plyranges} API introduces additional summary verbs,
-\texttt{reduce\_ranges()} and \texttt{disjoin\_ranges()}that return
-Ranges objects after being returned. The \texttt{reduce\_ranges()}
-operation merges overlapping and neighbour ranges, while
-\texttt{disjoin\_ranges()} expands ranges by taking the union of end
-points.
-
-The \texttt{reduce\_ranges()} operation could be used to find coverage
-peaks accross chromosome 10. We can manually set a threshold to restrict
+The \texttt{reduce\_ranges()} operation is used to find coverage peaks
+across chromosome 10. We can manually set a threshold to restrict
 genomic regions to have a coverage score of greater than 8, and then
-merge those regions and compute the maximum coverage over all the
-regions that were reduced.
+merge nearby regions. The maximum coverage is computed over all the
+coverage scores in regions that were reduced.
 
@@
 
-The combination of these verbs together encapsulate most operations that
-can be performed in the original \emph{IRanges} and \emph{GenomicRanges}
-packages without the user be exposed to new classes. In those packages
-to perform the above operation requires users to understand what an run
-length encoding object and views object means and perform additional
-manipulations to return the result back to a \emph{Ranges}.
-
-\hypertarget{arithimetic-on-ranges}{%
-\subsubsection{Arithimetic on Ranges}\label{arithimetic-on-ranges}}
-
-The API has an expressive algebra for performing arithmetic on Ranges
-via the verbs \texttt{set\_width()} and \texttt{stretch()}. As the names
-suggest \texttt{set\_width()} modifies the width of a Ranges object,
-while \texttt{stretch} extends the start and end of a Ranges object.
-These can be chained with the anchoring functions
-\texttt{anchor\_start()}, \texttt{anchor\_end()},
-\texttt{anchor\_center()}, \texttt{anchor\_3p()} or
-\texttt{anchor\_5p()}, which fix the coordinates of a Ranges object in
-place. Moreover, the \texttt{shift\_} and \texttt{flank\_} family of
-functions can be used to shift all coordinates in a Ranges object or
-generate flanking regions from a Ranges object to the left, right,
-upstream or downstream of the input. Unlike, the Bioconductor API,
-\emph{plyranges} makes it expclit via function calls whether to take
-into account the strand information of a \emph{Ranges} object.
-
-Returning to the Ranges object containing normalised coverage scores for
-the methylation data, we can filter to find the coordinates of the peak
-containing maximum coverage score. We can then find a 5000 nt region
-centered around the maximum position by anchoring and modifying the the
-width.
+Returning to the \emph{Ranges} object containing normalised coverage
+scores for the methylation data, we can filter to find the coordinates
+of the peak containing maximum coverage score. We can then find a 5000
+nt region centered around the maximum position by anchoring and
+modifying the the width.
 
@@
 
-\hypertarget{overlapping-ranges}{%
-\subsubsection{Overlapping Ranges}\label{overlapping-ranges}}
-
-A common operation to perform between two \emph{Ranges} objects is to
-find overlaps or nearest neighbours. The \emph{plyranges} API recasts
-these operations as `joins' or `pairing' operations. For overlaps, there
-are three join operations: \texttt{join\_overlap\_intersect()},
-\texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
-are shown in figure (\ref{fig:olap}).
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/olap-1}
-
-}
-
-\caption{The three overlap joins: the query and subject ranges are coloured by their metadata.
-When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.}\label{fig:olap}
-\end{figure}
-
-These operations consider any overlap between two input ranges and
-return any corresponding metadata from both Ranges objects as metadata.
-The intersect join takes the intersect of the start and end coordinates
-of overlapping intervals of the query and subject Ranges (for
-GenomicRanges it also accounts for sequence name), when there is a
-overlap the metadata corresponding to the query and subject Ranges are
-returned. Similarly, inner join takes the start and end coordinates of
-the query Ranges that overlap the subject Ranges and returns metadata of
-the overlapping query and subject Ranges. Finally, the left join
-performs a left outer join between the query and subject Ranges, it
-returns all genomic intervals from the query ranges, and returns missing
-values in metadata columns when there is no overlap.
-
-The overlap innner join could be used to restrict the chromosome 10
-normalised coverage scores that are within the 5000nt region that
+Finally, the overlap inner join could be used to restrict the chromosome
+10 normalised coverage scores that are within the 5000nt region that
 contains the max peak on chromosome 10 (visualised in figure
@@
 \NormalTok{peak_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(., }
-\NormalTok{                     chr10_max_score_region }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(chr10_max_score_region }\OperatorTok{%>%}\StringTok{ }
 \StringTok{                       }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{score))}
@@
 
-A user may also restrict or group by overlaps with the
-\texttt{filter\_by\_overlaps()}, \texttt{filter\_by\_non\_overlaps()}
-and \texttt{group\_by\_overlaps()}. All overlap methods can be modified
-with the \texttt{within} suffix (which changes the type of overlap from
-`any' to `within') or the \texttt{directed} suffix (which takes into
-account the strand of a GenomicRanges object.).
+\section{Availablilty and Future
+Work}\label{availablilty-and-future-work}
 
-For nearest neighbours, the \texttt{plyranges} API provides
-\texttt{join\_nearest()}, \texttt{join\_precede()}, and
-\texttt{join\_follow()} functions. These functions are similar to the
-overlapping functions, in that they return the query ranges that are
-nearest (or precede or follow) the subject ranges and add metadata from
-the subject ranges when the query is a nearest neighbour of the subject.
-Like the overlap joins, these functions can modified with suffixes to
-find nearest neighbours that are left, right, upstream or downstream of
-the subject.
-
-The pairing operations, \texttt{pair\_overlap()},
-\texttt{pair\_nearest()}, \texttt{pair\_follow()}, and
-\texttt{pair\_precede()} are similar to the join operation but instead
-of returning a Ranges, they pair up the subject and query Ranges objects
-into a DataFrame, alongside their metadata columns. This data structure
-is similar to the \emph{Pairs} data structure in the \emph{S4Vectors}
-package or the BED-PE file format.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\begin{itemize}
-\tightlist
-\item
-  Redo an analysis in https://www.nature.com/articles/nature14248
-\end{itemize}
-
-\hypertarget{availablilty-and-future-work}{%
-\section{Availablilty and Future
-Work}\label{availablilty-and-future-work}}
-
-The \emph{plyranges} package is avaiable on the Bioconductor project
+The \emph{plyranges} package is available on the Bioconductor project
 website \url{https://bioconductor.org} or can be accessed via Github
 \url{https://github.com/sa-lee/plyranges}. We aim to continue developing
-the \texttt{plyranges} package and extend it for use with more complex
+the \emph{plyranges} package and extend it for use with more complex
 data structures such as the \emph{SummarizedExperiment} class, which can
 be used for analysing transcriptomic and variant data. As the
-\texttt{plyranges} interface encourages tidy data practices it
-integrates well with the principles of the grammar of graphics, we aim
-to use it for the visualisation of multimodal biological datasets.
+\emph{plyranges} interface encourages tidy data practices it integrates
+well with the principles of the grammar of graphics, we aim to use it
+for the visualisation of multimodal biological datasets.
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Lawrence2013-wg}{}%
+\hypertarget{ref-Lawrence2013-wg}{}
 1. Lawrence M, Huber W, Pagès H, Aboyoun P, Carlson M, Gentleman R, et
@@
 
-\leavevmode\hypertarget{ref-Ovaska2013-gd}{}%
-2. Ovaska K, Lyly L, Sahu B, Jänne OA, Hautaniemi S. Genomic region
-operation kit for flexible processing of deep sequencing data. IEEE/ACM
-Trans Comput Biol Bioinform. 2013;10: 200--206.
-doi:\href{https://doi.org/10.1109/TCBB.2012.170}{10.1109/TCBB.2012.170}
+\hypertarget{ref-Huber2015-ei}{}
+2. Huber W, Carey VJ, Gentleman R, Anders S, Carlson M, Carvalho BS, et
+al. Orchestrating high-throughput genomic analysis with bioconductor.
+Nat Methods. Springer Nature; 2015;12: 115--121.
+doi:\href{https://doi.org/10.1038/nmeth.3252}{10.1038/nmeth.3252}
 
-\leavevmode\hypertarget{ref-Kozanitis2014-va}{}%
-3. Kozanitis C, Heiberg A, Varghese G, Bafna V. Using genome query
+\hypertarget{ref-Fowler2010-zd}{}
+3. Fowler M. Domain-Specific languages {[}Internet{]}. Pearson
+Education; 2010. Available:
+\url{https://market.android.com/details?id=book-ri1muolw_YwC}
+
+\hypertarget{ref-Kozanitis2014-va}{}
+4. Kozanitis C, Heiberg A, Varghese G, Bafna V. Using genome query
 language to uncover genetic variation. Bioinformatics. 2014;30: 1--8.
@@
 
-\leavevmode\hypertarget{ref-Kozanitis2016-bm}{}%
-4. Kozanitis C, Patterson DA. GenAp: A distributed SQL interface for
+\hypertarget{ref-Kozanitis2016-bm}{}
+5. Kozanitis C, Patterson DA. GenAp: A distributed SQL interface for
 genomic data. BMC Bioinformatics. 2016;17: 63.
@@
 
-\leavevmode\hypertarget{ref-Kaitoua2017-pw}{}%
-5. Kaitoua A, Pinoli P, Bertoni M, Ceri S. Framework for supporting
+\hypertarget{ref-Kaitoua2017-pw}{}
+6. Kaitoua A, Pinoli P, Bertoni M, Ceri S. Framework for supporting
 genomic operations. IEEE Trans Comput. 2017;66: 443--457.
@@
 
-\leavevmode\hypertarget{ref-Quinlan2010-gc}{}%
-6. Quinlan AR, Hall IM. BEDTools: A flexible suite of utilities for
+\hypertarget{ref-Quinlan2010-gc}{}
+7. Quinlan AR, Hall IM. BEDTools: A flexible suite of utilities for
 comparing genomic features. Bioinformatics. 2010;26: 841--842.
@@
 
-\leavevmode\hypertarget{ref-Dale2011-js}{}%
-7. Dale RK, Pedersen BS, Quinlan AR. Pybedtools: A flexible python
+\hypertarget{ref-Dale2011-js}{}
+8. Dale RK, Pedersen BS, Quinlan AR. Pybedtools: A flexible python
 library for manipulating genomic datasets and annotations.
@@
 
-\leavevmode\hypertarget{ref-Wickham2014-jc}{}%
-8. Wickham H. Tidy data. Journal of Statistical Software, Articles.
+\hypertarget{ref-Wickham2014-jc}{}
+9. Wickham H. Tidy data. Journal of Statistical Software, Articles.
 2014;59: 1--23.
@@
 
-\leavevmode\hypertarget{ref-Wickham2017-dplyr}{}%
-9. Wickham H, Francois R, Henry L, Müller K. Dplyr: A grammar of data
+\hypertarget{ref-Wickham2017-dplyr}{}
+10. Wickham H, Francois R, Henry L, Müller K. Dplyr: A grammar of data
 manipulation {[}Internet{]}. 2017. Available:
 \url{https://CRAN.R-project.org/package=dplyr}
+
+\hypertarget{ref-R-magrittr}{}
+11. Bache SM, Wickham H. Magrittr: A forward-pipe operator for r
+{[}Internet{]}. 2014. Available:
+\url{https://CRAN.R-project.org/package=magrittr}
+
+\hypertarget{ref-R-rlang}{}
+12. Henry L, Wickham H. Rlang: Functions for base types and core r and
+'tidyverse' features {[}Internet{]}. 2017. Available:
+\url{http://rlang.tidyverse.org}
+
+\hypertarget{ref-R-S4vectors}{}
+13. Pagès H, Lawrence M, Aboyoun P. S4Vectors: S4 implementation of
+vector-like and list-like objects. 2017.
+
+\hypertarget{ref-Lawrence2009-nt}{}
+14. Lawrence M, Gentleman R, Carey V. Rtracklayer: An R package for
+interfacing with genome browsers. Bioinformatics. 2009;25: 1841--1842.
+doi:\href{https://doi.org/10.1093/bioinformatics/btp328}{10.1093/bioinformatics/btp328}
+
+\hypertarget{ref-R-ahub}{}
+15. Morgan M. AnnotationHub: Client to access annotationhub resources.
+2017.
+
+\hypertarget{ref-Roadmap_Epigenomics_Consortium2015-pr}{}
+16. Roadmap Epigenomics Consortium, Kundaje A, Meuleman W, Ernst J,
+Bilenky M, Yen A, et al. Integrative analysis of 111 reference human
+epigenomes. Nature. 2015;518: 317--330.
+doi:\href{https://doi.org/10.1038/nature14248}{10.1038/nature14248}
 

### 2018-03-21 paper.Rmd
update abstract

@@
 ---
-title: "plyranges: a grammar for manipulating genomics data" 
+title: "plyranges: a grammar for transforming genomics data" 
 author:
@@
 abstract: |
-    The Bioconductor project has created many powerful abstractions for reasoning about genomics data, such as the _Ranges_ data structures for representing genomic intervals.
-By recognising that these data structures follow 'tidy' data principles we have created a grammar of genomic data manipulation that defines verbs for performing actions on and between genomic interval data.
-This grammar simplifies performing common genomic data analysis tasks via a consistent interface to existing Bioconductor infrastructure which results in creating human readable analysis pipelines.
+    The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments.
+However, there is a cognitive load placed on a user in learning a data abstraction and understanding its appropriate use.
+Throughout a standard workflow, a user must navigate and know many of these abstractions to perform an genomic analysis task, when a single data abstraction, a _Ranges_ object will suffice.
+The _Ranges_ class naturally represent genomic intervals and their associated measurements.
+By recognising that the _Ranges_ class follows 'tidy' data principles we have created a grammar of genomic data transformation.
+The grammar defines verbs for performing actions on and between genomic interval data.
+It provides a principled way
+    of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows.
 We have implemented this grammar as an Bioconductor/R package called plyranges.

### 2018-03-21 paper.Rmd
rejigged para 1

@@
 
-Genomic data may be naturally represented as sets of pairs of integers 
-corresponding to the start and end points of sequences. Further information
-such as strandedness and chromosome name may be added to these sets to 
-provide biological context. Because of the flexibility of this representation 
-supplemental information such as measurements obtained from an experimental 
-assay or annotations from a genome database can be joined to their relevant
-sequences. In the Bioconductor/R packages `IRanges` and `GenomicRanges` these 
+There is a large variety of genomics data available to the modern biologist
+from sequencing assays to public collections: annotations of genomic features
+such as genes or exons for reference genomes, coverage vectors in the BigWig
+format, peaks called from ChIP-Seq experiments and many more.
+All of these hetergenous data types can be naturally represented as annotated sets of 
+integer pairs, corresponding to the start and end points of sequences, along with 
+strandedness and chromosome location to provide biological context. Measurements
+or metadata can be included as additional features on their relevant sequences. 
+As an example agene can be represented with its coordinates, along with its 
+identifier and the identifiers of its exons; or an RNA-seq experiment may be
+represented as sets of genes with a matching count column.
+
+In the Bioconductor/R packages `IRanges` and `GenomicRanges` these 
 representations have been implemented as a suite of data structures called
 _Ranges_ @Lawrence2013-wg; @Huber2015-ei. These data structures cover 
-many common data types  encountered in bioinformatics analyses. For
-example, a gene can be represented with its 
-coordinates, along with its identifier and the identifiers of its exons; or
-an RNA-seq experiment may be represented as sets of genes with a matching count 
-column.
+many common data types  encountered in bioinformatics analyses.
 

### 2018-03-21 paper.Rmd
more changes to intro

@@
 
-In the Bioconductor/R packages `IRanges` and `GenomicRanges` these 
-representations have been implemented as a suite of data structures called
-_Ranges_ @Lawrence2013-wg; @Huber2015-ei. These data structures cover 
-many common data types  encountered in bioinformatics analyses.
+Many domain specific languages (DSLs) have been created to process and reason
+about heterogenous genomics data. Broadly, they are implemented as either
+query languages or as an embedded command line tool.
+An example of the former is the Genome Query Language (GQL) and its distributed 
+implementation GenAp which use an SQL-like syntax for fast retrieval of 
+information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. 
+Similarly, the Genometric Query Language (GMQL) implements a 
+relational algebra for combining genomic datasets @Kaitoua2017-pw.  
+The command line application BEDtools develops an extensive algebra for 
+performing arithmetic between two or more sets of genomic regions 
+@Quinlan2010-gc; @Dale2011-js. All of the aformentioned DSLs are designed to 
+be run either at the command line or embedded in scripts for batch processing. 
+As a consquence, they are not integrated into an environment designed to 
+perform interactive data  analysis that has access to state of the art 
+methods for statistics and visualisation like the R language.
 
-The Bioconductor infrastructure for computing with genomic ranges are highly 
-efficient and powerful, however the application programming interface (API) for 
-performing analysis tasks with _Ranges_ is complex due to its large number of
-methods and classes. It also makes resulting scripts written 
-difficult for a non-programmer to read and reason about. An alternative approach 
-would be to implement a domain specific language (DSL) as a fluent interface 
-built on top _Ranges_ [@Fowler2010-zd]. The goal of a fluent interface is to 
-enable users to write  human-readable code via method chaining and consistent function returns. Fluent  interfaces fit naturally in the context of bioinformatics workflows because they 
-enable writing succinct pipelines.
+The Bioconductor/R packages `IRanges` and `GenomicRanges` 
+define a DSL for analysing genomics data with R via
+an object-oriented (OO) interface @Lawrence2013-wg; @Huber2015-ei. This interface
+provides abstractions for representing genomics data and enables interoperability
+by allowing users and developers to use these abstractions in their own code
+and packages. However, due to the large number of abstractions there are costs
+involved. Due to the OO interface each abstraction will behave slightly 
+differently for any given generic method. There is a cognitive load on the user
+to understand the appropriate context for each abstraction. An alternative approach
+is to recognise that most genomic data analysis tasks can be performed with 
+data structures that naturally represent genomic intervals.
 
-Several other DSLs have been implemented to reason about genomics data. Broadly, 
-these are either implemented as query languages or as command line tools 
-embedded in the Unix environment.  An example of the former is the 
-Genome Query Language (GQL) and its distributed 
-implementation GenAp which use an SQL-like syntax for fast retrieval of 
-information from genomic databases and BAM files @Kozanitis2014-va; @Kozanitis2016-bm. 
-Another  example is the Genometric Query Language (GMQL) which implements a 
-relational algebra for combining big genomic datasets @Kaitoua2017-pw.  
-The command line application 
-BEDtools develops an extensive algebra for performing arithmetic between two or 
-more sets of genomic regions  @Quinlan2010-gc. It also has a python interface 
-which simplifies constructing scripts for performing analyses based on BEDTools 
-@Dale2011-js. 
-<!--- probably should also mention something about their cons, and more detail --->
-
-The abstraction provided by the _Ranges_ data structures aligns with the concept
-of tidy data @Wickham2014-jc. A tidy dataset is defined to be to a tabular
+This natural abstraction aligns with the concept of tidy data @Wickham2014-jc. 
+A tidy dataset is defined to be to a tabular
 data structure that has observations as rows and columns as variables, and each
@@
 us to see how the data relates to the design of an experiment
-and the variables measured. Consequently, it makes both the modelling and 
-manipulation of data more systematic. The _Ranges_ data structure follows 
-this abstraction: it is a rectangular table corresponding to a single biological 
+and the variables measured. The genomic interval data structure follows 
+this pattern: it is a rectangular table corresponding to a single biological 
 context. Each row contains a single observation and each column is a variable about
@@
 
-The _plyranges_ API implements a domain specific language using the existing
-_IRanges_ and _GenomicRanges_ packages in Bioconductor as a back-end. Consequently,
-our API still has the speed and efficiency of the aforementioned packages but
-with a more coherent interface. The API also
-extends the grammar elements in _dplyr_ [@Wickham2017-dplyr] 
-for performing genomic specific
-manipulations such as finding overlapping regions or nearest neighbour regions
-between many _Ranges_ . The _plyranges_ API is specifically designed to enable
-fast interactive analysis of _Ranges_ objects but can also be used for scripting 
+The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
+fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
+The package defines a consistent syntax and style for transforming tabular data sets via
+functions that are named as verbs. Each function performs an action on the data
+set and is cohesive and endomorphic. Cohesion means that each function performs
+a singular task and has no side-effects, while endomorphism means a function
+call returns the same type of object as its input.
+
+Then the last paragraph (already pretty close) states that:
+
+    we aim to simplify the interface to the most fundamental Ranges data structures, which we must reuse for the reasons stated above,
+    it is possible to align the Ranges data structures with the tabular notions of the dplyr API
+    so we will extend dplyr to incorporate genomic notions
+
+We have created a DSL called `plyranges` that uses the existing
+`IRanges` and `GenomicRanges` packages in Bioconductor as a back-end. Our API 
+still has the speed and efficiency of these packages but
+with a consistent interface. The API also extends the grammar elements in `dplyr` 
+for performing genomic specific manipulations such as finding overlapping 
+regions or nearest neighbour regions between genomic intervals. 
+The _plyranges_ DSL is specifically designed to enable fast interactive 
+analysis of _Ranges_ objects but can also be used for scripting 
 genomic data workflows.

### 2018-03-21 paper.Rmd
intro polish

@@
 
-There is a large variety of genomics data available to the modern biologist
-from sequencing assays to public collections: annotations of genomic features
-such as genes or exons for reference genomes, coverage vectors in the BigWig
-format, peaks called from ChIP-Seq experiments and many more.
-All of these hetergenous data types can be naturally represented as annotated sets of 
-integer pairs, corresponding to the start and end points of sequences, along with 
-strandedness and chromosome location to provide biological context. Measurements
-or metadata can be included as additional features on their relevant sequences. 
-As an example agene can be represented with its coordinates, along with its 
-identifier and the identifiers of its exons; or an RNA-seq experiment may be
-represented as sets of genes with a matching count column.
+There is a large variety of genomics data available to the scientific community,
+ranging from highly specific sequencing assays to publicly availble collections.
+These datasets similarly contain a wide variety of features: annotations of 
+genomic features such as genes or exons for reference genomes, large coverage 
+vectors, peaks called from ChIP-Seq experiments and many more. A modern biologist
+is expected to be able to reason about, combine and wrangle these heterogenous
+data sources in order to answer questions about their experimental hypotheses.
+Abstractions in the forms of data representation and software interfaces are
+key factors in inhibiting or enabling a biologist to construct reproducible data workflows
+to answer their research questions.
 
@@
 is to recognise that most genomic data analysis tasks can be performed with 
-data structures that naturally represent genomic intervals.
+data structures that naturally represent genomic intervals. That is, genomics
+data can be represented as sets of integer pairs, where each pair has a start
+and an end coordianate, along with strings representing strand and chromosome 
+location. Measurements or metadata can be included as additional annotations
+on each pair.  As an example. a gene can be represented with its coordinates, along with its 
+identifier and the identifiers of its exons; or an RNA-seq experiment may be
+represented as sets of genes with a matching count column.
 
@@
 a singular task and has no side-effects, while endomorphism means a function
-call returns the same type of object as its input.
+call returns the same type of object as its input. Taken together these
+features enable a user to write human readable analysis workflows.
 
-Then the last paragraph (already pretty close) states that:
-
-    we aim to simplify the interface to the most fundamental Ranges data structures, which we must reuse for the reasons stated above,
-    it is possible to align the Ranges data structures with the tabular notions of the dplyr API
-    so we will extend dplyr to incorporate genomic notions
-
-We have created a DSL called `plyranges` that uses the existing
-`IRanges` and `GenomicRanges` packages in Bioconductor as a back-end. Our API 
-still has the speed and efficiency of these packages but
-with a consistent interface. The API also extends the grammar elements in `dplyr` 
-for performing genomic specific manipulations such as finding overlapping 
-regions or nearest neighbour regions between genomic intervals. 
-The _plyranges_ DSL is specifically designed to enable fast interactive 
-analysis of _Ranges_ objects but can also be used for scripting 
-genomic data workflows.
+We have created a DSL called `plyranges` that is directly motivated by the design
+of `dplyr` but extends its  grammar to incorporate features of genomics data.
+Our interface uses the existing Bioconductor packages `IRanges` and 
+`GenomicRanges` but simplifies them by only exposing tidy genomic interval
+data structures to the user. Hence, we can encapsulate the features of the 
+of the `dplyr` API while performing genomic specific transformations 
+such as finding overlapping  regions or nearest neighbour regions between 
+genomic intervals.  The `plyranges` DSL is designed to enable fast interactive 
+analysis of genomics but also enables the creation of reproducible genomic 
+data workflows.
 
 # Design and Implementation
+
+<--- need to formally introduce the Ranges classes now --->
 

### 2018-03-22 paper.Rmd
a more abstract definition of the Ranges

@@
 
-<--- need to formally introduce the Ranges classes now --->
+The `plyranges` DSL is built on two fundamental data structures defined
+by the `IRanges` and `GenomicRanges` packages called _IntegerRanges_
+and _GenomicRanges_. Unless otherwise specified, throughout the paper
+we refer to either of these data structures as _Ranges_. In its simplest form 
+the data model for an _IntegerRanges_ object is defined as a table with
+$n$ rows, where each row contains a closed interval of two integers:
+
+$$
+\textit{IntegerRange} := \{ \left[\text{start}_i, \text{end}_{i}\right]  : i = 1\dots n \}.
+$$
+In this form an _IntegerRanges_ object has three core columns: a start, an
+end, and the width of the interval. The _GenomicRanges_ data
+model has additional semantics providing biological context, strand and 
+sequence name (seqname), as well as containing an _IntegerRanges_ object.
+
+$$
+\textit{GenomicRange} := \{  \{\text{seqname}_i, \left[\text{start}_i, \text{end}_{i}\right] , \text{strand}_i\}  : i = 1\dots n \}.
+$$
+A _GenomicRanges_ object can have an additional property that provides
+annotation for each sequence name over a genome. This contstrains the the 
+start and end points for a given sequence to lie inside the bounds
+of the genome space. Additional values can be bound to the bare _IntegerRanges_ or _GenomicRanges_ and form metadata columns.
+
+By definition the _Ranges_ class encapsulate tidy data prinicples because
+every row corresponds to a sequence and every column corresponds to a variable.
+Hence, we have designed the `plyranges` DSL to have the grammar
+and principles of `dplyr`: fluency, consistency, cohesion and endomorphism.
 

### 2018-03-22 paper.Rmd
reshuffling

@@
 Hence, we have designed the `plyranges` DSL to have the grammar
-and principles of `dplyr`: fluency, consistency, cohesion and endomorphism.
+and design principles of `dplyr`: fluency, consistency, cohesion and endomorphism.
 
-We have designed the API to be fluent. Every function call corresponds to an
-action on a _Ranges_ object (they are named verbs) and where possible
-functions have few arguments. Each verb is constructed
+To promote fluency we have designed every function call to perform an
+action on a _Ranges_ object and where possible
+functions have few arguments. Each function is named after a verb and is designed
 to enable a tab completion based workflow. Both of these aspects reduce the 
 cognitive load on a new user since most manipulations can be performed with a 
-vocabulary of several verbs, rather than having to memorise functions with many 
-arguments that are nouns (as is required in the existing Bioconductor packages). 
-This is also has the advantage of allowing users to write human readable code
-because verbs describe what the code is doing rather than how its doing it.
+vocabulary of several verbs, rather than having to memorise functions (that
+are named as nouns) with many arguments (which is required in the existing Bioconductor packages). Workflows can be composed by chaining verbs together into
+'sentences' via the forward pipe operator,`%>%` (exported from the R package `magrittr` @R-magrittr). Overall, this allows users to write human readable code
+because workflows describe what the code is doing rather than how its doing it.
 
-Workflows can be composed by chaining verbs together via the forward pipe operator,`%>%` 
-(exported from the R package _magrittr_ @R-magrittr). This is possible because every 
+In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
+use of non-standard evaluation in R via the _rlang_ package @R-rlang. Simply, this
+means that computations are performed and evaluated in the context of the 
+_Ranges_ objects, which emphasises the interactive nature of the API. This has
+the disadvantage that programming with _plyranges_ becomes more difficult 
+because a user needs to capture expressions inside function calls and then
+unquote them.
+
+ This is possible because every 
 function call is an endomorphism: when the input is _Ranges_ object the output
@@
 
-In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
-use of non-standard evaluation in R via the _rlang_ package @R-rlang. Simply, this
-means that computations are performed and evaluated in the context of the 
-_Ranges_ objects, which emphasises the interactive nature of the API. This has
-the disadvantage that programming with _plyranges_ becomes more difficult 
-because a user needs to capture expressions inside function calls and then
-unquote them.
 <!--- need to make this clearer? --->

### 2018-03-26 paper.Rmd
restructure of design section

@@
 start and end points for a given sequence to lie inside the bounds
-of the genome space. Additional values can be bound to the bare _IntegerRanges_ or _GenomicRanges_ and form metadata columns.
+of the genome space. Additional values can be bound to the bare 
+_IntegerRanges_ or _GenomicRanges_ to form metadata columns.
 
@@
 Hence, we have designed the `plyranges` DSL to have the grammar
-and design principles of `dplyr`: fluency, consistency, cohesion and endomorphism.
+and design principles of `dplyr`: fluency, consistency, cohesion and 
+endomorphism. All of these principles are discussed with respect to the 
+_Ranges_ data model below. Where possible we contrast our design to the existing
+Bioconductor infrastructure.
 
-To promote fluency we have designed every function call to perform an
-action on a _Ranges_ object and where possible
-functions have few arguments. Each function is named after a verb and is designed
-to enable a tab completion based workflow. Both of these aspects reduce the 
-cognitive load on a new user since most manipulations can be performed with a 
-vocabulary of several verbs, rather than having to memorise functions (that
-are named as nouns) with many arguments (which is required in the existing Bioconductor packages). Workflows can be composed by chaining verbs together into
-'sentences' via the forward pipe operator,`%>%` (exported from the R package `magrittr` @R-magrittr). Overall, this allows users to write human readable code
-because workflows describe what the code is doing rather than how its doing it.
-
-In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
-use of non-standard evaluation in R via the _rlang_ package @R-rlang. Simply, this
+A caveat to constructing a compatible API with `dplyr` is that 
+`plyranges` makes extensive use of non-standard evaluation in R
+(achieved via the `rlang` package @R-rlang). Simply, this
 means that computations are performed and evaluated in the context of the 
 _Ranges_ objects, which emphasises the interactive nature of the API. This has
-the disadvantage that programming with _plyranges_ becomes more difficult 
-because a user needs to capture expressions inside function calls and then
-unquote them.
+the disadvantage that programming with `plyranges` becomes more difficult 
+because a user needs to know about non-standard evaluation and user defined
+functions can be difficult to debug.
 
- This is possible because every 
-function call is an endomorphism: when the input is _Ranges_ object the output
-will also be a _Ranges_ object. One advantage of endomorphism is that it 
-does not  require any additional learning of classes beyond _Ranges_ and 
-the _DataFrame_ classes. This strongly deviates from the design of 
-the _Ranges_ Bioconductor packages, where many methods return a new class
-upon return. The Bioconductor design enables efficient computing as users
-are exposed to low-level features of its API which _plyranges_ abstracts
-away. Method chaining via the pipe operator can also be difficult to debug,
-as there multiple points of failure.
+## Fluency
 
-<!--- need to make this clearer? --->
+We have designed every function to perform a single
+action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the _dplyr_ package
+and implements context specific verbs for transforming _Ranges_ objects (described in
+table ). Each verb preserves the semantics of _Ranges_ object and via
+duck typing works with derivatives of the _Ranges_ class. 
+Both of these aspects reduce the cognitive load on a new user since most 
+manipulations can be performed with a  vocabulary of several verbs, rather 
+than having to memorise function names that are nouns. Workflows can be composed
+by chaining verbs together into 'sentences' via the forward pipe operator,`%>%` 
+(exported from the R package `magrittr` @R-magrittr), which can be read as the 
+word 'then'. Overall, this  allows users to write human readable code because 
+workflows describe what the code is doing rather than how its doing it.
 
-## Actions on Ranges
+```{r, echo = FALSE, message = FALSE, warning = FALSE}
+library(tibble)
+feature_tbl <- tribble(
+  ~Verb, ~Action, ~Description,
+  "`mutate()`", "modifies columns", "Takes a _Ranges_ object and a set of name-value pairs and generates a new _Ranges_ object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
+ "`filter()`", "subset rows", "Takes a _Ranges_ object and a set of logical expressions and restricts the _Ranges_ object to where all logical expressions are true.",
+ "`summarise()`", "aggregate one or more columns by a function", "Takes a _Ranges_ object and a set of name-value pairs and aggregates the _Ranges_ according to functions evaluated in the name-value pairs.",
+"`group_by()`", "partition a _Ranges_ object by a column", "Create an implicit grouping of _Ranges_ object by one or more columns.
+This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
+"`select()`", "select columns of a _Ranges_ objects", "Drop or reorder metadata columns.",
+"`arrange()`", "sort a _Ranges_ object by columns", "Sort a _Ranges_ object by named variables",
+"`disjoin_ranges()`", "aggregate one more columns of a _Ranges_ object over the union of end coordinates", "Takes a _Ranges_ object and a set of name-value pairs and expands the _Ranges_ object by taking the union of end coordinates.
+For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
+"`reduce_ranges()`", "aggregate one more columns of a _Ranges_ object by merging overlapping and neighbouring ranges.", "Takes a _Ranges_ object and a set of name-value pairs and reduces the _Ranges_ object by merging overlapping and neighouring ranges.
+For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
+)
+```
 
-The _plyranges_ API exports the six core verbs from the _dplyr_ package
-and modifies them for use with _Ranges_ objects. The
-verb `mutate()` takes a Ranges object and a set of name-value pairs
-and generates a new Ranges object that with modified or new metadata
-columns or modified core components (start, end, width, seqnames, strand).
-The use of `mutate()` means that a user no longer needs knowledge of the 
-accessors of the Ranges object, as they can modify them in place.
-The `filter()` function takes a Ranges object and logical expressions and
-restricts Ranges object to where the logical expression evaluates to true.
-The `summarise()` function takes a Ranges object and a set of name-value pairs
-and aggregates the Ranges according to functions evaluated in the name-value pairs.
-As `summarise()` is an aggregation it may break the structure of the of a Ranges
-object, hence it returns a _DataFrame_ object.
-The `select()` function determines
-which metadata columns are returned and the order they are returned in. The 
-`arrange()` function sorts a Ranges object by named variables. 
-The `group_by()` function creates an implicit grouping of Ranges object according
-to variables in the Ranges object. This modifies the actions of `mutate()`,
-`summarise()` and `filter()`, so they are performed on each partition created 
-by the grouping. The `group_by()` operation acts as a replacement for 
-the _GenomicRangesList_ class in the original _GenomicRanges_ API.
-
-The _plyranges_ API introduces additional summary verbs, `reduce_ranges()`
-and `disjoin_ranges()`that return Ranges objects after being returned. The
-`reduce_ranges()` operation merges overlapping and neighbour ranges, while
-`disjoin_ranges()` expands ranges by taking the union of end points.
-
-## Arithimetic on Ranges
+## Consistency
 
@@
 take into account the strand information of a _Ranges_ object.
+
+The methods `as_granges()` and `as_iranges()`
+for constructing _Ranges_ from tabular data structures, such as the _data.frame_ in
+base R. These methods use non-standard evaluation so columns in a _data.frame_
+can be modified before a _Ranges_ object is created. The API also has
+convenience methods for annotating or extracting annotations from _GRanges_
+objects with the `set_genome_info()` and `get_genome_info()` functions.
+
+There is a consistent framework for reading and writing 
+files from and to common genomic data formats, using the _rtracklayer_ package
+as a back-end @Lawrence2009-nt. The methods are implemented in the `read_/write_` family of 
+functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
+narrowPeaks, GFF/GTF, WIG and BigWig files.
+
+## Endomorphism
+
+ (that
+are named as nouns) with many arguments (which is required in the existing 
+Bioconductor packages)
+ This is possible because every 
+function call is an endomorphism: when the input is _Ranges_ object the output
+will also be a _Ranges_ object. One advantage of endomorphism is that it 
+does not  require any additional learning of classes beyond _Ranges_ and 
+the _DataFrame_ classes. This strongly deviates from the design of 
+the _Ranges_ Bioconductor packages, where many methods return a new class
+upon return. The Bioconductor design enables efficient computing as users
+are exposed to low-level features of its API which _plyranges_ abstracts
+away. Method chaining via the pipe operator can also be difficult to debug,
+as there multiple points of failure.
+
+The combination of verbs we have defined above cover most operations that
+can be performed in the original _IRanges_ and _GenomicRanges_ packages
+without the user being exposed to new classes. In those packages to perform
+most operations requires users to learn many classes and perform 
+additional manipulations to return the results of their computation back to a
+_Ranges_ object.
+
+## Cohesion
 
@@
 data structure in the _S4Vectors_ @R-S4vectors package or the BED-PE file format.
-
-The combination of verbs we have defined above encapsulate all operations that
-can be performed in the original _IRanges_ and _GenomicRanges_ packages
-without the user be exposed to new classes. In those packages to perform
-most operations requires users to learn many classes and perform 
-additional manipulations to return the results of their computation back to a
-_Ranges_ object.
-
-## Construction and Import/Output
-
-The methods `as_granges()` and `as_iranges()`
-for constructing _Ranges_ from tabular data structures, such as the _data.frame_ in
-base R. These methods use non-standard evaluation so columns in a _data.frame_
-can be modified before a _Ranges_ object is created. The API also has
-convenience methods for annotating or extracting annotations from _GRanges_
-objects with the `set_genome_info()` and `get_genome_info()` functions.
-
-There is a consistent framework for reading and writing 
-files from and to common genomic data formats, using the _rtracklayer_ package
-as a back-end @Lawrence2009-nt. The methods are implemented in the `read_/write_` family of 
-functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
-narrowPeaks, GFF/GTF, WIG and BigWig files.
 

### 2018-03-26 paper.Rmd
more words

@@
 start and end points for a given sequence to lie inside the bounds
-of the genome space. Additional values can be bound to the bare 
-_IntegerRanges_ or _GenomicRanges_ to form metadata columns.
+of the genome space. Also note that _GenomicRanges_ have one-based 
+coordinates rather than zero-based coordinates.  Additional measurements
+or annotations can be bound to the bare _IntegerRanges_ or _GenomicRanges_ to
+form metadata columns.
 
@@
 We have designed every function to perform a single
-action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the _dplyr_ package
+action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the `dplyr` package
 and implements context specific verbs for transforming _Ranges_ objects (described in
@@
 manipulations can be performed with a  vocabulary of several verbs, rather 
-than having to memorise function names that are nouns. Workflows can be composed
-by chaining verbs together into 'sentences' via the forward pipe operator,`%>%` 
-(exported from the R package `magrittr` @R-magrittr), which can be read as the 
-word 'then'. Overall, this  allows users to write human readable code because 
-workflows describe what the code is doing rather than how its doing it.
+than having to memorise function names that are nouns. Moreover, this approach 
+strongly constrasts the `GenomicRanges/IRanges` OO interface, which emphasises 
+the use of setter and getter methods. In that interface, core components and 
+metadata are updated via replacement methods (requiring knowledge of the
+class components), while our interface requires only a single call to `mutate()`
+to perform the modification.
+
+Workflows can be composed by chaining verbs together into 'sentences' via the 
+forward pipe operator,`%>%`  (exported from the R package `magrittr` 
+@R-magrittr), which can be read as the word 'then'. Overall, this  allows users 
+to write human readable code because workflows describe what the code is doing 
+rather than how its doing it.
 

### 2018-03-26 paper.Rmd
consistency of naming

@@
 via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
-We aim to continue developing the _plyranges_ package and extend it for use
+We aim to continue developing the `plyranges` package and extend it for use
 with more complex data structures  such as the 
 _SummarizedExperiment_ class, which can be used for analysing transcriptomic
-and variant data. As the _plyranges_ interface encourages tidy data 
+and variant data. As the `plyranges` interface encourages tidy data 
 practices it integrates well with the principles of the grammar of graphics, 
-we aim to use it for the visualisation of multimodal biological datasets.
+we aim to use it to prepare data for the visualisation of multimodal biological 
+datasets.
 

### 2018-03-27 paper.Rmd

@@
 
-There is a large variety of genomics data available to the scientific community,
-ranging from highly specific sequencing assays to publicly availble collections.
-These datasets similarly contain a wide variety of features: annotations of 
-genomic features such as genes or exons for reference genomes, large coverage 
-vectors, peaks called from ChIP-Seq experiments and many more. A modern biologist
-is expected to be able to reason about, combine and wrangle these heterogenous
-data sources in order to answer questions about their experimental hypotheses.
-Abstractions in the forms of data representation and software interfaces are
-key factors in inhibiting or enabling a biologist to construct reproducible data workflows
-to answer their research questions.
+High-throughput genomics promises to unlock new disease therapies,
+reveal basic biology, and solve other problems that plague the human
+condition. To deliver on those promises, scientists must derive a
+stream of knowledge from a deluge of data. Genomic data is challenging
+in both scale and complexity. Innovations in sequencing technology
+often outstrip our capacity to process the output. Beyond their common
+association with genomic coordinates, genomic data are heterogeneous,
+consisting of raw sequence read alignments, genomic feature
+annotations like genes and exons, and summaries like coverage vectors,
+ChIP-seq peak calls, variant calls, and per-feature read
+counts. Genomic scientists need software tools to wrangle the
+different types of data, process the data at scale, test hypotheses,
+and generate new ones, all while focusing on the biology, not the
+computation. For the tool developer, the challenge is to define ways
+to model and operate on the data that align with the mental model of
+scientists, and to provide an implementation that scales with their
+ambition.
 
-Many domain specific languages (DSLs) have been created to process and reason
-about heterogenous genomics data. Broadly, they are implemented as either
-query languages or as an embedded command line tool.
-An example of the former is the Genome Query Language (GQL) and its distributed 
-implementation GenAp which use an SQL-like syntax for fast retrieval of 
-information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. 
+Several domain specific languages (DSLs) enable scientists to process
+and reason about heterogenous genomics data by expresssing common
+operations, such as range manipulation and overlap-based joins, using
+the vocabulary of genomics. Their implementations either delegate
+computations to a database, or operate over collections of files in
+standard formats like BED.  An example of the former is the Genome
+Query Language (GQL) and its distributed implementation GenAp which
+use an SQL-like syntax for fast retrieval of information of
+unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm.
 Similarly, the Genometric Query Language (GMQL) implements a
-relational algebra for combining genomic datasets @Kaitoua2017-pw.  
-The command line application BEDtools develops an extensive algebra for 
+relational algebra for combining genomic datasets @Kaitoua2017-pw. The
+command line application BEDtools defines an extensive algebra for
 performing arithmetic between two or more sets of genomic regions
-@Quinlan2010-gc; @Dale2011-js. All of the aformentioned DSLs are designed to 
-be run either at the command line or embedded in scripts for batch processing. 
-As a consquence, they are not integrated into an environment designed to 
-perform interactive data  analysis that has access to state of the art 
-methods for statistics and visualisation like the R language.
+stored as files @Quinlan2010-gc; @Dale2011-js. All of the
+aformentioned DSLs are designed to be evaluated either at the command
+line or embedded in scripts for batch processing. They exist in a
+sparse ecosystem, mostly consisting of UNIX and database tools that
+lack biological semantics and operate at the level of files and
+database tables.
 
-The Bioconductor/R packages `IRanges` and `GenomicRanges` 
-define a DSL for analysing genomics data with R via
-an object-oriented (OO) interface @Lawrence2013-wg; @Huber2015-ei. This interface
-provides abstractions for representing genomics data and enables interoperability
-by allowing users and developers to use these abstractions in their own code
-and packages. However, due to the large number of abstractions there are costs
-involved. Due to the OO interface each abstraction will behave slightly 
-differently for any given generic method. There is a cognitive load on the user
-to understand the appropriate context for each abstraction. An alternative approach
-is to recognise that most genomic data analysis tasks can be performed with 
-data structures that naturally represent genomic intervals. That is, genomics
-data can be represented as sets of integer pairs, where each pair has a start
-and an end coordianate, along with strings representing strand and chromosome 
-location. Measurements or metadata can be included as additional annotations
-on each pair.  As an example. a gene can be represented with its coordinates, along with its 
-identifier and the identifiers of its exons; or an RNA-seq experiment may be
-represented as sets of genes with a matching count column.
+The Bioconductor packages `IRanges` and `GenomicRanges`
+(@Lawrence2013-wg; @Huber2015-ei) define a DSL for analysing genomics
+data with R, an interactive data analysis environment that encourages
+reproducibility and provides high-level abstractions for manipulating,
+modeling and plotting data, through state of the art methods in
+statistical computing. The packages define object-oriented (OO)
+abstractions for representing genomic data and enable
+interoperability by allowing users and developers to use these
+abstractions in their own code and packages. The Bioconductor
+infrastructure models the genomic data and operations from the
+perspective of the power user, one who understands and wants to take
+advantage of the subtle differences in data types. This design has
+enabled the development of sophisticated tools, as evidenced by the
+hundreds of packages depending the framework. Unfortunately, the
+myriad of data structures have overlapping purposes and important but
+obscure differences in behavior that often confuse the typical end
+user.
 
-This natural abstraction aligns with the concept of tidy data @Wickham2014-jc. 
-A tidy dataset is defined to be to a tabular
-data structure that has observations as rows and columns as variables, and each
-tidy dataset represents measurements from a single observational unit.
-The tidy data pattern is useful because it allows 
-us to see how the data relates to the design of an experiment
-and the variables measured. The genomic interval data structure follows 
-this pattern: it is a rectangular table corresponding to a single biological 
-context. Each row contains a single observation and each column is a variable about
-that observation.
+Recently, there has been a concerted, community effort to standardize
+R data structures and workflows around the notion of tidy data
+@Wickham2014-jc. A tidy dataset is defined to have a tabular structure
+that has observations as rows and columns as variables. Each tidy
+dataset represents measurements from a single observational unit.  The
+tidy data pattern is useful because it allows us to see how the data
+relates to the design of an experiment and the variables measured.
+The `dplyr` pacakge [@Wickham2017-dplyr] defines an API that maps
+notions from the general relational algebra to operations on tidy
+data. It expresses each operation as a cohesive, endomorphic
+verb. Taken together these features enable a user to write human
+readable analysis workflows.
 
-The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
-fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
-The package defines a consistent syntax and style for transforming tabular data sets via
-functions that are named as verbs. Each function performs an action on the data
-set and is cohesive and endomorphic. Cohesion means that each function performs
-a singular task and has no side-effects, while endomorphism means a function
-call returns the same type of object as its input. Taken together these
-features enable a user to write human readable analysis workflows.
+We have created a genomic DSL called `plyranges` that reformulates
+notions from existing genomic algebras and embeds them in R as a
+genomic extension of `dplyr`. By analogy, `plyranges` is to the
+genomic algebra, as `dplyr` is to the relational algebra.  The
+`plyranges` Bioconductor package implements the language on top of a
+key subset of Bioconductor data structures and thus fully integrates
+with the Bioconductor framework, gaining access to its scalable data
+representations and sophisticated statistical methods.
 
-We have created a DSL called `plyranges` that is directly motivated by the design
-of `dplyr` but extends its  grammar to incorporate features of genomics data.
-Our interface uses the existing Bioconductor packages `IRanges` and 
-`GenomicRanges` but simplifies them by only exposing tidy genomic interval
-data structures to the user. Hence, we can encapsulate the features of the 
-of the `dplyr` API while performing genomic specific transformations 
-such as finding overlapping  regions or nearest neighbour regions between 
-genomic intervals.  The `plyranges` DSL is designed to enable fast interactive 
-analysis of genomics but also enables the creation of reproducible genomic 
-data workflows.
+# Stuff pushed out of the intro (needs a home)
+
+The most general Bioconductor data structure, GRanges, is capable of
+representing all types of genomic data at a semantic level that
+roughly matches the intuition of most users.  A GRanges is essentially
+a table, with columns for the chromosome, stand and end coordinates,
+and the strand, along with an arbitrary set of additional columns,
+consisting of measurements or metadata specific to the data type or
+experiment.  For example, a gene can be represented by its chromosomal
+coordinates, along with its identifier and the identifiers of its
+exons; or an RNA-seq experiment may be represented as sets of genes
+with a matching count column.
+
+GRanges follows the tidy data pattern: it is a rectangular table
+corresponding to a single biological context. Each row contains a
+single observation and each column is a variable about that
+observation.
+
+Cohesion means that each function performs a singular task and has no
+side-effects, while endomorphism means a function call returns the
+same type of object as its input.
 

### 2018-04-09 paper.Rmd
shuffling the intro around, adding some of Michael's changes.

@@
 High-throughput genomics promises to unlock new disease therapies,
-reveal basic biology, and solve other problems that plague the human
-condition. To deliver on those promises, scientists must derive a
-stream of knowledge from a deluge of data. Genomic data is challenging
-in both scale and complexity. Innovations in sequencing technology
-often outstrip our capacity to process the output. Beyond their common
+and strengthen our knowledge of basic biology. To deliver on those promises, 
+scientists must derive a stream of knowledge from a deluge of data. Genomic 
+data is challenging in both scale and complexity. Innovations in sequencing 
+technology often outstrip our capacity to process the output. Beyond their common
 association with genomic coordinates, genomic data are heterogeneous,
@@
 
-Several domain specific languages (DSLs) enable scientists to process
-and reason about heterogenous genomics data by expresssing common
+Several domain specific languages (DSLs) enable scientists to process and reason
+about heterogenous genomics data heterogenous genomics data by expresssing common
 operations, such as range manipulation and overlap-based joins, using
@@
 computations to a database, or operate over collections of files in
-standard formats like BED.  An example of the former is the Genome
-Query Language (GQL) and its distributed implementation GenAp which
-use an SQL-like syntax for fast retrieval of information of
-unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm.
+standard formats like BED.
+An example of the former is the Genome Query Language (GQL) and its distributed 
+implementation GenAp which use an SQL-like syntax for fast retrieval of 
+information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. 
 Similarly, the Genometric Query Language (GMQL) implements a 
-relational algebra for combining genomic datasets @Kaitoua2017-pw. The
-command line application BEDtools defines an extensive algebra for
+relational algebra for combining genomic datasets @Kaitoua2017-pw.  
+The command line application BEDtools develops an extensive algebra for 
 performing arithmetic between two or more sets of genomic regions 
-stored as files @Quinlan2010-gc; @Dale2011-js. All of the
+@Quinlan2010-gc; @Dale2011-js. All of the
 aformentioned DSLs are designed to be evaluated either at the command
@@
 
-The Bioconductor packages `IRanges` and `GenomicRanges`
-(@Lawrence2013-wg; @Huber2015-ei) define a DSL for analysing genomics
-data with R, an interactive data analysis environment that encourages
+The Bioconductor/R packages `IRanges` and `GenomicRanges` 
+[@Lawrence2013-wg; @Huber2015-ei] define a DSL for analysing genomics data with 
+R, an interactive data analysis environment that encourages
 reproducibility and provides high-level abstractions for manipulating,
@@
 R data structures and workflows around the notion of tidy data
-@Wickham2014-jc. A tidy dataset is defined to have a tabular structure
-that has observations as rows and columns as variables. Each tidy
-dataset represents measurements from a single observational unit.  The
-tidy data pattern is useful because it allows us to see how the data
-relates to the design of an experiment and the variables measured.
-The `dplyr` pacakge [@Wickham2017-dplyr] defines an API that maps
-notions from the general relational algebra to operations on tidy
-data. It expresses each operation as a cohesive, endomorphic
-verb. Taken together these features enable a user to write human
-readable analysis workflows.
+@Wickham2014-jc. A tidy dataset is defined to be to a tabular
+data structure that has observations as rows and columns as variables, and each
+tidy dataset represents measurements from a single observational unit.
+The tidy data pattern is useful because it allows 
+us to see how the data relates to the design of an experiment
+and the variables measured. The genomic interval data structure follows 
+this pattern: it is a rectangular table corresponding to a single biological 
+context. Each row contains a single observation and each column is a variable about
+that observation.
 
@@
 
-# Stuff pushed out of the intro (needs a home)
+# Design and Implementation
 
-The most general Bioconductor data structure, GRanges, is capable of
-representing all types of genomic data at a semantic level that
-roughly matches the intuition of most users.  A GRanges is essentially
-a table, with columns for the chromosome, stand and end coordinates,
-and the strand, along with an arbitrary set of additional columns,
-consisting of measurements or metadata specific to the data type or
-experiment.  For example, a gene can be represented by its chromosomal
-coordinates, along with its identifier and the identifiers of its
-exons; or an RNA-seq experiment may be represented as sets of genes
-with a matching count column.
+via
+ This interface
+provides abstractions for representing genomics data and enables interoperability
+by allowing users and developers to use these abstractions in their own code
+and packages. However, due to the large number of abstractions there are costs
+involved. Due to the OO interface each abstraction will behave slightly 
+differently for any given generic method. There is a cognitive load on the user
+to understand the appropriate context for each abstraction. An alternative approach
+is to recognise that most genomic data analysis tasks can be performed with 
+data structures that naturally represent genomic intervals. That is, genomics
+data can be represented as sets of integer pairs, where each pair has a start
+and an end coordianate, along with strings representing strand and chromosome 
+location. Measurements or metadata can be included as additional annotations
+on each pair.  As an example. a gene can be represented with its coordinates, along with its 
+identifier and the identifiers of its exons; or an RNA-seq experiment may be
+represented as sets of genes with a matching count column.
 
-GRanges follows the tidy data pattern: it is a rectangular table
-corresponding to a single biological context. Each row contains a
-single observation and each column is a variable about that
-observation.
-
-Cohesion means that each function performs a singular task and has no
-side-effects, while endomorphism means a function call returns the
-same type of object as its input.
-
-# Design and Implementation
+The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
+fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
+The package defines a consistent syntax and style for transforming tabular data sets via
+functions that are named as verbs. Each function performs an action on the data
+set and is cohesive and endomorphic. Cohesion means that each function performs
+a singular task and has no side-effects, while endomorphism means a function
+call returns the same type of object as its input. Taken together these
+features enable a user to write human readable analysis workflows.
 

### 2018-04-09 paper.Rmd
more rearranging

@@
 # Design and Implementation
+
+The most general Bioconductor data structure, GRanges, is capable of
+representing all types of genomic data at a semantic level that
+roughly matches the intuition of most users.  A GRanges is essentially
+a table, with columns for the chromosome, stand and end coordinates,
+and the strand, along with an arbitrary set of additional columns,
+consisting of measurements or metadata specific to the data type or
+experiment.  For example, a gene can be represented by its chromosomal
+coordinates, along with its identifier and the identifiers of its
+exons; or an RNA-seq experiment may be represented as sets of genes
+with a matching count column.
+
+GRanges follows the tidy data pattern: it is a rectangular table
+corresponding to a single biological context. Each row contains a
+single observation and each column is a variable about that
+observation.
 

### 2018-04-09 paper.Rmd
add in valr citation closes #8

@@
 computations to a database, or operate over collections of files in
-standard formats like BED.
-An example of the former is the Genome Query Language (GQL) and its distributed 
-implementation GenAp which use an SQL-like syntax for fast retrieval of 
-information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. 
-Similarly, the Genometric Query Language (GMQL) implements a 
-relational algebra for combining genomic datasets @Kaitoua2017-pw.  
-The command line application BEDtools develops an extensive algebra for 
-performing arithmetic between two or more sets of genomic regions 
-@Quinlan2010-gc; @Dale2011-js. All of the
-aformentioned DSLs are designed to be evaluated either at the command
-line or embedded in scripts for batch processing. They exist in a
-sparse ecosystem, mostly consisting of UNIX and database tools that
+standard formats like BED.An example of the former is the Genome Query 
+Language (GQL) and its distributed implementation GenAp which use an 
+SQL-like syntax for fast retrieval of information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric Query Language 
+(GMQL) implements a relational algebra for combining genomic datasets 
+@Kaitoua2017-pw.  The command line application BEDtools develops an extensive algebra for performing arithmetic between two or more sets of genomic regions 
+@Quinlan2010-gc. All of the aformentioned DSLs are designed to be evaluated
+either at the commandline or embedded in scripts for batch processing. They 
+exist in asparse ecosystem, mostly consisting of UNIX and database tools that
 lack biological semantics and operate at the level of files and
@@
 The Bioconductor/R packages `IRanges` and `GenomicRanges` 
-[@Lawrence2013-wg; @Huber2015-ei] define a DSL for analysing genomics data with 
-R, an interactive data analysis environment that encourages
-reproducibility and provides high-level abstractions for manipulating,
+[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
+analysing genomics data with R, an interactive data analysis environment that 
+encouragesreproducibility and provides high-level abstractions for manipulating,
 modeling and plotting data, through state of the art methods in
@@
 interoperability by allowing users and developers to use these
-abstractions in their own code and packages. The Bioconductor
-infrastructure models the genomic data and operations from the
+abstractions in their own code and packages. Other genomic DSLs 
+that are embedded in programming languages include pybedtools and valr 
+[@Dale2011-js; @Kent2017], however these packages lack the interoperability
+provided by in the aformentioned Bioconductor packages and are not easily 
+extended.
+
+The Bioconductor infrastructure models the genomic data and operations from the
 perspective of the power user, one who understands and wants to take
@@
 us to see how the data relates to the design of an experiment
-and the variables measured. The genomic interval data structure follows 
-this pattern: it is a rectangular table corresponding to a single biological 
-context. Each row contains a single observation and each column is a variable about
-that observation.
+and the variables measured. The `dplyr` pacakge [@Wickham2017-dplyr] 
+defines an API that maps notions from the general relational algebra to 
+operations on tidy data. It expresses each operation as a cohesive, endomorphic
+verb. Taken together these features enable a user to write human
+readable analysis workflows.
 

### 2018-04-13 paper.Rmd
reordering design section

@@
 Language (GQL) and its distributed implementation GenAp which use an 
-SQL-like syntax for fast retrieval of information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric Query Language 
-(GMQL) implements a relational algebra for combining genomic datasets 
-@Kaitoua2017-pw.  The command line application BEDtools develops an extensive algebra for performing arithmetic between two or more sets of genomic regions 
+SQL-like syntax for fast retrieval of information of unprocessed 
+sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric 
+Query Language (GMQL) implements a relational algebra for combining genomic datasets 
+@Kaitoua2017-pw.  The command line application BEDtools develops an extensive 
+algebra for performing arithmetic between two or more sets of genomic regions 
 @Quinlan2010-gc. All of the aformentioned DSLs are designed to be evaluated
@@
 
-The most general Bioconductor data structure, GRanges, is capable of
+The `plyranges` DSL is built on the most general Bioconductor data structure, 
+GRanges, which is capable of
 representing all types of genomic data at a semantic level that
 roughly matches the intuition of most users.  A GRanges is essentially
-a table, with columns for the chromosome, stand and end coordinates,
+a table, with columns for the chromosome, start and end coordinates,
 and the strand, along with an arbitrary set of additional columns,
 consisting of measurements or metadata specific to the data type or
-experiment.  For example, a gene can be represented by its chromosomal
-coordinates, along with its identifier and the identifiers of its
-exons; or an RNA-seq experiment may be represented as sets of genes
-with a matching count column.
+experiment. For example a GRanges can be used to represent a gene with
+its chromosomal coordinates, and along with its exon identifiers; or 
+a GRanges can represent an RNA-seq experiment as a set of genes
+with a corresponding count column.
 
-GRanges follows the tidy data pattern: it is a rectangular table
-corresponding to a single biological context. Each row contains a
+By definition the GRanges follow  the tidy data pattern: it is a rectangular 
+table corresponding to a single biological context. Each row contains a
 single observation and each column is a variable about that
-observation.
-
-via
- This interface
-provides abstractions for representing genomics data and enables interoperability
-by allowing users and developers to use these abstractions in their own code
-and packages. However, due to the large number of abstractions there are costs
-involved. Due to the OO interface each abstraction will behave slightly 
-differently for any given generic method. There is a cognitive load on the user
-to understand the appropriate context for each abstraction. An alternative approach
-is to recognise that most genomic data analysis tasks can be performed with 
-data structures that naturally represent genomic intervals. That is, genomics
-data can be represented as sets of integer pairs, where each pair has a start
-and an end coordianate, along with strings representing strand and chromosome 
-location. Measurements or metadata can be included as additional annotations
-on each pair.  As an example. a gene can be represented with its coordinates, along with its 
-identifier and the identifiers of its exons; or an RNA-seq experiment may be
-represented as sets of genes with a matching count column.
-
-The tidy data abstraction enables the construction of fluent application programming interfaces (APIs) [@Fowler2010-zd] that are optimised for readability. An example of a 
-fluent interface built on the tidy data pattern is the R pacakge `dplyr` [@Wickham2017-dplyr]. 
-The package defines a consistent syntax and style for transforming tabular data sets via
-functions that are named as verbs. Each function performs an action on the data
-set and is cohesive and endomorphic. Cohesion means that each function performs
-a singular task and has no side-effects, while endomorphism means a function
-call returns the same type of object as its input. Taken together these
-features enable a user to write human readable analysis workflows.
-
-The `plyranges` DSL is built on two fundamental data structures defined
-by the `IRanges` and `GenomicRanges` packages called _IntegerRanges_
-and _GenomicRanges_. Unless otherwise specified, throughout the paper
-we refer to either of these data structures as _Ranges_. In its simplest form 
-the data model for an _IntegerRanges_ object is defined as a table with
-$n$ rows, where each row contains a closed interval of two integers:
-
-$$
-\textit{IntegerRange} := \{ \left[\text{start}_i, \text{end}_{i}\right]  : i = 1\dots n \}.
-$$
-In this form an _IntegerRanges_ object has three core columns: a start, an
-end, and the width of the interval. The _GenomicRanges_ data
-model has additional semantics providing biological context, strand and 
-sequence name (seqname), as well as containing an _IntegerRanges_ object.
-
-$$
-\textit{GenomicRange} := \{  \{\text{seqname}_i, \left[\text{start}_i, \text{end}_{i}\right] , \text{strand}_i\}  : i = 1\dots n \}.
-$$
-A _GenomicRanges_ object can have an additional property that provides
-annotation for each sequence name over a genome. This contstrains the the 
-start and end points for a given sequence to lie inside the bounds
-of the genome space. Also note that _GenomicRanges_ have one-based 
-coordinates rather than zero-based coordinates.  Additional measurements
-or annotations can be bound to the bare _IntegerRanges_ or _GenomicRanges_ to
-form metadata columns.
-
-By definition the _Ranges_ class encapsulate tidy data prinicples because
-every row corresponds to a sequence and every column corresponds to a variable.
-Hence, we have designed the `plyranges` DSL to have the grammar
-and design principles of `dplyr`: fluency, consistency, cohesion and 
-endomorphism. All of these principles are discussed with respect to the 
-_Ranges_ data model below. Where possible we contrast our design to the existing
-Bioconductor infrastructure.
-
-A caveat to constructing a compatible API with `dplyr` is that 
-`plyranges` makes extensive use of non-standard evaluation in R
-(achieved via the `rlang` package @R-rlang). Simply, this
-means that computations are performed and evaluated in the context of the 
-_Ranges_ objects, which emphasises the interactive nature of the API. This has
-the disadvantage that programming with `plyranges` becomes more difficult 
-because a user needs to know about non-standard evaluation and user defined
-functions can be difficult to debug.
-
-## Fluency
-
-We have designed every function to perform a single
-action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the `dplyr` package
-and implements context specific verbs for transforming _Ranges_ objects (described in
-table ). Each verb preserves the semantics of _Ranges_ object and via
-duck typing works with derivatives of the _Ranges_ class. 
-Both of these aspects reduce the cognitive load on a new user since most 
-manipulations can be performed with a  vocabulary of several verbs, rather 
-than having to memorise function names that are nouns. Moreover, this approach 
-strongly constrasts the `GenomicRanges/IRanges` OO interface, which emphasises 
-the use of setter and getter methods. In that interface, core components and 
-metadata are updated via replacement methods (requiring knowledge of the
-class components), while our interface requires only a single call to `mutate()`
-to perform the modification.
-
-Workflows can be composed by chaining verbs together into 'sentences' via the 
-forward pipe operator,`%>%`  (exported from the R package `magrittr` 
-@R-magrittr), which can be read as the word 'then'. Overall, this  allows users 
-to write human readable code because workflows describe what the code is doing 
-rather than how its doing it.
-
-```{r, echo = FALSE, message = FALSE, warning = FALSE}
-library(tibble)
-feature_tbl <- tribble(
-  ~Verb, ~Action, ~Description,
-  "`mutate()`", "modifies columns", "Takes a _Ranges_ object and a set of name-value pairs and generates a new _Ranges_ object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
- "`filter()`", "subset rows", "Takes a _Ranges_ object and a set of logical expressions and restricts the _Ranges_ object to where all logical expressions are true.",
- "`summarise()`", "aggregate one or more columns by a function", "Takes a _Ranges_ object and a set of name-value pairs and aggregates the _Ranges_ according to functions evaluated in the name-value pairs.",
-"`group_by()`", "partition a _Ranges_ object by a column", "Create an implicit grouping of _Ranges_ object by one or more columns.
-This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
-"`select()`", "select columns of a _Ranges_ objects", "Drop or reorder metadata columns.",
-"`arrange()`", "sort a _Ranges_ object by columns", "Sort a _Ranges_ object by named variables",
-"`disjoin_ranges()`", "aggregate one more columns of a _Ranges_ object over the union of end coordinates", "Takes a _Ranges_ object and a set of name-value pairs and expands the _Ranges_ object by taking the union of end coordinates.
-For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
-"`reduce_ranges()`", "aggregate one more columns of a _Ranges_ object by merging overlapping and neighbouring ranges.", "Takes a _Ranges_ object and a set of name-value pairs and reduces the _Ranges_ object by merging overlapping and neighouring ranges.
-For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
-)
-```
-
-## Consistency
-
-The _plyranges_ API has an expressive algebra for performing arithmetic on Ranges via
-the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
-modifies the width of a Ranges object, while `stretch` extends the start and
-end of a Ranges object. These can be chained with the anchoring functions
-`anchor_start()`, `anchor_end()`, `anchor_center()`, `anchor_3p()` or 
-`anchor_5p()`, which fix the coordinates of a Ranges object in place. 
-Moreover, the `shift_` and `flank_` family of functions can be used to shift
-all coordinates in a Ranges object or generate flanking regions from a Ranges 
-object to the left, right, upstream or downstream of the input. Unlike, the
-Bioconductor API, _plyranges_ makes it explicit via function calls whether to
-take into account the strand information of a _Ranges_ object.
-
-The methods `as_granges()` and `as_iranges()`
-for constructing _Ranges_ from tabular data structures, such as the _data.frame_ in
-base R. These methods use non-standard evaluation so columns in a _data.frame_
-can be modified before a _Ranges_ object is created. The API also has
-convenience methods for annotating or extracting annotations from _GRanges_
-objects with the `set_genome_info()` and `get_genome_info()` functions.
-
-There is a consistent framework for reading and writing 
-files from and to common genomic data formats, using the _rtracklayer_ package
-as a back-end @Lawrence2009-nt. The methods are implemented in the `read_/write_` family of 
-functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
-narrowPeaks, GFF/GTF, WIG and BigWig files.
-
-## Endomorphism
-
- (that
-are named as nouns) with many arguments (which is required in the existing 
-Bioconductor packages)
- This is possible because every 
-function call is an endomorphism: when the input is _Ranges_ object the output
-will also be a _Ranges_ object. One advantage of endomorphism is that it 
-does not  require any additional learning of classes beyond _Ranges_ and 
-the _DataFrame_ classes. This strongly deviates from the design of 
-the _Ranges_ Bioconductor packages, where many methods return a new class
-upon return. The Bioconductor design enables efficient computing as users
-are exposed to low-level features of its API which _plyranges_ abstracts
-away. Method chaining via the pipe operator can also be difficult to debug,
-as there multiple points of failure.
-
-The combination of verbs we have defined above cover most operations that
-can be performed in the original _IRanges_ and _GenomicRanges_ packages
-without the user being exposed to new classes. In those packages to perform
-most operations requires users to learn many classes and perform 
-additional manipulations to return the results of their computation back to a
-_Ranges_ object.
+observation. Hence, we have designed the `plyranges` DSL to extend the grammar
+and design principles of `dplyr`:  cohesion, consistency, endomorphism,
+and fluency. All of these principles are defined and discussed
+below in the context of the GRanges class. Where applicable we contrast our 
+design to the existing Bioconductor infrastructure.
 
@@
 
-## Overlapping Ranges
+A function is cohesive if it performs a singular task and does not produce
+any side-effects. In the `plyranges` DSL our algebra for performing
+genomic arithmetic is a key example of cohesion. We define anchoring operators
+that decorate a GRanges object with an 'anchor' and modify the semantics
+of performing genomic arithmetic. The anchoring operators amount to fixing a GRanges
+object by its start, center, or end coordinates (or fixing these coordinates
+by strand). This enables any arithmetic function that performs a 
+coordinate transformation to remain cohesive: that is they always perfom
+the same operation regardless of whether a GRanges object is anchored or not.
+The only difference is the result of performing the arithimetic changes when
+contextual information is added to the GRanges object.
+
+Another example of altering object semantics, while maintaining cohesion is
+through the 'group_by' operator. Like anchoring, this operator decorates
+a GRanges object with a column name (or names) that imply a partitioning of the 
+GRanges by the unique values in the column(s). Functions defined in the 
+`plyranges` DSL that perform restriction, aggregation, or column modfication
+still perform those singular tasks, however grouping changes how those tasks
+are performed.
 
@@
 
+## Consistency
+
+The _plyranges_ API has an expressive algebra for performing arithmetic on Ranges via
+the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
+modifies the width of a Ranges object, while `stretch` extends the start and
+end of a Ranges object. These can be chained with the anchoring functions
+`anchor_start()`, `anchor_end()`, `anchor_center()`, `anchor_3p()` or 
+`anchor_5p()`, which fix the coordinates of a Ranges object in place. 
+Moreover, the `shift_` and `flank_` family of functions can be used to shift
+all coordinates in a Ranges object or generate flanking regions from a Ranges 
+object to the left, right, upstream or downstream of the input. Unlike, the
+Bioconductor API, _plyranges_ makes it explicit via function calls whether to
+take into account the strand information of a _Ranges_ object.
+
+The methods `as_granges()` and `as_iranges()`
+for constructing _Ranges_ from tabular data structures, such as the _data.frame_ in
+base R. These methods use non-standard evaluation so columns in a _data.frame_
+can be modified before a _Ranges_ object is created. The API also has
+convenience methods for annotating or extracting annotations from _GRanges_
+objects with the `set_genome_info()` and `get_genome_info()` functions.
+
+There is a consistent framework for reading and writing 
+files from and to common genomic data formats, using the _rtracklayer_ package
+as a back-end @Lawrence2009-nt. The methods are implemented in the `read_/write_` family of 
+functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
+narrowPeaks, GFF/GTF, WIG and BigWig files.
+
+## Fluency
+
+We have designed every function to perform a single
+action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the `dplyr` package
+and implements context specific verbs for transforming _Ranges_ objects (described in
+table ). Each verb preserves the semantics of _Ranges_ object and via
+duck typing works with derivatives of the _Ranges_ class. 
+Both of these aspects reduce the cognitive load on a new user since most 
+manipulations can be performed with a  vocabulary of several verbs, rather 
+than having to memorise function names that are nouns. Moreover, this approach 
+strongly constrasts the `GenomicRanges/IRanges` OO interface, which emphasises 
+the use of setter and getter methods. In that interface, core components and 
+metadata are updated via replacement methods (requiring knowledge of the
+class components), while our interface requires only a single call to `mutate()`
+to perform the modification.
+
+Workflows can be composed by chaining verbs together into 'sentences' via the 
+forward pipe operator,`%>%`  (exported from the R package `magrittr` 
+@R-magrittr), which can be read as the word 'then'. Overall, this  allows users 
+to write human readable code because workflows describe what the code is doing 
+rather than how its doing it.
+
+```{r, echo = FALSE, message = FALSE, warning = FALSE}
+library(tibble)
+feature_tbl <- tribble(
+  ~Verb, ~Action, ~Description,
+  "`mutate()`", "modifies columns", "Takes a _Ranges_ object and a set of name-value pairs and generates a new _Ranges_ object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
+ "`filter()`", "subset rows", "Takes a _Ranges_ object and a set of logical expressions and restricts the _Ranges_ object to where all logical expressions are true.",
+ "`summarise()`", "aggregate one or more columns by a function", "Takes a _Ranges_ object and a set of name-value pairs and aggregates the _Ranges_ according to functions evaluated in the name-value pairs.",
+"`group_by()`", "partition a _Ranges_ object by a column", "Create an implicit grouping of _Ranges_ object by one or more columns.
+This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
+"`select()`", "select columns of a _Ranges_ objects", "Drop or reorder metadata columns.",
+"`arrange()`", "sort a _Ranges_ object by columns", "Sort a _Ranges_ object by named variables",
+"`disjoin_ranges()`", "aggregate one more columns of a _Ranges_ object over the union of end coordinates", "Takes a _Ranges_ object and a set of name-value pairs and expands the _Ranges_ object by taking the union of end coordinates.
+For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
+"`reduce_ranges()`", "aggregate one more columns of a _Ranges_ object by merging overlapping and neighbouring ranges.", "Takes a _Ranges_ object and a set of name-value pairs and reduces the _Ranges_ object by merging overlapping and neighouring ranges.
+For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
+)
+```
+
+## Endomorphism
+
+ (that
+are named as nouns) with many arguments (which is required in the existing 
+Bioconductor packages)
+ This is possible because every 
+function call is an endomorphism: when the input is _Ranges_ object the output
+will also be a _Ranges_ object. One advantage of endomorphism is that it 
+does not  require any additional learning of classes beyond _Ranges_ and 
+the _DataFrame_ classes. This strongly deviates from the design of 
+the _Ranges_ Bioconductor packages, where many methods return a new class
+upon return. The Bioconductor design enables efficient computing as users
+are exposed to low-level features of its API which _plyranges_ abstracts
+away. Method chaining via the pipe operator can also be difficult to debug,
+as there multiple points of failure.
+
+The combination of verbs we have defined above cover most operations that
+can be performed in the original _IRanges_ and _GenomicRanges_ packages
+without the user being exposed to new classes. In those packages to perform
+most operations requires users to learn many classes and perform 
+additional manipulations to return the results of their computation back to a
+_Ranges_ object.
+
+## Limitations
+
+A caveat to constructing a compatible API with `dplyr` is that 
+`plyranges` makes extensive use of non-standard evaluation in R
+(achieved via the `rlang` package @R-rlang). Simply, this
+means that computations are performed and evaluated in the context of the 
+_Ranges_ objects, which emphasises the interactive nature of the API. This has
+the disadvantage that programming with `plyranges` becomes more difficult 
+because a user needs to know about non-standard evaluation and user defined
+functions can be difficult to debug.
+
 # Results
 
-As an example, we use the Bioconductor package _AnnotationHub_ @R-ahub to search for 
-BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project @Roadmap_Epigenomics_Consortium2015-pr.
-We choose to focus on assays for primary T CD8+ memory cells from peripheral 
-blood. We can then read the BigWig file  corresponding to the H3 lysine 27 
- trimethylation (H3K27Me3) 
-methylation mark over chromosome 10.
+## Peak Finding
 
-First, we gather the BigWig file
-and extract its annotation information and filter it to chromosome 10.
+We use the Bioconductor package _AnnotationHub_ @R-ahub to search for 
+BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project 
+@Roadmap_Epigenomics_Consortium2015-pr. We choose to focus on assays for primary
+T CD8+ memory cells from peripheral blood. We can then read the BigWig file 
+corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation mark 
+over chromosome 10.
+
+First, we gather the BigWig file and extract its annotation information and 
+filter it to chromosome 10.
 
@@
 then merge nearby regions. The maximum coverage is computed over all the
-coverage scores in regions that were reduced.
+coverage scores in the regions that were reduced.
 
@@
 
-Returning to the _Ranges_ object containing normalised coverage scores for
+Returning to the GRanges object containing normalised coverage scores for
 the methylation data, we can filter to find the coordinates of the peak 
-containing maximum coverage score. We can then find 
+containing the maximum coverage score. We can then find 
 a 5000 nt region centered around the maximum position by anchoring and 
@@
 
-Finally, the overlap inner join could be used to restrict the chromosome 10
-normalised coverage scores that are within the 5000nt region that contains
+Finally, the overlap inner join is used to restrict the chromosome 10
+normalised coverage scores that are within the 5000nt region that contain
 the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).
@@
 peak_region <- chr10_scores %>%
-  join_overlap_inner(chr10_max_score_region %>% 
-                       select(-score))
+  join_overlap_inner(chr10_max_score_region)
 
@@
 ggbio::autoplot(peak_region, 
-                aes(y = score), 
+                aes(y = score.x), 
                 geom = "blank") + 
@@
 ```
+
+## Computing Windowed Statistics
+
+(example of group_by_overlaps + summarise)
+
+## Transition Transversion Ratio
+
+(example of group_by + mutate + summarise and working on an object that inherits
+from GRanges)
 

### 2018-04-13 paper.Rmd
cohesion section draft

@@
 However, there is a cognitive load placed on a user in learning a data abstraction and understanding its appropriate use.
-Throughout a standard workflow, a user must navigate and know many of these abstractions to perform an genomic analysis task, when a single data abstraction, a _Ranges_ object will suffice.
-The _Ranges_ class naturally represent genomic intervals and their associated measurements.
-By recognising that the _Ranges_ class follows 'tidy' data principles we have created a grammar of genomic data transformation.
+Throughout a standard workflow, a user must navigate and know many of these abstractions to perform an genomic analysis task, when a single data abstraction, a GRanges object will suffice.
+The GRanges class naturally represent genomic intervals and their associated measurements.
+By recognising that the GRanges class follows 'tidy' data principles we have created a grammar of genomic data transformation.
 The grammar defines verbs for performing actions on and between genomic interval data.
@@
 
-Another example of altering object semantics, while maintaining cohesion is
-through the 'group_by' operator. Like anchoring, this operator decorates
-a GRanges object with a column name (or names) that imply a partitioning of the 
-GRanges by the unique values in the column(s). Functions defined in the 
-`plyranges` DSL that perform restriction, aggregation, or column modfication
-still perform those singular tasks, however grouping changes how those tasks
-are performed.
+Another example of our algebra altering object semantics, while maintaining 
+cohesion is through the 'group_by' operator. Like anchoring, this operator 
+decorates a GRanges object with a column name (or names) that defines a 
+partitioning of the  GRanges by the unique values in the column(s). Functions 
+defined in the  `plyranges` DSL that perform restriction, aggregation, or 
+column modfication still perform those singular tasks, however grouping changes
+how those tasks are performed.
 
-A common operation to perform between two _Ranges_ objects is to find overlaps
-or nearest neighbours. The _plyranges_ API recasts these operations as 'joins'
-or 'pairing' operations. For overlaps, there are three join operations: 
-`join_overlap_intersect()`, `join_overlap_inner()` and `join_overlap_left()`
-which are shown in figure (\ref{fig:olap}).
+Our algebra recasts the actions of finding overlaps or nearest neightbours 
+between two genomic regions as a relational join operator. The join operator 
+acts on two GRanges object, a query and a subject. The join operator is 
+relational in the sense that metadata from the query and subject ranges is 
+retained in the joined range.  The commonality between
+all join operators in the `plyranges` DSL is to use the index of where in
+the query range the subject range 'hits' the query range as a primary key 
+(figure *need to make figure*). The notion of 'hits' and hence
+the resulting key is determined by the type of join (preceding, nearest, follows
+overlaps).
+
+We extend this idea by introducing three types of overlap joins: inner,
+intersect and left. The inner join considers a hit to be any overlap between
+subject and query and expands the query range to have length equal to the 
+number of hits. The intersect join uses the same definition of a hit as
+the inner join but returns a range where the start and end coordinates
+are the intersect of coordinates in the subject that hit the query. Finally,
+the overlap left join is akin to left outer join in Cobb's relational algebra:
+it performs an overlap inner join but also returns all query ranges that
+are not hit by the subject.
+
+The behaviour of a join operator can be further altered with additional
+suffixes by restricting or expanding how a subject 'hits' a query. For example,
+for overlap joins we can use suffixes to encapsulate Allan's Interval Algebra:
+to consider subject 'hits' that are entirely 'within' a query range, the 'within'
+prefix is used. We also use the 'directed' suffix to explicitly include 
+notions of strandedness in our algebra.
+
+comparsion to Bioconductor API here?
 
@@
 
-These operations consider any overlap between
-two input ranges and return any corresponding metadata from both Ranges objects
-as metadata. The intersect join takes the intersect of the start and end coordinates
-of overlapping intervals of the query and subject _Ranges_ 
-(for _GenomicRanges_ it also accounts for sequence name), when there is a overlap
-the metadata corresponding to the query and subject _Ranges_ are returned. 
-Similarly, inner join takes the start and end coordinates of the query _Ranges_
-that overlap the subject _Ranges_ and returns metadata of the overlapping query
-and subject _Ranges_. Finally, the left join performs a left outer join between
-the query and subject _Ranges_, it returns all genomic intervals from the query
-ranges, and returns missing values in metadata columns when there is no overlap.
-
-A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
-`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
-be modified with the `within` suffix (which changes the type of overlap 
-from 'any' to 'within') or the `directed` suffix (which takes into account
-the strand of a _GenomicRanges_ object.).
-
-For nearest neighbours, the _plyranges_ API provides `join_nearest()`, 
-`join_precede()`, and `join_follow()` functions. These functions are similar
-to the overlapping functions, in that they return the query ranges that are
-nearest (or precede or follow) the subject ranges and add metadata from
-the subject ranges when the query is a nearest neighbour of the subject. Like
-the overlap joins, these functions can modified with suffixes to find nearest
-neighbours that are left, right, upstream or downstream of the subject.
-
-The pairing operations, `pair_overlap()`, `pair_nearest()`, `pair_follow()`, 
-and `pair_precede()` are similar to the join operation but instead of returning
-a _Ranges_, they pair up the subject and query _Ranges_ objects into a _DataFrame_,
-alongside their metadata columns. This data structure is similar to the _Pairs_
-data structure in the _S4Vectors_ @R-S4vectors package or the BED-PE file format.
-
 ## Consistency
 
-The _plyranges_ API has an expressive algebra for performing arithmetic on Ranges via
-the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
-modifies the width of a Ranges object, while `stretch` extends the start and
-end of a Ranges object. These can be chained with the anchoring functions
-`anchor_start()`, `anchor_end()`, `anchor_center()`, `anchor_3p()` or 
-`anchor_5p()`, which fix the coordinates of a Ranges object in place. 
-Moreover, the `shift_` and `flank_` family of functions can be used to shift
-all coordinates in a Ranges object or generate flanking regions from a Ranges 
-object to the left, right, upstream or downstream of the input. Unlike, the
-Bioconductor API, _plyranges_ makes it explicit via function calls whether to
-take into account the strand information of a _Ranges_ object.
-
 The methods `as_granges()` and `as_iranges()`
-for constructing _Ranges_ from tabular data structures, such as the _data.frame_ in
+for constructing GRanges from tabular data structures, such as the _data.frame_ in
 base R. These methods use non-standard evaluation so columns in a _data.frame_
-can be modified before a _Ranges_ object is created. The API also has
+can be modified before a GRanges object is created. The API also has
 convenience methods for annotating or extracting annotations from _GRanges_
@@
 
+## Endomorphism
+
+ (that
+are named as nouns) with many arguments (which is required in the existing 
+Bioconductor packages)
+ This is possible because every 
+function call is an endomorphism: when the input is GRanges object the output
+will also be a GRanges object. One advantage of endomorphism is that it 
+does not  require any additional learning of classes beyond GRanges and 
+the _DataFrame_ classes. This strongly deviates from the design of 
+the GRanges Bioconductor packages, where many methods return a new class
+upon return. The Bioconductor design enables efficient computing as users
+are exposed to low-level features of its API which _plyranges_ abstracts
+away. Method chaining via the pipe operator can also be difficult to debug,
+as there multiple points of failure.
+
+The combination of verbs we have defined above cover most operations that
+can be performed in the original _IRanges_ and _GenomicRanges_ packages
+without the user being exposed to new classes. In those packages to perform
+most operations requires users to learn many classes and perform 
+additional manipulations to return the results of their computation back to a
+GRanges object.
+
 ## Fluency
@@
 We have designed every function to perform a single
-action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the `dplyr` package
-and implements context specific verbs for transforming _Ranges_ objects (described in
-table ). Each verb preserves the semantics of _Ranges_ object and via
-duck typing works with derivatives of the _Ranges_ class. 
+action on a GRanges object. The `plyranges` DSL implments the core verbs from the `dplyr` package
+and implements context specific verbs for transforming GRanges objects (described in
+table ). Each verb preserves the semantics of GRanges object and via
+duck typing works with derivatives of the GRanges class. 
 Both of these aspects reduce the cognitive load on a new user since most 
@@
   ~Verb, ~Action, ~Description,
-  "`mutate()`", "modifies columns", "Takes a _Ranges_ object and a set of name-value pairs and generates a new _Ranges_ object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
- "`filter()`", "subset rows", "Takes a _Ranges_ object and a set of logical expressions and restricts the _Ranges_ object to where all logical expressions are true.",
- "`summarise()`", "aggregate one or more columns by a function", "Takes a _Ranges_ object and a set of name-value pairs and aggregates the _Ranges_ according to functions evaluated in the name-value pairs.",
-"`group_by()`", "partition a _Ranges_ object by a column", "Create an implicit grouping of _Ranges_ object by one or more columns.
+  "`mutate()`", "modifies columns", "Takes a GRanges object and a set of name-value pairs and generates a new GRanges object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
+ "`filter()`", "subset rows", "Takes a GRanges object and a set of logical expressions and restricts the GRanges object to where all logical expressions are true.",
+ "`summarise()`", "aggregate one or more columns by a function", "Takes a GRanges object and a set of name-value pairs and aggregates the GRanges according to functions evaluated in the name-value pairs.",
+"`group_by()`", "partition a GRanges object by a column", "Create an implicit grouping of GRanges object by one or more columns.
 This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
-"`select()`", "select columns of a _Ranges_ objects", "Drop or reorder metadata columns.",
-"`arrange()`", "sort a _Ranges_ object by columns", "Sort a _Ranges_ object by named variables",
-"`disjoin_ranges()`", "aggregate one more columns of a _Ranges_ object over the union of end coordinates", "Takes a _Ranges_ object and a set of name-value pairs and expands the _Ranges_ object by taking the union of end coordinates.
+"`select()`", "select columns of a GRanges objects", "Drop or reorder metadata columns.",
+"`arrange()`", "sort a GRanges object by columns", "Sort a GRanges object by named variables",
+"`disjoin_ranges()`", "aggregate one more columns of a GRanges object over the union of end coordinates", "Takes a GRanges object and a set of name-value pairs and expands the GRanges object by taking the union of end coordinates.
 For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
-"`reduce_ranges()`", "aggregate one more columns of a _Ranges_ object by merging overlapping and neighbouring ranges.", "Takes a _Ranges_ object and a set of name-value pairs and reduces the _Ranges_ object by merging overlapping and neighouring ranges.
+"`reduce_ranges()`", "aggregate one more columns of a GRanges object by merging overlapping and neighbouring ranges.", "Takes a GRanges object and a set of name-value pairs and reduces the GRanges object by merging overlapping and neighouring ranges.
 For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
@@
 ```
-
-## Endomorphism
-
- (that
-are named as nouns) with many arguments (which is required in the existing 
-Bioconductor packages)
- This is possible because every 
-function call is an endomorphism: when the input is _Ranges_ object the output
-will also be a _Ranges_ object. One advantage of endomorphism is that it 
-does not  require any additional learning of classes beyond _Ranges_ and 
-the _DataFrame_ classes. This strongly deviates from the design of 
-the _Ranges_ Bioconductor packages, where many methods return a new class
-upon return. The Bioconductor design enables efficient computing as users
-are exposed to low-level features of its API which _plyranges_ abstracts
-away. Method chaining via the pipe operator can also be difficult to debug,
-as there multiple points of failure.
-
-The combination of verbs we have defined above cover most operations that
-can be performed in the original _IRanges_ and _GenomicRanges_ packages
-without the user being exposed to new classes. In those packages to perform
-most operations requires users to learn many classes and perform 
-additional manipulations to return the results of their computation back to a
-_Ranges_ object.
 
@@
 means that computations are performed and evaluated in the context of the 
-_Ranges_ objects, which emphasises the interactive nature of the API. This has
+GRanges objects, which emphasises the interactive nature of the API. This has
 the disadvantage that programming with `plyranges` becomes more difficult 

### 2018-04-13 paper.Rmd
moving things about

@@
 
-comparsion to Bioconductor API here?
+comparsion to Bioconductor API here? table of join ops?
 
@@
 
- (that
-are named as nouns) with many arguments (which is required in the existing 
-Bioconductor packages)
 This is possible because every 
@@
 are exposed to low-level features of its API which _plyranges_ abstracts
-away. Method chaining via the pipe operator can also be difficult to debug,
-as there multiple points of failure.
+away.
 
@@
 
-We have designed every function to perform a single
-action on a GRanges object. The `plyranges` DSL implments the core verbs from the `dplyr` package
-and implements context specific verbs for transforming GRanges objects (described in
-table ). Each verb preserves the semantics of GRanges object and via
-duck typing works with derivatives of the GRanges class. 
+Using the design principles defined above every function in `plyranges`
+performs a single action on GRanges objects. The `plyranges` DSL implments the 
+core verbs from the `dplyr` package and implements a genomic relational algebra
+for transforming GRanges objects (table x, y). Each verb preserves the 
+semantics of GRanges object and works with derivatives of the GRanges class. 
 Both of these aspects reduce the cognitive load on a new user since most 
 manipulations can be performed with a vocabulary of several verbs, rather 
-than having to memorise function names that are nouns. Moreover, this approach 
-strongly constrasts the `GenomicRanges/IRanges` OO interface, which emphasises 
-the use of setter and getter methods. In that interface, core components and 
-metadata are updated via replacement methods (requiring knowledge of the
-class components), while our interface requires only a single call to `mutate()`
-to perform the modification.
+than having to memorise function names that are nouns.
+
+This approach strongly constrasts the `GenomicRanges/IRanges` OO interface, 
+which emphasises the use of setter and getter methods. In that interface, 
+core components and metadata are updated via replacement methods 
+(requiring knowledge of the class components), while our interface requires 
+only a single call to `mutate()` to perform the modification.
 
@@
 
+Method chaining via the pipe operator can also be difficult to debug,
+as there multiple points of failure
+
 # Results
+
+Here we provide examples on how to use the `plyranges` DSL to construct
+genomic data workflows.
 

### 2018-04-13 paper.Rmd
endomorphism section
context for consistency section

@@
 
+* replacing arguments with functions
+* standardised inputs for reading/writing genomics data as GRanges objects
+
 The methods `as_granges()` and `as_iranges()`
@@
 
-This is possible because every 
-function call is an endomorphism: when the input is GRanges object the output
-will also be a GRanges object. One advantage of endomorphism is that it 
-does not  require any additional learning of classes beyond GRanges and 
-the _DataFrame_ classes. This strongly deviates from the design of 
-the GRanges Bioconductor packages, where many methods return a new class
-upon return. The Bioconductor design enables efficient computing as users
-are exposed to low-level features of its API which _plyranges_ abstracts
-away.
+Most function calls in `plyranges` are endomorphisms: when the input 
+is GRanges object the output will also be a GRanges object. The use of
+endmoprhism in the `plyranges` DSL enables a user to predict the structure of 
+the output of their computations and does not require them to learn any 
+additional classes beyond GRanges and DataFrames.
 
-The combination of verbs we have defined above cover most operations that
-can be performed in the original _IRanges_ and _GenomicRanges_ packages
-without the user being exposed to new classes. In those packages to perform
-most operations requires users to learn many classes and perform 
-additional manipulations to return the results of their computation back to a
-GRanges object.
+This design pattern strongly deviates from the design of the OO interface in 
+`GenomicRanges`, where many methods return a new class that is unfamiliar to 
+the user upon return. These low-level classes enable efficient computing at the cost
+of complexity and are abstracted away in `plyranges` interface.
 
@@
 
-Using the design principles defined above every function in `plyranges`
+As a consequence of the design principles defined above every function in `plyranges`
 performs a single action on GRanges objects. The `plyranges` DSL implments the 
@@
 
-A caveat to constructing a compatible API with `dplyr` is that 
+A caveat to constructing a compatible interface with `dplyr` is that 
 `plyranges` makes extensive use of non-standard evaluation in R
@@
 means that computations are performed and evaluated in the context of the 
-GRanges objects, which emphasises the interactive nature of the API. This has
-the disadvantage that programming with `plyranges` becomes more difficult 
+GRanges objects, which emphasises the interactive nature of the interface.
+This has the disadvantage that programming with `plyranges` becomes more difficult 
 because a user needs to know about non-standard evaluation and user defined
 functions can be difficult to debug.
-
-Method chaining via the pipe operator can also be difficult to debug,
-as there multiple points of failure
 

### 2018-04-16 paper.Rmd
design section redraft

@@
 Several domain specific languages (DSLs) enable scientists to process and reason
-about heterogenous genomics data heterogenous genomics data by expresssing common
+about heterogeneous genomics data heterogeneous genomics data by expressing common
 operations, such as range manipulation and overlap-based joins, using
@@
 algebra for performing arithmetic between two or more sets of genomic regions 
-@Quinlan2010-gc. All of the aformentioned DSLs are designed to be evaluated
-either at the commandline or embedded in scripts for batch processing. They 
-exist in asparse ecosystem, mostly consisting of UNIX and database tools that
+@Quinlan2010-gc. All of the aforementioned DSLs are designed to be evaluated
+either at the command line or embedded in scripts for batch processing. They 
+exist in sparse ecosystem, mostly consisting of UNIX and database tools that
 lack biological semantics and operate at the level of files and
@@
 analysing genomics data with R, an interactive data analysis environment that 
-encouragesreproducibility and provides high-level abstractions for manipulating,
-modeling and plotting data, through state of the art methods in
+encourages reproducibility and provides high-level abstractions for manipulating,
+modelling and plotting data, through state of the art methods in
 statistical computing. The packages define object-oriented (OO)
@@
 [@Dale2011-js; @Kent2017], however these packages lack the interoperability
-provided by in the aformentioned Bioconductor packages and are not easily 
+provided by in the aforementioned Bioconductor packages and are not easily 
 extended.
@@
 myriad of data structures have overlapping purposes and important but
-obscure differences in behavior that often confuse the typical end
+obscure differences in behaviour that often confuse the typical end
 user.
@@
 us to see how the data relates to the design of an experiment
-and the variables measured. The `dplyr` pacakge [@Wickham2017-dplyr] 
+and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
 defines an API that maps notions from the general relational algebra to 
@@
 by strand). This enables any arithmetic function that performs a 
-coordinate transformation to remain cohesive: that is they always perfom
+coordinate transformation to remain cohesive: that is they always perform
 the same operation regardless of whether a GRanges object is anchored or not.
-The only difference is the result of performing the arithimetic changes when
+The only difference is the result of performing the arithmetic changes when
 contextual information is added to the GRanges object.
@@
 defined in the  `plyranges` DSL that perform restriction, aggregation, or 
-column modfication still perform those singular tasks, however grouping changes
+column modification still perform those singular tasks, however grouping changes
 how those tasks are performed.
 
-Our algebra recasts the actions of finding overlaps or nearest neightbours 
+Our algebra recasts the actions of finding overlaps or nearest neighbours 
 between two genomic regions as a relational join operator. The join operator 
@@
 
-comparsion to Bioconductor API here? table of join ops?
-
-```{r olap, echo = FALSE, fig.cap="The three overlap joins: the query and subject ranges are coloured by their metadata.
-When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.", fig.width = 5, fig.height = 7}
-a <- GRanges(seqnames = "chr1",
-               strand = c("+", "-"),
-               ranges = IRanges(c(1, 9), c(7, 10)),
-               key.a = letters[1:2])
-
-b <- GRanges(seqnames = "chr1",
-               strand = c("-", "+"),
-               ranges = IRanges(c(2, 6), c(4, 8)),
-               key.b = LETTERS[1:2])
-
-c <- join_overlap_intersect(a,b)
-d <- join_overlap_inner(a, b)
-e <- join_overlap_left(a, b)
-
-p1 <- autoplot(a, aes(fill = key.a)) + 
-  scale_fill_brewer(type = "qual", palette = "Set2") + 
-  guides(fill = FALSE)
-p2 <- autoplot(b, aes(fill = key.b))  + 
-  scale_fill_brewer(type = "qual", palette = "Set1") +
-  guides(fill = FALSE)
-p3 <- autoplot(c, aes(fill = key.a, colour = key.b)) + 
-  scale_fill_brewer(type = "qual", palette = "Set2") +
-  scale_color_brewer(type = "qual", palette = "Set1") +
-  guides(fill = FALSE, color = FALSE) 
-p4 <- autoplot(d, aes(fill = key.a, colour = key.b)) + 
-  scale_fill_brewer(type = "qual", palette = "Set2") +
-  scale_color_brewer(type = "qual", palette = "Set1") +
-  guides(fill = FALSE, colour = FALSE)
-p5 <- autoplot(e, aes(fill = key.a, colour = key.b))  +
-  scale_fill_brewer(type = "qual", palette = "Set2") +
-  scale_color_brewer(type = "qual", palette = "Set1") +
-  guides(fill = FALSE, colour = FALSE)
-
-tracks(query = p1, subject = p2, 
-       `intersect` = p3, 
-       `inner` = p4, 
-       `left` = p5) 
-```
+more comparison to Bioconductor API here? table of join ops?
 
@@
 
-* replacing arguments with functions
-* standardised inputs for reading/writing genomics data as GRanges objects
+A core design principle of the `plyranges` DSL is interface consistency:
+a user should not be surprised by the input or output of `plyranges` code.
+A key example of consistency is how `plyranges` handles strand information.
+Every function that computes with strand information indicates its intentions
+by including suffixes such as 'directed', 'upstream' or 'downstream' in its
+name, otherwise strand is ignored. This strongly differs from 
+the Bioconductor packages, which produces surprising output by assuming the 
+user is always interested in using strand in the majority of circumstances 
+(unless a user is computing coverage or finding flanking ranges).
 
-The methods `as_granges()` and `as_iranges()`
-for constructing GRanges from tabular data structures, such as the _data.frame_ in
-base R. These methods use non-standard evaluation so columns in a _data.frame_
-can be modified before a GRanges object is created. The API also has
-convenience methods for annotating or extracting annotations from _GRanges_
-objects with the `set_genome_info()` and `get_genome_info()` functions.
-
-There is a consistent framework for reading and writing 
-files from and to common genomic data formats, using the _rtracklayer_ package
-as a back-end @Lawrence2009-nt. The methods are implemented in the `read_/write_` family of 
-functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
-narrowPeaks, GFF/GTF, WIG and BigWig files.
+Our use of suffixes in function names, also highlights are core tenet of 
+the `plyranges` design: avoid complex generic functions with many arguments
+and instead use cohesive functions with a minimal number of arguments. As an
+example, we have written a consistent framework for reading and writing 
+files from and to common genomic data formats as GRanges, 
+using the `rtracklayer` package as a back-end @Lawrence2009-nt. We have replaced
+that packages generic functions for importing files with a family of reader
+functions that all take the exact same arguments.
 
@@
 is GRanges object the output will also be a GRanges object. The use of
-endmoprhism in the `plyranges` DSL enables a user to predict the structure of 
+endomorphism in the `plyranges` DSL enables a user to predict the structure of 
 the output of their computations and does not require them to learn any 
@@
 As a consequence of the design principles defined above every function in `plyranges`
-performs a single action on GRanges objects. The `plyranges` DSL implments the 
+performs a single action on GRanges objects. The `plyranges` DSL implements the 
 core verbs from the `dplyr` package and implements a genomic relational algebra
@@
 
-This approach strongly constrasts the `GenomicRanges/IRanges` OO interface, 
+This approach strongly contrasts the `GenomicRanges/IRanges` OO interface, 
 which emphasises the use of setter and getter methods. In that interface, 

### 2018-04-16 paper.tex
design section redraft

@@
 {\Large
-\textbf\newline{plyranges: a grammar for manipulating genomics data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{plyranges: a grammar for transforming genomics data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \section*{Abstract}
-The Bioconductor project has created many powerful abstractions for
-reasoning about genomics data, such as the \emph{Ranges} data structures
-for representing genomic intervals. By recognising that these data
-structures follow `tidy' data principles we have created a grammar of
-genomic data manipulation that defines verbs for performing actions on
-and between genomic interval data. This grammar simplifies performing
-common genomic data analysis tasks via a consistent interface to
-existing Bioconductor infrastructure which results in creating human
-readable analysis pipelines. We have implemented this grammar as an
+The Bioconductor project has created many useful data abstractions for
+analysing high-throughput genomics experiments. However, there is a
+cognitive load placed on a user in learning a data abstraction and
+understanding its appropriate use. Throughout a standard workflow, a
+user must navigate and know many of these abstractions to perform an
+genomic analysis task, when a single data abstraction, a GRanges object
+will suffice. The GRanges class naturally represent genomic intervals
+and their associated measurements. By recognising that the GRanges class
+follows `tidy' data principles we have created a grammar of genomic data
+transformation. The grammar defines verbs for performing actions on and
+between genomic interval data. It provides a principled way of
+performing common genomic data analysis tasks through a coherent
+interface to existing Bioconductor infrastructure, resulting in human
+readable analysis workflows. We have implemented this grammar as an
 Bioconductor/R package called plyranges.
@@
 
-Genomic data may be naturally represented as sets of pairs of integers
-corresponding to the start and end points of sequences. Further
-information such as strandedness and chromosome name may be added to
-these sets to provide biological context. Because of the flexibility of
-this representation supplemental information such as measurements
-obtained from an experimental assay or annotations from a genome
-database can be joined to their relevant sequences. In the
-Bioconductor/R packages \texttt{IRanges} and \texttt{GenomicRanges}
-these representations have been implemented as a suite of data
-structures called \emph{Ranges} {[}1{]}; {[}2{]}. These data structures
-cover many common data types encountered in bioinformatics analyses. For
-example, a gene can be represented with its coordinates, along with its
-identifier and the identifiers of its exons; or an RNA-seq experiment
-may be represented as sets of genes with a matching count column.
+High-throughput genomics promises to unlock new disease therapies, and
+strengthen our knowledge of basic biology. To deliver on those promises,
+scientists must derive a stream of knowledge from a deluge of data.
+Genomic data is challenging in both scale and complexity. Innovations in
+sequencing technology often outstrip our capacity to process the output.
+Beyond their common association with genomic coordinates, genomic data
+are heterogeneous, consisting of raw sequence read alignments, genomic
+feature annotations like genes and exons, and summaries like coverage
+vectors, ChIP-seq peak calls, variant calls, and per-feature read
+counts. Genomic scientists need software tools to wrangle the different
+types of data, process the data at scale, test hypotheses, and generate
+new ones, all while focusing on the biology, not the computation. For
+the tool developer, the challenge is to define ways to model and operate
+on the data that align with the mental model of scientists, and to
+provide an implementation that scales with their ambition.
 
-The Bioconductor infrastructure for computing with genomic ranges are
-highly efficient and powerful, however the application programming
-interface (API) for performing analysis tasks with \emph{Ranges} is
-complex due to its large number of methods and classes. It also makes
-resulting scripts written difficult for a non-programmer to read and
-reason about. An alternative approach would be to implement a domain
-specific language (DSL) as a fluent interface built on top \emph{Ranges}
-{[}3{]}. The goal of a fluent interface is to enable users to write
-human-readable code via method chaining and consistent function returns.
-Fluent interfaces fit naturally in the context of bioinformatics
-workflows because they enable writing succinct pipelines.
+Several domain specific languages (DSLs) enable scientists to process
+and reason about heterogeneous genomics data heterogeneous genomics data
+by expressing common operations, such as range manipulation and
+overlap-based joins, using the vocabulary of genomics. Their
+implementations either delegate computations to a database, or operate
+over collections of files in standard formats like BED.An example of the
+former is the Genome Query Language (GQL) and its distributed
+implementation GenAp which use an SQL-like syntax for fast retrieval of
+information of unprocessed sequencing data {[}1{]}; {[}2{]}. Similarly,
+the Genometric Query Language (GMQL) implements a relational algebra for
+combining genomic datasets {[}3{]}. The command line application
+BEDtools develops an extensive algebra for performing arithmetic between
+two or more sets of genomic regions {[}4{]}. All of the aforementioned
+DSLs are designed to be evaluated either at the command line or embedded
+in scripts for batch processing. They exist in sparse ecosystem, mostly
+consisting of UNIX and database tools that lack biological semantics and
+operate at the level of files and database tables.
 
-Several other DSLs have been implemented to reason about genomics data.
-Broadly, these are either implemented as query languages or as command
-line tools embedded in the Unix environment. An example of the former is
-the Genome Query Language (GQL) and its distributed implementation GenAp
-which use an SQL-like syntax for fast retrieval of information from
-genomic databases and BAM files {[}4{]}; {[}5{]}. Another example is the
-Genometric Query Language (GMQL) which implements a relational algebra
-for combining big genomic datasets {[}6{]}.\\
-The command line application BEDtools develops an extensive algebra for
-performing arithmetic between two or more sets of genomic regions
-{[}7{]}. It also has a python interface which simplifies constructing
-scripts for performing analyses based on BEDTools {[}8{]}.
+The Bioconductor/R packages \texttt{IRanges} and \texttt{GenomicRanges}
+{[}5--7{]} define a DSL for analysing genomics data with R, an
+interactive data analysis environment that encourages reproducibility
+and provides high-level abstractions for manipulating, modelling and
+plotting data, through state of the art methods in statistical
+computing. The packages define object-oriented (OO) abstractions for
+representing genomic data and enable interoperability by allowing users
+and developers to use these abstractions in their own code and packages.
+Other genomic DSLs that are embedded in programming languages include
+pybedtools and valr {[}8,9{]}, however these packages lack the
+interoperability provided by in the aforementioned Bioconductor packages
+and are not easily extended.
 
-The abstraction provided by the \emph{Ranges} data structures aligns
-with the concept of tidy data {[}9{]}. A tidy dataset is defined to be
-to a tabular data structure that has observations as rows and columns as
-variables, and each tidy dataset represents measurements from a single
-observational unit. The tidy data pattern is useful because it allows us
-to see how the data relates to the design of an experiment and the
-variables measured. Consequently, it makes both the modelling and
-manipulation of data more systematic. The \emph{Ranges} data structure
-follows this abstraction: it is a rectangular table corresponding to a
-single biological context. Each row contains a single observation and
-each column is a variable about that observation.
+The Bioconductor infrastructure models the genomic data and operations
+from the perspective of the power user, one who understands and wants to
+take advantage of the subtle differences in data types. This design has
+enabled the development of sophisticated tools, as evidenced by the
+hundreds of packages depending the framework. Unfortunately, the myriad
+of data structures have overlapping purposes and important but obscure
+differences in behaviour that often confuse the typical end user.
 
-The \emph{plyranges} API implements a domain specific language using the
-existing \emph{IRanges} and \emph{GenomicRanges} packages in
-Bioconductor as a back-end. Consequently, our API still has the speed
-and efficiency of the aforementioned packages but with a more coherent
-interface. The API also extends the grammar elements in \emph{dplyr}
-{[}10{]} for performing genomic specific manipulations such as finding
-overlapping regions or nearest neighbour regions between many
-\emph{Ranges} . The \emph{plyranges} API is specifically designed to
-enable fast interactive analysis of \emph{Ranges} objects but can also
-be used for scripting genomic data workflows.
+Recently, there has been a concerted, community effort to standardize R
+data structures and workflows around the notion of tidy data {[}10{]}. A
+tidy dataset is defined to be to a tabular data structure that has
+observations as rows and columns as variables, and each tidy dataset
+represents measurements from a single observational unit. The tidy data
+pattern is useful because it allows us to see how the data relates to
+the design of an experiment and the variables measured. The
+\texttt{dplyr} package {[}11{]} defines an API that maps notions from
+the general relational algebra to operations on tidy data. It expresses
+each operation as a cohesive, endomorphic verb. Taken together these
+features enable a user to write human readable analysis workflows.
+
+We have created a genomic DSL called \texttt{plyranges} that
+reformulates notions from existing genomic algebras and embeds them in R
+as a genomic extension of \texttt{dplyr}. By analogy, \texttt{plyranges}
+is to the genomic algebra, as \texttt{dplyr} is to the relational
+algebra. The \texttt{plyranges} Bioconductor package implements the
+language on top of a key subset of Bioconductor data structures and thus
+fully integrates with the Bioconductor framework, gaining access to its
+scalable data representations and sophisticated statistical methods.
 
@@
 
-We have designed the API to be fluent. Every function call corresponds
-to an action on a \emph{Ranges} object (they are named verbs) and where
-possible functions have few arguments. Each verb is constructed to
-enable a tab completion based workflow. Both of these aspects reduce the
-cognitive load on a new user since most manipulations can be performed
-with a vocabulary of several verbs, rather than having to memorise
-functions with many arguments that are nouns (as is required in the
-existing Bioconductor packages). This is also has the advantage of
-allowing users to write human readable code because verbs describe what
-the code is doing rather than how its doing it.
+The \texttt{plyranges} DSL is built on the most general Bioconductor
+data structure, GRanges, which is capable of representing all types of
+genomic data at a semantic level that roughly matches the intuition of
+most users. A GRanges is essentially a table, with columns for the
+chromosome, start and end coordinates, and the strand, along with an
+arbitrary set of additional columns, consisting of measurements or
+metadata specific to the data type or experiment. For example a GRanges
+can be used to represent a gene with its chromosomal coordinates, and
+along with its exon identifiers; or a GRanges can represent an RNA-seq
+experiment as a set of genes with a corresponding count column.
 
-Workflows can be composed by chaining verbs together via the forward
-pipe operator,\texttt{\%\textgreater{}\%} (exported from the R package
-\emph{magrittr} {[}11{]}). This is possible because every function call
-is an endomorphism: when the input is \emph{Ranges} object the output
-will also be a \emph{Ranges} object. One advantage of endomorphism is
-that it does not require any additional learning of classes beyond
-\emph{Ranges} and the \emph{DataFrame} classes. This strongly deviates
-from the design of the \emph{Ranges} Bioconductor packages, where many
-methods return a new class upon return. The Bioconductor design enables
-efficient computing as users are exposed to low-level features of its
-API which \emph{plyranges} abstracts away. Method chaining via the pipe
-operator can also be difficult to debug, as there multiple points of
-failure.
+By definition the GRanges follow the tidy data pattern: it is a
+rectangular table corresponding to a single biological context. Each row
+contains a single observation and each column is a variable about that
+observation. Hence, we have designed the \texttt{plyranges} DSL to
+extend the grammar and design principles of \texttt{dplyr}: cohesion,
+consistency, endomorphism, and fluency. All of these principles are
+defined and discussed below in the context of the GRanges class. Where
+applicable we contrast our design to the existing Bioconductor
+infrastructure.
 
-In order to provide a compatible API with \emph{dplyr}, \emph{plyranges}
-makes extensive use of non-standard evaluation in R via the \emph{rlang}
-package {[}12{]}. Simply, this means that computations are performed and
-evaluated in the context of the \emph{Ranges} objects, which emphasises
-the interactive nature of the API. This has the disadvantage that
-programming with \emph{plyranges} becomes more difficult because a user
-needs to capture expressions inside function calls and then unquote
-them.
+\subsection{Cohesion}\label{cohesion}
 
-\subsection{Actions on Ranges}\label{actions-on-ranges}
+A function is cohesive if it performs a singular task and does not
+produce any side-effects. In the \texttt{plyranges} DSL our algebra for
+performing genomic arithmetic is a key example of cohesion. We define
+anchoring operators that decorate a GRanges object with an `anchor' and
+modify the semantics of performing genomic arithmetic. The anchoring
+operators amount to fixing a GRanges object by its start, center, or end
+coordinates (or fixing these coordinates by strand). This enables any
+arithmetic function that performs a coordinate transformation to remain
+cohesive: that is they always perform the same operation regardless of
+whether a GRanges object is anchored or not. The only difference is the
+result of performing the arithmetic changes when contextual information
+is added to the GRanges object.
 
-The \emph{plyranges} API exports the six core verbs from the
-\emph{dplyr} package and modifies them for use with \emph{Ranges}
-objects. The verb \texttt{mutate()} takes a Ranges object and a set of
-name-value pairs and generates a new Ranges object that with modified or
-new metadata columns or modified core components (start, end, width,
-seqnames, strand). The use of \texttt{mutate()} means that a user no
-longer needs knowledge of the accessors of the Ranges object, as they
-can modify them in place. The \texttt{filter()} function takes a Ranges
-object and logical expressions and restricts Ranges object to where the
-logical expression evaluates to true. The \texttt{summarise()} function
-takes a Ranges object and a set of name-value pairs and aggregates the
-Ranges according to functions evaluated in the name-value pairs. As
-\texttt{summarise()} is an aggregation it may break the structure of the
-of a Ranges object, hence it returns a \emph{DataFrame} object. The
-\texttt{select()} function determines which metadata columns are
-returned and the order they are returned in. The \texttt{arrange()}
-function sorts a Ranges object by named variables. The
-\texttt{group\_by()} function creates an implicit grouping of Ranges
-object according to variables in the Ranges object. This modifies the
-actions of \texttt{mutate()}, \texttt{summarise()} and
-\texttt{filter()}, so they are performed on each partition created by
-the grouping. The \texttt{group\_by()} operation acts as a replacement
-for the \emph{GenomicRangesList} class in the original
-\emph{GenomicRanges} API.
+Another example of our algebra altering object semantics, while
+maintaining cohesion is through the `group\_by' operator. Like
+anchoring, this operator decorates a GRanges object with a column name
+(or names) that defines a partitioning of the GRanges by the unique
+values in the column(s). Functions defined in the \texttt{plyranges} DSL
+that perform restriction, aggregation, or column modification still
+perform those singular tasks, however grouping changes how those tasks
+are performed.
 
-The \emph{plyranges} API introduces additional summary verbs,
-\texttt{reduce\_ranges()} and \texttt{disjoin\_ranges()}that return
-Ranges objects after being returned. The \texttt{reduce\_ranges()}
-operation merges overlapping and neighbour ranges, while
-\texttt{disjoin\_ranges()} expands ranges by taking the union of end
-points.
+Our algebra recasts the actions of finding overlaps or nearest
+neighbours between two genomic regions as a relational join operator.
+The join operator acts on two GRanges object, a query and a subject. The
+join operator is relational in the sense that metadata from the query
+and subject ranges is retained in the joined range. The commonality
+between all join operators in the \texttt{plyranges} DSL is to use the
+index of where in the query range the subject range `hits' the query
+range as a primary key (figure \emph{need to make figure}). The notion
+of `hits' and hence the resulting key is determined by the type of join
+(preceding, nearest, follows overlaps).
 
-\subsection{Arithimetic on Ranges}\label{arithimetic-on-ranges}
+We extend this idea by introducing three types of overlap joins: inner,
+intersect and left. The inner join considers a hit to be any overlap
+between subject and query and expands the query range to have length
+equal to the number of hits. The intersect join uses the same definition
+of a hit as the inner join but returns a range where the start and end
+coordinates are the intersect of coordinates in the subject that hit the
+query. Finally, the overlap left join is akin to left outer join in
+Cobb's relational algebra: it performs an overlap inner join but also
+returns all query ranges that are not hit by the subject.
 
-The \emph{plyranges} API has an expressive algebra for performing
-arithmetic on Ranges via the verbs \texttt{set\_width()} and
-\texttt{stretch()}. As the names suggest \texttt{set\_width()} modifies
-the width of a Ranges object, while \texttt{stretch} extends the start
-and end of a Ranges object. These can be chained with the anchoring
-functions \texttt{anchor\_start()}, \texttt{anchor\_end()},
-\texttt{anchor\_center()}, \texttt{anchor\_3p()} or
-\texttt{anchor\_5p()}, which fix the coordinates of a Ranges object in
-place. Moreover, the \texttt{shift\_} and \texttt{flank\_} family of
-functions can be used to shift all coordinates in a Ranges object or
-generate flanking regions from a Ranges object to the left, right,
-upstream or downstream of the input. Unlike, the Bioconductor API,
-\emph{plyranges} makes it explicit via function calls whether to take
-into account the strand information of a \emph{Ranges} object.
+The behaviour of a join operator can be further altered with additional
+suffixes by restricting or expanding how a subject `hits' a query. For
+example, for overlap joins we can use suffixes to encapsulate Allan's
+Interval Algebra: to consider subject `hits' that are entirely `within'
+a query range, the `within' prefix is used. We also use the `directed'
+suffix to explicitly include notions of strandedness in our algebra.
 
-\subsection{Overlapping Ranges}\label{overlapping-ranges}
+more comparison to Bioconductor API here? table of join ops?
 
-A common operation to perform between two \emph{Ranges} objects is to
-find overlaps or nearest neighbours. The \emph{plyranges} API recasts
-these operations as `joins' or `pairing' operations. For overlaps, there
-are three join operations: \texttt{join\_overlap\_intersect()},
-\texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
-are shown in figure (\ref{fig:olap}).
+\subsection{Consistency}\label{consistency}
 
-\begin{figure}
+A core design principle of the \texttt{plyranges} DSL is interface
+consistency: a user should not be surprised by the input or output of
+\texttt{plyranges} code. A key example of consistency is how
+\texttt{plyranges} handles strand information. Every function that
+computes with strand information indicates its intentions by including
+suffixes such as `directed', `upstream' or `downstream' in its name,
+otherwise strand is ignored. This strongly differs from the Bioconductor
+packages, which produces surprising output by assuming the user is
+always interested in using strand in the majority of circumstances
+(unless a user is computing coverage or finding flanking ranges).
 
-{\centering \includegraphics{paper_files/figure-latex/olap-1}
+Our use of suffixes in function names, also highlights are core tenet of
+the \texttt{plyranges} design: avoid complex generic functions with many
+arguments and instead use cohesive functions with a minimal number of
+arguments. As an example, we have written a consistent framework for
+reading and writing files from and to common genomic data formats as
+GRanges, using the \texttt{rtracklayer} package as a back-end {[}12{]}.
+We have replaced that packages generic functions for importing files
+with a family of reader functions that all take the exact same
+arguments.
 
-}
+\subsection{Endomorphism}\label{endomorphism}
 
-\caption{The three overlap joins: the query and subject ranges are coloured by their metadata.
-When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.}\label{fig:olap}
-\end{figure}
+Most function calls in \texttt{plyranges} are endomorphisms: when the
+input is GRanges object the output will also be a GRanges object. The
+use of endomorphism in the \texttt{plyranges} DSL enables a user to
+predict the structure of the output of their computations and does not
+require them to learn any additional classes beyond GRanges and
+DataFrames.
 
-These operations consider any overlap between two input ranges and
-return any corresponding metadata from both Ranges objects as metadata.
-The intersect join takes the intersect of the start and end coordinates
-of overlapping intervals of the query and subject \emph{Ranges} (for
-\emph{GenomicRanges} it also accounts for sequence name), when there is
-a overlap the metadata corresponding to the query and subject
-\emph{Ranges} are returned. Similarly, inner join takes the start and
-end coordinates of the query \emph{Ranges} that overlap the subject
-\emph{Ranges} and returns metadata of the overlapping query and subject
-\emph{Ranges}. Finally, the left join performs a left outer join between
-the query and subject \emph{Ranges}, it returns all genomic intervals
-from the query ranges, and returns missing values in metadata columns
-when there is no overlap.
+This design pattern strongly deviates from the design of the OO
+interface in \texttt{GenomicRanges}, where many methods return a new
+class that is unfamiliar to the user upon return. These low-level
+classes enable efficient computing at the cost of complexity and are
+abstracted away in \texttt{plyranges} interface.
 
-A user may also restrict or group by overlaps with the
-\texttt{filter\_by\_overlaps()}, \texttt{filter\_by\_non\_overlaps()}
-and \texttt{group\_by\_overlaps()}. All overlap methods can be modified
-with the \texttt{within} suffix (which changes the type of overlap from
-`any' to `within') or the \texttt{directed} suffix (which takes into
-account the strand of a \emph{GenomicRanges} object.).
+\subsection{Fluency}\label{fluency}
 
-For nearest neighbours, the \emph{plyranges} API provides
-\texttt{join\_nearest()}, \texttt{join\_precede()}, and
-\texttt{join\_follow()} functions. These functions are similar to the
-overlapping functions, in that they return the query ranges that are
-nearest (or precede or follow) the subject ranges and add metadata from
-the subject ranges when the query is a nearest neighbour of the subject.
-Like the overlap joins, these functions can modified with suffixes to
-find nearest neighbours that are left, right, upstream or downstream of
-the subject.
+As a consequence of the design principles defined above every function
+in \texttt{plyranges} performs a single action on GRanges objects. The
+\texttt{plyranges} DSL implements the core verbs from the \texttt{dplyr}
+package and implements a genomic relational algebra for transforming
+GRanges objects (table x, y). Each verb preserves the semantics of
+GRanges object and works with derivatives of the GRanges class. Both of
+these aspects reduce the cognitive load on a new user since most
+manipulations can be performed with a vocabulary of several verbs,
+rather than having to memorise function names that are nouns.
 
-The pairing operations, \texttt{pair\_overlap()},
-\texttt{pair\_nearest()}, \texttt{pair\_follow()}, and
-\texttt{pair\_precede()} are similar to the join operation but instead
-of returning a \emph{Ranges}, they pair up the subject and query
-\emph{Ranges} objects into a \emph{DataFrame}, alongside their metadata
-columns. This data structure is similar to the \emph{Pairs} data
-structure in the \emph{S4Vectors} {[}13{]} package or the BED-PE file
-format.
+This approach strongly contrasts the \texttt{GenomicRanges/IRanges} OO
+interface, which emphasises the use of setter and getter methods. In
+that interface, core components and metadata are updated via replacement
+methods (requiring knowledge of the class components), while our
+interface requires only a single call to \texttt{mutate()} to perform
+the modification.
 
-The combination of verbs we have defined above encapsulate all
-operations that can be performed in the original \emph{IRanges} and
-\emph{GenomicRanges} packages without the user be exposed to new
-classes. In those packages to perform most operations requires users to
-learn many classes and perform additional manipulations to return the
-results of their computation back to a \emph{Ranges} object.
+Workflows can be composed by chaining verbs together into `sentences'
+via the forward pipe operator,\texttt{\%\textgreater{}\%} (exported from
+the R package \texttt{magrittr} {[}13{]}), which can be read as the word
+`then'. Overall, this allows users to write human readable code because
+workflows describe what the code is doing rather than how its doing it.
 
-\subsection{Construction and
-Import/Output}\label{construction-and-importoutput}
+\subsection{Limitations}\label{limitations}
 
-The methods \texttt{as\_granges()} and \texttt{as\_iranges()} for
-constructing \emph{Ranges} from tabular data structures, such as the
-\emph{data.frame} in base R. These methods use non-standard evaluation
-so columns in a \emph{data.frame} can be modified before a \emph{Ranges}
-object is created. The API also has convenience methods for annotating
-or extracting annotations from \emph{GRanges} objects with the
-\texttt{set\_genome\_info()} and \texttt{get\_genome\_info()} functions.
-
-There is a consistent framework for reading and writing files from and
-to common genomic data formats, using the \emph{rtracklayer} package as
-a back-end {[}14{]}. The methods are implemented in the
-\texttt{read\_/write\_} family of functions, currently \emph{plyranges}
-can read and write BAM, BED, BEDPE, narrowPeaks, GFF/GTF, WIG and BigWig
-files.
+A caveat to constructing a compatible interface with \texttt{dplyr} is
+that \texttt{plyranges} makes extensive use of non-standard evaluation
+in R (achieved via the \texttt{rlang} package {[}14{]}). Simply, this
+means that computations are performed and evaluated in the context of
+the GRanges objects, which emphasises the interactive nature of the
+interface. This has the disadvantage that programming with
+\texttt{plyranges} becomes more difficult because a user needs to know
+about non-standard evaluation and user defined functions can be
+difficult to debug.
 
@@
 
-As an example, we use the Bioconductor package \emph{AnnotationHub}
-{[}15{]} to search for BigWig files from ChIP-Seq experiments from the
-Human Epigenome Roadmap project {[}16{]}. We choose to focus on assays
-for primary T CD8+ memory cells from peripheral blood. We can then read
-the BigWig file corresponding to the H3 lysine 27 trimethylation
-(H3K27Me3) methylation mark over chromosome 10.
+Here we provide examples on how to use the \texttt{plyranges} DSL to
+construct genomic data workflows.
+
+\subsection{Peak Finding}\label{peak-finding}
+
+We use the Bioconductor package \emph{AnnotationHub} {[}15{]} to search
+for BigWig files from ChIP-Seq experiments from the Human Epigenome
+Roadmap project {[}16{]}. We choose to focus on assays for primary T
+CD8+ memory cells from peripheral blood. We can then read the BigWig
+file corresponding to the H3 lysine 27 trimethylation (H3K27Me3)
+methylation mark over chromosome 10.
 
@@
 merge nearby regions. The maximum coverage is computed over all the
-coverage scores in regions that were reduced.
+coverage scores in the regions that were reduced.
 
@@
 
-Returning to the \emph{Ranges} object containing normalised coverage
-scores for the methylation data, we can filter to find the coordinates
-of the peak containing maximum coverage score. We can then find a 5000
-nt region centered around the maximum position by anchoring and
-modifying the the width.
+Returning to the GRanges object containing normalised coverage scores
+for the methylation data, we can filter to find the coordinates of the
+peak containing the maximum coverage score. We can then find a 5000 nt
+region centered around the maximum position by anchoring and modifying
+the the width.
 
@@
 
-Finally, the overlap inner join could be used to restrict the chromosome
-10 normalised coverage scores that are within the 5000nt region that
-contains the max peak on chromosome 10 (visualised in figure
+Finally, the overlap inner join is used to restrict the chromosome 10
+normalised coverage scores that are within the 5000nt region that
+contain the max peak on chromosome 10 (visualised in figure
 \ref{fig:peak-viz}).
@@
 \NormalTok{peak_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(chr10_max_score_region }\OperatorTok{%>%}\StringTok{ }
-\StringTok{                       }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{score))}
+\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(chr10_max_score_region)}
 \end{Highlighting}
@@
 
+\subsection{Computing Windowed
+Statistics}\label{computing-windowed-statistics}
+
+(example of group\_by\_overlaps + summarise)
+
+\subsection{Transition Transversion
+Ratio}\label{transition-transversion-ratio}
+
+(example of group\_by + mutate + summarise and working on an object that
+inherits from GRanges)
+
 \section{Availablilty and Future
@@
 \url{https://github.com/sa-lee/plyranges}. We aim to continue developing
-the \emph{plyranges} package and extend it for use with more complex
+the \texttt{plyranges} package and extend it for use with more complex
 data structures such as the \emph{SummarizedExperiment} class, which can
 be used for analysing transcriptomic and variant data. As the
-\emph{plyranges} interface encourages tidy data practices it integrates
-well with the principles of the grammar of graphics, we aim to use it
-for the visualisation of multimodal biological datasets.
+\texttt{plyranges} interface encourages tidy data practices it
+integrates well with the principles of the grammar of graphics, we aim
+to use it to prepare data for the visualisation of multimodal biological
+datasets.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-Kozanitis2014-va}{}
+1. Kozanitis C, Heiberg A, Varghese G, Bafna V. Using genome query
+language to uncover genetic variation. Bioinformatics. 2014;30: 1--8.
+doi:\href{https://doi.org/10.1093/bioinformatics/btt250}{10.1093/bioinformatics/btt250}
+
+\hypertarget{ref-Kozanitis2016-bm}{}
+2. Kozanitis C, Patterson DA. GenAp: A distributed SQL interface for
+genomic data. BMC Bioinformatics. 2016;17: 63.
+doi:\href{https://doi.org/10.1186/s12859-016-0904-1}{10.1186/s12859-016-0904-1}
+
+\hypertarget{ref-Kaitoua2017-pw}{}
+3. Kaitoua A, Pinoli P, Bertoni M, Ceri S. Framework for supporting
+genomic operations. IEEE Trans Comput. 2017;66: 443--457.
+doi:\href{https://doi.org/10.1109/TC.2016.2603980}{10.1109/TC.2016.2603980}
+
+\hypertarget{ref-Quinlan2010-gc}{}
+4. Quinlan AR, Hall IM. BEDTools: A flexible suite of utilities for
+comparing genomic features. Bioinformatics. 2010;26: 841--842.
+doi:\href{https://doi.org/10.1093/bioinformatics/btq033}{10.1093/bioinformatics/btq033}
+
+\hypertarget{ref-r-core}{}
+5. R Core Team. R: A language and environment for statistical computing
+{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
+2018. Available: \url{https://www.R-project.org/}
+
 \hypertarget{ref-Lawrence2013-wg}{}
-1. Lawrence M, Huber W, Pagès H, Aboyoun P, Carlson M, Gentleman R, et
+6. Lawrence M, Huber W, Pagès H, Aboyoun P, Carlson M, Gentleman R, et
 al. Software for computing and annotating genomic ranges. PLoS Comput
@@
 \hypertarget{ref-Huber2015-ei}{}
-2. Huber W, Carey VJ, Gentleman R, Anders S, Carlson M, Carvalho BS, et
+7. Huber W, Carey VJ, Gentleman R, Anders S, Carlson M, Carvalho BS, et
 al. Orchestrating high-throughput genomic analysis with bioconductor.
@@
 doi:\href{https://doi.org/10.1038/nmeth.3252}{10.1038/nmeth.3252}
-
-\hypertarget{ref-Fowler2010-zd}{}
-3. Fowler M. Domain-Specific languages {[}Internet{]}. Pearson
-Education; 2010. Available:
-\url{https://market.android.com/details?id=book-ri1muolw_YwC}
-
-\hypertarget{ref-Kozanitis2014-va}{}
-4. Kozanitis C, Heiberg A, Varghese G, Bafna V. Using genome query
-language to uncover genetic variation. Bioinformatics. 2014;30: 1--8.
-doi:\href{https://doi.org/10.1093/bioinformatics/btt250}{10.1093/bioinformatics/btt250}
-
-\hypertarget{ref-Kozanitis2016-bm}{}
-5. Kozanitis C, Patterson DA. GenAp: A distributed SQL interface for
-genomic data. BMC Bioinformatics. 2016;17: 63.
-doi:\href{https://doi.org/10.1186/s12859-016-0904-1}{10.1186/s12859-016-0904-1}
-
-\hypertarget{ref-Kaitoua2017-pw}{}
-6. Kaitoua A, Pinoli P, Bertoni M, Ceri S. Framework for supporting
-genomic operations. IEEE Trans Comput. 2017;66: 443--457.
-doi:\href{https://doi.org/10.1109/TC.2016.2603980}{10.1109/TC.2016.2603980}
-
-\hypertarget{ref-Quinlan2010-gc}{}
-7. Quinlan AR, Hall IM. BEDTools: A flexible suite of utilities for
-comparing genomic features. Bioinformatics. 2010;26: 841--842.
-doi:\href{https://doi.org/10.1093/bioinformatics/btq033}{10.1093/bioinformatics/btq033}
 
@@
 
+\hypertarget{ref-Kent2017}{}
+9. Riemondy KA, Sheridan RM, Gillen A, Yu Y, Bennett CG, Hesselberth JR.
+Valr: Reproducible genome interval arithmetic in r. F1000Research. 2017;
+doi:\href{https://doi.org/10.12688/f1000research.11997.1}{10.12688/f1000research.11997.1}
+
 \hypertarget{ref-Wickham2014-jc}{}
-9. Wickham H. Tidy data. Journal of Statistical Software, Articles.
+10. Wickham H. Tidy data. Journal of Statistical Software, Articles.
 2014;59: 1--23.
@@
 \hypertarget{ref-Wickham2017-dplyr}{}
-10. Wickham H, Francois R, Henry L, Müller K. Dplyr: A grammar of data
+11. Wickham H, Francois R, Henry L, Müller K. Dplyr: A grammar of data
 manipulation {[}Internet{]}. 2017. Available:
@@
 
+\hypertarget{ref-Lawrence2009-nt}{}
+12. Lawrence M, Gentleman R, Carey V. Rtracklayer: An R package for
+interfacing with genome browsers. Bioinformatics. 2009;25: 1841--1842.
+doi:\href{https://doi.org/10.1093/bioinformatics/btp328}{10.1093/bioinformatics/btp328}
+
 \hypertarget{ref-R-magrittr}{}
-11. Bache SM, Wickham H. Magrittr: A forward-pipe operator for r
+13. Bache SM, Wickham H. Magrittr: A forward-pipe operator for r
 {[}Internet{]}. 2014. Available:
@@
 \hypertarget{ref-R-rlang}{}
-12. Henry L, Wickham H. Rlang: Functions for base types and core r and
+14. Henry L, Wickham H. Rlang: Functions for base types and core r and
 'tidyverse' features {[}Internet{]}. 2017. Available:
 \url{http://rlang.tidyverse.org}
-
-\hypertarget{ref-R-S4vectors}{}
-13. Pagès H, Lawrence M, Aboyoun P. S4Vectors: S4 implementation of
-vector-like and list-like objects. 2017.
-
-\hypertarget{ref-Lawrence2009-nt}{}
-14. Lawrence M, Gentleman R, Carey V. Rtracklayer: An R package for
-interfacing with genome browsers. Bioinformatics. 2009;25: 1841--1842.
-doi:\href{https://doi.org/10.1093/bioinformatics/btp328}{10.1093/bioinformatics/btp328}
 

### 2018-04-17 paper.Rmd
fleshing out example 2

@@
 q <- AnnotationHub()
+# T-cell BW file
 bw_file <- q[["AH33458"]] 
@@
 )
+
+knitr::kable(feature_tbl, booktabs = TRUE)
 ```
@@
 Here we provide examples on how to use the `plyranges` DSL to construct
-genomic data workflows.
+genomic data workflows and highlight its interoperability with existing
+Bionconductor packages.
 
@@
 
-We use the Bioconductor package _AnnotationHub_ @R-ahub to search for 
+The Bioconductor package `AnnotationHub` @R-ahub  can be used to search for 
 BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project 
-@Roadmap_Epigenomics_Consortium2015-pr. We choose to focus on assays for primary
-T CD8+ memory cells from peripheral blood. We can then read the BigWig file 
-corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation mark 
-over chromosome 10.
+@Roadmap_Epigenomics_Consortium2015-pr. Here we focus on assays for primary
+T CD8+ memory cells from peripheral blood. Using `plyranges` we will read the 
+BigWig file  corresponding to the H3 lysine 27 trimethylation (H3K27Me3) 
+methylation mark over chromosome 10.
 
@@
   set_genome_info(genome = "hg19")
-chr10_scores
 ```
@@
 
-Returning to the GRanges object containing normalised coverage scores for
-the methylation data, we can filter to find the coordinates of the peak 
+Returning to the GRanges object containing normalised coverage scores, 
+we filter to find the coordinates of the peak 
 containing the maximum coverage score. We can then find 
@@
 normalised coverage scores that are within the 5000nt region that contain
-the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).
+the max peak on chromosome 10 (figure \ref{fig:peak-viz}).
 
@@
 
-(example of group_by_overlaps + summarise)
+Another common operation in genomics data analysis is to compute data
+summaries over genomic windows. In `plyranges` this can be achieved via
+the `group_by_overlaps` operator. Continuing with the data from the 
+Human Epigenome Roadmap Consortium data, we can count the number of reads
+that fall into a fixed bins of size 10000bp over a BAM file of H3K27Me3 methylation
+marks from the H1 cell line. We extract reads that have a mapping quality
+score greater than 20:
+
+```{r}
+h1_bam <- read_bam("h1.bam") %>%
+  filter(mapq > 20) 
+```
+
+Note that the BAM file is only read into memory once we perform an operation
+on it. Next we generate our genomic bins using the `tile` function
+from `GenomicRanges`:
+
+```{r}
+bins <- tile(h1_bam, width = 10000)
+```
+
+Finally, we can use `group_by_overlaps` with `summarise` to compute the
+total number of reads with in each window.
+
+```{r}
+h1_bam_read_summary <- h1_bam %>%
+  group_by_overlaps(bins) %>%
+  summarise(n_reads = n())
+```
 
@@
 
-The _plyranges_ package is available on the Bioconductor project website 
+The `plyranges` package is available on the Bioconductor project website 
 [https://bioconductor.org](https://bioconductor.org) or can be accessed
@@
 with more complex data structures  such as the 
-_SummarizedExperiment_ class, which can be used for analysing transcriptomic
+SummarizedExperiment class, which can be used for analysing transcriptomic
 and variant data. As the `plyranges` interface encourages tidy data 

### 2018-04-17 paper.Rmd
fix some typos

@@
 Several domain specific languages (DSLs) enable scientists to process and reason
-about heterogeneous genomics data heterogeneous genomics data by expressing common
+about heterogeneous genomics data by expressing common
 operations, such as range manipulation and overlap-based joins, using
@@
 computations to a database, or operate over collections of files in
-standard formats like BED.An example of the former is the Genome Query 
+standard formats like BED. An example of the former is the Genome Query 
 Language (GQL) and its distributed implementation GenAp which use an 
@@
 either at the command line or embedded in scripts for batch processing. They 
-exist in sparse ecosystem, mostly consisting of UNIX and database tools that
+exist in a sparse ecosystem, mostly consisting of UNIX and database tools that
 lack biological semantics and operate at the level of files and
@@
 enabled the development of sophisticated tools, as evidenced by the
-hundreds of packages depending the framework. Unfortunately, the
+hundreds of packages depending on the framework. Unfortunately, the
 myriad of data structures have overlapping purposes and important but
@@
 R data structures and workflows around the notion of tidy data
-@Wickham2014-jc. A tidy dataset is defined to be to a tabular
-data structure that has observations as rows and columns as variables, and each
-tidy dataset represents measurements from a single observational unit.
+@Wickham2014-jc. A tidy dataset is defined as a tabular
+data structure that has observations as rows and columns as variables,
+and all measurements pertain to a single observational unit.
 The tidy data pattern is useful because it allows 
-us to see how the data relates to the design of an experiment
+us to see how the data relate to the design of an experiment
 and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
@@
 consisting of measurements or metadata specific to the data type or
-experiment. For example a GRanges can be used to represent a gene with
-its chromosomal coordinates, and along with its exon identifiers; or 
-a GRanges can represent an RNA-seq experiment as a set of genes
-with a corresponding count column.
+experiment. For example, a GRanges can represent a gene with
+its chromosomal coordinates, exonic structure, and a count column
+summarized from an RNA-seq experiment.
 
-By definition the GRanges follow  the tidy data pattern: it is a rectangular 
+By definition GRanges follows the tidy data pattern: it is a rectangular 
 table corresponding to a single biological context. Each row contains a

### 2018-04-18 paper.Rmd
algebra stuff from wiki - will rework

@@
 design to the existing Bioconductor infrastructure.
+
+## Genomic Relational Algebra
+
+* Arithmetic
+** Unary
+   IRanges has already explored this space and defined its own
+   "algebra", but we could revisit it. bedtools is also inspiring.
+
+   We need to make adjustments to the start, end and width. Extending
+   dplyr, this would just be =mutate()=, e.g.:
+   #+begin_src R
+   mutate(x, start=start+1L)
+   #+end_src
+   And while that could be supported, we typically want to make
+   relative adjustments.
+
+   One complication is that the start, end and width are mutually
+   dependent. Changing the start, as above, needs to change either
+   the width or the end to preserve integrity. We assume that the
+   user thinks of the range as a pair of endpoints (this is natural
+   in genomics anyway), so setting the start will modify the width
+   but preserve the end. The opposite, preserving the start but not
+   the other endpoint, is =IRanges::shift()=. Maybe we should be more
+   explicit about whether the width or the other endpoint is changed
+   by side effect.
+
+   Here are some possible verbs for changing the width:
+   * =stretch_start/end= :: Extend the range by 'x' by pulling the
+	start or the end (and leaving the other one anchored),
+   * =squeeze_start/end= :: Opposite of stretch (could just use
+	negative value for stretch, but being explicit seems
+	preferable in many cases),
+
+   Not sure whether =squeeze= is actually useful, as we typically
+   want to stretch a range to consider context, not squeeze it.
+
+   Or maybe it would be more obvious if we supported the notion of
+   anchoring, similar to grouping. This would require decorating the
+   object. For example,
+   #+begin_src R
+     stretch(anchor_end(x), y)
+   #+end_src
+   This pushes information from the syntax to the object state. While
+   it simplifies the grammar, object behavior is more complex and
+   implicit. But if the code is part of the same chain, it seems
+   intuitive.
+
+   There would be three ways to anchor:
+   * =anchor_start()=
+   * =anchor_end()=
+   * =anchor_center()=
+
+   Anchoring the midpoint would enable symmetric zooming.
+
+   This is basically a deconstruction of =IRanges::resize()=, except
+   =resize()= is absolute. Perhaps =set_width()= could play
+   the role of =resize()=. It is not clear whether we need
+   =set_start()= and =set_end()=.
+
+   When strand is involved, we will need something like:
+   * =anchor_3p= (start for negative strand)
+   * =anchor_5p= (end for negative strand)
+
+   Anchoring the width would not make sense, because (1) it is not a
+   point that can be intuitively anchored and (2) stretching and
+   squeezing obviously modify the width, so it would not apply to
+   those.
+
+   For width-invariant changes, there is =shift()= from the IRanges
+   API. That is a symmetric operation (positive for right, negative
+   for left). We could introduce:
+   * =shift_left()=
+   * =shift_right()=
+   If we wanted the direction to be explicit.
+
+   Note that with GRanges, strand can modulate direction, and we need
+   to distinguish left/right (chromosomal coordinates) from the
+   direction of transcription:
+   * =shift_upstream()=
+   * =shift_downstream()=
+
+     | Fluent                   | IRanges       | bedtools |
+     |--------------------------+---------------+----------|
+     | stretch(anchor_start())  | -             | slop -r  |
+     | stretch(anchor_end())    | -             | slop -l  |
+     | stretch(anchor_center()) | +()           | slop -b  |
+     | set_width(anchor_*)      | resize(fix=*) | -        |
+     | shift_right()            | shift()       | shift -s |
+     | shift_left()             | -             | -        |
+     | shift_up/downstream()    | -             | -        |
+
+  Note that =resize,GRanges()= and =slop -s= use the strand for
+  direction, i.e, =anchor_3p()= and =anchor_5p()=. Bedtools supports
+  width-relative adjustment with =slop -pct=; we will let the user
+  be explicit, e.g.: =stretch(x, 2*width(x))=.
+
+  One problem is that can be stretched or shifted out of bounds, or squeezed out of existence.
+  Currently, GenomicRanges will warn the user when a range is pushed
+  out of bounds and suggest calling =trim()=. We should probably
+  throw an error in this case, because it is never a good idea for
+  an object to be in a (semi)-invalid state. The resolution needs
+  to be explicit, as part of the function call. By default,
+  we throw an error. There could be variants of all of these
+  operations with the post fix =_up_to= that imply the trim, 
+  e.g., =stretch_up_to()= or =set_width_up_to()=. Another
+  resolution would be to drop the ranges that would become
+  out of bounds. Those could use the =_or_drop()= postfix.
+
+  We could also use this anchoring notion to define a new verb
+  =flip()= that generates flanking regions, but that will cause
+  semantic issues with strandedness. Instead, we probably need
+  =flank_left()= and =flank_right()=. With strand, there would be
+  =flank_before()= and =flank_after()=.
+
+** Binary
+   Binary operations between two ranges mostly align with set
+   operations that treat ranges as sets of integers. These are
+   parallel/vectorized operations between two vectors/tables of
+   ranges, so are not to be confused with the aggregating set
+   operations described later.
+
+   Currently, these set operations have the =p= prefix, e.g.,
+   =punion()=. We could make the binary nature more obvious by
+   aliasing them as operators:
+   * =%union%=
+   * =%intersect%=
+   * =%setdiff%=
+
+   There are also operations that relate to the relative positions of
+   two ranges:
+   * =between()=: the range in the gap between the two ranges
+     * To get the separation, just do =width(between(x, y))=
+   * =span()=: a range that spans both ranges (and the gap)
+
+   As a note, to express a binary operation involving the subject of a
+   =mutate()=, =summarize()= or whatever call, we could use the pronoun:
+   #+begin_src R
+   mutate(x, foo = width(. %intersect% other_ranges))
+   #+end_src
 

### 2018-04-19 paper.Rmd
reworking opportunities --> limitations

@@
 
-## Limitations
+## Opportunities
 
@@
 means that computations are performed and evaluated in the context of the 
-GRanges objects, which emphasises the interactive nature of the interface.
-This has the disadvantage that programming with `plyranges` becomes more difficult 
-because a user needs to know about non-standard evaluation and user defined
-functions can be difficult to debug.
+GRanges objects; emphasising the interactive nature of our API.
+Consequently, when programming with `plyranges` a user needs to be aware of
+how non-standard evaluation in R works and how to adapt their code 
+accordingly. However, with the rise of R packages like `rlang` this process
+is becoming less difficult.
+
+While GRanges are an intuitive representation for data measured on genomic regions,
+more flexible data structures are required to represent data from multiple sample
+experiments. The Bionconductor class SummarizedExperiment is the canonical 
+data structure for representing data for combining multiomic measurements 
+from multiple samples. The grammar and design of the `plryanges` DSL can be
+naturally extended to the SummarizedExperiment.
 

### 2018-04-19 paper.Rmd
ideas for endomorphism section

@@
 
+* compute coverage as an example here
+
 This design pattern strongly deviates from the design of the OO interface in 
@@
 Finally, we can use `group_by_overlaps` with `summarise` to compute the
-total number of reads with in each window.
+total number of reads within each window.
 
@@
 ```
+
+https://support.bioconductor.org/p/71601/ also useful example?
 

### 2018-04-20 paper.Rmd
adding data for other examples

@@
 ```{r setup, include = FALSE}
-knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>", 
-                      fig.width = 5, fig.height = 3, fig.align = "center")
-options(digits = 2)
+library(tibble)
+library(knitr)
+opts_chunk$set(message = FALSE, 
+               warning = FALSE, 
+               comment = "#>", 
+               fig.width = 5, 
+               fig.height = 3, 
+               fig.align = "center")
+# for peak figures
 library(ggbio)
-library(ggplot2)
+
 library(plyranges)
 
+# -- Data from the Human Epigenomics RoadMap consortium
+# retrieving a BigWigFile from AnnotationHub
 library(AnnotationHub)
@@
 bw_file <- q[["AH33458"]]
+
+# a BamFile from the H1 cell line
+h1_bam <- "data/GSM433167_BI.H3K27me3.bam"
+
+# Genotype Files accross h1 cell line methylation marks
+h1_geno <- "data/GSM1463263_JS-ESCH1.txt.gz"
+
 ```
@@
 ```{r, echo = FALSE, message = FALSE, warning = FALSE}
-library(tibble)
 feature_tbl <- tribble(

### 2018-04-20 paper.Rmd
add acknowledgments

--- 2018-04-19 paper.Rmd
@@
 
-     | Fluent                   | IRanges       | bedtools |
-     |--------------------------+---------------+----------|
-     | stretch(anchor_start())  | -             | slop -r  |
-     | stretch(anchor_end())    | -             | slop -l  |
-     | stretch(anchor_center()) | +()           | slop -b  |
-     | set_width(anchor_*)      | resize(fix=*) | -        |
-     | shift_right()            | shift()       | shift -s |
-     | shift_left()             | -             | -        |
-     | shift_up/downstream()    | -             | -        |
-
   Note that =resize,GRanges()= and =slop -s= use the strand for
@@
 
-```{r, echo = FALSE, message = FALSE, warning = FALSE}
+```{r, echo = FALSE, eval = FALSE}
 library(tibble)
@@
 
-knitr::kable(feature_tbl, booktabs = TRUE)
+knitr::kable(feature_tbl, format = "latex")
 ```
@@
 
-```{r}
+```{r, eval = FALSE}
 h1_bam <- read_bam("h1.bam") %>%
@@
 
-```{r}
+```{r, eval = FALSE}
 bins <- tile(h1_bam, width = 10000)
@@
 
-```{r}
+```{r, eval = FALSE}
 h1_bam_read_summary <- h1_bam %>%
@@
 
+# Acknowledgements
+
+We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
+Institute and Dr Paul Harrison for their feedback on earlier drafts of
+this work. We would also like to thank Lori Shepherd and Hèrve Pages
+for the code review they performed. This report was written with
+`knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio.
+
 # References {#references .unnumbered}

### 2018-04-20 paper.Rmd
algebra section

@@
 
-* Arithmetic
-** Unary
-   IRanges has already explored this space and defined its own
-   "algebra", but we could revisit it. bedtools is also inspiring.
+The `plyranges` DSL defines an expressive algebra for performing genomic
+arithemtic withn and between GRanges objects. Often in analyses of
+genomic regions we are required to make adjustments to the start, end,
+and width columns. This is supported directly in `plyranges` via extending
+the `mutate()` function in `dplyr` (table 1). However, often adjustments
+need to be relative: there is a mutual dependence between the start, end and
+width columns over a genomic region. Changing the start,column needs to change
+either the width or the end to preserve integrity of the object. The GRanges 
+object by design, causes a user to think of the range as a pair of endpoints, 
+and hence we have developed operators that perform coordinate transfomations with 
+the notion of an 'anchoring' combinator.
 
-   We need to make adjustments to the start, end and width. Extending
-   dplyr, this would just be =mutate()=, e.g.:
-   #+begin_src R
-   mutate(x, start=start+1L)
-   #+end_src
-   And while that could be supported, we typically want to make
-   relative adjustments.
+For width modification, we can stretch the range by a fixed amount or
+resize the width of a range. The 'anchoring' combinator preserves one of the 
+endpoints or the  midpoint of a range when altering performing
+width modification. It can also be used to preserve coordinates over strand, 
+for example, anchoring over the three prime end of a genomic region, will 
+preserve the start on the negative strand and end on the positive strand. 
+Anchoring the width would not make sense, because it is not a point that can be
+intuitively anchored and stretching and resizing obviously modify the width, 
+so it would not apply to those.
 
-   One complication is that the start, end and width are mutually
-   dependent. Changing the start, as above, needs to change either
-   the width or the end to preserve integrity. We assume that the
-   user thinks of the range as a pair of endpoints (this is natural
-   in genomics anyway), so setting the start will modify the width
-   but preserve the end. The opposite, preserving the start but not
-   the other endpoint, is =IRanges::shift()=. Maybe we should be more
-   explicit about whether the width or the other endpoint is changed
-   by side effect.
+Transformations that are width invariant are designed to be explicit
+about the direction to modify and are performed via the shift operator. 
+For GRanges objects strand modulates direction and we distinguish left/right 
+(chromosomal coordinates) from the direction of transcription. To generate
+flanking regions, we use the flank operator which is also explicit about the
+direction of the input range.
 
-   Here are some possible verbs for changing the width:
-   * =stretch_start/end= :: Extend the range by 'x' by pulling the
-	start or the end (and leaving the other one anchored),
-   * =squeeze_start/end= :: Opposite of stretch (could just use
-	negative value for stretch, but being explicit seems
-	preferable in many cases),
-
-   Not sure whether =squeeze= is actually useful, as we typically
-   want to stretch a range to consider context, not squeeze it.
-
-   Or maybe it would be more obvious if we supported the notion of
-   anchoring, similar to grouping. This would require decorating the
-   object. For example,
-   #+begin_src R
-     stretch(anchor_end(x), y)
-   #+end_src
-   This pushes information from the syntax to the object state. While
-   it simplifies the grammar, object behavior is more complex and
-   implicit. But if the code is part of the same chain, it seems
-   intuitive.
-
-   There would be three ways to anchor:
-   * =anchor_start()=
-   * =anchor_end()=
-   * =anchor_center()=
-
-   Anchoring the midpoint would enable symmetric zooming.
-
-   This is basically a deconstruction of =IRanges::resize()=, except
-   =resize()= is absolute. Perhaps =set_width()= could play
-   the role of =resize()=. It is not clear whether we need
-   =set_start()= and =set_end()=.
-
-   When strand is involved, we will need something like:
-   * =anchor_3p= (start for negative strand)
-   * =anchor_5p= (end for negative strand)
-
-   Anchoring the width would not make sense, because (1) it is not a
-   point that can be intuitively anchored and (2) stretching and
-   squeezing obviously modify the width, so it would not apply to
-   those.
-
-   For width-invariant changes, there is =shift()= from the IRanges
-   API. That is a symmetric operation (positive for right, negative
-   for left). We could introduce:
-   * =shift_left()=
-   * =shift_right()=
-   If we wanted the direction to be explicit.
-
-   Note that with GRanges, strand can modulate direction, and we need
-   to distinguish left/right (chromosomal coordinates) from the
-   direction of transcription:
-   * =shift_upstream()=
-   * =shift_downstream()=
-
-  Note that =resize,GRanges()= and =slop -s= use the strand for
-  direction, i.e, =anchor_3p()= and =anchor_5p()=. Bedtools supports
-  width-relative adjustment with =slop -pct=; we will let the user
-  be explicit, e.g.: =stretch(x, 2*width(x))=.
-
-  One problem is that can be stretched or shifted out of bounds, or squeezed out of existence.
-  Currently, GenomicRanges will warn the user when a range is pushed
-  out of bounds and suggest calling =trim()=. We should probably
-  throw an error in this case, because it is never a good idea for
-  an object to be in a (semi)-invalid state. The resolution needs
-  to be explicit, as part of the function call. By default,
-  we throw an error. There could be variants of all of these
-  operations with the post fix =_up_to= that imply the trim, 
-  e.g., =stretch_up_to()= or =set_width_up_to()=. Another
-  resolution would be to drop the ranges that would become
-  out of bounds. Those could use the =_or_drop()= postfix.
-
-  We could also use this anchoring notion to define a new verb
-  =flip()= that generates flanking regions, but that will cause
-  semantic issues with strandedness. Instead, we probably need
-  =flank_left()= and =flank_right()=. With strand, there would be
-  =flank_before()= and =flank_after()=.
-
-** Binary
 Binary operations between two ranges mostly align with set
@@
 parallel/vectorized operations between two vectors/tables of
-   ranges, so are not to be confused with the aggregating set
-   operations described later.
-
-   Currently, these set operations have the =p= prefix, e.g.,
-   =punion()=. We could make the binary nature more obvious by
-   aliasing them as operators:
-   * =%union%=
-   * =%intersect%=
-   * =%setdiff%=
-
-   There are also operations that relate to the relative positions of
-   two ranges:
-   * =between()=: the range in the gap between the two ranges
-     * To get the separation, just do =width(between(x, y))=
-   * =span()=: a range that spans both ranges (and the gap)
-
-   As a note, to express a binary operation involving the subject of a
-   =mutate()=, =summarize()= or whatever call, we could use the pronoun:
-   #+begin_src R
-   mutate(x, foo = width(. %intersect% other_ranges))
-   #+end_src
-
-## Cohesion
-
-A function is cohesive if it performs a singular task and does not produce
-any side-effects. In the `plyranges` DSL our algebra for performing
-genomic arithmetic is a key example of cohesion. We define anchoring operators
-that decorate a GRanges object with an 'anchor' and modify the semantics
-of performing genomic arithmetic. The anchoring operators amount to fixing a GRanges
-object by its start, center, or end coordinates (or fixing these coordinates
-by strand). This enables any arithmetic function that performs a 
-coordinate transformation to remain cohesive: that is they always perform
-the same operation regardless of whether a GRanges object is anchored or not.
-The only difference is the result of performing the arithmetic changes when
-contextual information is added to the GRanges object.
-
-Another example of our algebra altering object semantics, while maintaining 
-cohesion is through the 'group_by' operator. Like anchoring, this operator 
-decorates a GRanges object with a column name (or names) that defines a 
-partitioning of the  GRanges by the unique values in the column(s). Functions 
-defined in the  `plyranges` DSL that perform restriction, aggregation, or 
-column modification still perform those singular tasks, however grouping changes
-how those tasks are performed.
+ranges, and are not to be confused with the aggregating 
+operations defined in table x. We make the binary nature of these operators
+by having in-fix functions in R for the parallel union, intersect and
+asymetric set difference. There are also operations that relate to the 
+relative positions of two ranges such as the 'between' operator which
+returns the range in the gap between two ranges and the 'span' operator 
+a range that spans both ranges. The binary operations can be expressed within
+the functions in table x.
 
@@
 
-more comparison to Bioconductor API here? table of join ops?
+## Cohesion
+
+A function is cohesive if it performs a singular task and does not produce
+any side-effects. In the `plyranges` DSL our algebra for performing
+genomic arithmetic is a key example of cohesion. We define anchoring operators
+that decorate a GRanges object with an 'anchor' and modify the semantics
+of performing genomic arithmetic. The anchoring operators amount to fixing a GRanges
+object by its start, center, or end coordinates (or fixing these coordinates
+by strand). This enables any arithmetic function that performs a 
+coordinate transformation to remain cohesive: that is they always perform
+the same operation regardless of whether a GRanges object is anchored or not.
+The only difference is the result of performing the arithmetic changes when
+contextual information is added to the GRanges object.
+
+Another example of our algebra altering object semantics, while maintaining 
+cohesion is through the 'group_by' operator. Like anchoring, this operator 
+decorates a GRanges object with a column name (or names) that defines a 
+partitioning of the  GRanges by the unique values in the column(s). Functions 
+defined in the  `plyranges` DSL that perform restriction, aggregation, or 
+column modification still perform those singular tasks, however grouping changes
+how those tasks are performed.
 
@@
 `GenomicRanges`, where many methods return a new class that is unfamiliar to 
-the user upon return. These low-level classes enable efficient computing at the cost
-of complexity and are abstracted away in `plyranges` interface.
+the user upon return. These low-level classes enable efficient computing at 
+the cost of complexity and are abstracted away in `plyranges` interface.
 
@@
 ```{r, echo = FALSE, eval = FALSE}
-library(tibble)
 feature_tbl <- tribble(
@@
 ```{r, eval = FALSE}
-h1_bam <- read_bam("h1.bam") %>%
+alignments <- read_bam(h1_bam) %>%
   filter(mapq > 20) 

### 2018-04-23 paper.Rmd
array example

@@
 h1_bam <- "data/GSM433167_BI.H3K27me3.bam"
-
-# Genotype Files accross h1 cell line methylation marks
-h1_geno <- "data/GSM1463263_JS-ESCH1.txt.gz"
-
 ```
@@
 
+## Quality Control Metrics
+
+```{r, echo = FALSE, warning = FALSE, cache = TRUE}
+# prepare a GRanges object from array data
+library(readr)
+# annotation information available here
+# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952
+# chromosome zero are 'problem' probes and should be filtered
+anno <- read_tsv("data/GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz",
+                 trim_ws = TRUE,
+                 col_types = c("cccc----")) %>%
+  dplyr::filter(Chr != "0")
+
+# genotypes for H1 cell line, data starts on line 10
+# data obtained from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1463263
+geno <- read_tsv("data/GSM1463263_JS-ESCH1.txt.gz", skip = 10)
+
+# rename columns
+geno <- geno %>% 
+  dplyr::select(name = `SNP Name`, 
+         score = `GC Score`, 
+         baf = `B Allele Freq`, 
+         logR = `Log R Ratio`)
+
+# merge everything
+complete_vranges_df <- dplyr::inner_join(geno, anno, by = c("name" = "Name")) %>%
+  dplyr::rename(alleles = Alleles, seqnames = Chr, start = MapInfo)
+
+h1_snp_array <- as_granges(complete_vranges_df,
+                           start = as.integer(start),
+                           end = as.integer(start))
+h1_snp_array <- h1_snp_array %>% 
+  mutate(ref = stringr::str_replace(alleles, "\\[(.)/.*", "\\1"),
+         alt = stringr::str_replace(alleles, ".*/(.)\\]", "\\1"))
+```
+
+We have created a GRanges object from genotyping performed on the
+H1 cell line, consisting of approximately two million single nucleotide
+polymorphisms (SNP) and short insertion/deletions (indel). The GRanges object consists
+of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
+log relative intensity of the probes, GC content score over a probe, and the
+name of the probe. We can use this information to compute the transition-transversion
+ratio, a quality control metric, within each chromosome in GRanges object.
+
+First we filter out any insertion or deletion alleles and 
+the SNPs present on the mitochondria then create a logical vector corresponding
+to whether there is a transition event. 
+```{r}
+h1_snp_array <- h1_snp_array %>%
+  filter(!(ref %in% c("I", "D")), seqnames != "M") %>%
+  mutate(transition = (ref %in% c("A", "G") & alt %in% c("G","A")) |
+                      (ref %in% c("C","T") & alt %in% c("T", "C")))
+```
+
+We can then compute the transition-transversion ratio using the `group_by`
+and `summarise` pattern, it is computed as the total number of transition
+SNPs divided by the total number of transversion SNPs within each chromsome
+(figure \ref{fig:titv-viz}).
+
+```{r}
+ti_tv_results <- h1_snp_array %>% 
+  group_by(seqnames) %>%
+  summarise(n_snps = n(),
+            ti_tv = sum(transition) / sum(!transition)) 
+```
+
+```{r titv-viz, echo = FALSE, fig.cap = "Dot plot of chromosomes ordered by estimated transition-transversion ratio. A white reference line is drawn at the expected ratio for a human exome."}
+ti_tv_results %>% 
+  as.data.frame() %>% 
+  ggplot(aes(forcats::fct_reorder(seqnames, ti_tv), ti_tv)) +
+  geom_point() + 
+  geom_hline(yintercept = 3, colour = "white", size = 4) +
+  labs(x = "Chromosome", y = "Transition Transverion Ratio") +
+  coord_flip()
+
+```
+
 ## Computing Windowed Statistics
@@
 
-## Transition Transversion Ratio
-
-(example of group_by + mutate + summarise and working on an object that inherits
-from GRanges)
-
 # Availablilty and Future Work
@@
 for the code review they performed. This report was written with
-`knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio.
+`knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio. All code
+required to reproduce this report is available at 
+https://github.com/sa-lee/plyranges-paper.
 

### 2018-04-23 paper.Rmd
get rid of line

@@
   coord_flip() 
-
 ```

### 2018-04-23 paper.Rmd
overlap figure now in place.

@@
 the query range the subject range 'hits' the query range as a primary key 
-(figure *need to make figure*). The notion of 'hits' and hence
+(figure \ref{fig:olaps-fig}). The notion of 'hits' and hence
 the resulting key is determined by the type of join (preceding, nearest, follows
@@
 
+```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators.
+Each join takes query and subject range as input (black and grey rectangles, respectively) with associated metadata (pink and orange rectangles respectively).
+An index for the join is computed via returning a Hits object, which contains the indices of where the subject overlaps the query range.
+This index is used to expand the query or (take the intersect) of the query ranges with subject ranges and return metadata associated with each input.
+This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
+knitr::include_graphics("olaps-figure.png")
+```
 We extend this idea by introducing three types of overlap joins: inner,
-intersect and left. The inner join considers a hit to be any overlap between
-subject and query and expands the query range to have length equal to the 
-number of hits. The intersect join uses the same definition of a hit as
-the inner join but returns a range where the start and end coordinates
+intersect and left (figure \ref{fig:olaps-fig}). The inner join considers a
+hit to be any overlap between subject and query and expands the query range to 
+have length equal to the  number of hits. The intersect join uses the same 
+definition of a hit as the inner join but returns a range where the start and end coordinates
 are the intersect of coordinates in the subject that hit the query. Finally,

### 2018-04-23 paper.tex
overlap figure now in place.

@@
 Several domain specific languages (DSLs) enable scientists to process
-and reason about heterogeneous genomics data heterogeneous genomics data
-by expressing common operations, such as range manipulation and
-overlap-based joins, using the vocabulary of genomics. Their
-implementations either delegate computations to a database, or operate
-over collections of files in standard formats like BED.An example of the
-former is the Genome Query Language (GQL) and its distributed
-implementation GenAp which use an SQL-like syntax for fast retrieval of
-information of unprocessed sequencing data {[}1{]}; {[}2{]}. Similarly,
-the Genometric Query Language (GMQL) implements a relational algebra for
-combining genomic datasets {[}3{]}. The command line application
-BEDtools develops an extensive algebra for performing arithmetic between
-two or more sets of genomic regions {[}4{]}. All of the aforementioned
-DSLs are designed to be evaluated either at the command line or embedded
-in scripts for batch processing. They exist in sparse ecosystem, mostly
-consisting of UNIX and database tools that lack biological semantics and
-operate at the level of files and database tables.
+and reason about heterogeneous genomics data by expressing common
+operations, such as range manipulation and overlap-based joins, using
+the vocabulary of genomics. Their implementations either delegate
+computations to a database, or operate over collections of files in
+standard formats like BED. An example of the former is the Genome Query
+Language (GQL) and its distributed implementation GenAp which use an
+SQL-like syntax for fast retrieval of information of unprocessed
+sequencing data {[}1{]}; {[}2{]}. Similarly, the Genometric Query
+Language (GMQL) implements a relational algebra for combining genomic
+datasets {[}3{]}. The command line application BEDtools develops an
+extensive algebra for performing arithmetic between two or more sets of
+genomic regions {[}4{]}. All of the aforementioned DSLs are designed to
+be evaluated either at the command line or embedded in scripts for batch
+processing. They exist in a sparse ecosystem, mostly consisting of UNIX
+and database tools that lack biological semantics and operate at the
+level of files and database tables.
 
@@
 enabled the development of sophisticated tools, as evidenced by the
-hundreds of packages depending the framework. Unfortunately, the myriad
-of data structures have overlapping purposes and important but obscure
-differences in behaviour that often confuse the typical end user.
+hundreds of packages depending on the framework. Unfortunately, the
+myriad of data structures have overlapping purposes and important but
+obscure differences in behaviour that often confuse the typical end
+user.
 
@@
 data structures and workflows around the notion of tidy data {[}10{]}. A
-tidy dataset is defined to be to a tabular data structure that has
-observations as rows and columns as variables, and each tidy dataset
-represents measurements from a single observational unit. The tidy data
-pattern is useful because it allows us to see how the data relates to
-the design of an experiment and the variables measured. The
-\texttt{dplyr} package {[}11{]} defines an API that maps notions from
-the general relational algebra to operations on tidy data. It expresses
-each operation as a cohesive, endomorphic verb. Taken together these
-features enable a user to write human readable analysis workflows.
+tidy dataset is defined as a tabular data structure that has
+observations as rows and columns as variables, and all measurements
+pertain to a single observational unit. The tidy data pattern is useful
+because it allows us to see how the data relate to the design of an
+experiment and the variables measured. The \texttt{dplyr} package
+{[}11{]} defines an API that maps notions from the general relational
+algebra to operations on tidy data. It expresses each operation as a
+cohesive, endomorphic verb. Taken together these features enable a user
+to write human readable analysis workflows.
 
@@
 arbitrary set of additional columns, consisting of measurements or
-metadata specific to the data type or experiment. For example a GRanges
-can be used to represent a gene with its chromosomal coordinates, and
-along with its exon identifiers; or a GRanges can represent an RNA-seq
-experiment as a set of genes with a corresponding count column.
+metadata specific to the data type or experiment. For example, a GRanges
+can represent a gene with its chromosomal coordinates, exonic structure,
+and a count column summarized from an RNA-seq experiment.
 
-By definition the GRanges follow the tidy data pattern: it is a
-rectangular table corresponding to a single biological context. Each row
-contains a single observation and each column is a variable about that
-observation. Hence, we have designed the \texttt{plyranges} DSL to
-extend the grammar and design principles of \texttt{dplyr}: cohesion,
-consistency, endomorphism, and fluency. All of these principles are
-defined and discussed below in the context of the GRanges class. Where
-applicable we contrast our design to the existing Bioconductor
-infrastructure.
+By definition GRanges follows the tidy data pattern: it is a rectangular
+table corresponding to a single biological context. Each row contains a
+single observation and each column is a variable about that observation.
+Hence, we have designed the \texttt{plyranges} DSL to extend the grammar
+and design principles of \texttt{dplyr}: cohesion, consistency,
+endomorphism, and fluency. All of these principles are defined and
+discussed below in the context of the GRanges class. Where applicable we
+contrast our design to the existing Bioconductor infrastructure.
+
+\subsection{Genomic Relational
+Algebra}\label{genomic-relational-algebra}
+
+The \texttt{plyranges} DSL defines an expressive algebra for performing
+genomic arithemtic withn and between GRanges objects. Often in analyses
+of genomic regions we are required to make adjustments to the start,
+end, and width columns. This is supported directly in \texttt{plyranges}
+via extending the \texttt{mutate()} function in \texttt{dplyr} (table
+1). However, often adjustments need to be relative: there is a mutual
+dependence between the start, end and width columns over a genomic
+region. Changing the start,column needs to change either the width or
+the end to preserve integrity of the object. The GRanges object by
+design, causes a user to think of the range as a pair of endpoints, and
+hence we have developed operators that perform coordinate transfomations
+with the notion of an `anchoring' combinator.
+
+For width modification, we can stretch the range by a fixed amount or
+resize the width of a range. The `anchoring' combinator preserves one of
+the endpoints or the midpoint of a range when altering performing width
+modification. It can also be used to preserve coordinates over strand,
+for example, anchoring over the three prime end of a genomic region,
+will preserve the start on the negative strand and end on the positive
+strand. Anchoring the width would not make sense, because it is not a
+point that can be intuitively anchored and stretching and resizing
+obviously modify the width, so it would not apply to those.
+
+Transformations that are width invariant are designed to be explicit
+about the direction to modify and are performed via the shift operator.
+For GRanges objects strand modulates direction and we distinguish
+left/right (chromosomal coordinates) from the direction of
+transcription. To generate flanking regions, we use the flank operator
+which is also explicit about the direction of the input range.
+
+Binary operations between two ranges mostly align with set operations
+that treat ranges as sets of integers. These are parallel/vectorized
+operations between two vectors/tables of ranges, and are not to be
+confused with the aggregating operations defined in table x. We make the
+binary nature of these operators by having in-fix functions in R for the
+parallel union, intersect and asymetric set difference. There are also
+operations that relate to the relative positions of two ranges such as
+the `between' operator which returns the range in the gap between two
+ranges and the `span' operator a range that spans both ranges. The
+binary operations can be expressed within the functions in table x.
+
+Our algebra recasts the actions of finding overlaps or nearest
+neighbours between two genomic regions as a relational join operator.
+The join operator acts on two GRanges object, a query and a subject. The
+join operator is relational in the sense that metadata from the query
+and subject ranges is retained in the joined range. The commonality
+between all join operators in the \texttt{plyranges} DSL is to use the
+index of where in the query range the subject range `hits' the query
+range as a primary key (figure \ref{fig:olaps-fig}). The notion of
+`hits' and hence the resulting key is determined by the type of join
+(preceding, nearest, follows overlaps).
+
+\begin{figure}
+
+{\centering \includegraphics[width=400pt]{olaps-figure}
+
+}
+
+\caption{Illustration of the three overlap join operators.
+Each join takes query and subject range as input (black and grey rectangles, respectively) with associated metadata (pink and orange rectangles respectively).
+An index for the join is computed via returning a Hits object, which contains the indices of where the subject overlaps the query range.
+This index is used to expand the query or (take the intersect) of the query ranges with subject ranges and return metadata associated with each input.
+This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations.}\label{fig:olaps-fig}
+\end{figure}
+
+We extend this idea by introducing three types of overlap joins: inner,
+intersect and left. The inner join considers a hit to be any overlap
+between subject and query and expands the query range to have length
+equal to the number of hits. The intersect join uses the same definition
+of a hit as the inner join but returns a range where the start and end
+coordinates are the intersect of coordinates in the subject that hit the
+query. Finally, the overlap left join is akin to left outer join in
+Cobb's relational algebra: it performs an overlap inner join but also
+returns all query ranges that are not hit by the subject.
+
+The behaviour of a join operator can be further altered with additional
+suffixes by restricting or expanding how a subject `hits' a query. For
+example, for overlap joins we can use suffixes to encapsulate Allan's
+Interval Algebra: to consider subject `hits' that are entirely `within'
+a query range, the `within' prefix is used. We also use the `directed'
+suffix to explicitly include notions of strandedness in our algebra.
 
@@
 are performed.
-
-Our algebra recasts the actions of finding overlaps or nearest
-neighbours between two genomic regions as a relational join operator.
-The join operator acts on two GRanges object, a query and a subject. The
-join operator is relational in the sense that metadata from the query
-and subject ranges is retained in the joined range. The commonality
-between all join operators in the \texttt{plyranges} DSL is to use the
-index of where in the query range the subject range `hits' the query
-range as a primary key (figure \emph{need to make figure}). The notion
-of `hits' and hence the resulting key is determined by the type of join
-(preceding, nearest, follows overlaps).
-
-We extend this idea by introducing three types of overlap joins: inner,
-intersect and left. The inner join considers a hit to be any overlap
-between subject and query and expands the query range to have length
-equal to the number of hits. The intersect join uses the same definition
-of a hit as the inner join but returns a range where the start and end
-coordinates are the intersect of coordinates in the subject that hit the
-query. Finally, the overlap left join is akin to left outer join in
-Cobb's relational algebra: it performs an overlap inner join but also
-returns all query ranges that are not hit by the subject.
-
-The behaviour of a join operator can be further altered with additional
-suffixes by restricting or expanding how a subject `hits' a query. For
-example, for overlap joins we can use suffixes to encapsulate Allan's
-Interval Algebra: to consider subject `hits' that are entirely `within'
-a query range, the `within' prefix is used. We also use the `directed'
-suffix to explicitly include notions of strandedness in our algebra.
-
-more comparison to Bioconductor API here? table of join ops?
 
@@
 
+\begin{itemize}
+\tightlist
+\item
+  compute coverage as an example here
+\end{itemize}
+
 This design pattern strongly deviates from the design of the OO
@@
 
-\subsection{Limitations}\label{limitations}
+\subsection{Opportunities}\label{opportunities}
 
@@
 means that computations are performed and evaluated in the context of
-the GRanges objects, which emphasises the interactive nature of the
-interface. This has the disadvantage that programming with
-\texttt{plyranges} becomes more difficult because a user needs to know
-about non-standard evaluation and user defined functions can be
-difficult to debug.
+the GRanges objects; emphasising the interactive nature of our API.
+Consequently, when programming with \texttt{plyranges} a user needs to
+be aware of how non-standard evaluation in R works and how to adapt
+their code accordingly. However, with the rise of R packages like
+\texttt{rlang} this process is becoming less difficult.
+
+While GRanges are an intuitive representation for data measured on
+genomic regions, more flexible data structures are required to represent
+data from multiple sample experiments. The Bionconductor class
+SummarizedExperiment is the canonical data structure for representing
+data for combining multiomic measurements from multiple samples. The
+grammar and design of the \texttt{plryanges} DSL can be naturally
+extended to the SummarizedExperiment.
 
@@
 Here we provide examples on how to use the \texttt{plyranges} DSL to
-construct genomic data workflows.
+construct genomic data workflows and highlight its interoperability with
+existing Bionconductor packages.
 
@@
 
-We use the Bioconductor package \emph{AnnotationHub} {[}15{]} to search
-for BigWig files from ChIP-Seq experiments from the Human Epigenome
-Roadmap project {[}16{]}. We choose to focus on assays for primary T
-CD8+ memory cells from peripheral blood. We can then read the BigWig
-file corresponding to the H3 lysine 27 trimethylation (H3K27Me3)
-methylation mark over chromosome 10.
+The Bioconductor package \texttt{AnnotationHub} {[}15{]} can be used to
+search for BigWig files from ChIP-Seq experiments from the Human
+Epigenome Roadmap project {[}16{]}. Here we focus on assays for primary
+T CD8+ memory cells from peripheral blood. Using \texttt{plyranges} we
+will read the BigWig file corresponding to the H3 lysine 27
+trimethylation (H3K27Me3) methylation mark over chromosome 10.
 
@@
 \StringTok{  }\KeywordTok{set_genome_info}\NormalTok{(}\DataTypeTok{genome =} \StringTok{"hg19"}\NormalTok{)}
-\NormalTok{chr10_scores}
 \end{Highlighting}
 \end{Shaded}
-
-\begin{verbatim}
-#> GRanges object with 5789841 ranges and 1 metadata column:
-#>             seqnames                 ranges strand |              score
-#>                <Rle>              <IRanges>  <Rle> |          <numeric>
-#>         [1]    chr10         [    1, 60602]      * | 0.0422799997031689
-#>         [2]    chr10         [60603, 60781]      * |  0.163240000605583
-#>         [3]    chr10         [60782, 60816]      * |  0.372139990329742
-#>         [4]    chr10         [60817, 60995]      * |  0.163240000605583
-#>         [5]    chr10         [60996, 61625]      * | 0.0422799997031689
-#>         ...      ...                    ...    ... .                ...
-#>   [5789837]    chr10 [135524723, 135524734]      * |  0.144319996237755
-#>   [5789838]    chr10 [135524735, 135524775]      * |  0.250230014324188
-#>   [5789839]    chr10 [135524776, 135524784]      * |  0.427789986133575
-#>   [5789840]    chr10 [135524785, 135524806]      * |  0.730019986629486
-#>   [5789841]    chr10 [135524807, 135524837]      * |   1.03103005886078
-#>   -------
-#>   seqinfo: 25 sequences from hg19 genome
-\end{verbatim}
 
@@
 
-Returning to the GRanges object containing normalised coverage scores
-for the methylation data, we can filter to find the coordinates of the
-peak containing the maximum coverage score. We can then find a 5000 nt
-region centered around the maximum position by anchoring and modifying
-the the width.
+Returning to the GRanges object containing normalised coverage scores,
+we filter to find the coordinates of the peak containing the maximum
+coverage score. We can then find a 5000 nt region centered around the
+maximum position by anchoring and modifying the the width.
 
@@
 normalised coverage scores that are within the 5000nt region that
-contain the max peak on chromosome 10 (visualised in figure
-\ref{fig:peak-viz}).
+contain the max peak on chromosome 10 (figure \ref{fig:peak-viz}).
 
@@
 
+\subsection{Quality Control Metrics}\label{quality-control-metrics}
+
+We have created a GRanges object from genotyping performed on the H1
+cell line, consisting of approximately two million single nucleotide
+polymorphisms (SNP) and short insertion/deletions (indel). The GRanges
+object consists of 7 columns, relating to the alleles of a SNP or indel,
+the B-allele frequency, log relative intensity of the probes, GC content
+score over a probe, and the name of the probe. We can use this
+information to compute the transition-transversion ratio, a quality
+control metric, within each chromosome in GRanges object.
+
+First we filter out any insertion or deletion alleles and the SNPs
+present on the mitochondria then create a logical vector corresponding
+to whether there is a transition event.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{h1_snp_array <-}\StringTok{ }\NormalTok{h1_snp_array }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{filter}\NormalTok{(}\OperatorTok{!}\NormalTok{(ref }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"I"}\NormalTok{, }\StringTok{"D"}\NormalTok{)), seqnames }\OperatorTok{!=}\StringTok{ "M"}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{transition =}\NormalTok{ (ref }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"A"}\NormalTok{, }\StringTok{"G"}\NormalTok{) }\OperatorTok{&}\StringTok{ }\NormalTok{alt }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"G"}\NormalTok{,}\StringTok{"A"}\NormalTok{)) }\OperatorTok{|}
+\StringTok{                      }\NormalTok{(ref }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"C"}\NormalTok{,}\StringTok{"T"}\NormalTok{) }\OperatorTok{&}\StringTok{ }\NormalTok{alt }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"T"}\NormalTok{, }\StringTok{"C"}\NormalTok{)))}
+\end{Highlighting}
+\end{Shaded}
+
+We can then compute the transition-transversion ratio using the
+\texttt{group\_by} and \texttt{summarise} pattern, it is computed as the
+total number of transition SNPs divided by the total number of
+transversion SNPs within each chromsome (figure \ref{fig:titv-viz}).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{ti_tv_results <-}\StringTok{ }\NormalTok{h1_snp_array }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{group_by}\NormalTok{(seqnames) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{n_snps =} \KeywordTok{n}\NormalTok{(),}
+            \DataTypeTok{ti_tv =} \KeywordTok{sum}\NormalTok{(transition) }\OperatorTok{/}\StringTok{ }\KeywordTok{sum}\NormalTok{(}\OperatorTok{!}\NormalTok{transition)) }
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/titv-viz-1}
+
+}
+
+\caption{Dot plot of chromosomes ordered by estimated transition-transversion ratio. A white reference line is drawn at the expected ratio for a human exome.}\label{fig:titv-viz}
+\end{figure}
+
 \subsection{Computing Windowed
@@
 
-(example of group\_by\_overlaps + summarise)
+Another common operation in genomics data analysis is to compute data
+summaries over genomic windows. In \texttt{plyranges} this can be
+achieved via the \texttt{group\_by\_overlaps} operator. Continuing with
+the data from the Human Epigenome Roadmap Consortium data, we can count
+the number of reads that fall into a fixed bins of size 10000bp over a
+BAM file of H3K27Me3 methylation marks from the H1 cell line. We extract
+reads that have a mapping quality score greater than 20:
 
-\subsection{Transition Transversion
-Ratio}\label{transition-transversion-ratio}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{alignments <-}\StringTok{ }\KeywordTok{read_bam}\NormalTok{(h1_bam) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{filter}\NormalTok{(mapq }\OperatorTok{>}\StringTok{ }\DecValTok{20}\NormalTok{) }
+\end{Highlighting}
+\end{Shaded}
 
-(example of group\_by + mutate + summarise and working on an object that
-inherits from GRanges)
+Note that the BAM file is only read into memory once we perform an
+operation on it. Next we generate our genomic bins using the
+\texttt{tile} function from \texttt{GenomicRanges}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{bins <-}\StringTok{ }\KeywordTok{tile}\NormalTok{(h1_bam, }\DataTypeTok{width =} \DecValTok{10000}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Finally, we can use \texttt{group\_by\_overlaps} with \texttt{summarise}
+to compute the total number of reads within each window.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{h1_bam_read_summary <-}\StringTok{ }\NormalTok{h1_bam }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{group_by_overlaps}\NormalTok{(bins) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{n_reads =} \KeywordTok{n}\NormalTok{())}
+\end{Highlighting}
+\end{Shaded}
+
+https://support.bioconductor.org/p/71601/ also useful example?
 
@@
 
-The \emph{plyranges} package is available on the Bioconductor project
+The \texttt{plyranges} package is available on the Bioconductor project
 website \url{https://bioconductor.org} or can be accessed via Github
@@
 the \texttt{plyranges} package and extend it for use with more complex
-data structures such as the \emph{SummarizedExperiment} class, which can
-be used for analysing transcriptomic and variant data. As the
+data structures such as the SummarizedExperiment class, which can be
+used for analysing transcriptomic and variant data. As the
 \texttt{plyranges} interface encourages tidy data practices it
@@
 datasets.
+
+\section{Acknowledgements}\label{acknowledgements}
+
+We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall
+Institute and Dr Paul Harrison for their feedback on earlier drafts of
+this work. We would also like to thank Lori Shepherd and Hèrve Pages for
+the code review they performed. This report was written with
+\texttt{knitr} {[}17{]} and the figures were made with \texttt{ggbio}
+{[}18{]}. All code required to reproduce this report is available at
+https://github.com/sa-lee/plyranges-paper.
 
@@
 
+\hypertarget{ref-R-knitr}{}
+17. Xie Y. Dynamic documents with R and knitr {[}Internet{]}. 2nd ed.
+Boca Raton, Florida: Chapman; Hall/CRC; 2015. Available:
+\url{https://yihui.name/knitr/}
+
+\hypertarget{ref-R-ggbio}{}
+18. Yin T, Cook D, Lawrence M. Ggbio: An r package for extending the
+grammar of graphics for genomic data. Genome Biology. BioMed Central
+Ltd; 2012;13: R77.
+
 \nolinenumbers

### 2018-04-24 paper.Rmd
comment on example 2 ...
complete draft (minus final example)

@@
     of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows.
-We have implemented this grammar as an Bioconductor/R package called plyranges.
+We have implemented this grammar as a Bioconductor/R package called plyranges.
 
@@
 and fluency. All of these principles are defined and discussed
-below in the context of the GRanges class. Where applicable we contrast our 
-design to the existing Bioconductor infrastructure.
+below in the context of the GRanges class and an overview of 
+the grammar is provided in table \ref{tab:grammar}. 
+Where applicable we contrast our design to the existing Bioconductor infrastructure.
+
+\begin{table}[ht]
+\label{tab:grammar}
+\centering
+\begin{tabular}{|l|l|p{4cm}|}
+  \hline
+Operation Type & Verb &  Description \\ 
+  \hline
+  Arithmetic &  `mutate()` & modifies any column \\
+  Arithmetic & `select()` & select columns \\
+  Arithmetic & `arrange()` & sort by columns \\
+  Arithmetic & `set\_width()` & modifies width \\
+  Arithmetic & `stretch()` & extend range by fixed amount \\
+  Arithmetic & `shift\_(direction)` & shift coordinates \\
+  Arithmetic & `flank\_(direction)` & generate flanking regions \\
+  Arithmetic & `\%setop\%` & row-wise set operations \\
+  Restriction & `filter()` & subset rows \\
+  Restriction & `filter\_by\_overlaps()` & subset by overlap \\
+  Restriction & `filter\_by\_non\_overlaps()` & subset by no overlap \\
+  Merging & `join\_overlap\_()` & merge by overlapping ranges \\
+  Merging & `join\_(follow/precedes/nearest)` & merge nearest neighbours \\
+  Merging & `(setop)\_ranges` & range-wise set operations \\
+  Aggregation & `summarise()` & aggregate over column(s) \\ 
+  Aggregation & `disjoin\_ranges()` & aggregate column(s) over the union of end coordinates \\
+  Aggregation & `reduce\_ranges()` & aggregate column(s) by merging overlapping and neighbouring ranges \\
+  Modifier & `group\_by()` & partition by column(s)  \\ 
+  Modifier & `group\_by\_overlaps()` & partition by overlaps \\
+  Modifier & `anchor\_direction()` & fix coordinates \\
+   \hline
+\end{tabular}
+\caption{Overview of `plyranges` grammar.}
+\end{table}
 
@@
 The `plyranges` DSL defines an expressive algebra for performing genomic
-arithemtic withn and between GRanges objects. Often in analyses of
+arithmetic with and between GRanges objects. Often in analyses of
 genomic regions we are required to make adjustments to the start, end,
 and width columns. This is supported directly in `plyranges` via extending
-the `mutate()` function in `dplyr` (table 1). However, often adjustments
+the `mutate()` function in `dplyr` (table \ref{tab:grammar}). 
+However, often adjustments
 need to be relative: there is a mutual dependence between the start, end and
@@
 object by design, causes a user to think of the range as a pair of endpoints, 
-and hence we have developed operators that perform coordinate transfomations with 
+and hence we have developed operators that perform coordinate transformations with 
 the notion of an 'anchoring' combinator.
@@
 parallel/vectorized operations between two vectors/tables of
-ranges, and are not to be confused with the aggregating 
-operations defined in table x. We make the binary nature of these operators
+ranges, and are not to be confused with the merging 
+operations defined in table \ref{tab:grammar}. We make the binary nature of 
+these operators
 by having in-fix functions in R for the parallel union, intersect and
-asymetric set difference. There are also operations that relate to the 
+asymmetric set difference. There are also operations that relate to the 
 relative positions of two ranges such as the 'between' operator which
@@
 a range that spans both ranges. The binary operations can be expressed within
-the functions in table x.
+the arithmetic, restriction or aggregation functions in table \ref{tab:grammar}.
 
@@
 ```
+
 We extend this idea by introducing three types of overlap joins: inner,
@@
 
-* compute coverage as an example here
-
-This design pattern strongly deviates from the design of the OO interface in 
-`GenomicRanges`, where many methods return a new class that is unfamiliar to 
-the user upon return. These low-level classes enable efficient computing at 
-the cost of complexity and are abstracted away in `plyranges` interface.
+As an example, in `plyranges` when we compute coverage over a GRanges object,
+the result of the operation is an expanded GRanges object with a score column,
+corresponding the estimated coverage over a genomic region. Hence, the resulting
+coverage score can easily be composed with other expressions in our algebra.
+This pattern strongly deviates from the design of the OO interface in 
+`GenomicRanges`, which in the case of computing coverage would return an RleList,
+an object that has many new methods and that is potentially unfamiliar to 
+a new user. While, these low-level classes enable efficient computing,
+the increase complexity, and as a consequence are abstracted away in the 
+`plyranges` DSL.
 
@@
 rather than how its doing it.
-
-```{r, echo = FALSE, eval = FALSE}
-feature_tbl <- tribble(
-  ~Verb, ~Action, ~Description,
-  "`mutate()`", "modifies columns", "Takes a GRanges object and a set of name-value pairs and generates a new GRanges object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
- "`filter()`", "subset rows", "Takes a GRanges object and a set of logical expressions and restricts the GRanges object to where all logical expressions are true.",
- "`summarise()`", "aggregate one or more columns by a function", "Takes a GRanges object and a set of name-value pairs and aggregates the GRanges according to functions evaluated in the name-value pairs.",
-"`group_by()`", "partition a GRanges object by a column", "Create an implicit grouping of GRanges object by one or more columns.
-This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
-"`select()`", "select columns of a GRanges objects", "Drop or reorder metadata columns.",
-"`arrange()`", "sort a GRanges object by columns", "Sort a GRanges object by named variables",
-"`disjoin_ranges()`", "aggregate one more columns of a GRanges object over the union of end coordinates", "Takes a GRanges object and a set of name-value pairs and expands the GRanges object by taking the union of end coordinates.
-For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
-"`reduce_ranges()`", "aggregate one more columns of a GRanges object by merging overlapping and neighbouring ranges.", "Takes a GRanges object and a set of name-value pairs and reduces the GRanges object by merging overlapping and neighouring ranges.
-For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
-)
-
-knitr::kable(feature_tbl, format = "latex")
-```
 
@@
 to whether there is a transition event.
+
 ```{r}
@@
 and `summarise` pattern, it is computed as the total number of transition
-SNPs divided by the total number of transversion SNPs within each chromsome
+SNPs divided by the total number of transversion SNPs within each chromosome
 (figure \ref{fig:titv-viz}).
@@
 
+<!-- 
+Trivial example? Could do something with a smooth summarised over
+genomic coordinates? 
+-->
+
 ## Computing Windowed Statistics
@@
 Note that the BAM file is only read into memory once we perform an operation
-on it. Next we generate our genomic bins using the `tile` function
-from `GenomicRanges`:
+on it. Because of interoperability over Bioconductor, we can generate
+genomic bins using the `tile` function from `GenomicRanges`:
 
@@
 `knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio. All code
-required to reproduce this report is available at 
+required to reproduce this article is available at 
 https://github.com/sa-lee/plyranges-paper.

### 2018-04-24 paper.tex
complete draft (minus final example)

@@
 endomorphism, and fluency. All of these principles are defined and
-discussed below in the context of the GRanges class. Where applicable we
+discussed below in the context of the GRanges class and an overview of
+the grammar is provided in table \ref{tab:grammar}. Where applicable we
 contrast our design to the existing Bioconductor infrastructure.
+
+\begin{table}[ht]
+\label{tab:grammar}
+\centering
+\begin{tabular}{|l|l|p{4cm}|}
+  \hline
+Operation Type & Verb &  Description \\ 
+  \hline
+  Arithmetic &  `mutate()` & modifies any column \\
+  Arithmetic & `select()` & select columns \\
+  Arithmetic & `arrange()` & sort by columns \\
+  Arithmetic & `set\_width()` & modifies width \\
+  Arithmetic & `stretch()` & extend range by fixed amount \\
+  Arithmetic & `shift\_(direction)` & shift coordinates \\
+  Arithmetic & `flank\_(direction)` & generate flanking regions \\
+  Arithmetic & `\%setop\%` & row-wise set operations \\
+  Restriction & `filter()` & subset rows \\
+  Restriction & `filter\_by\_overlaps()` & subset by overlap \\
+  Restriction & `filter\_by\_non\_overlaps()` & subset by no overlap \\
+  Merging & `join\_overlap\_()` & merge by overlapping ranges \\
+  Merging & `join\_(follow/precedes/nearest)` & merge nearest neighbours \\
+  Merging & `(setop)\_ranges` & range-wise set operations \\
+  Aggregation & `summarise()` & aggregate over column(s) \\ 
+  Aggregation & `disjoin\_ranges()` & aggregate column(s) over the union of end coordinates \\
+  Aggregation & `reduce\_ranges()` & aggregate column(s) by merging overlapping and neighbouring ranges \\
+  Modifier & `group\_by()` & partition by column(s)  \\ 
+  Modifier & `group\_by\_overlaps()` & partition by overlaps \\
+  Modifier & `anchor\_direction()` & fix coordinates \\
+   \hline
+\end{tabular}
+\caption{Overview of `plyranges` grammar.}
+\end{table}
 
@@
 The \texttt{plyranges} DSL defines an expressive algebra for performing
-genomic arithemtic withn and between GRanges objects. Often in analyses
+genomic arithmetic with and between GRanges objects. Often in analyses
 of genomic regions we are required to make adjustments to the start,
@@
 via extending the \texttt{mutate()} function in \texttt{dplyr} (table
-1). However, often adjustments need to be relative: there is a mutual
-dependence between the start, end and width columns over a genomic
-region. Changing the start,column needs to change either the width or
-the end to preserve integrity of the object. The GRanges object by
-design, causes a user to think of the range as a pair of endpoints, and
-hence we have developed operators that perform coordinate transfomations
-with the notion of an `anchoring' combinator.
+\ref{tab:grammar}). However, often adjustments need to be relative:
+there is a mutual dependence between the start, end and width columns
+over a genomic region. Changing the start,column needs to change either
+the width or the end to preserve integrity of the object. The GRanges
+object by design, causes a user to think of the range as a pair of
+endpoints, and hence we have developed operators that perform coordinate
+transformations with the notion of an `anchoring' combinator.
 
@@
 operations between two vectors/tables of ranges, and are not to be
-confused with the aggregating operations defined in table x. We make the
-binary nature of these operators by having in-fix functions in R for the
-parallel union, intersect and asymetric set difference. There are also
-operations that relate to the relative positions of two ranges such as
-the `between' operator which returns the range in the gap between two
-ranges and the `span' operator a range that spans both ranges. The
-binary operations can be expressed within the functions in table x.
+confused with the merging operations defined in table \ref{tab:grammar}.
+We make the binary nature of these operators by having in-fix functions
+in R for the parallel union, intersect and asymmetric set difference.
+There are also operations that relate to the relative positions of two
+ranges such as the `between' operator which returns the range in the gap
+between two ranges and the `span' operator a range that spans both
+ranges. The binary operations can be expressed within the arithmetic,
+restriction or aggregation functions in table \ref{tab:grammar}.
 
@@
 We extend this idea by introducing three types of overlap joins: inner,
-intersect and left. The inner join considers a hit to be any overlap
-between subject and query and expands the query range to have length
-equal to the number of hits. The intersect join uses the same definition
-of a hit as the inner join but returns a range where the start and end
-coordinates are the intersect of coordinates in the subject that hit the
-query. Finally, the overlap left join is akin to left outer join in
-Cobb's relational algebra: it performs an overlap inner join but also
-returns all query ranges that are not hit by the subject.
+intersect and left (figure \ref{fig:olaps-fig}). The inner join
+considers a hit to be any overlap between subject and query and expands
+the query range to have length equal to the number of hits. The
+intersect join uses the same definition of a hit as the inner join but
+returns a range where the start and end coordinates are the intersect of
+coordinates in the subject that hit the query. Finally, the overlap left
+join is akin to left outer join in Cobb's relational algebra: it
+performs an overlap inner join but also returns all query ranges that
+are not hit by the subject.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  compute coverage as an example here
-\end{itemize}
-
-This design pattern strongly deviates from the design of the OO
-interface in \texttt{GenomicRanges}, where many methods return a new
-class that is unfamiliar to the user upon return. These low-level
-classes enable efficient computing at the cost of complexity and are
-abstracted away in \texttt{plyranges} interface.
+As an example, in \texttt{plyranges} when we compute coverage over a
+GRanges object, the result of the operation is an expanded GRanges
+object with a score column, corresponding the estimated coverage over a
+genomic region. Hence, the resulting coverage score can easily be
+composed with other expressions in our algebra. This pattern strongly
+deviates from the design of the OO interface in \texttt{GenomicRanges},
+which in the case of computing coverage would return an RleList, an
+object that has many new methods and that is potentially unfamiliar to a
+new user. While, these low-level classes enable efficient computing, the
+increase complexity, and as a consequence are abstracted away in the
+\texttt{plyranges} DSL.
 
@@
 total number of transition SNPs divided by the total number of
-transversion SNPs within each chromsome (figure \ref{fig:titv-viz}).
+transversion SNPs within each chromosome (figure \ref{fig:titv-viz}).
 
@@
 Note that the BAM file is only read into memory once we perform an
-operation on it. Next we generate our genomic bins using the
-\texttt{tile} function from \texttt{GenomicRanges}:
+operation on it. Because of interoperability over Bioconductor, we can
+generate genomic bins using the \texttt{tile} function from
+\texttt{GenomicRanges}:
 
@@
 \texttt{knitr} {[}17{]} and the figures were made with \texttt{ggbio}
-{[}18{]}. All code required to reproduce this report is available at
+{[}18{]}. All code required to reproduce this article is available at
 https://github.com/sa-lee/plyranges-paper.

### 2018-04-26 paper.Rmd
rework of algebra part

@@
 myriad of data structures have overlapping purposes and important but
-obscure differences in behaviour that often confuse the typical end
+obscure differences in behavior that often confuse the typical end
 user.
@@
 
-# Design and Implementation
+# Genomic relational algebra
 
-The `plyranges` DSL is built on the most general Bioconductor data structure, 
+## Data model
+
+The `plyranges` DSL is built on the core Bioconductor data structure
 GRanges, which is capable of
@@
 
-## Genomic Relational Algebra
+## Algebraic operations
 
-The `plyranges` DSL defines an expressive algebra for performing genomic
-arithmetic with and between GRanges objects. Often in analyses of
-genomic regions we are required to make adjustments to the start, end,
-and width columns. This is supported directly in `plyranges` via extending
-the `mutate()` function in `dplyr` (table \ref{tab:grammar}). 
-However, often adjustments
-need to be relative: there is a mutual dependence between the start, end and
-width columns over a genomic region. Changing the start,column needs to change
-either the width or the end to preserve integrity of the object. The GRanges 
-object by design, causes a user to think of the range as a pair of endpoints, 
-and hence we have developed operators that perform coordinate transformations with 
-the notion of an 'anchoring' combinator.
+The `plyranges` DSL defines an expressive algebra for performing
+genomic operations with and between GRanges objects (see table
+\ref{tab:grammar}). [ML: add a couple overview sentences indicating the
+classes of operations like arithmetic, merging, aggregation]
 
-For width modification, we can stretch the range by a fixed amount or
-resize the width of a range. The 'anchoring' combinator preserves one of the 
-endpoints or the  midpoint of a range when altering performing
-width modification. It can also be used to preserve coordinates over strand, 
-for example, anchoring over the three prime end of a genomic region, will 
-preserve the start on the negative strand and end on the positive strand. 
-Anchoring the width would not make sense, because it is not a point that can be
-intuitively anchored and stretching and resizing obviously modify the width, 
-so it would not apply to those.
+GRanges describes the within-sequence coordinates of a range by its
+_start_, _end_ and _width_. Those three variables are mutually
+dependent and partially redundant, so direct manipulation of them is
+problematic. For example, changing the _width_ column needs to change
+either the _start_, _end_ or both to preserve integrity of the
+object. We introduce the _anchor_ combinator to disambiguate these
+adjustments. Supported anchor points include the start, end and
+midpoint, as well as the 3' and 5' ends for strand-directed
+ranges. For example, if we anchor the start, then setting the width
+will adjust the end while leaving the start stationary.
 
-Transformations that are width invariant are designed to be explicit
-about the direction to modify and are performed via the shift operator. 
-For GRanges objects strand modulates direction and we distinguish left/right 
-(chromosomal coordinates) from the direction of transcription. To generate
-flanking regions, we use the flank operator which is also explicit about the
-direction of the input range.
+[ML: where does the word 'combinator' come from here? maybe 'modifier'
+ is better? that's what the table uses]
 
-Binary operations between two ranges mostly align with set
-operations that treat ranges as sets of integers. These are
-parallel/vectorized operations between two vectors/tables of
-ranges, and are not to be confused with the merging 
-operations defined in table \ref{tab:grammar}. We make the binary nature of 
-these operators
-by having in-fix functions in R for the parallel union, intersect and
-asymmetric set difference. There are also operations that relate to the 
-relative positions of two ranges such as the 'between' operator which
-returns the range in the gap between two ranges and the 'span' operator 
-a range that spans both ranges. The binary operations can be expressed within
-the arithmetic, restriction or aggregation functions in table \ref{tab:grammar}.
+The algebra also defines conveniences for relative coordinate
+adjustments: _shift_ (unanchored adjustment to both start and end) and
+_stretch_ (anchored adjustment of width). We can perform any relative
+adjustment by some combination of those two operations.  The _stretch_
+operation requires an anchor and assumes the midpoint by
+default. Since _shift_ is unanchored, the user specifies a suffix for
+indicating the direction: left/right or, for stranded features,
+upstream/downstream. For example, _shift\_right_ shifts a range to the
+right.
 
-Our algebra recasts the actions of finding overlaps or nearest neighbours 
-between two genomic regions as a relational join operator. The join operator 
-acts on two GRanges object, a query and a subject. The join operator is 
-relational in the sense that metadata from the query and subject ranges is 
-retained in the joined range.  The commonality between
-all join operators in the `plyranges` DSL is to use the index of where in
-the query range the subject range 'hits' the query range as a primary key 
-(figure \ref{fig:olaps-fig}). The notion of 'hits' and hence
-the resulting key is determined by the type of join (preceding, nearest, follows
-overlaps).
+The _flank_ operation generates new ranges that are adjacent to
+existing ones. This is useful, for example, when generating upstream
+promoter regions for genes. Analogous to _shift_, a suffix indicates
+the side of the input range to flank.
+
+As with other genomic grammars, we define set operations that treat
+ranges as sets of integers, including _intersect_, _union_,
+_difference_, and _complement_. There are two sets of these: parallel
+and merging. The parallel operations map to infix operators, which we
+surround with _%_ symbols in accordance with R syntax rules.  For
+example, the parallel intersection (_x %intersect% y_) finds the
+intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
+is the length of both _x_ and _y_. In constrast, the merging
+intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
+ranges representing whereever there was overlap between a range in _x_
+and a range in _y_. We use the infix syntax for the parallel
+operations, since it is the conventional syntax for parallel, binary
+operations in R. Finding the parallel union will fail when two ranges
+have a gap, so we introduce a _span_ operator that takes the union
+while filling any gap. The _complement_ operation is unique in that it
+is unary. It finds the regions not covered by any of the ranges in a
+single set. Closely related is the _between_ parallel operation, which
+finds the gap separating _xi_ and _yi_. The binary operations are
+callable from within the arithmetic, restriction and aggregation
+expressions.
+
+Our algebra recasts finding overlaps or nearest neighbours between two
+genomic regions as variants of the relational join operator. A join
+acts on two GRanges objects, a query and a subject. The join operator
+is relational in the sense that metadata from the query and subject
+ranges is retained in the joined range.  All join operators in the
+`plyranges` DSL generate a set of hits based on overlap or proximity
+of ranges and use those hits to merge the two datasets in different
+ways. There are four supported matching algorithms: _overlap_,
+_nearest_, _precede_, and _follow_. We can further restrict the
+matching by whether the query is completely _within_ the subject, and
+adding the _directed_ suffix ensures that matching ranges have the
+same direction (strand).
+
+For merging based on the hits, we have three modes: _inner_,
+_intersect_ and _left_. The _inner_ overlap join is similar to the
+conventional inner join in that there is a row in the result for every
+match. A major difference is that the matching is not by identity, so
+we have to choose one of the ranges from each pair. We always choose
+the left range. The _intersect_ join uses the intersection instead of
+the left range. Finally, the overlap _left_ join is akin to left outer
+join in Cobb's relational algebra: it performs an overlap inner join
+but also returns all query ranges that are not hit by the subject.
 
@@
 
-We extend this idea by introducing three types of overlap joins: inner,
-intersect and left (figure \ref{fig:olaps-fig}). The inner join considers a
-hit to be any overlap between subject and query and expands the query range to 
-have length equal to the  number of hits. The intersect join uses the same 
-definition of a hit as the inner join but returns a range where the start and end coordinates
-are the intersect of coordinates in the subject that hit the query. Finally,
-the overlap left join is akin to left outer join in Cobb's relational algebra:
-it performs an overlap inner join but also returns all query ranges that
-are not hit by the subject.
-
-The behaviour of a join operator can be further altered with additional
-suffixes by restricting or expanding how a subject 'hits' a query. For example,
-for overlap joins we can use suffixes to encapsulate Allan's Interval Algebra:
-to consider subject 'hits' that are entirely 'within' a query range, the 'within'
-prefix is used. We also use the 'directed' suffix to explicitly include 
-notions of strandedness in our algebra.
+[ add a simple paragraph about how we can filter, sort, group and
+  aggregate just like other tabular structures ]
 

### 2018-04-27 paper.tex
rebuild of paper

@@
 interface to existing Bioconductor infrastructure, resulting in human
-readable analysis workflows. We have implemented this grammar as an
+readable analysis workflows. We have implemented this grammar as a
 Bioconductor/R package called plyranges.
@@
 myriad of data structures have overlapping purposes and important but
-obscure differences in behaviour that often confuse the typical end
-user.
+obscure differences in behavior that often confuse the typical end user.
 
@@
 
-\section{Design and Implementation}\label{design-and-implementation}
+\section{Genomic relational algebra}\label{genomic-relational-algebra}
 
-The \texttt{plyranges} DSL is built on the most general Bioconductor
-data structure, GRanges, which is capable of representing all types of
-genomic data at a semantic level that roughly matches the intuition of
-most users. A GRanges is essentially a table, with columns for the
+\subsection{Data model}\label{data-model}
+
+The \texttt{plyranges} DSL is built on the core Bioconductor data
+structure GRanges, which is capable of representing all types of genomic
+data at a semantic level that roughly matches the intuition of most
+users. A GRanges is essentially a table, with columns for the
 chromosome, start and end coordinates, and the strand, along with an
@@
 
-\subsection{Genomic Relational
-Algebra}\label{genomic-relational-algebra}
+\subsection{Algebraic operations}\label{algebraic-operations}
 
 The \texttt{plyranges} DSL defines an expressive algebra for performing
-genomic arithmetic with and between GRanges objects. Often in analyses
-of genomic regions we are required to make adjustments to the start,
-end, and width columns. This is supported directly in \texttt{plyranges}
-via extending the \texttt{mutate()} function in \texttt{dplyr} (table
-\ref{tab:grammar}). However, often adjustments need to be relative:
-there is a mutual dependence between the start, end and width columns
-over a genomic region. Changing the start,column needs to change either
-the width or the end to preserve integrity of the object. The GRanges
-object by design, causes a user to think of the range as a pair of
-endpoints, and hence we have developed operators that perform coordinate
-transformations with the notion of an `anchoring' combinator.
+genomic operations with and between GRanges objects (see table
+\ref{tab:grammar}). {[}ML: add a couple overview sentences indicating
+the classes of operations like arithmetic, merging, aggregation{]}
 
-For width modification, we can stretch the range by a fixed amount or
-resize the width of a range. The `anchoring' combinator preserves one of
-the endpoints or the midpoint of a range when altering performing width
-modification. It can also be used to preserve coordinates over strand,
-for example, anchoring over the three prime end of a genomic region,
-will preserve the start on the negative strand and end on the positive
-strand. Anchoring the width would not make sense, because it is not a
-point that can be intuitively anchored and stretching and resizing
-obviously modify the width, so it would not apply to those.
+GRanges describes the within-sequence coordinates of a range by its
+\emph{start}, \emph{end} and \emph{width}. Those three variables are
+mutually dependent and partially redundant, so direct manipulation of
+them is problematic. For example, changing the \emph{width} column needs
+to change either the \emph{start}, \emph{end} or both to preserve
+integrity of the object. We introduce the \emph{anchor} combinator to
+disambiguate these adjustments. Supported anchor points include the
+start, end and midpoint, as well as the 3' and 5' ends for
+strand-directed ranges. For example, if we anchor the start, then
+setting the width will adjust the end while leaving the start
+stationary.
 
-Transformations that are width invariant are designed to be explicit
-about the direction to modify and are performed via the shift operator.
-For GRanges objects strand modulates direction and we distinguish
-left/right (chromosomal coordinates) from the direction of
-transcription. To generate flanking regions, we use the flank operator
-which is also explicit about the direction of the input range.
+{[}ML: where does the word `combinator' come from here? maybe `modifier'
+is better? that's what the table uses{]}
 
-Binary operations between two ranges mostly align with set operations
-that treat ranges as sets of integers. These are parallel/vectorized
-operations between two vectors/tables of ranges, and are not to be
-confused with the merging operations defined in table \ref{tab:grammar}.
-We make the binary nature of these operators by having in-fix functions
-in R for the parallel union, intersect and asymmetric set difference.
-There are also operations that relate to the relative positions of two
-ranges such as the `between' operator which returns the range in the gap
-between two ranges and the `span' operator a range that spans both
-ranges. The binary operations can be expressed within the arithmetic,
-restriction or aggregation functions in table \ref{tab:grammar}.
+The algebra also defines conveniences for relative coordinate
+adjustments: \emph{shift} (unanchored adjustment to both start and end)
+and \emph{stretch} (anchored adjustment of width). We can perform any
+relative adjustment by some combination of those two operations. The
+\emph{stretch} operation requires an anchor and assumes the midpoint by
+default. Since \emph{shift} is unanchored, the user specifies a suffix
+for indicating the direction: left/right or, for stranded features,
+upstream/downstream. For example, \emph{shift\_right} shifts a range to
+the right.
 
-Our algebra recasts the actions of finding overlaps or nearest
-neighbours between two genomic regions as a relational join operator.
-The join operator acts on two GRanges object, a query and a subject. The
-join operator is relational in the sense that metadata from the query
-and subject ranges is retained in the joined range. The commonality
-between all join operators in the \texttt{plyranges} DSL is to use the
-index of where in the query range the subject range `hits' the query
-range as a primary key (figure \ref{fig:olaps-fig}). The notion of
-`hits' and hence the resulting key is determined by the type of join
-(preceding, nearest, follows overlaps).
+The \emph{flank} operation generates new ranges that are adjacent to
+existing ones. This is useful, for example, when generating upstream
+promoter regions for genes. Analogous to \emph{shift}, a suffix
+indicates the side of the input range to flank.
+
+As with other genomic grammars, we define set operations that treat
+ranges as sets of integers, including \emph{intersect}, \emph{union},
+\emph{difference}, and \emph{complement}. There are two sets of these:
+parallel and merging. The parallel operations map to infix operators,
+which we surround with \emph{\%} symbols in accordance with R syntax
+rules. For example, the parallel intersection (\emph{x \%intersect\% y})
+finds the intersecting range between \emph{xi} and \emph{yi} for
+\emph{i} in \emph{1\ldots{}n}, where \emph{n} is the length of both
+\emph{x} and \emph{y}. In constrast, the merging intersection
+(\emph{intersect\_ranges(x, y)}) returns a new set of disjoint ranges
+representing whereever there was overlap between a range in \emph{x} and
+a range in \emph{y}. We use the infix syntax for the parallel
+operations, since it is the conventional syntax for parallel, binary
+operations in R. Finding the parallel union will fail when two ranges
+have a gap, so we introduce a \emph{span} operator that takes the union
+while filling any gap. The \emph{complement} operation is unique in that
+it is unary. It finds the regions not covered by any of the ranges in a
+single set. Closely related is the \emph{between} parallel operation,
+which finds the gap separating \emph{xi} and \emph{yi}. The binary
+operations are callable from within the arithmetic, restriction and
+aggregation expressions.
+
+Our algebra recasts finding overlaps or nearest neighbours between two
+genomic regions as variants of the relational join operator. A join acts
+on two GRanges objects, a query and a subject. The join operator is
+relational in the sense that metadata from the query and subject ranges
+is retained in the joined range. All join operators in the
+\texttt{plyranges} DSL generate a set of hits based on overlap or
+proximity of ranges and use those hits to merge the two datasets in
+different ways. There are four supported matching algorithms:
+\emph{overlap}, \emph{nearest}, \emph{precede}, and \emph{follow}. We
+can further restrict the matching by whether the query is completely
+\emph{within} the subject, and adding the \emph{directed} suffix ensures
+that matching ranges have the same direction (strand).
+
+For merging based on the hits, we have three modes: \emph{inner},
+\emph{intersect} and \emph{left}. The \emph{inner} overlap join is
+similar to the conventional inner join in that there is a row in the
+result for every match. A major difference is that the matching is not
+by identity, so we have to choose one of the ranges from each pair. We
+always choose the left range. The \emph{intersect} join uses the
+intersection instead of the left range. Finally, the overlap \emph{left}
+join is akin to left outer join in Cobb's relational algebra: it
+performs an overlap inner join but also returns all query ranges that
+are not hit by the subject.
 
@@
 
-We extend this idea by introducing three types of overlap joins: inner,
-intersect and left (figure \ref{fig:olaps-fig}). The inner join
-considers a hit to be any overlap between subject and query and expands
-the query range to have length equal to the number of hits. The
-intersect join uses the same definition of a hit as the inner join but
-returns a range where the start and end coordinates are the intersect of
-coordinates in the subject that hit the query. Finally, the overlap left
-join is akin to left outer join in Cobb's relational algebra: it
-performs an overlap inner join but also returns all query ranges that
-are not hit by the subject.
-
-The behaviour of a join operator can be further altered with additional
-suffixes by restricting or expanding how a subject `hits' a query. For
-example, for overlap joins we can use suffixes to encapsulate Allan's
-Interval Algebra: to consider subject `hits' that are entirely `within'
-a query range, the `within' prefix is used. We also use the `directed'
-suffix to explicitly include notions of strandedness in our algebra.
+{[} add a simple paragraph about how we can filter, sort, group and
+aggregate just like other tabular structures {]}
 

### 2018-04-30 paper.Rmd
update figures

@@
 [@Dale2011-js; @Kent2017], however these packages lack the interoperability
-provided by in the aforementioned Bioconductor packages and are not easily 
+provided by the aforementioned Bioconductor packages and are not easily 
 extended.
@@
 consisting of measurements or metadata specific to the data type or
-experiment. For example, a GRanges can represent a gene with
-its chromosomal coordinates, exonic structure, and a count column
-summarized from an RNA-seq experiment.
+experiment (figure \ref{fig:GRanges}).
+
+```{r GRanges, echo = FALSE, out.width="400pt", fig.cap="An illustration of the GRanges data model for a sample from an RNA-seq experiment.
+The core components of the data model inlcude a seqnames column (representing the chromosome), a ranges column which consists of start and end coordinates for a genomic region, and a strand identifier (either positive, negative, or unstranded).
+Metadata are included as columns to the right of the dotted line as annotations (gene_id) or range level covariates (GC score and count)."}
+knitr::include_graphics("diagrams/diagrams.001.png")
+```
 
@@
 
+*
+
 \begin{table}[ht]
@@
   \hline
-Operation Type & Verb &  Description \\ 
+  & Verb &  Description \\ 
   \hline
-  Arithmetic &  `mutate()` & modifies any column \\
+  Arithmetic &  {\tt mutate()} & modifies any column \\
   Arithmetic & `select()` & select columns \\
@@
 \end{tabular}
-\caption{Overview of `plyranges` grammar.}
+\caption{Overview of `plyranges` grammar. Expand here (bold to emphasise
+dplyr adaptations)}
 \end{table}
@@
 This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
-knitr::include_graphics("olaps-figure.png")
+knitr::include_graphics("diagrams/diagrams.002.png")
 ```
@@
 
+(change this)
 Here we provide examples on how to use the `plyranges` DSL to construct
@@
   anchor_center() %>%
-  set_width(5000)
+  mutate(width = 5000)
 ```

### 2018-04-30 paper.Rmd
tidy up of table

@@
 
+\begin{table}[ht]
+\label{tab:grammar}
+\centering
+\begin{tabular}{|l|l|p{4cm}|}
+  \hline
+  & Verb &  Description \\ 
+  \hline
+   & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
+   Aggregation & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
+   &  \emph{\reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
+   \hline
+   &  \textbf{\emph{mutate()}} & modifies any column \\
+   & \textbf{\emph{select()}} & select columns \\
+  Arithmetic (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
+   & \emph{stretch()} & extend range by fixed amount \\
+   &  \emph{shift\_(direction)} & shift coordinates \\
+   & \emph{flank\_(direction)} & generate flanking regions \\
+   & \emph{\%intersection\% } & row-wise intersection \\
+   & \emph{\%union\%} & row-wise union \\
+  Arithmetic (Binary) & { \tt \%setdiff\% } & row-wise set difference \\
+   & \emph{between()} & row-wise gap range \\
+   & \emph{span()} & row-wise spanning range \\
+   \hline
+    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
+    & \emph{join\_nearest} & merge by nearest neighbour ranges \\
+    & \emph{join\_follow} & merge by following ranges \\
+    Merging & \emph{join\_precedes} & merge by preceding ranges \\
+    & \emph{union_ranges} & range-wise union \\
+    & \emph{intersect_ranges} & range-wise intersect \\
+    & \emph{setdiff_ranges} & range-wise set difference \\
+    & \emph{complement_ranges} & range-wise union \\
+  \hline
+   & \textbf{\emph{anchor\_direction()}} & fix coordinates at direction \\
+  Modifier & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
+   & \emph{group\_by\_overlaps()} & partition by overlaps \\\
+   \hline
+   & \textbf{\emph{filter()}} & subset rows \\
+  Restriction & \emph{filter\_by\_overlaps()} & subset by overlap \\
+    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
+   \hline
+\end{tabular}
+\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
+briefly described and categorised into one of: aggregation, unary or binary
+arithmetic, merging, modifier, or restriction. A verb is given bold text if
+its origin is from the \texttt{dplyr} grammar.}
+\end{table}
+
 By definition GRanges follows the tidy data pattern: it is a rectangular 
@@
 below in the context of the GRanges class and an overview of 
-the grammar is provided in table \ref{tab:grammar}. 
-Where applicable we contrast our design to the existing Bioconductor infrastructure.
-
-*
-
-\begin{table}[ht]
-\label{tab:grammar}
-\centering
-\begin{tabular}{|l|l|p{4cm}|}
-  \hline
-  & Verb &  Description \\ 
-  \hline
-  Arithmetic &  {\tt mutate()} & modifies any column \\
-  Arithmetic & `select()` & select columns \\
-  Arithmetic & `arrange()` & sort by columns \\
-  Arithmetic & `set\_width()` & modifies width \\
-  Arithmetic & `stretch()` & extend range by fixed amount \\
-  Arithmetic & `shift\_(direction)` & shift coordinates \\
-  Arithmetic & `flank\_(direction)` & generate flanking regions \\
-  Arithmetic & `\%setop\%` & row-wise set operations \\
-  Restriction & `filter()` & subset rows \\
-  Restriction & `filter\_by\_overlaps()` & subset by overlap \\
-  Restriction & `filter\_by\_non\_overlaps()` & subset by no overlap \\
-  Merging & `join\_overlap\_()` & merge by overlapping ranges \\
-  Merging & `join\_(follow/precedes/nearest)` & merge nearest neighbours \\
-  Merging & `(setop)\_ranges` & range-wise set operations \\
-  Aggregation & `summarise()` & aggregate over column(s) \\ 
-  Aggregation & `disjoin\_ranges()` & aggregate column(s) over the union of end coordinates \\
-  Aggregation & `reduce\_ranges()` & aggregate column(s) by merging overlapping and neighbouring ranges \\
-  Modifier & `group\_by()` & partition by column(s)  \\ 
-  Modifier & `group\_by\_overlaps()` & partition by overlaps \\
-  Modifier & `anchor\_direction()` & fix coordinates \\
-   \hline
-\end{tabular}
-\caption{Overview of `plyranges` grammar. Expand here (bold to emphasise
-dplyr adaptations)}
-\end{table}
+the grammar is provided in table \ref{tab:grammar}. Where applicable we contrast
+our design to the existing Bioconductor infrastructure.
 

### 2018-04-30 paper.Rmd
rework caption for overlaps figure.

@@
 ```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators.
-Each join takes query and subject range as input (black and grey rectangles, respectively) with associated metadata (pink and orange rectangles respectively).
-An index for the join is computed via returning a Hits object, which contains the indices of where the subject overlaps the query range.
-This index is used to expand the query or (take the intersect) of the query ranges with subject ranges and return metadata associated with each input.
+Each join takes query and subject range as input (black and light gray rectangles, respectively).
+An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
+This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
+The join semantics alter what is returned: for an *inner* join the query range is returned for each match, for an intersect *join* the intersection is taken between overlapping ranges, and for a *left* join all query ranges are returned even if the subject range does not overlap them.
 This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}

### 2018-05-02 paper.Rmd
next draft up; resizing figures

@@
 # retrieving a BigWigFile from AnnotationHub
+
+if (!dir.exists("./data")) dir.create("./data")
+
 library(AnnotationHub)
@@
 # a BamFile from the H1 cell line
-h1_bam <- "data/GSM433167_BI.H3K27me3.bam"
+h1_bam <- "./data/GSM433167_BI.H3K27me3.bam"
+if (!file.exists(h1_bam))
+  download.file(
+    url = "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM433nnn/GSM433167/suppl/GSM433167%5FBI%2EH3K27me3%2Ebam",
+    destfile = h1_bam
+  )
+
+# array data for h1 cell line
+# illumina annotations
+array_info <- "data/GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz"
+if (!file.exists(array_info)) 
+  download.file(
+    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GPL18952&format=file&file=GPL18952%5FHumanOmni2%2D5%2D8%2Dv1%2D1%2DC%2Ecsv%2Egz",
+    destfile = array_info
+  )
+
+snp_info <- "data/GSM1463263_JS-ESCH1.txt.gz"
+if (!file.exists(snp_info))
+  download.file(
+    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM1463263&format=file&file=GSM1463263%5FJS%2DESCH1%2Etxt%2Egz",
+    destfile = snp_info
+  )
+
 ```
@@
 
+\begin{figure}
+{\centering \includegraphics[width=400pt]{diagrams/diagrams-001}}
+\caption{An illustration of the GRanges data model for a
+sample from an RNA-seq experiment. The core components of the data model
+inlcude a seqnames column (representing the chromosome), a ranges column
+which consists of start and end coordinates for a genomic region, and a
+strand identifier (either positive, negative, or unstranded). Metadata
+are included as columns to the right of the dotted line as annotations
+(gene\_id) or range level covariates (GC score and count).}
+\label{fig:GRanges} 
+\end{figure}
+
 The `plyranges` DSL is built on the core Bioconductor data structure
@@
 
-```{r GRanges, echo = FALSE, out.width="400pt", fig.cap="An illustration of the GRanges data model for a sample from an RNA-seq experiment.
-The core components of the data model inlcude a seqnames column (representing the chromosome), a ranges column which consists of start and end coordinates for a genomic region, and a strand identifier (either positive, negative, or unstranded).
-Metadata are included as columns to the right of the dotted line as annotations (gene_id) or range level covariates (GC score and count)."}
-knitr::include_graphics("diagrams/diagrams.001.png")
-```
-
-\begin{table}[ht]
-\label{tab:grammar}
+\begin{table}[!htbp]
 \centering
@@
    Aggregation & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
-   &  \emph{\reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
+   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
    \hline
@@
    & \emph{\%union\%} & row-wise union \\
-  Arithmetic (Binary) & { \tt \%setdiff\% } & row-wise set difference \\
+   & \emph{compute\_coverage} & coverage over all ranges \\
+  Arithmetic (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
    & \emph{between()} & row-wise gap range \\
@@
     Merging & \emph{join\_precedes} & merge by preceding ranges \\
-    & \emph{union_ranges} & range-wise union \\
-    & \emph{intersect_ranges} & range-wise intersect \\
-    & \emph{setdiff_ranges} & range-wise set difference \\
-    & \emph{complement_ranges} & range-wise union \\
+    & \emph{union\_ranges} & range-wise union \\
+    & \emph{intersect\_ranges} & range-wise intersect \\
+    & \emph{setdiff\_ranges} & range-wise set difference \\
+    & \emph{complement\_ranges} & range-wise union \\
   \hline
@@
   Modifier & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
-   & \emph{group\_by\_overlaps()} & partition by overlaps \\\
+   & \emph{group\_by\_overlaps()} & partition by overlaps \\
    \hline
@@
 arithmetic, merging, modifier, or restriction. A verb is given bold text if
-its origin is from the \texttt{dplyr} grammar.}
+its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
 \end{table}
@@
 genomic operations with and between GRanges objects (see table
-\ref{tab:grammar}). [ML: add a couple overview sentences indicating the
-classes of operations like arithmetic, merging, aggregation]
+\ref{tab:grammar}). The grammar includes several classes of operation
+that cover most use cases in genomics data analysis. There are 
+operators for performing arithmetic such as modifying width or performing
+set operations, merging such as finding overlaps between GRanges, 
+subsetting by genomic regions, and aggregating over columns within in a GRanges
+object.
 
@@
 either the _start_, _end_ or both to preserve integrity of the
-object. We introduce the _anchor_ combinator to disambiguate these
+object. We introduce the _anchor_ modifier to disambiguate these
 adjustments. Supported anchor points include the start, end and
@@
 will adjust the end while leaving the start stationary.
-
-[ML: where does the word 'combinator' come from here? maybe 'modifier'
- is better? that's what the table uses]
 
@@
 This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
-knitr::include_graphics("diagrams/diagrams.002.png")
+knitr::include_graphics("diagrams/diagrams-002.png")
 ```
 
-[ add a simple paragraph about how we can filter, sort, group and
-  aggregate just like other tabular structures ]
+Since the GRanges object is a tabular data structure our grammar includes
+operators to filter, sort and aggregate by columns in a GRanges. These 
+operations can be performed over partitions within GRanges object
+using the *group_by* modifier. Together with our algebra for arithmetic and
+merging, these operations enable flexible and reproducible analysis using
+the desgin principles of the `dplyr` grammar.
 
@@
 
-(change this)
-Here we provide examples on how to use the `plyranges` DSL to construct
-genomic data workflows and highlight its interoperability with existing
-Bionconductor packages.
+Here we provide illustrative examples by using the `plyranges` DSL to show
+how our grammar could be integrated into genomic data workflows. We also
+highlight how interoperability with existing Bioconductor infrastructure, 
+enables easy access to public datasets and methods for analysis and 
+visualisation.
 
@@
 
+## Computing Windowed Statistics
+
+Another common operation in genomics data analysis is to compute data
+summaries over genomic windows. In `plyranges` this can be achieved via
+the `group_by_overlaps` operator. Continuing with the data from the 
+Human Epigenome Roadmap Consortium data, we can count the number of reads
+that fall into a fixed bins of size 10000bp over a BAM file of H3K27Me3 methylation
+marks from the H1 cell line. We extract reads that have a mapping quality
+score greater than 20:
+
+```{r, eval = FALSE}
+alignments <- read_bam(h1_bam) %>%
+  filter(mapq > 20) 
+```
+
+Note that the BAM file is only read into memory once we perform an operation
+on it. Because of interoperability over Bioconductor, we can generate
+genomic bins using the `tile` function from `GenomicRanges`:
+
+```{r, eval = FALSE}
+bins <- tile(h1_bam, width = 10000)
+```
+
+Finally, we can use `group_by_overlaps` with `summarise` to compute the
+total number of reads within each window.
+
+```{r, eval = FALSE}
+h1_bam_read_summary <- h1_bam %>%
+  group_by_overlaps(bins) %>%
+  summarise(n_reads = n())
+```
+
+<!-- 
+https://support.bioconductor.org/p/71601/ also useful example?
+-->
+
 ## Quality Control Metrics
@@
 # chromosome zero are 'problem' probes and should be filtered
-anno <- read_tsv("data/GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz",
+anno <- read_tsv(array_info,
                  trim_ws = TRUE,
@@
 # data obtained from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1463263
-geno <- read_tsv("data/GSM1463263_JS-ESCH1.txt.gz", skip = 10)
+geno <- read_tsv(snp_info, skip = 10)
 
@@
 
-## Computing Windowed Statistics
-
-Another common operation in genomics data analysis is to compute data
-summaries over genomic windows. In `plyranges` this can be achieved via
-the `group_by_overlaps` operator. Continuing with the data from the 
-Human Epigenome Roadmap Consortium data, we can count the number of reads
-that fall into a fixed bins of size 10000bp over a BAM file of H3K27Me3 methylation
-marks from the H1 cell line. We extract reads that have a mapping quality
-score greater than 20:
-
-```{r, eval = FALSE}
-alignments <- read_bam(h1_bam) %>%
-  filter(mapq > 20) 
-```
-
-Note that the BAM file is only read into memory once we perform an operation
-on it. Because of interoperability over Bioconductor, we can generate
-genomic bins using the `tile` function from `GenomicRanges`:
-
-```{r, eval = FALSE}
-bins <- tile(h1_bam, width = 10000)
-```
-
-Finally, we can use `group_by_overlaps` with `summarise` to compute the
-total number of reads within each window.
-
-```{r, eval = FALSE}
-h1_bam_read_summary <- h1_bam %>%
-  group_by_overlaps(bins) %>%
-  summarise(n_reads = n())
-```
-
-https://support.bioconductor.org/p/71601/ also useful example?
-
 # Availablilty and Future Work

### 2018-05-02 paper.tex
next draft up; resizing figures

@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
 \newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
 \newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
 \newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
 \newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
 
@@
 % Use "Eq" instead of "Equation" for equation citations.
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 pybedtools and valr {[}8,9{]}, however these packages lack the
-interoperability provided by in the aforementioned Bioconductor packages
+interoperability provided by the aforementioned Bioconductor packages
 and are not easily extended.
@@
 
-\section{Genomic relational algebra}\label{genomic-relational-algebra}
+\hypertarget{genomic-relational-algebra}{%
+\section{Genomic relational algebra}\label{genomic-relational-algebra}}
 
-\subsection{Data model}\label{data-model}
+\hypertarget{data-model}{%
+\subsection{Data model}\label{data-model}}
+
+\begin{figure}
+{\centering \includegraphics[width=400pt]{diagrams/diagrams-001}}
+\caption{An illustration of the GRanges data model for a
+sample from an RNA-seq experiment. The core components of the data model
+inlcude a seqnames column (representing the chromosome), a ranges column
+which consists of start and end coordinates for a genomic region, and a
+strand identifier (either positive, negative, or unstranded). Metadata
+are included as columns to the right of the dotted line as annotations
+(gene\_id) or range level covariates (GC score and count).}
+\label{fig:GRanges} 
+\end{figure}
 
@@
 arbitrary set of additional columns, consisting of measurements or
-metadata specific to the data type or experiment. For example, a GRanges
-can represent a gene with its chromosomal coordinates, exonic structure,
-and a count column summarized from an RNA-seq experiment.
+metadata specific to the data type or experiment (figure
+\ref{fig:GRanges}).
+
+\begin{table}[!htbp]
+\centering
+\begin{tabular}{|l|l|p{4cm}|}
+  \hline
+  & Verb &  Description \\ 
+  \hline
+   & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
+   Aggregation & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
+   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
+   \hline
+   &  \textbf{\emph{mutate()}} & modifies any column \\
+   & \textbf{\emph{select()}} & select columns \\
+  Arithmetic (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
+   & \emph{stretch()} & extend range by fixed amount \\
+   &  \emph{shift\_(direction)} & shift coordinates \\
+   & \emph{flank\_(direction)} & generate flanking regions \\
+   & \emph{\%intersection\% } & row-wise intersection \\
+   & \emph{\%union\%} & row-wise union \\
+   & \emph{compute\_coverage} & coverage over all ranges \\
+  Arithmetic (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
+   & \emph{between()} & row-wise gap range \\
+   & \emph{span()} & row-wise spanning range \\
+   \hline
+    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
+    & \emph{join\_nearest} & merge by nearest neighbour ranges \\
+    & \emph{join\_follow} & merge by following ranges \\
+    Merging & \emph{join\_precedes} & merge by preceding ranges \\
+    & \emph{union\_ranges} & range-wise union \\
+    & \emph{intersect\_ranges} & range-wise intersect \\
+    & \emph{setdiff\_ranges} & range-wise set difference \\
+    & \emph{complement\_ranges} & range-wise union \\
+  \hline
+   & \textbf{\emph{anchor\_direction()}} & fix coordinates at direction \\
+  Modifier & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
+   & \emph{group\_by\_overlaps()} & partition by overlaps \\
+   \hline
+   & \textbf{\emph{filter()}} & subset rows \\
+  Restriction & \emph{filter\_by\_overlaps()} & subset by overlap \\
+    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
+   \hline
+\end{tabular}
+\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
+briefly described and categorised into one of: aggregation, unary or binary
+arithmetic, merging, modifier, or restriction. A verb is given bold text if
+its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
+\end{table}
 
@@
 
-\begin{table}[ht]
-\label{tab:grammar}
-\centering
-\begin{tabular}{|l|l|p{4cm}|}
-  \hline
-Operation Type & Verb &  Description \\ 
-  \hline
-  Arithmetic &  `mutate()` & modifies any column \\
-  Arithmetic & `select()` & select columns \\
-  Arithmetic & `arrange()` & sort by columns \\
-  Arithmetic & `set\_width()` & modifies width \\
-  Arithmetic & `stretch()` & extend range by fixed amount \\
-  Arithmetic & `shift\_(direction)` & shift coordinates \\
-  Arithmetic & `flank\_(direction)` & generate flanking regions \\
-  Arithmetic & `\%setop\%` & row-wise set operations \\
-  Restriction & `filter()` & subset rows \\
-  Restriction & `filter\_by\_overlaps()` & subset by overlap \\
-  Restriction & `filter\_by\_non\_overlaps()` & subset by no overlap \\
-  Merging & `join\_overlap\_()` & merge by overlapping ranges \\
-  Merging & `join\_(follow/precedes/nearest)` & merge nearest neighbours \\
-  Merging & `(setop)\_ranges` & range-wise set operations \\
-  Aggregation & `summarise()` & aggregate over column(s) \\ 
-  Aggregation & `disjoin\_ranges()` & aggregate column(s) over the union of end coordinates \\
-  Aggregation & `reduce\_ranges()` & aggregate column(s) by merging overlapping and neighbouring ranges \\
-  Modifier & `group\_by()` & partition by column(s)  \\ 
-  Modifier & `group\_by\_overlaps()` & partition by overlaps \\
-  Modifier & `anchor\_direction()` & fix coordinates \\
-   \hline
-\end{tabular}
-\caption{Overview of `plyranges` grammar.}
-\end{table}
-
-\subsection{Algebraic operations}\label{algebraic-operations}
+\hypertarget{algebraic-operations}{%
+\subsection{Algebraic operations}\label{algebraic-operations}}
 
@@
 genomic operations with and between GRanges objects (see table
-\ref{tab:grammar}). {[}ML: add a couple overview sentences indicating
-the classes of operations like arithmetic, merging, aggregation{]}
+\ref{tab:grammar}). The grammar includes several classes of operation
+that cover most use cases in genomics data analysis. There are operators
+for performing arithmetic such as modifying width or performing set
+operations, merging such as finding overlaps between GRanges, subsetting
+by genomic regions, and aggregating over columns within in a GRanges
+object.
 
@@
 to change either the \emph{start}, \emph{end} or both to preserve
-integrity of the object. We introduce the \emph{anchor} combinator to
+integrity of the object. We introduce the \emph{anchor} modifier to
 disambiguate these adjustments. Supported anchor points include the
@@
 stationary.
-
-{[}ML: where does the word `combinator' come from here? maybe `modifier'
-is better? that's what the table uses{]}
 
@@
 
-{\centering \includegraphics[width=400pt]{olaps-figure}
+{\centering \includegraphics[width=400pt]{diagrams/diagrams-002}
 
@@
 \caption{Illustration of the three overlap join operators.
-Each join takes query and subject range as input (black and grey rectangles, respectively) with associated metadata (pink and orange rectangles respectively).
-An index for the join is computed via returning a Hits object, which contains the indices of where the subject overlaps the query range.
-This index is used to expand the query or (take the intersect) of the query ranges with subject ranges and return metadata associated with each input.
+Each join takes query and subject range as input (black and light gray rectangles, respectively).
+An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
+This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
+The join semantics alter what is returned: for an *inner* join the query range is returned for each match, for an intersect *join* the intersection is taken between overlapping ranges, and for a *left* join all query ranges are returned even if the subject range does not overlap them.
 This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations.}\label{fig:olaps-fig}
@@
 
-{[} add a simple paragraph about how we can filter, sort, group and
-aggregate just like other tabular structures {]}
+Since the GRanges object is a tabular data structure our grammar
+includes operators to filter, sort and aggregate by columns in a
+GRanges. These operations can be performed over partitions within
+GRanges object using the \emph{group\_by} modifier. Together with our
+algebra for arithmetic and merging, these operations enable flexible and
+reproducible analysis using the desgin principles of the \texttt{dplyr}
+grammar.
 
-\subsection{Cohesion}\label{cohesion}
+\hypertarget{cohesion}{%
+\subsection{Cohesion}\label{cohesion}}
 
@@
 
-\subsection{Consistency}\label{consistency}
+\hypertarget{consistency}{%
+\subsection{Consistency}\label{consistency}}
 
@@
 
-\subsection{Endomorphism}\label{endomorphism}
+\hypertarget{endomorphism}{%
+\subsection{Endomorphism}\label{endomorphism}}
 
@@
 
-\subsection{Fluency}\label{fluency}
+\hypertarget{fluency}{%
+\subsection{Fluency}\label{fluency}}
 
@@
 
-\subsection{Opportunities}\label{opportunities}
+\hypertarget{opportunities}{%
+\subsection{Opportunities}\label{opportunities}}
 
@@
 
-\section{Results}\label{results}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-Here we provide examples on how to use the \texttt{plyranges} DSL to
-construct genomic data workflows and highlight its interoperability with
-existing Bionconductor packages.
+Here we provide illustrative examples by using the \texttt{plyranges}
+DSL to show how our grammar could be integrated into genomic data
+workflows. We also highlight how interoperability with existing
+Bioconductor infrastructure, enables easy access to public datasets and
+methods for analysis and visualisation.
 
-\subsection{Peak Finding}\label{peak-finding}
+\hypertarget{peak-finding}{%
+\subsection{Peak Finding}\label{peak-finding}}
 
@@
 \StringTok{  }\KeywordTok{anchor_center}\NormalTok{() }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{set_width}\NormalTok{(}\DecValTok{5000}\NormalTok{)}
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{width =} \DecValTok{5000}\NormalTok{)}
 \end{Highlighting}
@@
 
-\subsection{Quality Control Metrics}\label{quality-control-metrics}
+\hypertarget{computing-windowed-statistics}{%
+\subsection{Computing Windowed
+Statistics}\label{computing-windowed-statistics}}
+
+Another common operation in genomics data analysis is to compute data
+summaries over genomic windows. In \texttt{plyranges} this can be
+achieved via the \texttt{group\_by\_overlaps} operator. Continuing with
+the data from the Human Epigenome Roadmap Consortium data, we can count
+the number of reads that fall into a fixed bins of size 10000bp over a
+BAM file of H3K27Me3 methylation marks from the H1 cell line. We extract
+reads that have a mapping quality score greater than 20:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{alignments <-}\StringTok{ }\KeywordTok{read_bam}\NormalTok{(h1_bam) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{filter}\NormalTok{(mapq }\OperatorTok{>}\StringTok{ }\DecValTok{20}\NormalTok{) }
+\end{Highlighting}
+\end{Shaded}
+
+Note that the BAM file is only read into memory once we perform an
+operation on it. Because of interoperability over Bioconductor, we can
+generate genomic bins using the \texttt{tile} function from
+\texttt{GenomicRanges}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{bins <-}\StringTok{ }\KeywordTok{tile}\NormalTok{(h1_bam, }\DataTypeTok{width =} \DecValTok{10000}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Finally, we can use \texttt{group\_by\_overlaps} with \texttt{summarise}
+to compute the total number of reads within each window.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{h1_bam_read_summary <-}\StringTok{ }\NormalTok{h1_bam }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{group_by_overlaps}\NormalTok{(bins) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{n_reads =} \KeywordTok{n}\NormalTok{())}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{quality-control-metrics}{%
+\subsection{Quality Control Metrics}\label{quality-control-metrics}}
 
@@
 
-\subsection{Computing Windowed
-Statistics}\label{computing-windowed-statistics}
-
-Another common operation in genomics data analysis is to compute data
-summaries over genomic windows. In \texttt{plyranges} this can be
-achieved via the \texttt{group\_by\_overlaps} operator. Continuing with
-the data from the Human Epigenome Roadmap Consortium data, we can count
-the number of reads that fall into a fixed bins of size 10000bp over a
-BAM file of H3K27Me3 methylation marks from the H1 cell line. We extract
-reads that have a mapping quality score greater than 20:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{alignments <-}\StringTok{ }\KeywordTok{read_bam}\NormalTok{(h1_bam) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{filter}\NormalTok{(mapq }\OperatorTok{>}\StringTok{ }\DecValTok{20}\NormalTok{) }
-\end{Highlighting}
-\end{Shaded}
-
-Note that the BAM file is only read into memory once we perform an
-operation on it. Because of interoperability over Bioconductor, we can
-generate genomic bins using the \texttt{tile} function from
-\texttt{GenomicRanges}:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{bins <-}\StringTok{ }\KeywordTok{tile}\NormalTok{(h1_bam, }\DataTypeTok{width =} \DecValTok{10000}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Finally, we can use \texttt{group\_by\_overlaps} with \texttt{summarise}
-to compute the total number of reads within each window.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{h1_bam_read_summary <-}\StringTok{ }\NormalTok{h1_bam }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{group_by_overlaps}\NormalTok{(bins) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{n_reads =} \KeywordTok{n}\NormalTok{())}
-\end{Highlighting}
-\end{Shaded}
-
-https://support.bioconductor.org/p/71601/ also useful example?
-
+\hypertarget{availablilty-and-future-work}{%
 \section{Availablilty and Future
-Work}\label{availablilty-and-future-work}
+Work}\label{availablilty-and-future-work}}
 
@@
 
-\section{Acknowledgements}\label{acknowledgements}
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
 
@@
 
-\section*{References}\label{references}
+\hypertarget{references}{%
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Kozanitis2014-va}{}
+\leavevmode\hypertarget{ref-Kozanitis2014-va}{}%
 1. Kozanitis C, Heiberg A, Varghese G, Bafna V. Using genome query
@@
 
-\hypertarget{ref-Kozanitis2016-bm}{}
+\leavevmode\hypertarget{ref-Kozanitis2016-bm}{}%
 2. Kozanitis C, Patterson DA. GenAp: A distributed SQL interface for
@@
 
-\hypertarget{ref-Kaitoua2017-pw}{}
+\leavevmode\hypertarget{ref-Kaitoua2017-pw}{}%
 3. Kaitoua A, Pinoli P, Bertoni M, Ceri S. Framework for supporting
@@
 
-\hypertarget{ref-Quinlan2010-gc}{}
+\leavevmode\hypertarget{ref-Quinlan2010-gc}{}%
 4. Quinlan AR, Hall IM. BEDTools: A flexible suite of utilities for
@@
 
-\hypertarget{ref-r-core}{}
+\leavevmode\hypertarget{ref-r-core}{}%
 5. R Core Team. R: A language and environment for statistical computing
@@
 
-\hypertarget{ref-Lawrence2013-wg}{}
+\leavevmode\hypertarget{ref-Lawrence2013-wg}{}%
 6. Lawrence M, Huber W, Pagès H, Aboyoun P, Carlson M, Gentleman R, et
@@
 
-\hypertarget{ref-Huber2015-ei}{}
+\leavevmode\hypertarget{ref-Huber2015-ei}{}%
 7. Huber W, Carey VJ, Gentleman R, Anders S, Carlson M, Carvalho BS, et
@@
 
-\hypertarget{ref-Dale2011-js}{}
+\leavevmode\hypertarget{ref-Dale2011-js}{}%
 8. Dale RK, Pedersen BS, Quinlan AR. Pybedtools: A flexible python
@@
 
-\hypertarget{ref-Kent2017}{}
+\leavevmode\hypertarget{ref-Kent2017}{}%
 9. Riemondy KA, Sheridan RM, Gillen A, Yu Y, Bennett CG, Hesselberth JR.
@@
 
-\hypertarget{ref-Wickham2014-jc}{}
+\leavevmode\hypertarget{ref-Wickham2014-jc}{}%
 10. Wickham H. Tidy data. Journal of Statistical Software, Articles.
@@
 
-\hypertarget{ref-Wickham2017-dplyr}{}
+\leavevmode\hypertarget{ref-Wickham2017-dplyr}{}%
 11. Wickham H, Francois R, Henry L, Müller K. Dplyr: A grammar of data
@@
 
-\hypertarget{ref-Lawrence2009-nt}{}
+\leavevmode\hypertarget{ref-Lawrence2009-nt}{}%
 12. Lawrence M, Gentleman R, Carey V. Rtracklayer: An R package for
@@
 
-\hypertarget{ref-R-magrittr}{}
+\leavevmode\hypertarget{ref-R-magrittr}{}%
 13. Bache SM, Wickham H. Magrittr: A forward-pipe operator for r
@@
 
-\hypertarget{ref-R-rlang}{}
+\leavevmode\hypertarget{ref-R-rlang}{}%
 14. Henry L, Wickham H. Rlang: Functions for base types and core r and
@@
 
-\hypertarget{ref-R-ahub}{}
+\leavevmode\hypertarget{ref-R-ahub}{}%
 15. Morgan M. AnnotationHub: Client to access annotationhub resources.
@@
 
-\hypertarget{ref-Roadmap_Epigenomics_Consortium2015-pr}{}
+\leavevmode\hypertarget{ref-Roadmap_Epigenomics_Consortium2015-pr}{}%
 16. Roadmap Epigenomics Consortium, Kundaje A, Meuleman W, Ernst J,
@@
 
-\hypertarget{ref-R-knitr}{}
+\leavevmode\hypertarget{ref-R-knitr}{}%
 17. Xie Y. Dynamic documents with R and knitr {[}Internet{]}. 2nd ed.
@@
 
-\hypertarget{ref-R-ggbio}{}
+\leavevmode\hypertarget{ref-R-ggbio}{}%
 18. Yin T, Cook D, Lawrence M. Ggbio: An r package for extending the

### 2018-05-03 paper.Rmd
update figures

@@
 \begin{figure}
-{\centering \includegraphics[width=400pt]{diagrams/diagrams-001}}
+{\centering \includegraphics[width=400pt]{diagrams/GRanges}}
 \caption{An illustration of the GRanges data model for a
 sample from an RNA-seq experiment. The core components of the data model
-inlcude a seqnames column (representing the chromosome), a ranges column
+inlcude a seqname column (representing the chromosome), a ranges column
 which consists of start and end coordinates for a genomic region, and a
@@
 ways. There are four supported matching algorithms: _overlap_,
-_nearest_, _precede_, and _follow_. We can further restrict the
+_nearest_, _precede_, and _follow_ (figures 
+\ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further restrict the
 matching by whether the query is completely _within_ the subject, and
@@
 same direction (strand).
+
+```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators.
+Each join takes query and subject range as input (black and light gray rectangles, respectively).
+An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
+This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
+The join semantics alter what is returned: for an \textbf{inner} join the query range is returned for each match, for an \textbf{intersect} the intersection is taken between overlapping ranges, and for a \textbf{left} join all query ranges are returned even if the subject range does not overlap them.
+This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
+knitr::include_graphics("diagrams/diagrams-002.png")
+```
 
@@
 
-```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators.
-Each join takes query and subject range as input (black and light gray rectangles, respectively).
-An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
-This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
-The join semantics alter what is returned: for an *inner* join the query range is returned for each match, for an intersect *join* the intersection is taken between overlapping ranges, and for a *left* join all query ranges are returned even if the subject range does not overlap them.
-This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
-knitr::include_graphics("diagrams/diagrams-002.png")
+```{r nn-fig, echo = FALSE, out.width = "400pt", fig.cap="Illustration of neighbour finding joins.
+Each join takes a query and subject range and computes a 'Hits' object.
+For the \textbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range.
+For the \textbf{follow} join there is only one query range that follows any subject range.
+Likewise for the \textbf{precede} join, there is only one query range that precedes a subject range."}
+knitr::include_graphics("diagrams/diagrams-003.png")
 ```
@@
 
-Our use of suffixes in function names, also highlights are core tenet of 
+Our use of suffixes in function names, also highlights a core tenet of 
 the `plyranges` design: avoid complex generic functions with many arguments

### 2018-05-03 paper.tex
update figures

@@
 different ways. There are four supported matching algorithms:
-\emph{overlap}, \emph{nearest}, \emph{precede}, and \emph{follow}. We
-can further restrict the matching by whether the query is completely
-\emph{within} the subject, and adding the \emph{directed} suffix ensures
-that matching ranges have the same direction (strand).
+\emph{overlap}, \emph{nearest}, \emph{precede}, and \emph{follow}
+(figures \ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further
+restrict the matching by whether the query is completely \emph{within}
+the subject, and adding the \emph{directed} suffix ensures that matching
+ranges have the same direction (strand).
+
+\begin{figure}
+
+{\centering \includegraphics[width=400pt]{diagrams/diagrams-002}
+
+}
+
+\caption{Illustration of the three overlap join operators.
+Each join takes query and subject range as input (black and light gray rectangles, respectively).
+An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
+This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
+The join semantics alter what is returned: for an     extbf{inner} join the query range is returned for each match, for an    extbf{intersect} the intersection is taken between overlapping ranges, and for a    extbf{left} join all query ranges are returned even if the subject range does not overlap them.
+This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations.}\label{fig:olaps-fig}
+\end{figure}
 
@@
 
-{\centering \includegraphics[width=400pt]{diagrams/diagrams-002}
+{\centering \includegraphics[width=400pt]{diagrams/diagrams-003}
 
@@
 
-\caption{Illustration of the three overlap join operators.
-Each join takes query and subject range as input (black and light gray rectangles, respectively).
-An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
-This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
-The join semantics alter what is returned: for an *inner* join the query range is returned for each match, for an intersect *join* the intersection is taken between overlapping ranges, and for a *left* join all query ranges are returned even if the subject range does not overlap them.
-This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations.}\label{fig:olaps-fig}
+\caption{Illustration of neighbour finding joins.
+Each join takes a query and subject range and computes a 'Hits' object.
+For the   extbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range.
+For the   extbf{follow} join there is only one query range that follows any subject range.
+Likewise for the   extbf{precede} join, there is only one query range that precedes a subject range.}\label{fig:nn-fig}
 \end{figure}
@@
 
-Our use of suffixes in function names, also highlights are core tenet of
+Our use of suffixes in function names, also highlights a core tenet of
 the \texttt{plyranges} design: avoid complex generic functions with many

### 2018-05-04 paper.Rmd
expand on description of GRanges and how it is tidy
the goal is to give a more complete argument as to why we chose GRanges

@@
 The `plyranges` DSL is built on the core Bioconductor data structure
-GRanges, which is capable of
-representing all types of genomic data at a semantic level that
-roughly matches the intuition of most users.  A GRanges is essentially
-a table, with columns for the chromosome, start and end coordinates,
-and the strand, along with an arbitrary set of additional columns,
-consisting of measurements or metadata specific to the data type or
-experiment (figure \ref{fig:GRanges}).
+GRanges, which is essentially a constrained table, with fixed columns
+for the chromosome, start and end coordinates, and the strand, along
+with an arbitrary set of additional columns, consisting of
+measurements or metadata specific to the data type or experiment
+(figure \ref{fig:GRanges}).  GRanges balances flexibility with formal
+constraints, so that it is applicable to virtually any genomic
+workflow, while also being semantically rich enough to support
+high-level operations on genomic ranges. As a core data structure,
+GRanges enables interoperability between plyranges and the rest of
+Bioconductor. Adhering to a single data structure simplifies the API
+and makes it easier to learn and understand, in part because
+operations become endomorphic, i.e., they return the same type as
+their input.
+
+GRanges follows the intuitive tidy data pattern: it is a rectangular
+table corresponding to a single biological context. Each row contains
+a single observation and each column is a variable describing the
+observations.  GRanges specializes the tidy pattern in that the
+observations always pertain to some genomic feature, but it largely
+remains compatible with the general relational operations defined by
+dplyr. Thus, we define our algebra as an extension of the dplyr
+algebra, and borrow its syntax conventions and design principles.
 
@@
 
-By definition GRanges follows the tidy data pattern: it is a rectangular 
-table corresponding to a single biological context. Each row contains a
-single observation and each column is a variable about that
-observation. Hence, we have designed the `plyranges` DSL to extend the grammar
-and design principles of `dplyr`:  cohesion, consistency, endomorphism,
-and fluency. All of these principles are defined and discussed
-below in the context of the GRanges class and an overview of 
-the grammar is provided in table \ref{tab:grammar}. Where applicable we contrast
-our design to the existing Bioconductor infrastructure.
-
 ## Algebraic operations
@@
 that cover most use cases in genomics data analysis. There are
-operators for performing arithmetic such as modifying width or performing
-set operations, merging such as finding overlaps between GRanges, 
-subsetting by genomic regions, and aggregating over columns within in a GRanges
-object.
+range arithmetic operators, such as for resizing ranges or finding their
+intersection, and operators for merging, filtering and aggregating by
+range-specific notions like overlap and proximity.
 
-GRanges describes the within-sequence coordinates of a range by its
-_start_, _end_ and _width_. Those three variables are mutually
+Arithmetic operations transform range coordinates, as defined by their
+_start_, _end_ and _width_. The three dimensions are mutually
 dependent and partially redundant, so direct manipulation of them is

### 2018-05-04 paper.Rmd
some tweaks to the wording in the algebra operation section

@@
 and merging. The parallel operations map to infix operators, which we
-surround with _%_ symbols in accordance with R syntax rules.  For
+surround with _%_ symbols, a convention borrowed from R syntax.  For
 example, the parallel intersection (_x %intersect% y_) finds the
@@
 finds the gap separating _xi_ and _yi_. The binary operations are
-callable from within the arithmetic, restriction and aggregation
+callable from within arithmetic, restriction and aggregation
 expressions.
 
-Our algebra recasts finding overlaps or nearest neighbours between two
+To support merging, our algebra recasts finding overlaps or 
+nearest neighbours between two
 genomic regions as variants of the relational join operator. A join

### 2018-05-08 paper.Rmd

@@
 
-Since the GRanges object is a tabular data structure our grammar includes
+Since the GRanges object is a tabular data structure, our grammar includes
 operators to filter, sort and aggregate by columns in a GRanges. These 
-operations can be performed over partitions within GRanges object
+operations can be performed over partitions formed
 using the *group_by* modifier. Together with our algebra for arithmetic and
-merging, these operations enable flexible and reproducible analysis using
-the desgin principles of the `dplyr` grammar.
+merging, these operations conform to the semantics and syntax of the 
+`dplyr` grammar.
 
-## Cohesion
+## Design principles
 
-A function is cohesive if it performs a singular task and does not produce
-any side-effects. In the `plyranges` DSL our algebra for performing
-genomic arithmetic is a key example of cohesion. We define anchoring operators
-that decorate a GRanges object with an 'anchor' and modify the semantics
-of performing genomic arithmetic. The anchoring operators amount to fixing a GRanges
-object by its start, center, or end coordinates (or fixing these coordinates
-by strand). This enables any arithmetic function that performs a 
-coordinate transformation to remain cohesive: that is they always perform
-the same operation regardless of whether a GRanges object is anchored or not.
-The only difference is the result of performing the arithmetic changes when
-contextual information is added to the GRanges object.
+The design of plyranges adheres to well understood principles of
+language and API design: cognitive consistency, cohesion,
+expressiveness and endomorphism [cite: Green's cognitive
+dimensions]. To varying degrees, these principles also underly the
+design of dplyr and the Bioconductor infrastructure.
 
-Another example of our algebra altering object semantics, while maintaining 
-cohesion is through the 'group_by' operator. Like anchoring, this operator 
-decorates a GRanges object with a column name (or names) that defines a 
-partitioning of the  GRanges by the unique values in the column(s). Functions 
-defined in the  `plyranges` DSL that perform restriction, aggregation, or 
-column modification still perform those singular tasks, however grouping changes
-how those tasks are performed.
+### Cognitive consistency and fluency
 
-## Consistency
+We aim for our interfaces to have a simple and direct mapping to the
+user's cognitive model, i.e., how the user thinks about the data. This
+requires careful selection of the level of abstraction so that the
+user can express workflows in the language of genomics. This motivates
+the adoption of the tidy GRanges object as our central data
+structure. The basic data.frame and dplyr tibble lack any notion of
+genomic ranges and so could not easily support our genomic grammar,
+with its specific verbs for range-oriented data manipulation. Another
+example of cognitive consistency is how plyranges is insensitive to
+direction/strand by default when, e.g., detecting
+overlaps. GenomicRanges has the opposite behavior. We believe that
+defaulting to purely spatial overlap is most intuitive to most users.
 
-A core design principle of the `plyranges` DSL is interface consistency:
-a user should not be surprised by the input or output of `plyranges` code.
-A key example of consistency is how `plyranges` handles strand information.
-Every function that computes with strand information indicates its intentions
-by including suffixes such as 'directed', 'upstream' or 'downstream' in its
-name, otherwise strand is ignored. This strongly differs from 
-the Bioconductor packages, which produces surprising output by assuming the 
-user is always interested in using strand in the majority of circumstances 
-(unless a user is computing coverage or finding flanking ranges).
+<!--
+ML:
+Hadley actually presents dplyr as a 'cognitive tool', i.e., something
+that helps users think about the problem. In my opinion, the best we
+can do is attempt to match the intuition of the user, so when the user
+decides what to do, there is an obvious way of doing it. I'm not ready
+to claim that the user should start from our grammar.
+-->
 
-Our use of suffixes in function names, also highlights a core tenet of 
-the `plyranges` design: avoid complex generic functions with many arguments
-and instead use cohesive functions with a minimal number of arguments. As an
-example, we have written a consistent framework for reading and writing 
-files from and to common genomic data formats as GRanges, 
-using the `rtracklayer` package as a back-end @Lawrence2009-nt. We have replaced
-that packages generic functions for importing files with a family of reader
-functions that all take the exact same arguments.
+Like dplyr, plyranges verbs are functional: they are free of side
+effects and return their result. This enables chaining of verbs
+through syntax like the forward pipe operator (`%>%`, read as "then")
+of the magrittr package @R-magrittr. This has syntax has a direct
+cognitive mapping to natural language and the intuitive notion of
+pipelines. The low-level object-oriented APIs of Bioconductor tend to
+manipulate data via subreplacement functions, like `start(gr) <-
+x`. These ultimately produce the side effect of replacing a symbol
+mapping in the current environment and thus are not amenable to
+so-called fluent syntax.
 
-## Endomorphism
+### Cohesion
 
-Most function calls in `plyranges` are endomorphisms: when the input 
-is GRanges object the output will also be a GRanges object. The use of
-endomorphism in the `plyranges` DSL enables a user to predict the structure of 
-the output of their computations and does not require them to learn any 
-additional classes beyond GRanges and DataFrames.
+A function is cohesive if it performs a singular task without
+producing any side-effects. Singular tasks are not necessarily atomic;
+they can always be broken down further at lower levels of abstraction.
+For example, to resize a range, the user needs to specify which
+position (start, end, midpoint) should be invariant over the
+transformation. The `resize()` function from the GenomicRanges package
+has a `fix` argument that sets the anchor, so calling `resize()`
+coalesces anchoring and width modification. The coupling at the
+function call level is justified since the effect of setting the width
+depends on the anchor. However, plyranges increases cohesion and
+decouples the anchoring into its own function call.
 
-As an example, in `plyranges` when we compute coverage over a GRanges object,
-the result of the operation is an expanded GRanges object with a score column,
-corresponding the estimated coverage over a genomic region. Hence, the resulting
-coverage score can easily be composed with other expressions in our algebra.
-This pattern strongly deviates from the design of the OO interface in 
-`GenomicRanges`, which in the case of computing coverage would return an RleList,
-an object that has many new methods and that is potentially unfamiliar to 
-a new user. While, these low-level classes enable efficient computing,
-the increase complexity, and as a consequence are abstracted away in the 
-`plyranges` DSL.
+Increasing cohesion simplifies the interface to each operation, makes
+the meaning of arguments more intuitive, and relies on function names
+as the primary means of expression, instead of a more complex mixture
+of function and argument names. Since functions are superordinate to
+their arguments, flattening the API at the function level enables the
+user to conceptualize the API as a catalog of functions, without
+having to descend further. A flat function catalog also enhances API
+discoverability, particularly through auto-completion in IDEs. One
+downside of pushing cohesion to this extreme is that function calls
+become coupled, and care is necessary to treat them as a group when
+modifying code.
 
-## Fluency
+### Expressiveness
 
-As a consequence of the design principles defined above every function in `plyranges`
-performs a single action on GRanges objects. The `plyranges` DSL implements the 
-core verbs from the `dplyr` package and implements a genomic relational algebra
-for transforming GRanges objects (table x, y). Each verb preserves the 
-semantics of GRanges object and works with derivatives of the GRanges class. 
-Both of these aspects reduce the cognitive load on a new user since most 
-manipulations can be performed with a vocabulary of several verbs, rather 
-than having to memorise function names that are nouns.
-
-This approach strongly contrasts the `GenomicRanges/IRanges` OO interface, 
-which emphasises the use of setter and getter methods. In that interface, 
-core components and metadata are updated via replacement methods 
-(requiring knowledge of the class components), while our interface requires 
-only a single call to `mutate()` to perform the modification.
-
-Workflows can be composed by chaining verbs together into 'sentences' via the 
-forward pipe operator,`%>%`  (exported from the R package `magrittr` 
-@R-magrittr), which can be read as the word 'then'. Overall, this allows users 
-to write human readable code because workflows describe what the code is doing 
-rather than how its doing it.
-
-## Opportunities
-
-A caveat to constructing a compatible interface with `dplyr` is that 
-`plyranges` makes extensive use of non-standard evaluation in R
-(achieved via the `rlang` package @R-rlang). Simply, this
-means that computations are performed and evaluated in the context of the 
-GRanges objects; emphasising the interactive nature of our API.
-Consequently, when programming with `plyranges` a user needs to be aware of
-how non-standard evaluation in R works and how to adapt their code 
-accordingly. However, with the rise of R packages like `rlang` this process
-is becoming less difficult.
-
-While GRanges are an intuitive representation for data measured on genomic regions,
-more flexible data structures are required to represent data from multiple sample
-experiments. The Bionconductor class SummarizedExperiment is the canonical 
-data structure for representing data for combining multiomic measurements 
-from multiple samples. The grammar and design of the `plryanges` DSL can be
-naturally extended to the SummarizedExperiment.
+Expressiveness relates to the information content in code: the
+programmer should be able to clarify intent without unnecessary
+verbosity. For example, our overlap-based join operations are more
+concise than the multiple steps necessary to achieve the same effect
+in the original GenomicRanges API. In other cases, the plyranges API
+increases verbosity for the sake of clarity and cohesion. Explicitly
+calling `anchor()` can require more typing, but the code is easier to
+comprehend. Another example is the set of routines for importing
+genomic annotations, including `read\_gff()`, `read\_bed()`, and
+`read_bam()`. Compared to the generic `import()` in rtracklayer, the
+explicit format-based naming in plyranges clarifies intent and the
+type of data being returned. Similarly, every plyranges function that
+computes with strand information indicates its intentions by including
+suffixes such as `directed`, `upstream` or `downstream` in its name,
+otherwise strand is ignored. The GenomicRanges API does not make this
+distinction explicit in its function naming, instead relying on a
+parameter that defaults to strand sensitivity, an arguably unintuitive
+behavior.
 
@@
 
+# Discussion
+
+A caveat to constructing a compatible interface with `dplyr` is that 
+`plyranges` makes extensive use of non-standard evaluation in R
+(achieved via the `rlang` package @R-rlang). Simply, this
+means that computations are performed and evaluated in the context of the 
+GRanges objects; emphasising the interactive nature of our API.
+Consequently, when programming with `plyranges` a user needs to be aware of
+how non-standard evaluation in R works and how to adapt their code 
+accordingly. However, with the rise of R packages like `rlang` this process
+is becoming less difficult.
+
+While GRanges are an intuitive representation for data measured on genomic regions,
+more flexible data structures are required to represent data from multiple sample
+experiments. The Bionconductor class SummarizedExperiment is the canonical 
+data structure for representing data for combining multiomic measurements 
+from multiple samples. The grammar and design of the `plryanges` DSL can be
+naturally extended to the SummarizedExperiment.
+
 # Availablilty and Future Work

### 2018-05-08 paper.Rmd
more detail for bam example

@@
 library(ggbio)
-
 library(plyranges)
@@
   )
+
+h1_bam_sorted <- "./data/GSM433167_BI.H3K27me3.sorted.bam"
+
+# sort and index bam
+if (!file.exists(h1_bam_sorted)) {
+  Rsamtools::sortBam(h1_bam, "./data/GSM433167_BI.H3K27me3.sorted")
+  Rsamtools::indexBam(h1_bam_sorted)
+}
 
@@
 Human Epigenome Roadmap Consortium data, we can count the number of reads
-that fall into a fixed bins of size 10000bp over a BAM file of H3K27Me3 methylation
-marks from the H1 cell line. We extract reads that have a mapping quality
-score greater than 20:
+that fall into a fixed bins of size 10000bp over a sorted and indexed 
+BAM file of H3K27Me3 methylation marks from the H1 cell line.
+
+We can resuse the annotation information from our BigWig file example, as the
+BAM file was aligned to the hg19 genome build. Alignments over
+the mitochondrial genome are dropped using the `dropSeqlevels` function
+from the Bioconductor package `GenomeInfoDb` @R-gdb.
+```{r locations, eval = FALSE}
+locations <- bw_file %>%
+  get_genome_info() %>% 
+  dropSeqlevels("chrM", pruning.mode = "coarse") %>%
+  set_genome_info(genome = "hg19") 
+```
+
+To generate bins of width 10000bp we can use the `tile` function from 
+`GenomicRanges`:
 
 ```{r, eval = FALSE}
-alignments <- read_bam(h1_bam) %>%
-  filter(mapq > 20) 
+bins <- tile(locations, width = 10000)
 ```
 
-Note that the BAM file is only read into memory once we perform an operation
-on it. Because of interoperability over Bioconductor, we can generate
-genomic bins using the `tile` function from `GenomicRanges`:
-
+Next we only read in alignments if they have a map quality score
+greater than 20.
 ```{r, eval = FALSE}
-bins <- tile(h1_bam, width = 10000)
+alignments <- read_bam(h1_bam_sorted) %>%
+    filter(mapq > 20) 
 ```
@@
 ```{r, eval = FALSE}
-h1_bam_read_summary <- h1_bam %>%
-  group_by_overlaps(bins) %>%
+alignments_summary <- bins %>%
+  group_by_overlaps(alginments) %>%
   summarise(n_reads = n())

### 2018-05-09 paper.Rmd
some comments on first example

@@
 
+[ML: please explain why we are doing this. Something like: we need to
+     get the range for chromosome 10, so that we can use it as a
+     filter when reading. The first time I read this, I was not sure what
+     'annotation information' and 'gather' meant here. Why say
+     "annotation information" when the API uses "genome information"?]
+
 ```{r load-bw, cache = TRUE}
@@
 case the hg19 genome build).
+
+[ML: getting the genome info does not look automatic, because we are calling
+	`set_genome_info()`; is that supposed to be here?]
 
@@
 coverage scores in the regions that were reduced.
+
+[ML: I think the above text could do a better job of describing what
+	`reduce_ranges()` does generally. You could say something like, after
+	filtering for regions with a score greater than 8, we can reduce the
+	individual runs to ranges representing the islands of coverage.]
 

### 2018-05-10 paper.Rmd
comments and tweaks for the rest of the paper

@@
 Another common operation in genomics data analysis is to compute data
-summaries over genomic windows. In `plyranges` this can be achieved via
-the `group_by_overlaps` operator. Continuing with the data from the 
-Human Epigenome Roadmap Consortium data, we can count the number of reads
-that fall into a fixed bins of size 10000bp over a sorted and indexed 
-BAM file of H3K27Me3 methylation marks from the H1 cell line.
+summaries over genomic windows. In `plyranges` this can be achieved
+via the `group_by_overlaps` operator.  We bin and count reads from a
+H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
+Consortium. The bins have a fixed width of 10000bp.
 
-We can resuse the annotation information from our BigWig file example, as the
+[ML: why not get the seqinfo from the BAM itself?]
+
+We reuse the genome information from our BigWig file, as the
 BAM file was aligned to the hg19 genome build. Alignments over
@@
 from the Bioconductor package `GenomeInfoDb` @R-gdb.
+[ML: seems like we should have a `select_seqlevels()` or something, so
+     that we could do `select_seqlevels(-chrM)`.. somehow we have to
+     explain to the reader why we are doing all of this seqlevel munging.]
 ```{r locations, eval = FALSE}
@@
 
-To generate bins of width 10000bp we can use the `tile` function from 
+To generate bins of width 10000bp we use the `tile` function from 
 `GenomicRanges`:
@@
 
-Finally, we can use `group_by_overlaps` with `summarise` to compute the
+[ML: seems like a great opportunity to point out the deferred reading]
+
+Finally, we use `group_by_overlaps` with `summarize` to compute the
 total number of reads within each window.
+
+[ML: there was a typo in the code: have you actually run this?]
 
@@
 alignments_summary <- bins %>%
-  group_by_overlaps(alginments) %>%
-  summarise(n_reads = n())
+  group_by_overlaps(alignments) %>%
+  summarize(n_reads = n())
 ```
@@
 H1 cell line, consisting of approximately two million single nucleotide
-polymorphisms (SNP) and short insertion/deletions (indel). The GRanges object consists
+polymorphisms (SNPs) and short insertion/deletions (indels). The GRanges object consists
 of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
@@
 
-First we filter out any insertion or deletion alleles and 
-the SNPs present on the mitochondria then create a logical vector corresponding
-to whether there is a transition event.
+First we filter out the indels and mitochondrial variants. Then we
+create a logical vector corresponding to whether there is a transition
+event.
 
@@
 
-We can then compute the transition-transversion ratio using the `group_by`
-and `summarise` pattern, it is computed as the total number of transition
-SNPs divided by the total number of transversion SNPs within each chromosome
-(figure \ref{fig:titv-viz}).
+We compute the per-chromosome transition-transversion ratio using the
+`group_by` and `summarize` pattern (figure \ref{fig:titv-viz}).
 
@@
   group_by(seqnames) %>%
-  summarise(n_snps = n(),
+  summarize(n_snps = n(),
             ti_tv = sum(transition) / sum(!transition)) 
@@
 
-# Discussion
+# Discussion and conclusion
+
+[ ML: there should be some quick summary paragraph here ]
 
 A caveat to constructing a compatible interface with `dplyr` is that
-`plyranges` makes extensive use of non-standard evaluation in R
-(achieved via the `rlang` package @R-rlang). Simply, this
-means that computations are performed and evaluated in the context of the 
-GRanges objects; emphasising the interactive nature of our API.
-Consequently, when programming with `plyranges` a user needs to be aware of
-how non-standard evaluation in R works and how to adapt their code 
-accordingly. However, with the rise of R packages like `rlang` this process
-is becoming less difficult.
+`plyranges` makes extensive use of non-standard evaluation in R via
+the `rlang` package @R-rlang. Simply, this means that computations are
+evaluated in the context of the GRanges objects. Both dplyr and
+plyranges are based on the `rlang` language, because it allows for more
+expressive code that is free of repeated references to the
+container. Implicitly referencing the container is particularly
+convenient when programming interactively.  Consequently, when
+programming with `plyranges`, a user needs to generally understand
+the `rlang` language and how to adapt their code accordingly. Users
+familiar with the tidyverse should already have such knowledge.
 
-While GRanges are an intuitive representation for data measured on genomic regions,
-more flexible data structures are required to represent data from multiple sample
-experiments. The Bionconductor class SummarizedExperiment is the canonical 
-data structure for representing data for combining multiomic measurements 
-from multiple samples. The grammar and design of the `plryanges` DSL can be
-naturally extended to the SummarizedExperiment.
+We aim to continue developing the `plyranges` package and to extend it
+for use with more complex data structures, such as the
+SummarizedExperiment class, the core Bioconductor data structure for
+representing experimental results (e.g., counts) from multiple sample
+experiments in conjunction with feature and sample metadata. The
+grammar and design of the `plryanges` DSL are naturally extensible to
+SummarizedExperiment.
 
-# Availablilty and Future Work
+As the `plyranges` interface encourages tidy data practices, it
+integrates well with the grammar of graphics [cite ggplot2].  To
+achieve responsive performance, interactive graphics rely on lazy data
+access and computing patterns, so the deferred mechanisms within
+plyranges should help support interactive genomics applications.
 
-The `plyranges` package is available on the Bioconductor project website 
-[https://bioconductor.org](https://bioconductor.org) or can be accessed
-via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
-We aim to continue developing the `plyranges` package and extend it for use
-with more complex data structures  such as the 
-SummarizedExperiment class, which can be used for analysing transcriptomic
-and variant data. As the `plyranges` interface encourages tidy data 
-practices it integrates well with the principles of the grammar of graphics, 
-we aim to use it to prepare data for the visualisation of multimodal biological 
-datasets.
+The `plyranges` package is available on the Bioconductor project
+website [https://bioconductor.org](https://bioconductor.org) or can be
+accessed via
+Github
+[https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
 

### 2018-05-14 paper.Rmd
tidying up - almost there

@@
 
+First, we extract the genome information from the BigWig file and filter
+to get the range for chromosome 10.
+
 First, we gather the BigWig file and extract its annotation information and 
-filter it to chromosome 10.
-
-[ML: please explain why we are doing this. Something like: we need to
-     get the range for chromosome 10, so that we can use it as a
-     filter when reading. The first time I read this, I was not sure what
-     'annotation information' and 'gather' meant here. Why say
-     "annotation information" when the API uses "genome information"?]
+filter it to chromosome 10. This range will be used as  a filter when reading
+the file.
 
@@
 Then we read the BigWig file only extracting scores if they overlap chromosome
-10. The annotation information from the file is automatically included (in this
-case the hg19 genome build).
-
-[ML: getting the genome info does not look automatic, because we are calling
-	`set_genome_info()`; is that supposed to be here?]
+10. We also add the genome build information to the resulting ranges.
 
@@
 
-The `reduce_ranges()` operation is used to find coverage peaks 
-across chromosome 10. We can manually set a threshold to restrict genomic
-regions to have a coverage score of greater than 8, and
-then merge nearby regions. The maximum coverage is computed over all the
-coverage scores in the regions that were reduced.
-
-[ML: I think the above text could do a better job of describing what
-	`reduce_ranges()` does generally. You could say something like, after
-	filtering for regions with a score greater than 8, we can reduce the
-	individual runs to ranges representing the islands of coverage.]
+After filtering for regions with a coverage score greater than 8, we can
+reduce individual runs to ranges representing the islands of coverage,
+using the `reduce_ranges()` function. This function allows a summary
+to be computer over each island: in this case we take the maximum of 
+the scores to find the coverage peaks over chromosome 10.
 
@@
 summaries over genomic windows. In `plyranges` this can be achieved
-via the `group_by_overlaps` operator.  We bin and count reads from a
+via the `group_by_overlaps()` operator.  We bin and count reads from a
 H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
@@
 
-[ML: why not get the seqinfo from the BAM itself?]
+We can directly obtain the genome information from the header of the
+BAM file: in this case the reads were aligned to the hg19 genome build
+and there are no reads overlapping the mitochondrial genome. To generate 
+bins of width 10000bp we apply `tile` function from `GenomicRanges`
+to the genomic coordinates:
+```{r bins, eval = TRUE}
+locations <- h1_bam_sorted %>%
+  read_bam() %>% 
+  get_genome_info()
 
-We reuse the genome information from our BigWig file, as the
-BAM file was aligned to the hg19 genome build. Alignments over
-the mitochondrial genome are dropped using the `dropSeqlevels` function
-from the Bioconductor package `GenomeInfoDb` @R-gdb.
-[ML: seems like we should have a `select_seqlevels()` or something, so
-     that we could do `select_seqlevels(-chrM)`.. somehow we have to
-     explain to the reader why we are doing all of this seqlevel munging.]
-```{r locations, eval = FALSE}
-locations <- bw_file %>%
-  get_genome_info() %>% 
-  dropSeqlevels("chrM", pruning.mode = "coarse") %>%
-  set_genome_info(genome = "hg19") 
+bins <- locations %>% 
+  tile(width = 10000L)
 ```
 
-To generate bins of width 10000bp we use the `tile` function from 
-`GenomicRanges`:
+In this case using `read_bam()` in conjunction with `get_genome_info()`
+only loads the genome information and nothing else.
 
-```{r, eval = FALSE}
-bins <- tile(locations, width = 10000)
+Next we only read in alginments that overlap the genomic locations
+we are interested in. The reading is deferred until after the filter is 
+constructed: that is only alignments that pass the filter are loaded into memory.
+
+```{r, eval = TRUE, cache = TRUE}
+alignments <- h1_bam_sorted %>% 
+    read_bam() %>%
+    select(qwidth) %>%
+
+    filter_by_overlaps(locations) %>%
 ```
-
-Next we only read in alignments if they have a map quality score
-greater than 20.
-```{r, eval = FALSE}
-alignments <- read_bam(h1_bam_sorted) %>%
-    filter(mapq > 20) 
-```
-
-[ML: seems like a great opportunity to point out the deferred reading]
 
@@
 
-[ML: there was a typo in the code: have you actually run this?]
-
 ```{r, eval = FALSE}
 alignments_summary <- bins %>%
-  group_by_overlaps(alignments) %>%
+  join_ %>%
   summarize(n_reads = n())
@@
 
-[ ML: there should be some quick summary paragraph here ]
+We have shown how to create expressive and 
+reproducible genomic workflows using the `plyranges` DSL. By
+realising that the GRanges data model is tidy we have highlighted how to 
+implement a grammar for performing genomic arithemtic, aggregation, restriction 
+and merging. Our examples show that `plyranges` code is succinct and human
+readable.
 

### 2018-05-15 README.Rmd
to generic style over plos - ready for biorxiv

@@
 )
+library(tibble)
+library(knitr)
+library(ggbio)
+library(plyranges)
+library(stringr)
+library(dplyr)
+library(readr)
+library(Biostrings)
+library(AnnotationHub)
 ```
 # plyranges-paper
+The setup chunk in the [paper](paper.Rmd) shows how to download the data with R.
 
-Overview of structure, references and layout for the plyranges package paper.
+Requires the following packages (and versions) to reproduce the paper:
 
-# Abstract
-
-The Bioconductor project has created many powerful abstractions for
-reasoning about genomics data, such as the _Ranges_ data structures 
-for representing genomic intervals. By recognising that these data structures
-follow 'tidy' data principles we have created a grammar of genomic data 
-manipulation that defines verbs for performing actions on and between genomic 
-interval data. This grammar simplifies performing common genomic data analysis 
-tasks via method chaining, type consistency and results in creating human 
-readable pipelines. We have implemented this grammar as an Bioconductor/R 
-package  plyranges.
-
-# Introduction
-
-Putting the work into context.
-
-## What do we mean by genomic data?
-
-  - How does that relate to 'tidy' data?
-  - Why tidy data is a useful abstraction.
-  - Why _Ranges_ follow this pattern.
-
-Key references: @Wickham2014-jc, @Lawrence2013-wg
-
-## How can we analyse genomic data?
-
-  - the Ranges data structures
-  - other abstractions for analysing genomic data
-    - the query language approaches
-    - the bedtools approach
-    - relational algebras
-
-Key references: @Quinlan2010-gc, @Dale2011-js, @Kaitoua2017-pw, @Ovaska2013-gd, @Mungall2014-dg @Kozanitis2016-bm
-
-## What does plyranges contribute?
-
-`plyranges` API provides
-
-- a consistent grammar for analysing genomic data.
-- fluent interface enables readable pipelines 
-- reproducibility via code clarity, declaraitve programming,  and return consistency
-
-Key references:
-
-# Design and Implementation
-
-Discussion of the API and its key features
-
-- every action on a Range is a verb (functional composition, method chaining)
-- fluent (human readable, code describes what to do rather than 'how')
-- every function returns a class familiar to the user
-- an expressive algebra for performing arithmetic (anchoring functions)
-- split-apply-combine strategies (group_by + reduce/summarise/mutate/filter)
-- recasting overlapping/nearest etc as joins
-
-# Results
-
-A case study showing plyranges makes life easier...
-Perhaps could use long reads data from Matt and his post-doc Charity (interesting
-use case of looking for intron-exon junctions)
-
-# Conclusion (future directions)
-
-- availability and maintenance of package
-- extending to SummarisedExperiment class
-- natural fit with ggbio2
-
-# References
+```{r, warning=FALSE}
+devtools::session_info()
+```

### 2018-05-15 paper.Rmd
to generic style over plos - ready for biorxiv

@@
     affiliation: Monash University
-address:
-  - code: Monash University
-    address: Department of Econometrics and Business Statistics, Clayton, Victoria, Australia
-  - code: Genentech
-    address: Bioinformatics and Computational Biology, Genentech, Inc., South San Francisco, California, United States of America 
-abstract: |
-    The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments.
+abstract: "The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments.
 However, there is a cognitive load placed on a user in learning a data abstraction and understanding its appropriate use.
@@
 The grammar defines verbs for performing actions on and between genomic interval data.
-It provides a principled way
-    of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows.
-We have implemented this grammar as a Bioconductor/R package called plyranges.
-
+It provides a principled way of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows.
+We have implemented this grammar as a Bioconductor/R package called plyranges."
+keywords: "Bioconductor, Grammar, Genomes, Data Analysis"
 bibliography: references.bib
 output: 
-  rticles::plos_article:
-csl: plos.csl
+  pdf_document:
+    template: svm-latex-ms.tex
+    keep_tex: true
+    fig_caption: true
+    latex_engine: pdflatex
+    citation_package: biblatex
+fontfamily: mathpazo
 ---
+
 ```{r setup, include = FALSE}
@@
 library(ggbio)
+
+# for bam example
+library(Biostrings)
 library(plyranges)
@@
 SQL-like syntax for fast retrieval of information of unprocessed 
-sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric 
+sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the Genometric 
 Query Language (GMQL) implements a relational algebra for combining genomic datasets 
-@Kaitoua2017-pw.  The command line application BEDtools develops an extensive 
+\cite{Kaitoua2017-pw}.  The command line application BEDtools develops an extensive 
 algebra for performing arithmetic between two or more sets of genomic regions 
-@Quinlan2010-gc. All of the aforementioned DSLs are designed to be evaluated
+\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be evaluated
 either at the command line or embedded in scripts for batch processing. They 
@@
 R data structures and workflows around the notion of tidy data
-@Wickham2014-jc. A tidy dataset is defined as a tabular
+\cite{Wickham2014-jc}. A tidy dataset is defined as a tabular
 data structure that has observations as rows and columns as variables,
@@
 us to see how the data relate to the design of an experiment
-and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
+and the variables measured. The `dplyr` package \cite{Wickham2017-dplyr} 
 defines an API that maps notions from the general relational algebra to 
@@
 \begin{figure}
-{\centering \includegraphics[width=400pt]{diagrams/GRanges}}
+\includegraphics[width=\textwidth]{diagrams/GRanges}
 \caption{An illustration of the GRanges data model for a
@@
 are included as columns to the right of the dotted line as annotations
-(gene\_id) or range level covariates (GC score and count).}
+(gene\_id) or range level covariates (score).}
 \label{fig:GRanges} 
@@
 \centering
-\begin{tabular}{|l|l|p{4cm}|}
+\begin{tabular}{|l|l|p{6cm}|}
   \hline
@@
   \hline
-   & \textbf{\emph{anchor\_direction()}} & fix coordinates at direction \\
+   & \emph{anchor\_direction()} & fix coordinates at direction \\
   Modifier & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
@@
 
-```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators.
+```{r olaps-fig, echo = FALSE, out.width="\\textwidth", fig.cap="Illustration of the three overlap join operators.
 Each join takes query and subject range as input (black and light gray rectangles, respectively).
@@
 
-```{r nn-fig, echo = FALSE, out.width = "400pt", fig.cap="Illustration of neighbour finding joins.
+```{r nn-fig, echo = FALSE, out.width = "\\textwidth", fig.cap="Illustration of neighbour finding joins.
 Each join takes a query and subject range and computes a 'Hits' object.
@@
 language and API design: cognitive consistency, cohesion,
-expressiveness and endomorphism [cite: Green's cognitive
-dimensions]. To varying degrees, these principles also underly the
+expressiveness and endomorphism \cite{Green1996-qg}. 
+To varying degrees, these principles also underly the
 design of dplyr and the Bioconductor infrastructure.
@@
 -->
-
 Like dplyr, plyranges verbs are functional: they are free of side
@@
 through syntax like the forward pipe operator (`%>%`, read as "then")
-of the magrittr package @R-magrittr. This has syntax has a direct
+of the magrittr package \cite{R-magrittr}. This has syntax has a direct
 cognitive mapping to natural language and the intuitive notion of
@@
 
-Here we provide illustrative examples by using the `plyranges` DSL to show
+Here we provide illustrative examples of using the `plyranges` DSL to show
 how our grammar could be integrated into genomic data workflows. We also
@@
 
-The Bioconductor package `AnnotationHub` @R-ahub  can be used to search for 
+The Bioconductor package `AnnotationHub` \cite{R-ahub}  can be used to search for 
 BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project 
-@Roadmap_Epigenomics_Consortium2015-pr. Here we focus on assays for primary
+\cite{Roadmap-Epigenomics-Consortium2015-pr}. Here we focus on assays for primary
 T CD8+ memory cells from peripheral blood. Using `plyranges` we will read the 
@@
 First, we extract the genome information from the BigWig file and filter
-to get the range for chromosome 10.
-
-First, we gather the BigWig file and extract its annotation information and 
-filter it to chromosome 10. This range will be used as  a filter when reading
-the file.
+to get the range for chromosome 10. This range will be used as a filter when 
+reading the file.
 
@@
 Returning to the GRanges object containing normalised coverage scores, 
-we filter to find the coordinates of the peak 
-containing the maximum coverage score. We can then find 
-a 5000 nt region centered around the maximum position by anchoring and 
-modifying the the width.
+we filter to find the coordinates of the peak  containing the maximum coverage 
+score. We can then find a 5000 nt region centered around the maximum position 
+by anchoring and modifying the width.
 
@@
                 geom = "blank") + 
-  geom_area(aes(x = start))
+  geom_area(aes(x = start)) +
+  ylab("score")
 ```
@@
 summaries over genomic windows. In `plyranges` this can be achieved
-via the `group_by_overlaps()` operator.  We bin and count reads from a
+via the `group_by_overlaps()` operator.  We bin and count and
+find the average GC content of reads from a
 H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
-Consortium. The bins have a fixed width of 10000bp.
+Consortium.
 
@@
 and there are no reads overlapping the mitochondrial genome. To generate 
-bins of width 10000bp we apply `tile` function from `GenomicRanges`
-to the genomic coordinates:
-```{r bins, eval = TRUE}
+bins of fixed width 10000bp we apply `tile()` function from `GenomicRanges`
+to the genomic coordinates (with `unlist` to return a GRanges object)
+```{r bins}
 locations <- h1_bam_sorted %>%
@@
 bins <- locations %>% 
-  tile(width = 10000L)
+  tile(width = 10000L) %>%
+  unlist(use.names = FALSE)
 ```
 
-In this case using `read_bam()` in conjunction with `get_genome_info()`
-only loads the genome information and nothing else.
+Next we only read in alginments that overlap the genomic locations
+we are interested in and select the query seqeunce. Note that
+the reading of the BAM file is deferred: only alignments that pass the filter 
+are loaded into memory. We can add another column representing the GC proportion 
+for each alignment using the `letterFrequency()` function from the `Biostrings` 
+package \cite{R-biostrings}.
 
-Next we only read in alginments that overlap the genomic locations
-we are interested in. The reading is deferred until after the filter is 
-constructed: that is only alignments that pass the filter are loaded into memory.
-
-```{r, eval = TRUE, cache = TRUE}
+```{r, cache = TRUE}
 alignments <- h1_bam_sorted %>% 
   read_bam() %>%
-    select(qwidth) %>%
-
   filter_by_overlaps(locations) %>%
+  select(seq) %>%
+  mutate(score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE)))
 ```
 
-Finally, we use `group_by_overlaps` with `summarize` to compute the
-total number of reads within each window.
+Finally, we use `group_by_overlaps()` with `summarize()` to compute the
+total number of reads and averge GC content within each genomic window.
 
-```{r, eval = FALSE}
+```{r, cache = TRUE}
 alignments_summary <- bins %>%
-  join_ %>%
-  summarize(n_reads = n())
+  group_by_overlaps(alignments %>% select(score)) %>% 
+  summarize(n_reads = n(), avg_gc = mean(score))
 ```
@@
 
-```{r}
+```{r, cache = TRUE}
 h1_snp_array <- h1_snp_array %>%
   filter(!(ref %in% c("I", "D")), seqnames != "M") %>%
-  mutate(transition = (ref %in% c("A", "G") & alt %in% c("G","A")) |
+  mutate(transition = (ref %in% c("A", "G") & alt %in% c("G","A"))|
                       (ref %in% c("C","T") & alt %in% c("T", "C")))
@@
 
-We compute the per-chromosome transition-transversion ratio using the
+We can then compute the per-chromosome transition-transversion ratio using the
 `group_by` and `summarize` pattern (figure \ref{fig:titv-viz}).
 
-```{r}
+```{r, cache = TRUE}
 ti_tv_results <- h1_snp_array %>% 
@@
 implement a grammar for performing genomic arithemtic, aggregation, restriction 
-and merging. Our examples show that `plyranges` code is succinct and human
-readable.
+and merging. Our examples show that `plyranges` code is succinct, human
+readable and can take advantage of the interoperability provided by
+the Bioconductor ecosytem and the R language.
 
@@
 `plyranges` makes extensive use of non-standard evaluation in R via
-the `rlang` package @R-rlang. Simply, this means that computations are
+the `rlang` package \cite{R-rlang}. Simply, this means that computations are
 evaluated in the context of the GRanges objects. Both dplyr and
@@
 As the `plyranges` interface encourages tidy data practices, it
-integrates well with the grammar of graphics [cite ggplot2].  To
+integrates well with the grammar of graphics \cite{Wickham2016-gz}.  To
 achieve responsive performance, interactive graphics rely on lazy data
@@
 for the code review they performed. This report was written with
-`knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio. All code
+`knitr` \cite{R-knitr} and the figures were made with `ggbio` \cite{R-ggbio}. All code
 required to reproduce this article is available at 

### 2018-05-15 paper.tex
to generic style over plos - ready for biorxiv

@@
-% Template for PLoS
-% Version 3.4 January 2017
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+% modified from https://github.com/svmiller/svm-r-markdown-templates
+\documentclass[]{article}
+\usepackage[top=0.85in,right=2in]{geometry}
+\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
+\usepackage[]{mathpazo}
 
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
 
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{1em}
 
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
+\usepackage{abstract}
+\renewcommand{\abstractname}{}    % clear the title
+\renewcommand{\absnamepos}{empty} % originally center
 
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
+\renewenvironment{abstract}
+ {{%
+    \setlength{\leftmargin}{0mm}
+    \setlength{\rightmargin}{\leftmargin}%
+  }%
+  \relax}
+ {\endlist}
 
-% cite package, to clean up citations in the main text. Do not remove.
-% \usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
+\makeatletter
+\def\@maketitle{%
+  \newpage
+%  \null
+%  \vskip 2em%
+%  \begin{center}%
+  \let \footnote \thanks
+    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
 }
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-% \bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
+%\fi
 \makeatother
 
-% Leave date blank
-\date{}
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
-
-%% Include all macros below
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
+\setcounter{secnumdepth}{0}
 
@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
 \newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
 \newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 
-\usepackage{forarray}
-\usepackage{xstring}
-\newcommand{\getIndex}[2]{
-  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
+\usepackage{graphicx,grffile}
+\usepackage[labelfont=bf]{caption}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+
+\title{plyranges: a grammar for transforming genomics data  }
+
+\author{\Large Stuart Lee\vspace{0.05in} \newline\normalsize\emph{Monash University}   \and \Large Michael Lawrence\vspace{0.05in} \newline\normalsize\emph{Genentech}   \and \Large Di Cook\vspace{0.05in} \newline\normalsize\emph{Monash University}  }
+
+\date{}
+
+\usepackage{titlesec}
+
+\titleformat*{\section}{\normalsize\bfseries}
+\titleformat*{\subsection}{\normalsize\itshape}
+\titleformat*{\subsubsection}{\normalsize\itshape}
+\titleformat*{\paragraph}{\normalsize\itshape}
+\titleformat*{\subparagraph}{\normalsize\itshape}
+
+% % 
+% % \usepackage{biblatex}
+% 
+\usepackage[backend=biber, style=numeric-comp, url=false, sorting=none]{biblatex}
+\addbibresource{references.bib}
+
+\newtheorem{hypothesis}{Hypothesis}
+\usepackage{setspace}
+
+\makeatletter
+\@ifpackageloaded{hyperref}{}{%
+\ifxetex
+  \PassOptionsToPackage{hyphens}{url}\usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \PassOptionsToPackage{hyphens}{url}\usepackage[unicode=true]{hyperref}
+\fi
 }
 
-\setcounter{secnumdepth}{0}
+\@ifpackageloaded{color}{
+    \PassOptionsToPackage{usenames,dvipsnames}{color}
+}{%
+    \usepackage[usenames,dvipsnames]{color}
+}
+\makeatother
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={Stuart Lee (Monash University) and Michael Lawrence (Genentech) and Di Cook (Monash University)},
+             pdfkeywords = {Bioconductor, Grammar, Genomes, Data Analysis},  
+            pdftitle={plyranges: a grammar for transforming genomics data},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
 
-\newcommand{\getAff}[1]{
-  \getIndex{#1}{Monash University,Genentech}
-}
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
+% add tightlist ----------
 \providecommand{\tightlist}{%
@@
 \begin{document}
-\vspace*{0.2in}
 
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{plyranges: a grammar for transforming genomics data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+% \pagenumbering{arabic}% resets `page` counter to 1 
+%
+% \maketitle
+
+{% \usefont{T1}{pnc}{m}{n}
+\setlength{\parindent}{0pt}
+\thispagestyle{plain}
+{\fontsize{18}{20}\selectfont\raggedright 
+\maketitle  % title \par
+
 }
-\newline
-\\
-Stuart Lee\textsuperscript{\getAff{Monash University}},
-Michael Lawrence\textsuperscript{\getAff{Genentech}},
-Di Cook\textsuperscript{\getAff{Monash University}}\\
-\bigskip
-\textbf{\getAff{Monash University}}Department of Econometrics and Business Statistics, Clayton, Victoria,
-Australia\\
-\textbf{\getAff{Genentech}}Bioinformatics and Computational Biology, Genentech, Inc., South San
-Francisco, California, United States of America\\
-\bigskip
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-The Bioconductor project has created many useful data abstractions for
+
+{
+   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
+\textbf{\authorfont Stuart Lee} \hskip 15pt \emph{\small Monash University}   \par \textbf{\authorfont Michael Lawrence} \hskip 15pt \emph{\small Genentech}   \par \textbf{\authorfont Di Cook} \hskip 15pt \emph{\small Monash University}
+
+}
+
+}
+
+\begin{abstract}
+
+    \hbox{\vrule height .2pt width 39.14pc}
+
+    \vskip 8.5pt % \small
+
+\noindent The Bioconductor project has created many useful data abstractions for
 analysing high-throughput genomics experiments. However, there is a
@@
 
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.
+\vskip 8.5pt \noindent \emph{Keywords}: Bioconductor, Grammar, Genomes, Data Analysis \par
 
-\linenumbers
+    \hbox{\vrule height .2pt width 39.14pc}
 
-% Use "Eq" instead of "Equation" for equation citations.
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\end{abstract}
+
+\vskip 6.5pt
+
+\noindent  \section{Introduction}\label{introduction}
 
@@
 SQL-like syntax for fast retrieval of information of unprocessed
-sequencing data {[}1{]}; {[}2{]}. Similarly, the Genometric Query
-Language (GMQL) implements a relational algebra for combining genomic
-datasets {[}3{]}. The command line application BEDtools develops an
-extensive algebra for performing arithmetic between two or more sets of
-genomic regions {[}4{]}. All of the aforementioned DSLs are designed to
-be evaluated either at the command line or embedded in scripts for batch
+sequencing data \autocites{Kozanitis2014-va}{Kozanitis2016-bm}.
+Similarly, the Genometric Query Language (GMQL) implements a relational
+algebra for combining genomic datasets \cite{Kaitoua2017-pw}. The
+command line application BEDtools develops an extensive algebra for
+performing arithmetic between two or more sets of genomic regions
+\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be
+evaluated either at the command line or embedded in scripts for batch
 processing. They exist in a sparse ecosystem, mostly consisting of UNIX
@@
 The Bioconductor/R packages \texttt{IRanges} and \texttt{GenomicRanges}
-{[}5--7{]} define a DSL for analysing genomics data with R, an
-interactive data analysis environment that encourages reproducibility
-and provides high-level abstractions for manipulating, modelling and
-plotting data, through state of the art methods in statistical
-computing. The packages define object-oriented (OO) abstractions for
-representing genomic data and enable interoperability by allowing users
-and developers to use these abstractions in their own code and packages.
-Other genomic DSLs that are embedded in programming languages include
-pybedtools and valr {[}8,9{]}, however these packages lack the
+\autocites{r-core}{Lawrence2013-wg}{Huber2015-ei} define a DSL for
+analysing genomics data with R, an interactive data analysis environment
+that encourages reproducibility and provides high-level abstractions for
+manipulating, modelling and plotting data, through state of the art
+methods in statistical computing. The packages define object-oriented
+(OO) abstractions for representing genomic data and enable
+interoperability by allowing users and developers to use these
+abstractions in their own code and packages. Other genomic DSLs that are
+embedded in programming languages include pybedtools and valr
+\autocites{Dale2011-js}{Kent2017}, however these packages lack the
 interoperability provided by the aforementioned Bioconductor packages
@@
 Recently, there has been a concerted, community effort to standardize R
-data structures and workflows around the notion of tidy data {[}10{]}. A
-tidy dataset is defined as a tabular data structure that has
-observations as rows and columns as variables, and all measurements
-pertain to a single observational unit. The tidy data pattern is useful
-because it allows us to see how the data relate to the design of an
-experiment and the variables measured. The \texttt{dplyr} package
-{[}11{]} defines an API that maps notions from the general relational
-algebra to operations on tidy data. It expresses each operation as a
-cohesive, endomorphic verb. Taken together these features enable a user
-to write human readable analysis workflows.
+data structures and workflows around the notion of tidy data
+\cite{Wickham2014-jc}. A tidy dataset is defined as a tabular data
+structure that has observations as rows and columns as variables, and
+all measurements pertain to a single observational unit. The tidy data
+pattern is useful because it allows us to see how the data relate to the
+design of an experiment and the variables measured. The \texttt{dplyr}
+package \cite{Wickham2017-dplyr} defines an API that maps notions from
+the general relational algebra to operations on tidy data. It expresses
+each operation as a cohesive, endomorphic verb. Taken together these
+features enable a user to write human readable analysis workflows.
 
@@
 
-\hypertarget{genomic-relational-algebra}{%
-\section{Genomic relational algebra}\label{genomic-relational-algebra}}
+\section{Genomic relational algebra}\label{genomic-relational-algebra}
 
-\hypertarget{data-model}{%
-\subsection{Data model}\label{data-model}}
+\subsection{Data model}\label{data-model}
 
 \begin{figure}
-{\centering \includegraphics[width=400pt]{diagrams/GRanges}}
+\includegraphics[width=\textwidth]{diagrams/GRanges}
 \caption{An illustration of the GRanges data model for a
@@
 The \texttt{plyranges} DSL is built on the core Bioconductor data
-structure GRanges, which is capable of representing all types of genomic
-data at a semantic level that roughly matches the intuition of most
-users. A GRanges is essentially a table, with columns for the
-chromosome, start and end coordinates, and the strand, along with an
-arbitrary set of additional columns, consisting of measurements or
-metadata specific to the data type or experiment (figure
-\ref{fig:GRanges}).
+structure GRanges, which is essentially a constrained table, with fixed
+columns for the chromosome, start and end coordinates, and the strand,
+along with an arbitrary set of additional columns, consisting of
+measurements or metadata specific to the data type or experiment (figure
+\ref{fig:GRanges}). GRanges balances flexibility with formal
+constraints, so that it is applicable to virtually any genomic workflow,
+while also being semantically rich enough to support high-level
+operations on genomic ranges. As a core data structure, GRanges enables
+interoperability between plyranges and the rest of Bioconductor.
+Adhering to a single data structure simplifies the API and makes it
+easier to learn and understand, in part because operations become
+endomorphic, i.e., they return the same type as their input.
+
+GRanges follows the intuitive tidy data pattern: it is a rectangular
+table corresponding to a single biological context. Each row contains a
+single observation and each column is a variable describing the
+observations. GRanges specializes the tidy pattern in that the
+observations always pertain to some genomic feature, but it largely
+remains compatible with the general relational operations defined by
+dplyr. Thus, we define our algebra as an extension of the dplyr algebra,
+and borrow its syntax conventions and design principles.
 
@@
 
-By definition GRanges follows the tidy data pattern: it is a rectangular
-table corresponding to a single biological context. Each row contains a
-single observation and each column is a variable about that observation.
-Hence, we have designed the \texttt{plyranges} DSL to extend the grammar
-and design principles of \texttt{dplyr}: cohesion, consistency,
-endomorphism, and fluency. All of these principles are defined and
-discussed below in the context of the GRanges class and an overview of
-the grammar is provided in table \ref{tab:grammar}. Where applicable we
-contrast our design to the existing Bioconductor infrastructure.
-
-\hypertarget{algebraic-operations}{%
-\subsection{Algebraic operations}\label{algebraic-operations}}
+\subsection{Algebraic operations}\label{algebraic-operations}
 
@@
 \ref{tab:grammar}). The grammar includes several classes of operation
-that cover most use cases in genomics data analysis. There are operators
-for performing arithmetic such as modifying width or performing set
-operations, merging such as finding overlaps between GRanges, subsetting
-by genomic regions, and aggregating over columns within in a GRanges
-object.
+that cover most use cases in genomics data analysis. There are range
+arithmetic operators, such as for resizing ranges or finding their
+intersection, and operators for merging, filtering and aggregating by
+range-specific notions like overlap and proximity.
 
-GRanges describes the within-sequence coordinates of a range by its
-\emph{start}, \emph{end} and \emph{width}. Those three variables are
+Arithmetic operations transform range coordinates, as defined by their
+\emph{start}, \emph{end} and \emph{width}. The three dimensions are
 mutually dependent and partially redundant, so direct manipulation of
@@
 parallel and merging. The parallel operations map to infix operators,
-which we surround with \emph{\%} symbols in accordance with R syntax
-rules. For example, the parallel intersection (\emph{x \%intersect\% y})
-finds the intersecting range between \emph{xi} and \emph{yi} for
+which we surround with \emph{\%} symbols, a convention borrowed from R
+syntax. For example, the parallel intersection (\emph{x \%intersect\%
+y}) finds the intersecting range between \emph{xi} and \emph{yi} for
 \emph{i} in \emph{1\ldots{}n}, where \emph{n} is the length of both
@@
 which finds the gap separating \emph{xi} and \emph{yi}. The binary
-operations are callable from within the arithmetic, restriction and
+operations are callable from within arithmetic, restriction and
 aggregation expressions.
 
-Our algebra recasts finding overlaps or nearest neighbours between two
-genomic regions as variants of the relational join operator. A join acts
-on two GRanges objects, a query and a subject. The join operator is
-relational in the sense that metadata from the query and subject ranges
-is retained in the joined range. All join operators in the
-\texttt{plyranges} DSL generate a set of hits based on overlap or
-proximity of ranges and use those hits to merge the two datasets in
-different ways. There are four supported matching algorithms:
-\emph{overlap}, \emph{nearest}, \emph{precede}, and \emph{follow}
-(figures \ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further
-restrict the matching by whether the query is completely \emph{within}
-the subject, and adding the \emph{directed} suffix ensures that matching
-ranges have the same direction (strand).
+To support merging, our algebra recasts finding overlaps or nearest
+neighbours between two genomic regions as variants of the relational
+join operator. A join acts on two GRanges objects, a query and a
+subject. The join operator is relational in the sense that metadata from
+the query and subject ranges is retained in the joined range. All join
+operators in the \texttt{plyranges} DSL generate a set of hits based on
+overlap or proximity of ranges and use those hits to merge the two
+datasets in different ways. There are four supported matching
+algorithms: \emph{overlap}, \emph{nearest}, \emph{precede}, and
+\emph{follow} (figures \ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can
+further restrict the matching by whether the query is completely
+\emph{within} the subject, and adding the \emph{directed} suffix ensures
+that matching ranges have the same direction (strand).
 
@@
 
-{\centering \includegraphics[width=400pt]{diagrams/diagrams-002}
+{\centering \includegraphics[width=\textwidth]{diagrams/diagrams-002}
 
@@
 
-{\centering \includegraphics[width=400pt]{diagrams/diagrams-003}
+{\centering \includegraphics[width=\textwidth]{diagrams/diagrams-003}
 
@@
 
-Since the GRanges object is a tabular data structure our grammar
+Since the GRanges object is a tabular data structure, our grammar
 includes operators to filter, sort and aggregate by columns in a
-GRanges. These operations can be performed over partitions within
-GRanges object using the \emph{group\_by} modifier. Together with our
-algebra for arithmetic and merging, these operations enable flexible and
-reproducible analysis using the desgin principles of the \texttt{dplyr}
-grammar.
+GRanges. These operations can be performed over partitions formed using
+the \emph{group\_by} modifier. Together with our algebra for arithmetic
+and merging, these operations conform to the semantics and syntax of the
+\texttt{dplyr} grammar.
 
-\hypertarget{cohesion}{%
-\subsection{Cohesion}\label{cohesion}}
+\subsection{Design principles}\label{design-principles}
 
-A function is cohesive if it performs a singular task and does not
-produce any side-effects. In the \texttt{plyranges} DSL our algebra for
-performing genomic arithmetic is a key example of cohesion. We define
-anchoring operators that decorate a GRanges object with an `anchor' and
-modify the semantics of performing genomic arithmetic. The anchoring
-operators amount to fixing a GRanges object by its start, center, or end
-coordinates (or fixing these coordinates by strand). This enables any
-arithmetic function that performs a coordinate transformation to remain
-cohesive: that is they always perform the same operation regardless of
-whether a GRanges object is anchored or not. The only difference is the
-result of performing the arithmetic changes when contextual information
-is added to the GRanges object.
+The design of plyranges adheres to well understood principles of
+language and API design: cognitive consistency, cohesion, expressiveness
+and endomorphism \cite{Green1996-qg}. To varying degrees, these
+principles also underly the design of dplyr and the Bioconductor
+infrastructure.
 
-Another example of our algebra altering object semantics, while
-maintaining cohesion is through the `group\_by' operator. Like
-anchoring, this operator decorates a GRanges object with a column name
-(or names) that defines a partitioning of the GRanges by the unique
-values in the column(s). Functions defined in the \texttt{plyranges} DSL
-that perform restriction, aggregation, or column modification still
-perform those singular tasks, however grouping changes how those tasks
-are performed.
+\subsubsection{Cognitive consistency and
+fluency}\label{cognitive-consistency-and-fluency}
 
-\hypertarget{consistency}{%
-\subsection{Consistency}\label{consistency}}
+We aim for our interfaces to have a simple and direct mapping to the
+user's cognitive model, i.e., how the user thinks about the data. This
+requires careful selection of the level of abstraction so that the user
+can express workflows in the language of genomics. This motivates the
+adoption of the tidy GRanges object as our central data structure. The
+basic data.frame and dplyr tibble lack any notion of genomic ranges and
+so could not easily support our genomic grammar, with its specific verbs
+for range-oriented data manipulation. Another example of cognitive
+consistency is how plyranges is insensitive to direction/strand by
+default when, e.g., detecting overlaps. GenomicRanges has the opposite
+behavior. We believe that defaulting to purely spatial overlap is most
+intuitive to most users.
 
-A core design principle of the \texttt{plyranges} DSL is interface
-consistency: a user should not be surprised by the input or output of
-\texttt{plyranges} code. A key example of consistency is how
-\texttt{plyranges} handles strand information. Every function that
-computes with strand information indicates its intentions by including
-suffixes such as `directed', `upstream' or `downstream' in its name,
-otherwise strand is ignored. This strongly differs from the Bioconductor
-packages, which produces surprising output by assuming the user is
-always interested in using strand in the majority of circumstances
-(unless a user is computing coverage or finding flanking ranges).
+Like dplyr, plyranges verbs are functional: they are free of side
+effects and return their result. This enables chaining of verbs through
+syntax like the forward pipe operator (\texttt{\%\textgreater{}\%}, read
+as ``then'') of the magrittr package \cite{R-magrittr}. This has syntax
+has a direct cognitive mapping to natural language and the intuitive
+notion of pipelines. The low-level object-oriented APIs of Bioconductor
+tend to manipulate data via subreplacement functions, like
+\texttt{start(gr)\ \textless{}-\ x}. These ultimately produce the side
+effect of replacing a symbol mapping in the current environment and thus
+are not amenable to so-called fluent syntax.
 
-Our use of suffixes in function names, also highlights a core tenet of
-the \texttt{plyranges} design: avoid complex generic functions with many
-arguments and instead use cohesive functions with a minimal number of
-arguments. As an example, we have written a consistent framework for
-reading and writing files from and to common genomic data formats as
-GRanges, using the \texttt{rtracklayer} package as a back-end {[}12{]}.
-We have replaced that packages generic functions for importing files
-with a family of reader functions that all take the exact same
-arguments.
+\subsubsection{Cohesion}\label{cohesion}
 
-\hypertarget{endomorphism}{%
-\subsection{Endomorphism}\label{endomorphism}}
+A function is cohesive if it performs a singular task without producing
+any side-effects. Singular tasks are not necessarily atomic; they can
+always be broken down further at lower levels of abstraction. For
+example, to resize a range, the user needs to specify which position
+(start, end, midpoint) should be invariant over the transformation. The
+\texttt{resize()} function from the GenomicRanges package has a
+\texttt{fix} argument that sets the anchor, so calling \texttt{resize()}
+coalesces anchoring and width modification. The coupling at the function
+call level is justified since the effect of setting the width depends on
+the anchor. However, plyranges increases cohesion and decouples the
+anchoring into its own function call.
 
-Most function calls in \texttt{plyranges} are endomorphisms: when the
-input is GRanges object the output will also be a GRanges object. The
-use of endomorphism in the \texttt{plyranges} DSL enables a user to
-predict the structure of the output of their computations and does not
-require them to learn any additional classes beyond GRanges and
-DataFrames.
+Increasing cohesion simplifies the interface to each operation, makes
+the meaning of arguments more intuitive, and relies on function names as
+the primary means of expression, instead of a more complex mixture of
+function and argument names. Since functions are superordinate to their
+arguments, flattening the API at the function level enables the user to
+conceptualize the API as a catalog of functions, without having to
+descend further. A flat function catalog also enhances API
+discoverability, particularly through auto-completion in IDEs. One
+downside of pushing cohesion to this extreme is that function calls
+become coupled, and care is necessary to treat them as a group when
+modifying code.
 
-As an example, in \texttt{plyranges} when we compute coverage over a
-GRanges object, the result of the operation is an expanded GRanges
-object with a score column, corresponding the estimated coverage over a
-genomic region. Hence, the resulting coverage score can easily be
-composed with other expressions in our algebra. This pattern strongly
-deviates from the design of the OO interface in \texttt{GenomicRanges},
-which in the case of computing coverage would return an RleList, an
-object that has many new methods and that is potentially unfamiliar to a
-new user. While, these low-level classes enable efficient computing, the
-increase complexity, and as a consequence are abstracted away in the
-\texttt{plyranges} DSL.
+\subsubsection{Expressiveness}\label{expressiveness}
 
-\hypertarget{fluency}{%
-\subsection{Fluency}\label{fluency}}
+Expressiveness relates to the information content in code: the
+programmer should be able to clarify intent without unnecessary
+verbosity. For example, our overlap-based join operations are more
+concise than the multiple steps necessary to achieve the same effect in
+the original GenomicRanges API. In other cases, the plyranges API
+increases verbosity for the sake of clarity and cohesion. Explicitly
+calling \texttt{anchor()} can require more typing, but the code is
+easier to comprehend. Another example is the set of routines for
+importing genomic annotations, including
+\texttt{read\textbackslash{}\_gff()},
+\texttt{read\textbackslash{}\_bed()}, and \texttt{read\_bam()}. Compared
+to the generic \texttt{import()} in rtracklayer, the explicit
+format-based naming in plyranges clarifies intent and the type of data
+being returned. Similarly, every plyranges function that computes with
+strand information indicates its intentions by including suffixes such
+as \texttt{directed}, \texttt{upstream} or \texttt{downstream} in its
+name, otherwise strand is ignored. The GenomicRanges API does not make
+this distinction explicit in its function naming, instead relying on a
+parameter that defaults to strand sensitivity, an arguably unintuitive
+behavior.
 
-As a consequence of the design principles defined above every function
-in \texttt{plyranges} performs a single action on GRanges objects. The
-\texttt{plyranges} DSL implements the core verbs from the \texttt{dplyr}
-package and implements a genomic relational algebra for transforming
-GRanges objects (table x, y). Each verb preserves the semantics of
-GRanges object and works with derivatives of the GRanges class. Both of
-these aspects reduce the cognitive load on a new user since most
-manipulations can be performed with a vocabulary of several verbs,
-rather than having to memorise function names that are nouns.
+\section{Results}\label{results}
 
-This approach strongly contrasts the \texttt{GenomicRanges/IRanges} OO
-interface, which emphasises the use of setter and getter methods. In
-that interface, core components and metadata are updated via replacement
-methods (requiring knowledge of the class components), while our
-interface requires only a single call to \texttt{mutate()} to perform
-the modification.
-
-Workflows can be composed by chaining verbs together into `sentences'
-via the forward pipe operator,\texttt{\%\textgreater{}\%} (exported from
-the R package \texttt{magrittr} {[}13{]}), which can be read as the word
-`then'. Overall, this allows users to write human readable code because
-workflows describe what the code is doing rather than how its doing it.
-
-\hypertarget{opportunities}{%
-\subsection{Opportunities}\label{opportunities}}
-
-A caveat to constructing a compatible interface with \texttt{dplyr} is
-that \texttt{plyranges} makes extensive use of non-standard evaluation
-in R (achieved via the \texttt{rlang} package {[}14{]}). Simply, this
-means that computations are performed and evaluated in the context of
-the GRanges objects; emphasising the interactive nature of our API.
-Consequently, when programming with \texttt{plyranges} a user needs to
-be aware of how non-standard evaluation in R works and how to adapt
-their code accordingly. However, with the rise of R packages like
-\texttt{rlang} this process is becoming less difficult.
-
-While GRanges are an intuitive representation for data measured on
-genomic regions, more flexible data structures are required to represent
-data from multiple sample experiments. The Bionconductor class
-SummarizedExperiment is the canonical data structure for representing
-data for combining multiomic measurements from multiple samples. The
-grammar and design of the \texttt{plryanges} DSL can be naturally
-extended to the SummarizedExperiment.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-Here we provide illustrative examples by using the \texttt{plyranges}
+Here we provide illustrative examples of using the \texttt{plyranges}
 DSL to show how our grammar could be integrated into genomic data
@@
 
-\hypertarget{peak-finding}{%
-\subsection{Peak Finding}\label{peak-finding}}
+\subsection{Peak Finding}\label{peak-finding}
 
-The Bioconductor package \texttt{AnnotationHub} {[}15{]} can be used to
-search for BigWig files from ChIP-Seq experiments from the Human
-Epigenome Roadmap project {[}16{]}. Here we focus on assays for primary
-T CD8+ memory cells from peripheral blood. Using \texttt{plyranges} we
-will read the BigWig file corresponding to the H3 lysine 27
-trimethylation (H3K27Me3) methylation mark over chromosome 10.
+The Bioconductor package \texttt{AnnotationHub} \cite{R-ahub} can be
+used to search for BigWig files from ChIP-Seq experiments from the Human
+Epigenome Roadmap project \cite{Roadmap-Epigenomics-Consortium2015-pr}.
+Here we focus on assays for primary T CD8+ memory cells from peripheral
+blood. Using \texttt{plyranges} we will read the BigWig file
+corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation
+mark over chromosome 10.
 
-First, we gather the BigWig file and extract its annotation information
-and filter it to chromosome 10.
+First, we extract the genome information from the BigWig file and filter
+to get the range for chromosome 10. This range will be used as a filter
+when reading the file.
 
@@
 Then we read the BigWig file only extracting scores if they overlap
-chromosome 10. The annotation information from the file is automatically
-included (in this case the hg19 genome build).
+chromosome 10. We also add the genome build information to the resulting
+ranges.
 
@@
 
-The \texttt{reduce\_ranges()} operation is used to find coverage peaks
-across chromosome 10. We can manually set a threshold to restrict
-genomic regions to have a coverage score of greater than 8, and then
-merge nearby regions. The maximum coverage is computed over all the
-coverage scores in the regions that were reduced.
+After filtering for regions with a coverage score greater than 8, we can
+reduce individual runs to ranges representing the islands of coverage,
+using the \texttt{reduce\_ranges()} function. This function allows a
+summary to be computer over each island: in this case we take the
+maximum of the scores to find the coverage peaks over chromosome 10.
 
@@
 coverage score. We can then find a 5000 nt region centered around the
-maximum position by anchoring and modifying the the width.
+maximum position by anchoring and modifying the width.
 
@@
 
-\hypertarget{computing-windowed-statistics}{%
 \subsection{Computing Windowed
-Statistics}\label{computing-windowed-statistics}}
+Statistics}\label{computing-windowed-statistics}
 
@@
 summaries over genomic windows. In \texttt{plyranges} this can be
-achieved via the \texttt{group\_by\_overlaps} operator. Continuing with
-the data from the Human Epigenome Roadmap Consortium data, we can count
-the number of reads that fall into a fixed bins of size 10000bp over a
-BAM file of H3K27Me3 methylation marks from the H1 cell line. We extract
-reads that have a mapping quality score greater than 20:
+achieved via the \texttt{group\_by\_overlaps()} operator. We bin and
+count and find the average GC content of reads from a H3K27Me3 ChIP-seq
+experiment by the Human Epigenome Roadmap Consortium.
+
+We can directly obtain the genome information from the header of the BAM
+file: in this case the reads were aligned to the hg19 genome build and
+there are no reads overlapping the mitochondrial genome. To generate
+bins of fixed width 10000bp we apply \texttt{tile()} function from
+\texttt{GenomicRanges} to the genomic coordinates (with \texttt{unlist}
+to return a GRanges object)
 
@@
 \begin{Highlighting}[]
-\NormalTok{alignments <-}\StringTok{ }\KeywordTok{read_bam}\NormalTok{(h1_bam) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{filter}\NormalTok{(mapq }\OperatorTok{>}\StringTok{ }\DecValTok{20}\NormalTok{) }
+\NormalTok{locations <-}\StringTok{ }\NormalTok{h1_bam_sorted }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{read_bam}\NormalTok{() }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{get_genome_info}\NormalTok{() }
+
+\NormalTok{bins <-}\StringTok{ }\NormalTok{locations }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{tile}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{unlist}\NormalTok{(}\DataTypeTok{use.names =} \OtherTok{FALSE}\NormalTok{)}
 \end{Highlighting}
@@
 
-Note that the BAM file is only read into memory once we perform an
-operation on it. Because of interoperability over Bioconductor, we can
-generate genomic bins using the \texttt{tile} function from
-\texttt{GenomicRanges}:
+Next we only read in alginments that overlap the genomic locations we
+are interested in and select the query seqeunce. Note that the reading
+of the BAM file is deferred: only alignments that pass the filter are
+loaded into memory. We can add another column representing the GC
+proportion for each alignment using the \texttt{letterFrequency()}
+function from the \texttt{Biostrings} package \cite{R-biostrings}.
 
@@
 \begin{Highlighting}[]
-\NormalTok{bins <-}\StringTok{ }\KeywordTok{tile}\NormalTok{(h1_bam, }\DataTypeTok{width =} \DecValTok{10000}\NormalTok{)}
+\NormalTok{alignments <-}\StringTok{ }\NormalTok{h1_bam_sorted }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{read_bam}\NormalTok{() }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{filter_by_overlaps}\NormalTok{(locations) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{select}\NormalTok{(seq) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{score =} \KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{letterFrequency}\NormalTok{(seq, }\StringTok{"GC"}\NormalTok{, }\DataTypeTok{as.prob =} \OtherTok{TRUE}\NormalTok{)))}
 \end{Highlighting}
@@
 
-Finally, we can use \texttt{group\_by\_overlaps} with \texttt{summarise}
-to compute the total number of reads within each window.
+Finally, we use \texttt{group\_by\_overlaps()} with \texttt{summarize()}
+to compute the total number of reads and averge GC content within each
+genomic window.
 
@@
 \begin{Highlighting}[]
-\NormalTok{h1_bam_read_summary <-}\StringTok{ }\NormalTok{h1_bam }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{group_by_overlaps}\NormalTok{(bins) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{n_reads =} \KeywordTok{n}\NormalTok{())}
+\NormalTok{alignments_summary <-}\StringTok{ }\NormalTok{bins }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{group_by_overlaps}\NormalTok{(alignments }\OperatorTok{%>%}\StringTok{ }\KeywordTok{select}\NormalTok{(score)) }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{n_reads =} \KeywordTok{n}\NormalTok{(), }\DataTypeTok{avg_gc =} \KeywordTok{mean}\NormalTok{(score))}
 \end{Highlighting}
@@
 
-\hypertarget{quality-control-metrics}{%
-\subsection{Quality Control Metrics}\label{quality-control-metrics}}
+\subsection{Quality Control Metrics}\label{quality-control-metrics}
 
@@
 cell line, consisting of approximately two million single nucleotide
-polymorphisms (SNP) and short insertion/deletions (indel). The GRanges
+polymorphisms (SNPs) and short insertion/deletions (indels). The GRanges
 object consists of 7 columns, relating to the alleles of a SNP or indel,
@@
 
-First we filter out any insertion or deletion alleles and the SNPs
-present on the mitochondria then create a logical vector corresponding
-to whether there is a transition event.
+First we filter out the indels and mitochondrial variants. Then we
+create a logical vector corresponding to whether there is a transition
+event.
 
@@
 \StringTok{  }\KeywordTok{filter}\NormalTok{(}\OperatorTok{!}\NormalTok{(ref }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"I"}\NormalTok{, }\StringTok{"D"}\NormalTok{)), seqnames }\OperatorTok{!=}\StringTok{ "M"}\NormalTok{) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{transition =}\NormalTok{ (ref }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"A"}\NormalTok{, }\StringTok{"G"}\NormalTok{) }\OperatorTok{&}\StringTok{ }\NormalTok{alt }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"G"}\NormalTok{,}\StringTok{"A"}\NormalTok{)) }\OperatorTok{|}
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{transition =}\NormalTok{ (ref }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"A"}\NormalTok{, }\StringTok{"G"}\NormalTok{) }\OperatorTok{&}\StringTok{ }\NormalTok{alt }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"G"}\NormalTok{,}\StringTok{"A"}\NormalTok{))}\OperatorTok{|}
 \StringTok{                      }\NormalTok{(ref }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"C"}\NormalTok{,}\StringTok{"T"}\NormalTok{) }\OperatorTok{&}\StringTok{ }\NormalTok{alt }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"T"}\NormalTok{, }\StringTok{"C"}\NormalTok{)))}
@@
 
-We can then compute the transition-transversion ratio using the
-\texttt{group\_by} and \texttt{summarise} pattern, it is computed as the
-total number of transition SNPs divided by the total number of
-transversion SNPs within each chromosome (figure \ref{fig:titv-viz}).
+We can then compute the per-chromosome transition-transversion ratio
+using the \texttt{group\_by} and \texttt{summarize} pattern (figure
+\ref{fig:titv-viz}).
 
@@
 \StringTok{  }\KeywordTok{group_by}\NormalTok{(seqnames) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{n_snps =} \KeywordTok{n}\NormalTok{(),}
+\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{n_snps =} \KeywordTok{n}\NormalTok{(),}
             \DataTypeTok{ti_tv =} \KeywordTok{sum}\NormalTok{(transition) }\OperatorTok{/}\StringTok{ }\KeywordTok{sum}\NormalTok{(}\OperatorTok{!}\NormalTok{transition)) }
@@
 
-\hypertarget{availablilty-and-future-work}{%
-\section{Availablilty and Future
-Work}\label{availablilty-and-future-work}}
+\section{Discussion and conclusion}\label{discussion-and-conclusion}
+
+We have shown how to create expressive and reproducible genomic
+workflows using the \texttt{plyranges} DSL. By realising that the
+GRanges data model is tidy we have highlighted how to implement a
+grammar for performing genomic arithemtic, aggregation, restriction and
+merging. Our examples show that \texttt{plyranges} code is succinct,
+human readable and can take advantage of the interoperability provided
+by the Bioconductor ecosytem and the R language.
+
+A caveat to constructing a compatible interface with \texttt{dplyr} is
+that \texttt{plyranges} makes extensive use of non-standard evaluation
+in R via the \texttt{rlang} package \cite{R-rlang}. Simply, this means
+that computations are evaluated in the context of the GRanges objects.
+Both dplyr and plyranges are based on the \texttt{rlang} language,
+because it allows for more expressive code that is free of repeated
+references to the container. Implicitly referencing the container is
+particularly convenient when programming interactively. Consequently,
+when programming with \texttt{plyranges}, a user needs to generally
+understand the \texttt{rlang} language and how to adapt their code
+accordingly. Users familiar with the tidyverse should already have such
+knowledge.
+
+We aim to continue developing the \texttt{plyranges} package and to
+extend it for use with more complex data structures, such as the
+SummarizedExperiment class, the core Bioconductor data structure for
+representing experimental results (e.g., counts) from multiple sample
+experiments in conjunction with feature and sample metadata. The grammar
+and design of the \texttt{plryanges} DSL are naturally extensible to
+SummarizedExperiment.
+
+As the \texttt{plyranges} interface encourages tidy data practices, it
+integrates well with the grammar of graphics \cite{Wickham2016-gz}. To
+achieve responsive performance, interactive graphics rely on lazy data
+access and computing patterns, so the deferred mechanisms within
+plyranges should help support interactive genomics applications.
 
@@
 website \url{https://bioconductor.org} or can be accessed via Github
-\url{https://github.com/sa-lee/plyranges}. We aim to continue developing
-the \texttt{plyranges} package and extend it for use with more complex
-data structures such as the SummarizedExperiment class, which can be
-used for analysing transcriptomic and variant data. As the
-\texttt{plyranges} interface encourages tidy data practices it
-integrates well with the principles of the grammar of graphics, we aim
-to use it to prepare data for the visualisation of multimodal biological
-datasets.
+\url{https://github.com/sa-lee/plyranges}.
 
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\section{Acknowledgements}\label{acknowledgements}
 
@@
 the code review they performed. This report was written with
-\texttt{knitr} {[}17{]} and the figures were made with \texttt{ggbio}
-{[}18{]}. All code required to reproduce this article is available at
-https://github.com/sa-lee/plyranges-paper.
+\texttt{knitr} \cite{R-knitr} and the figures were made with
+\texttt{ggbio} \cite{R-ggbio}. All code required to reproduce this
+article is available at \url{https://github.com/sa-lee/plyranges-paper}.
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Kozanitis2014-va}{}%
-1. Kozanitis C, Heiberg A, Varghese G, Bafna V. Using genome query
-language to uncover genetic variation. Bioinformatics. 2014;30: 1--8.
-doi:\href{https://doi.org/10.1093/bioinformatics/btt250}{10.1093/bioinformatics/btt250}
-
-\leavevmode\hypertarget{ref-Kozanitis2016-bm}{}%
-2. Kozanitis C, Patterson DA. GenAp: A distributed SQL interface for
-genomic data. BMC Bioinformatics. 2016;17: 63.
-doi:\href{https://doi.org/10.1186/s12859-016-0904-1}{10.1186/s12859-016-0904-1}
-
-\leavevmode\hypertarget{ref-Kaitoua2017-pw}{}%
-3. Kaitoua A, Pinoli P, Bertoni M, Ceri S. Framework for supporting
-genomic operations. IEEE Trans Comput. 2017;66: 443--457.
-doi:\href{https://doi.org/10.1109/TC.2016.2603980}{10.1109/TC.2016.2603980}
-
-\leavevmode\hypertarget{ref-Quinlan2010-gc}{}%
-4. Quinlan AR, Hall IM. BEDTools: A flexible suite of utilities for
-comparing genomic features. Bioinformatics. 2010;26: 841--842.
-doi:\href{https://doi.org/10.1093/bioinformatics/btq033}{10.1093/bioinformatics/btq033}
-
-\leavevmode\hypertarget{ref-r-core}{}%
-5. R Core Team. R: A language and environment for statistical computing
-{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
-2018. Available: \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-Lawrence2013-wg}{}%
-6. Lawrence M, Huber W, Pagès H, Aboyoun P, Carlson M, Gentleman R, et
-al. Software for computing and annotating genomic ranges. PLoS Comput
-Biol. 2013;9.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1003118}{10.1371/journal.pcbi.1003118}
-
-\leavevmode\hypertarget{ref-Huber2015-ei}{}%
-7. Huber W, Carey VJ, Gentleman R, Anders S, Carlson M, Carvalho BS, et
-al. Orchestrating high-throughput genomic analysis with bioconductor.
-Nat Methods. Springer Nature; 2015;12: 115--121.
-doi:\href{https://doi.org/10.1038/nmeth.3252}{10.1038/nmeth.3252}
-
-\leavevmode\hypertarget{ref-Dale2011-js}{}%
-8. Dale RK, Pedersen BS, Quinlan AR. Pybedtools: A flexible python
-library for manipulating genomic datasets and annotations.
-Bioinformatics. 2011;27: 3423--3424.
-doi:\href{https://doi.org/10.1093/bioinformatics/btr539}{10.1093/bioinformatics/btr539}
-
-\leavevmode\hypertarget{ref-Kent2017}{}%
-9. Riemondy KA, Sheridan RM, Gillen A, Yu Y, Bennett CG, Hesselberth JR.
-Valr: Reproducible genome interval arithmetic in r. F1000Research. 2017;
-doi:\href{https://doi.org/10.12688/f1000research.11997.1}{10.12688/f1000research.11997.1}
-
-\leavevmode\hypertarget{ref-Wickham2014-jc}{}%
-10. Wickham H. Tidy data. Journal of Statistical Software, Articles.
-2014;59: 1--23.
-doi:\href{https://doi.org/10.18637/jss.v059.i10}{10.18637/jss.v059.i10}
-
-\leavevmode\hypertarget{ref-Wickham2017-dplyr}{}%
-11. Wickham H, Francois R, Henry L, Müller K. Dplyr: A grammar of data
-manipulation {[}Internet{]}. 2017. Available:
-\url{https://CRAN.R-project.org/package=dplyr}
-
-\leavevmode\hypertarget{ref-Lawrence2009-nt}{}%
-12. Lawrence M, Gentleman R, Carey V. Rtracklayer: An R package for
-interfacing with genome browsers. Bioinformatics. 2009;25: 1841--1842.
-doi:\href{https://doi.org/10.1093/bioinformatics/btp328}{10.1093/bioinformatics/btp328}
-
-\leavevmode\hypertarget{ref-R-magrittr}{}%
-13. Bache SM, Wickham H. Magrittr: A forward-pipe operator for r
-{[}Internet{]}. 2014. Available:
-\url{https://CRAN.R-project.org/package=magrittr}
-
-\leavevmode\hypertarget{ref-R-rlang}{}%
-14. Henry L, Wickham H. Rlang: Functions for base types and core r and
-'tidyverse' features {[}Internet{]}. 2017. Available:
-\url{http://rlang.tidyverse.org}
-
-\leavevmode\hypertarget{ref-R-ahub}{}%
-15. Morgan M. AnnotationHub: Client to access annotationhub resources.
-2017.
-
-\leavevmode\hypertarget{ref-Roadmap_Epigenomics_Consortium2015-pr}{}%
-16. Roadmap Epigenomics Consortium, Kundaje A, Meuleman W, Ernst J,
-Bilenky M, Yen A, et al. Integrative analysis of 111 reference human
-epigenomes. Nature. 2015;518: 317--330.
-doi:\href{https://doi.org/10.1038/nature14248}{10.1038/nature14248}
-
-\leavevmode\hypertarget{ref-R-knitr}{}%
-17. Xie Y. Dynamic documents with R and knitr {[}Internet{]}. 2nd ed.
-Boca Raton, Florida: Chapman; Hall/CRC; 2015. Available:
-\url{https://yihui.name/knitr/}
-
-\leavevmode\hypertarget{ref-R-ggbio}{}%
-18. Yin T, Cook D, Lawrence M. Ggbio: An r package for extending the
-grammar of graphics for genomic data. Genome Biology. BioMed Central
-Ltd; 2012;13: R77.
-
-\nolinenumbers
+\newpage
+\singlespacing 
+\printbibliography[title=References]
 

### 2018-05-16 paper.Rmd
- spelling fixes - making examples flow better - consistently use \texttt for package names

@@
 sample from an RNA-seq experiment. The core components of the data model
-inlcude a seqname column (representing the chromosome), a ranges column
+include a seqname column (representing the chromosome), a ranges column
 which consists of start and end coordinates for a genomic region, and a
@@
 intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
-is the length of both _x_ and _y_. In constrast, the merging
+is the length of both _x_ and _y_. In contrast, the merging
 intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
-ranges representing whereever there was overlap between a range in _x_
+ranges representing wherever there was overlap between a range in _x_
 and a range in _y_. We use the infix syntax for the parallel
@@
 
-The design of plyranges adheres to well understood principles of
+The design of `plyranges` adheres to well understood principles of
 language and API design: cognitive consistency, cohesion,
 expressiveness and endomorphism \cite{Green1996-qg}. 
-To varying degrees, these principles also underly the
-design of dplyr and the Bioconductor infrastructure.
+To varying degrees, these principles also underlie the
+design of `dplyr` and the Bioconductor infrastructure.
 
@@
 the adoption of the tidy GRanges object as our central data
-structure. The basic data.frame and dplyr tibble lack any notion of
+structure. The basic data.frame and `dplyr` tibble lack any notion of
 genomic ranges and so could not easily support our genomic grammar,
 with its specific verbs for range-oriented data manipulation. Another
-example of cognitive consistency is how plyranges is insensitive to
+example of cognitive consistency is how `plyranges` is insensitive to
 direction/strand by default when, e.g., detecting
-overlaps. GenomicRanges has the opposite behavior. We believe that
+overlaps. `GenomicRanges` has the opposite behavior. We believe that
 defaulting to purely spatial overlap is most intuitive to most users.
@@
 -->
-Like dplyr, plyranges verbs are functional: they are free of side
+Like `dplyr`, `plyranges` verbs are functional: they are free of side
 effects and return their result. This enables chaining of verbs
 through syntax like the forward pipe operator (`%>%`, read as "then")
-of the magrittr package \cite{R-magrittr}. This has syntax has a direct
+of the `magrittr` package \cite{R-magrittr}. This has syntax has a direct
 cognitive mapping to natural language and the intuitive notion of
 pipelines. The low-level object-oriented APIs of Bioconductor tend to
-manipulate data via subreplacement functions, like `start(gr) <-
+manipulate data via sub-replacement functions, like `start(gr) <-
 x`. These ultimately produce the side effect of replacing a symbol
@@
 position (start, end, midpoint) should be invariant over the
-transformation. The `resize()` function from the GenomicRanges package
+transformation. The `resize()` function from the `GenomicRanges` package
 has a `fix` argument that sets the anchor, so calling `resize()`
@@
 concise than the multiple steps necessary to achieve the same effect
-in the original GenomicRanges API. In other cases, the plyranges API
+in the original `GenomicRanges` API. In other cases, the `plyranges` API
 increases verbosity for the sake of clarity and cohesion. Explicitly
@@
 genomic annotations, including `read\_gff()`, `read\_bed()`, and
-`read_bam()`. Compared to the generic `import()` in rtracklayer, the
+`read_bam()`. Compared to the generic `import()` in `rtracklayer`, the
 explicit format-based naming in plyranges clarifies intent and the
@@
 suffixes such as `directed`, `upstream` or `downstream` in its name,
-otherwise strand is ignored. The GenomicRanges API does not make this
+otherwise strand is ignored. The `GenomicRanges` API does not make this
 distinction explicit in its function naming, instead relying on a
@@
 
-The Bioconductor package `AnnotationHub` \cite{R-ahub}  can be used to search for 
-BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project 
-\cite{Roadmap-Epigenomics-Consortium2015-pr}. Here we focus on assays for primary
-T CD8+ memory cells from peripheral blood. Using `plyranges` we will read the 
-BigWig file  corresponding to the H3 lysine 27 trimethylation (H3K27Me3) 
-methylation mark over chromosome 10.
+In the workflow of ChIP-seq data analysis, we are interested in finding peaks 
+from islands of coverage over chromosome. Here we will use `plyranges`
+to call peaks from islands of coverage above 8 then plot the region
+surrounding the tallest peak.
+
+Using `plyranges` and the the Bioconductor package `AnnotationHub` \cite{R-ahub}
+we can download and read BigWig files from ChIP-Seq experiments from 
+the Human Epigenome Roadmap project \cite{Roadmap-Epigenomics-Consortium2015-pr}. 
+Here we analyse a BigWig file corresponding to H3 lysine 27 trimethylation 
+(H3K27Me3) of primary T CD8+ memory cells from peripheral blood, focussing
+on coverage islands over chromosome 10.
 
@@
 Then we read the BigWig file only extracting scores if they overlap chromosome
-10. We also add the genome build information to the resulting ranges.
+10. We also add the genome build information to the resulting ranges. This
+book-keeping is good practice as it ensures the integrity of any 
+downstream operations such as finding overlaps.
 
@@
 After filtering for regions with a coverage score greater than 8, we can
-reduce individual runs to ranges representing the islands of coverage,
+reduce individual runs to ranges representing the islands of coverage by
 using the `reduce_ranges()` function. This function allows a summary
@@
 and there are no reads overlapping the mitochondrial genome. To generate 
-bins of fixed width 10000bp we apply `tile()` function from `GenomicRanges`
-to the genomic coordinates (with `unlist` to return a GRanges object)
+bins of fixed width 10000nt we apply the `tile_ranges()` function
+to the genomic coordinates in locations.
 ```{r bins}
@@
 bins <- locations %>% 
-  tile(width = 10000L) %>%
-  unlist(use.names = FALSE)
+  tile_ranges(width = 10000L)
 ```
 
-Next we only read in alginments that overlap the genomic locations
-we are interested in and select the query seqeunce. Note that
+Next we only read in alignments that overlap the genomic locations
+we are interested in and select the query sequence. Note that
 the reading of the BAM file is deferred: only alignments that pass the filter 
@@
 
-Finally, we use `group_by_overlaps()` with `summarize()` to compute the
-total number of reads and averge GC content within each genomic window.
+Finally, we use `group_by_overlaps()` to see where the alignments overlap the 
+genomic windows and then apply `summarize()` to compute the
+total number of reads and average GC content within each window.
 
@@
 
-We can then compute the per-chromosome transition-transversion ratio using the
-`group_by` and `summarize` pattern (figure \ref{fig:titv-viz}).
+We then compute the transition-transversion ratio over each chromosome using
+`group_by()` in combination with `summarize()` (figure \ref{fig:titv-viz}).
 
@@
 realising that the GRanges data model is tidy we have highlighted how to 
-implement a grammar for performing genomic arithemtic, aggregation, restriction 
+implement a grammar for performing genomic arithmetic, aggregation, restriction 
 and merging. Our examples show that `plyranges` code is succinct, human
 readable and can take advantage of the interoperability provided by
-the Bioconductor ecosytem and the R language.
+the Bioconductor ecosystem and the R language.
 
@@
 The `plyranges` package is available on the Bioconductor project
-website [https://bioconductor.org](https://bioconductor.org) or can be
-accessed via
-Github
+website [https://bioconductor.org](https://bioconductor.org) 
+or can be accessed via Github
 [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).

### 2018-05-16 paper.tex
- spelling fixes - making examples flow better - consistently use \texttt for package names

@@
 \emph{i} in \emph{1\ldots{}n}, where \emph{n} is the length of both
-\emph{x} and \emph{y}. In constrast, the merging intersection
+\emph{x} and \emph{y}. In contrast, the merging intersection
 (\emph{intersect\_ranges(x, y)}) returns a new set of disjoint ranges
-representing whereever there was overlap between a range in \emph{x} and
+representing wherever there was overlap between a range in \emph{x} and
 a range in \emph{y}. We use the infix syntax for the parallel
@@
 
-The design of plyranges adheres to well understood principles of
-language and API design: cognitive consistency, cohesion, expressiveness
-and endomorphism \cite{Green1996-qg}. To varying degrees, these
-principles also underly the design of dplyr and the Bioconductor
-infrastructure.
+The design of \texttt{plyranges} adheres to well understood principles
+of language and API design: cognitive consistency, cohesion,
+expressiveness and endomorphism \cite{Green1996-qg}. To varying degrees,
+these principles also underlie the design of \texttt{dplyr} and the
+Bioconductor infrastructure.
 
@@
 adoption of the tidy GRanges object as our central data structure. The
-basic data.frame and dplyr tibble lack any notion of genomic ranges and
-so could not easily support our genomic grammar, with its specific verbs
-for range-oriented data manipulation. Another example of cognitive
-consistency is how plyranges is insensitive to direction/strand by
-default when, e.g., detecting overlaps. GenomicRanges has the opposite
-behavior. We believe that defaulting to purely spatial overlap is most
-intuitive to most users.
+basic data.frame and \texttt{dplyr} tibble lack any notion of genomic
+ranges and so could not easily support our genomic grammar, with its
+specific verbs for range-oriented data manipulation. Another example of
+cognitive consistency is how \texttt{plyranges} is insensitive to
+direction/strand by default when, e.g., detecting overlaps.
+\texttt{GenomicRanges} has the opposite behavior. We believe that
+defaulting to purely spatial overlap is most intuitive to most users.
 
-Like dplyr, plyranges verbs are functional: they are free of side
-effects and return their result. This enables chaining of verbs through
-syntax like the forward pipe operator (\texttt{\%\textgreater{}\%}, read
-as ``then'') of the magrittr package \cite{R-magrittr}. This has syntax
-has a direct cognitive mapping to natural language and the intuitive
-notion of pipelines. The low-level object-oriented APIs of Bioconductor
-tend to manipulate data via subreplacement functions, like
-\texttt{start(gr)\ \textless{}-\ x}. These ultimately produce the side
-effect of replacing a symbol mapping in the current environment and thus
-are not amenable to so-called fluent syntax.
+Like \texttt{dplyr}, \texttt{plyranges} verbs are functional: they are
+free of side effects and return their result. This enables chaining of
+verbs through syntax like the forward pipe operator
+(\texttt{\%\textgreater{}\%}, read as ``then'') of the \texttt{magrittr}
+package \cite{R-magrittr}. This has syntax has a direct cognitive
+mapping to natural language and the intuitive notion of pipelines. The
+low-level object-oriented APIs of Bioconductor tend to manipulate data
+via sub-replacement functions, like \texttt{start(gr)\ \textless{}-\ x}.
+These ultimately produce the side effect of replacing a symbol mapping
+in the current environment and thus are not amenable to so-called fluent
+syntax.
 
@@
 (start, end, midpoint) should be invariant over the transformation. The
-\texttt{resize()} function from the GenomicRanges package has a
+\texttt{resize()} function from the \texttt{GenomicRanges} package has a
 \texttt{fix} argument that sets the anchor, so calling \texttt{resize()}
@@
 concise than the multiple steps necessary to achieve the same effect in
-the original GenomicRanges API. In other cases, the plyranges API
-increases verbosity for the sake of clarity and cohesion. Explicitly
-calling \texttt{anchor()} can require more typing, but the code is
-easier to comprehend. Another example is the set of routines for
-importing genomic annotations, including
+the original \texttt{GenomicRanges} API. In other cases, the
+\texttt{plyranges} API increases verbosity for the sake of clarity and
+cohesion. Explicitly calling \texttt{anchor()} can require more typing,
+but the code is easier to comprehend. Another example is the set of
+routines for importing genomic annotations, including
 \texttt{read\textbackslash{}\_gff()},
 \texttt{read\textbackslash{}\_bed()}, and \texttt{read\_bam()}. Compared
-to the generic \texttt{import()} in rtracklayer, the explicit
+to the generic \texttt{import()} in \texttt{rtracklayer}, the explicit
 format-based naming in plyranges clarifies intent and the type of data
@@
 as \texttt{directed}, \texttt{upstream} or \texttt{downstream} in its
-name, otherwise strand is ignored. The GenomicRanges API does not make
-this distinction explicit in its function naming, instead relying on a
-parameter that defaults to strand sensitivity, an arguably unintuitive
-behavior.
+name, otherwise strand is ignored. The \texttt{GenomicRanges} API does
+not make this distinction explicit in its function naming, instead
+relying on a parameter that defaults to strand sensitivity, an arguably
+unintuitive behavior.
 
@@
 
-The Bioconductor package \texttt{AnnotationHub} \cite{R-ahub} can be
-used to search for BigWig files from ChIP-Seq experiments from the Human
-Epigenome Roadmap project \cite{Roadmap-Epigenomics-Consortium2015-pr}.
-Here we focus on assays for primary T CD8+ memory cells from peripheral
-blood. Using \texttt{plyranges} we will read the BigWig file
-corresponding to the H3 lysine 27 trimethylation (H3K27Me3) methylation
-mark over chromosome 10.
+In the workflow of ChIP-seq data analysis, we are interested in finding
+peaks from islands of coverage over chromosome. Here we will use
+\texttt{plyranges} to call peaks from islands of coverage above 8 then
+plot the region surrounding the tallest peak.
+
+Using \texttt{plyranges} and the the Bioconductor package
+\texttt{AnnotationHub} \cite{R-ahub} we can download and read BigWig
+files from ChIP-Seq experiments from the Human Epigenome Roadmap project
+\cite{Roadmap-Epigenomics-Consortium2015-pr}. Here we analyse a BigWig
+file corresponding to H3 lysine 27 trimethylation (H3K27Me3) of primary
+T CD8+ memory cells from peripheral blood, focussing on coverage islands
+over chromosome 10.
 
@@
 chromosome 10. We also add the genome build information to the resulting
-ranges.
+ranges. This book-keeping is good practice as it ensures the integrity
+of any downstream operations such as finding overlaps.
 
@@
 After filtering for regions with a coverage score greater than 8, we can
-reduce individual runs to ranges representing the islands of coverage,
+reduce individual runs to ranges representing the islands of coverage by
 using the \texttt{reduce\_ranges()} function. This function allows a
@@
 there are no reads overlapping the mitochondrial genome. To generate
-bins of fixed width 10000bp we apply \texttt{tile()} function from
-\texttt{GenomicRanges} to the genomic coordinates (with \texttt{unlist}
-to return a GRanges object)
+bins of fixed width 10000nt we apply the \texttt{tile\_ranges()}
+function to the genomic coordinates in locations.
 
@@
 \NormalTok{bins <-}\StringTok{ }\NormalTok{locations }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{tile}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{unlist}\NormalTok{(}\DataTypeTok{use.names =} \OtherTok{FALSE}\NormalTok{)}
+\StringTok{  }\KeywordTok{tile_ranges}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L)}
 \end{Highlighting}
@@
 
-Next we only read in alginments that overlap the genomic locations we
-are interested in and select the query seqeunce. Note that the reading
+Next we only read in alignments that overlap the genomic locations we
+are interested in and select the query sequence. Note that the reading
 of the BAM file is deferred: only alignments that pass the filter are
@@
 
-Finally, we use \texttt{group\_by\_overlaps()} with \texttt{summarize()}
-to compute the total number of reads and averge GC content within each
-genomic window.
+Finally, we use \texttt{group\_by\_overlaps()} to see where the
+alignments overlap the genomic windows and then apply
+\texttt{summarize()} to compute the total number of reads and average GC
+content within each window.
 
@@
 
-We can then compute the per-chromosome transition-transversion ratio
-using the \texttt{group\_by} and \texttt{summarize} pattern (figure
-\ref{fig:titv-viz}).
+We then compute the transition-transversion ratio over each chromosome
+using \texttt{group\_by()} in combination with \texttt{summarize()}
+(figure \ref{fig:titv-viz}).
 
@@
 GRanges data model is tidy we have highlighted how to implement a
-grammar for performing genomic arithemtic, aggregation, restriction and
+grammar for performing genomic arithmetic, aggregation, restriction and
 merging. Our examples show that \texttt{plyranges} code is succinct,
 human readable and can take advantage of the interoperability provided
-by the Bioconductor ecosytem and the R language.
+by the Bioconductor ecosystem and the R language.
 

### 2018-05-16 README.Rmd
- more detail in readme

@@
 # plyranges-paper
+
 The setup chunk in the [paper](paper.Rmd) shows how to download the data with R.
 
-Requires the following packages (and versions) to reproduce the paper:
+The paper uses the current development version of `plyranges` which
+you can install from github using:
+
+```{r, eval = FALSE}
+source("https://bioconductor.org/biocLite.R")
+biocLite("sa-lee/plyranges")
+```
+
+The following packages (and versions) are required to reproduce the paper:
 

### 2018-05-22 paper.Rmd
add more detailed author affiliations

@@
 ---
-title: "plyranges: a grammar for transforming genomics data" 
+title: "plyranges: a grammar of genomic data transformation" 
 author:
@@
     email: stuart.lee1@monash.edu
-    affiliation: Monash University
+    affiliation: |
+     | Department of Econometrics and Business Statistics, Monash University
+     |                        Molecular Medicine Division, Walter and Eliza Hall Institute
+  - name: Di Cook
+    email: di.cook@monash.edu
+    affiliation: Department of Econometrics and Business Statistics, Monash University
   - name: Michael Lawrence
     email: michael.lawrence@gene.com
-    affiliation: Genentech
-  - name: Di Cook
-    email: di.cook@monash.edu
-    affiliation: Monash University
+    affiliation: Bioinformatics and Computational Biology, Genentech, Inc.
 abstract: "The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments.
@@
 It provides a principled way of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows.
-We have implemented this grammar as a Bioconductor/R package called plyranges."
+We have implemented this grammar as a Bioconductor/R package called \\texttt{plyranges}."
 keywords: "Bioconductor, Grammar, Genomes, Data Analysis"
@@
 ---
-
 ```{r setup, include = FALSE}

### 2018-05-22 paper.tex
add more detailed author affiliations

@@
 
-\title{plyranges: a grammar for transforming genomics data  }
+\title{plyranges: a grammar of genomic data transformation  }
 
-\author{\Large Stuart Lee\vspace{0.05in} \newline\normalsize\emph{Monash University}   \and \Large Michael Lawrence\vspace{0.05in} \newline\normalsize\emph{Genentech}   \and \Large Di Cook\vspace{0.05in} \newline\normalsize\emph{Monash University}  }
+\author{\Large Stuart Lee\vspace{0.05in} \newline\normalsize\emph{Department of Econometrics and Business Statistics, Monash University\\
+\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
+Medicine Division, Walter and Eliza Hall Institute}   \and \Large Di Cook\vspace{0.05in} \newline\normalsize\emph{Department of Econometrics and Business Statistics, Monash University}   \and \Large Michael Lawrence\vspace{0.05in} \newline\normalsize\emph{Bioinformatics and Computational Biology, Genentech, Inc.}  }
 
@@
             bookmarks=true,
-            pdfauthor={Stuart Lee (Monash University) and Michael Lawrence (Genentech) and Di Cook (Monash University)},
+            pdfauthor={Stuart Lee (Department of Econometrics and Business Statistics, Monash University\\
+\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
+Medicine Division, Walter and Eliza Hall Institute) and Di Cook (Department of Econometrics and Business Statistics, Monash University) and Michael Lawrence (Bioinformatics and Computational Biology, Genentech, Inc.)},
              pdfkeywords = {Bioconductor, Grammar, Genomes, Data Analysis},  
-            pdftitle={plyranges: a grammar for transforming genomics data},
+            pdftitle={plyranges: a grammar of genomic data transformation},
             colorlinks=true,
@@
    \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
-\textbf{\authorfont Stuart Lee} \hskip 15pt \emph{\small Monash University}   \par \textbf{\authorfont Michael Lawrence} \hskip 15pt \emph{\small Genentech}   \par \textbf{\authorfont Di Cook} \hskip 15pt \emph{\small Monash University}
+\textbf{\authorfont Stuart Lee} \hskip 15pt \emph{\small Department of Econometrics and Business Statistics, Monash University\\
+\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
+Medicine Division, Walter and Eliza Hall Institute}   \par \textbf{\authorfont Di Cook} \hskip 15pt \emph{\small Department of Econometrics and Business Statistics, Monash University}   \par \textbf{\authorfont Michael Lawrence} \hskip 15pt \emph{\small Bioinformatics and Computational Biology, Genentech, Inc.}
 
@@
 readable analysis workflows. We have implemented this grammar as a
-Bioconductor/R package called plyranges.
+Bioconductor/R package called \texttt{plyranges}.
 
@@
 
-\noindent  \section{Introduction}\label{introduction}
+\noindent  \hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-\section{Genomic relational algebra}\label{genomic-relational-algebra}
+\hypertarget{genomic-relational-algebra}{%
+\section{Genomic relational algebra}\label{genomic-relational-algebra}}
 
-\subsection{Data model}\label{data-model}
+\hypertarget{data-model}{%
+\subsection{Data model}\label{data-model}}
 
@@
 
-\subsection{Algebraic operations}\label{algebraic-operations}
+\hypertarget{algebraic-operations}{%
+\subsection{Algebraic operations}\label{algebraic-operations}}
 
@@
 
-\subsection{Design principles}\label{design-principles}
+\hypertarget{design-principles}{%
+\subsection{Design principles}\label{design-principles}}
 
@@
 
+\hypertarget{cognitive-consistency-and-fluency}{%
 \subsubsection{Cognitive consistency and
-fluency}\label{cognitive-consistency-and-fluency}
+fluency}\label{cognitive-consistency-and-fluency}}
 
@@
 
-\subsubsection{Cohesion}\label{cohesion}
+\hypertarget{cohesion}{%
+\subsubsection{Cohesion}\label{cohesion}}
 
@@
 
-\subsubsection{Expressiveness}\label{expressiveness}
+\hypertarget{expressiveness}{%
+\subsubsection{Expressiveness}\label{expressiveness}}
 
@@
 
-\section{Results}\label{results}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 
-\subsection{Peak Finding}\label{peak-finding}
+\hypertarget{peak-finding}{%
+\subsection{Peak Finding}\label{peak-finding}}
 
@@
 
+\hypertarget{computing-windowed-statistics}{%
 \subsection{Computing Windowed
-Statistics}\label{computing-windowed-statistics}
+Statistics}\label{computing-windowed-statistics}}
 
@@
 
-\subsection{Quality Control Metrics}\label{quality-control-metrics}
+\hypertarget{quality-control-metrics}{%
+\subsection{Quality Control Metrics}\label{quality-control-metrics}}
 
@@
 
-\section{Discussion and conclusion}\label{discussion-and-conclusion}
+\hypertarget{discussion-and-conclusion}{%
+\section{Discussion and conclusion}\label{discussion-and-conclusion}}
 

### 2018-05-23 paper.Rmd
tidy up text and figures

@@
 ---
-title: "plyranges: a grammar of genomic data transformation" 
+title: "plyranges: A grammar of genomic data transformation" 
 author:
@@
 We have implemented this grammar as a Bioconductor/R package called \\texttt{plyranges}."
-keywords: "Bioconductor, Grammar, Genomes, Data Analysis"
+keywords: "bioconductor, grammar, genomes, data analysis"
 bibliography: references.bib
@@
 This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
-The join semantics alter what is returned: for an \textbf{inner} join the query range is returned for each match, for an \textbf{intersect} the intersection is taken between overlapping ranges, and for a \textbf{left} join all query ranges are returned even if the subject range does not overlap them.
+The join semantics alter what is returned: for an \\textbf{inner} join the query range is returned for each match, for an \\textbf{intersect} the intersection is taken between overlapping ranges, and for a \\textbf{left} join all query ranges are returned even if the subject range does not overlap them.
 This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
@@
 Each join takes a query and subject range and computes a 'Hits' object.
-For the \textbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range.
-For the \textbf{follow} join there is only one query range that follows any subject range.
-Likewise for the \textbf{precede} join, there is only one query range that precedes a subject range."}
+For the \\textbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range.
+For the \\textbf{follow} join there is only one query range that follows any subject range.
+Likewise for the \\textbf{precede} join, there is only one query range that precedes a subject range."}
 knitr::include_graphics("diagrams/diagrams-003.png")
@@
 Finally, the overlap inner join is used to restrict the chromosome 10
-normalised coverage scores that are within the 5000nt region that contain
-the max peak on chromosome 10 (figure \ref{fig:peak-viz}).
+coverage islands, to the islands that are contained in the 5000nt region that surrounds the max peak (figure \ref{fig:peak-viz}).
 
@@
 
-```{r peak-viz, echo = FALSE, fig.cap = "Visualisation of normalised coverage scores accross a 5000nt region of chromosome 10 from H3K27Me3 ChIP-Seq assay from the Human Epigenome Roadmap project."}
+```{r peak-viz, echo = FALSE, fig.cap = "The final result of the \\texttt{plyranges} operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot."}
 ggbio::autoplot(peak_region, 
@@
 
-```{r titv-viz, echo = FALSE, fig.cap = "Dot plot of chromosomes ordered by estimated transition-transversion ratio. A white reference line is drawn at the expected ratio for a human exome."}
+```{r titv-viz, echo = FALSE, fig.cap = "The final result of computing quality control metrics over the SNP array data with \\texttt{plyranges}, displayed as a dot plot.
+Chromosomes are ordered by their estimated transition-transversion ratio.
+A white reference line is drawn at the expected ratio for a human exome."}
 ti_tv_results %>% 
@@
   geom_hline(yintercept = 3, colour = "white", size = 4) +
-  labs(x = "Chromosome", y = "Transition Transverion Ratio") +
+  labs(x = "Chromosome", y = "Transition Transversion Ratio") +
   coord_flip() 
@@
 
-The `plyranges` package is available on the Bioconductor project
-website [https://bioconductor.org](https://bioconductor.org) 
-or can be accessed via Github
+The `plyranges` package can be obtained via the Bioconductor project
+website [https://bioconductor.org](https://bioconductor.org) or
+accessed via Github
 [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
@@
 this work. We would also like to thank Lori Shepherd and Hèrve Pages
-for the code review they performed. This report was written with
-`knitr` \cite{R-knitr} and the figures were made with `ggbio` \cite{R-ggbio}. All code
-required to reproduce this article is available at 
+for the code review they performed. This article was written with
+`knitr` \cite{R-knitr} and the figures were made with `ggbio` \cite{R-ggbio}. All code required to reproduce this article is available at 
 https://github.com/sa-lee/plyranges-paper.

### 2018-05-23 paper.tex
tidy up text and figures

@@
 
-\title{plyranges: a grammar of genomic data transformation  }
+\title{plyranges: A grammar of genomic data transformation  }
 
@@
 Medicine Division, Walter and Eliza Hall Institute) and Di Cook (Department of Econometrics and Business Statistics, Monash University) and Michael Lawrence (Bioinformatics and Computational Biology, Genentech, Inc.)},
-             pdfkeywords = {Bioconductor, Grammar, Genomes, Data Analysis},  
-            pdftitle={plyranges: a grammar of genomic data transformation},
+             pdfkeywords = {bioconductor, grammar, genomes, data analysis},  
+            pdftitle={plyranges: A grammar of genomic data transformation},
             colorlinks=true,
@@
 
-\vskip 8.5pt \noindent \emph{Keywords}: Bioconductor, Grammar, Genomes, Data Analysis \par
+\vskip 8.5pt \noindent \emph{Keywords}: bioconductor, grammar, genomes, data analysis \par
 
@@
 This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
-The join semantics alter what is returned: for an     extbf{inner} join the query range is returned for each match, for an    extbf{intersect} the intersection is taken between overlapping ranges, and for a    extbf{left} join all query ranges are returned even if the subject range does not overlap them.
+The join semantics alter what is returned: for an \textbf{inner} join the query range is returned for each match, for an \textbf{intersect} the intersection is taken between overlapping ranges, and for a \textbf{left} join all query ranges are returned even if the subject range does not overlap them.
 This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations.}\label{fig:olaps-fig}
@@
 Each join takes a query and subject range and computes a 'Hits' object.
-For the   extbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range.
-For the   extbf{follow} join there is only one query range that follows any subject range.
-Likewise for the   extbf{precede} join, there is only one query range that precedes a subject range.}\label{fig:nn-fig}
+For the \textbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range.
+For the \textbf{follow} join there is only one query range that follows any subject range.
+Likewise for the \textbf{precede} join, there is only one query range that precedes a subject range.}\label{fig:nn-fig}
 \end{figure}
@@
 Finally, the overlap inner join is used to restrict the chromosome 10
-normalised coverage scores that are within the 5000nt region that
-contain the max peak on chromosome 10 (figure \ref{fig:peak-viz}).
+coverage islands, to the islands that are contained in the 5000nt region
+that surrounds the max peak (figure \ref{fig:peak-viz}).
 
@@
 
-\caption{Visualisation of normalised coverage scores accross a 5000nt region of chromosome 10 from H3K27Me3 ChIP-Seq assay from the Human Epigenome Roadmap project.}\label{fig:peak-viz}
+\caption{The final result of the \texttt{plyranges} operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.}\label{fig:peak-viz}
 \end{figure}
@@
 
-\caption{Dot plot of chromosomes ordered by estimated transition-transversion ratio. A white reference line is drawn at the expected ratio for a human exome.}\label{fig:titv-viz}
+\caption{The final result of computing quality control metrics over the SNP array data with \texttt{plyranges}, displayed as a dot plot.
+Chromosomes are ordered by their estimated transition-transversion ratio.
+A white reference line is drawn at the expected ratio for a human exome.}\label{fig:titv-viz}
 \end{figure}
@@
 
-The \texttt{plyranges} package is available on the Bioconductor project
-website \url{https://bioconductor.org} or can be accessed via Github
+The \texttt{plyranges} package can be obtained via the Bioconductor
+project website \url{https://bioconductor.org} or accessed via Github
 \url{https://github.com/sa-lee/plyranges}.
@@
 this work. We would also like to thank Lori Shepherd and Hèrve Pages for
-the code review they performed. This report was written with
+the code review they performed. This article was written with
 \texttt{knitr} \cite{R-knitr} and the figures were made with

### 2018-06-15 paper.Rmd
changes for GB sub

@@
 author:
-  - name: Stuart Lee
+  - name: Stuart Lee \footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}
     email: stuart.lee1@monash.edu
-    affiliation: |
-     | Department of Econometrics and Business Statistics, Monash University
-     |                        Molecular Medicine Division, Walter and Eliza Hall Institute
-  - name: Di Cook
+  - name: Dianne Cook $^{\ref{monash}}$
     email: di.cook@monash.edu 
-    affiliation: Department of Econometrics and Business Statistics, Monash University
-  - name: Michael Lawrence
+  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author}
     email: michael.lawrence@gene.com
-    affiliation: Bioinformatics and Computational Biology, Genentech, Inc.
 abstract: "The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments.
@@
 We have implemented this grammar as a Bioconductor/R package called \\texttt{plyranges}."
-keywords: "bioconductor, grammar, genomes, data analysis"
+keywords: "Bioconductor, Grammar, Genomes, Data Analysis"
 bibliography: references.bib
@@
 ---
+
 ```{r setup, include = FALSE}
@@
 ```
-# Introduction
+# Background
 
@@
 
-# Genomic relational algebra
+# Results
 
-## Data model
+## Genomic Relation Algebra
+
+### Data Model
 
@@
 high-level operations on genomic ranges. As a core data structure,
-GRanges enables interoperability between plyranges and the rest of
+GRanges enables interoperability between `plyranges` and the rest of
 Bioconductor. Adhering to a single data structure simplifies the API
@@
 remains compatible with the general relational operations defined by
-dplyr. Thus, we define our algebra as an extension of the dplyr
+dplyr. Thus, we define our algebra as an extension of the `dplyr`
 algebra, and borrow its syntax conventions and design principles.
@@
 
-## Algebraic operations
+### Algebraic operations
 
@@
 _difference_, and _complement_. There are two sets of these: parallel
-and merging. The parallel operations map to infix operators, which we
-surround with _%_ symbols, a convention borrowed from R syntax.  For
-example, the parallel intersection (_x %intersect% y_) finds the
+and merging. For example, the parallel intersection (_x %intersect% y_) finds the
 intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
@@
 ranges representing wherever there was overlap between a range in _x_
-and a range in _y_. We use the infix syntax for the parallel
-operations, since it is the conventional syntax for parallel, binary
-operations in R. Finding the parallel union will fail when two ranges
+and a range in _y_. Finding the parallel union will fail when two ranges
 have a gap, so we introduce a _span_ operator that takes the union
@@
 
-## Design principles
-
-The design of `plyranges` adheres to well understood principles of
-language and API design: cognitive consistency, cohesion,
-expressiveness and endomorphism \cite{Green1996-qg}. 
-To varying degrees, these principles also underlie the
-design of `dplyr` and the Bioconductor infrastructure.
-
-### Cognitive consistency and fluency
-
-We aim for our interfaces to have a simple and direct mapping to the
-user's cognitive model, i.e., how the user thinks about the data. This
-requires careful selection of the level of abstraction so that the
-user can express workflows in the language of genomics. This motivates
-the adoption of the tidy GRanges object as our central data
-structure. The basic data.frame and `dplyr` tibble lack any notion of
-genomic ranges and so could not easily support our genomic grammar,
-with its specific verbs for range-oriented data manipulation. Another
-example of cognitive consistency is how `plyranges` is insensitive to
-direction/strand by default when, e.g., detecting
-overlaps. `GenomicRanges` has the opposite behavior. We believe that
-defaulting to purely spatial overlap is most intuitive to most users.
-
-<!--
-ML:
-Hadley actually presents dplyr as a 'cognitive tool', i.e., something
-that helps users think about the problem. In my opinion, the best we
-can do is attempt to match the intuition of the user, so when the user
-decides what to do, there is an obvious way of doing it. I'm not ready
-to claim that the user should start from our grammar.
--->
-Like `dplyr`, `plyranges` verbs are functional: they are free of side
-effects and return their result. This enables chaining of verbs
-through syntax like the forward pipe operator (`%>%`, read as "then")
-of the `magrittr` package \cite{R-magrittr}. This has syntax has a direct
-cognitive mapping to natural language and the intuitive notion of
-pipelines. The low-level object-oriented APIs of Bioconductor tend to
-manipulate data via sub-replacement functions, like `start(gr) <-
-x`. These ultimately produce the side effect of replacing a symbol
-mapping in the current environment and thus are not amenable to
-so-called fluent syntax.
-
-### Cohesion
-
-A function is cohesive if it performs a singular task without
-producing any side-effects. Singular tasks are not necessarily atomic;
-they can always be broken down further at lower levels of abstraction.
-For example, to resize a range, the user needs to specify which
-position (start, end, midpoint) should be invariant over the
-transformation. The `resize()` function from the `GenomicRanges` package
-has a `fix` argument that sets the anchor, so calling `resize()`
-coalesces anchoring and width modification. The coupling at the
-function call level is justified since the effect of setting the width
-depends on the anchor. However, plyranges increases cohesion and
-decouples the anchoring into its own function call.
-
-Increasing cohesion simplifies the interface to each operation, makes
-the meaning of arguments more intuitive, and relies on function names
-as the primary means of expression, instead of a more complex mixture
-of function and argument names. Since functions are superordinate to
-their arguments, flattening the API at the function level enables the
-user to conceptualize the API as a catalog of functions, without
-having to descend further. A flat function catalog also enhances API
-discoverability, particularly through auto-completion in IDEs. One
-downside of pushing cohesion to this extreme is that function calls
-become coupled, and care is necessary to treat them as a group when
-modifying code.
-
-### Expressiveness
-
-Expressiveness relates to the information content in code: the
-programmer should be able to clarify intent without unnecessary
-verbosity. For example, our overlap-based join operations are more
-concise than the multiple steps necessary to achieve the same effect
-in the original `GenomicRanges` API. In other cases, the `plyranges` API
-increases verbosity for the sake of clarity and cohesion. Explicitly
-calling `anchor()` can require more typing, but the code is easier to
-comprehend. Another example is the set of routines for importing
-genomic annotations, including `read\_gff()`, `read\_bed()`, and
-`read_bam()`. Compared to the generic `import()` in `rtracklayer`, the
-explicit format-based naming in plyranges clarifies intent and the
-type of data being returned. Similarly, every plyranges function that
-computes with strand information indicates its intentions by including
-suffixes such as `directed`, `upstream` or `downstream` in its name,
-otherwise strand is ignored. The `GenomicRanges` API does not make this
-distinction explicit in its function naming, instead relying on a
-parameter that defaults to strand sensitivity, an arguably unintuitive
-behavior.
-
-# Results
+## Developing workflows with `plyranges`
 
@@
 
-## Peak Finding
+### Peak Finding
 
@@
 
-```{r peak-viz, echo = FALSE, fig.cap = "The final result of the \\texttt{plyranges} operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot."}
+```{r peak-viz, echo = FALSE, fig.cap = "The final result of the \\texttt{plyranges} operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.", cache = TRUE}
 ggbio::autoplot(peak_region, 
@@
                 geom = "blank") + 
-  geom_area(aes(x = start)) +
+  geom_area() +
   ylab("score")
@@
 
-## Computing Windowed Statistics
+### Computing Windowed Statistics
 
@@
 ```
-
 <!-- 
@@
 -->
-
-## Quality Control Metrics
+### Quality Control Metrics
 
@@
 
-# Discussion and conclusion
+# Discussion
+
+The design of `plyranges` adheres to well understood principles of
+language and API design: cognitive consistency, cohesion,
+expressiveness and endomorphism \cite{Green1996-qg}. 
+To varying degrees, these principles also underlie the
+design of `dplyr` and the Bioconductor infrastructure.
+
+We have aimed for `plyranges` to have a simple and direct mapping to the
+user's cognitive model, i.e., how the user thinks about the data. This
+requires careful selection of the level of abstraction so that the
+user can express workflows in the language of genomics. This motivates
+the adoption of the tidy GRanges object as our central data
+structure. The basic data.frame and `dplyr` tibble lack any notion of
+genomic ranges and so could not easily support our genomic grammar,
+with its specific verbs for range-oriented data manipulation. Another
+example of cognitive consistency is how `plyranges` is insensitive to
+direction/strand by default when, e.g., detecting
+overlaps. `GenomicRanges` has the opposite behavior. We believe that
+defaulting to purely spatial overlap is most intuitive to most users.
+
+<!--
+ML:
+Hadley actually presents dplyr as a 'cognitive tool', i.e., something
+that helps users think about the problem. In my opinion, the best we
+can do is attempt to match the intuition of the user, so when the user
+decides what to do, there is an obvious way of doing it. I'm not ready
+to claim that the user should start from our grammar.
+-->
+
+Like `dplyr`, `plyranges` verbs are functional: they are free of side
+effects and return their result. This enables chaining of verbs
+through syntax like the forward pipe operator (`%>%`, read as "then")
+of the `magrittr` package \cite{R-magrittr}. This has syntax has a direct
+cognitive mapping to natural language and the intuitive notion of
+pipelines. The low-level object-oriented APIs of Bioconductor tend to
+manipulate data via sub-replacement functions, like `start(gr) <-
+x`. These ultimately produce the side effect of replacing a symbol
+mapping in the current environment and thus are not amenable to
+so-called fluent syntax.
+
+A function is cohesive if it performs a singular task without
+producing any side-effects. Singular tasks can always be broken 
+down further at lower levels of abstraction.
+For example, to resize a range, the user needs to specify which
+position (start, end, midpoint) should be invariant over the
+transformation. The `resize()` function from the `GenomicRanges` package
+has a `fix` argument that sets the anchor, so calling `resize()`
+coalesces anchoring and width modification. The coupling at the
+function call level is justified since the effect of setting the width
+depends on the anchor. However, `plyranges` increases cohesion and
+decouples the anchoring into its own function call.
+
+Increasing cohesion simplifies the interface to each operation, makes
+the meaning of arguments more intuitive, and relies on function names
+as the primary means of expression, instead of a more complex mixture
+of function and argument names. This results in the
+user being able to conceptualize the `plyranges` DSL as a flat catalog of 
+functions, without having to descend further into documentation to understand
+a function's arguments. A flat function catalog also enhances API
+discoverability, particularly through auto-completion in IDEs. One
+downside of pushing cohesion to this extreme is that function calls
+become coupled, and care is necessary to treat them as a group when
+modifying code.
+
+Expressiveness relates to the information content in code: the
+programmer should be able to clarify intent without unnecessary
+verbosity. For example, our overlap-based join operations are more
+concise than the multiple steps necessary to achieve the same effect
+in the original `GenomicRanges` API. In other cases, the `plyranges` API
+increases verbosity for the sake of clarity and cohesion. Explicitly
+calling _anchor()_ can require more typing, but the code is easier to
+comprehend. Another example is the set of routines for importing
+genomic annotations, including _read\_gff()_, _read\_bed()_, and
+_read\_bam()_. Compared to the generic _import()_ in `rtracklayer`, the
+explicit format-based naming in `plyranges` clarifies intent and the
+type of data being returned. Similarly, every `plyranges` function that
+computes with strand information indicates its intentions by including
+suffixes such as _directed_, _upstream_ or _downstream_ in its name,
+otherwise strand is ignored. The `GenomicRanges` API does not make this
+distinction explicit in its function naming, instead relying on a
+parameter that defaults to strand sensitivity, an arguably unintuitive
+behavior.
+
+A caveat to constructing a compatible interface with `dplyr` is that
+`plyranges` makes extensive use of non-standard evaluation in R via
+the `rlang` package \cite{R-rlang}. Simply, this means that computations are
+evaluated in the context of the GRanges objects. Both `dplyr` and
+`plyranges` are based on the `rlang` language, because it allows for more
+expressive code that is free of repeated references to the
+container. Implicitly referencing the container is particularly
+convenient when programming interactively.  Consequently, when
+programming with `plyranges`, a user needs to generally understand
+the `rlang` language and how to adapt their code accordingly. Users
+familiar with the tidyverse should already have such knowledge.
+
+# Conclusion
 
@@
 the Bioconductor ecosystem and the R language.
-
-A caveat to constructing a compatible interface with `dplyr` is that
-`plyranges` makes extensive use of non-standard evaluation in R via
-the `rlang` package \cite{R-rlang}. Simply, this means that computations are
-evaluated in the context of the GRanges objects. Both dplyr and
-plyranges are based on the `rlang` language, because it allows for more
-expressive code that is free of repeated references to the
-container. Implicitly referencing the container is particularly
-convenient when programming interactively.  Consequently, when
-programming with `plyranges`, a user needs to generally understand
-the `rlang` language and how to adapt their code accordingly. Users
-familiar with the tidyverse should already have such knowledge.
 
@@
 access and computing patterns, so the deferred mechanisms within
-plyranges should help support interactive genomics applications.
+`plyranges` should help support interactive genomics applications.
 
@@
 [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
+
+# Methods
+## Data Availability
+The BigWig file for the H3K27Me3 primary T CD8+ memory cells from peripheral 
+blood ChIP-seq data was downloaded from the `AnnotationHub` package (2.12.0)
+under accession AH33458. The BAM file corresponding to the
+H1 cell line ChIP-seq data is available at GEO under accession GSM433167. 
+The SNP array data for the H1 cell line data is available at GEO under accession
+GSM1463263.
+
+## Software Versions
+To produce the workflows as described in results section we used 
+R version 3.5 with the development version of `plyranges` (1.1.4) 
+and the  `BioStrings` package (2.48.0) installed.
+
+All code required to reproduce this article is available at https://github.com/sa-lee/plyranges-paper.
 
@@
 for the code review they performed. This article was written with
-`knitr` \cite{R-knitr} and the figures were made with `ggbio` \cite{R-ggbio}. All code required to reproduce this article is available at 
-https://github.com/sa-lee/plyranges-paper.
+`knitr` \cite{R-knitr} and the figures were made with `ggbio` \cite{R-ggbio}.
 

### 2018-06-15 paper.tex
changes for GB sub

@@
 \setlength{\parskip}{1em}
+
+\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
 
@@
 
-\author{\Large Stuart Lee\vspace{0.05in} \newline\normalsize\emph{Department of Econometrics and Business Statistics, Monash University\\
-\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
-Medicine Division, Walter and Eliza Hall Institute}   \and \Large Di Cook\vspace{0.05in} \newline\normalsize\emph{Department of Econometrics and Business Statistics, Monash University}   \and \Large Michael Lawrence\vspace{0.05in} \newline\normalsize\emph{Bioinformatics and Computational Biology, Genentech, Inc.}  }
+\author{\Large Stuart Lee
+\footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}\vspace{0.05in} \newline\normalsize\emph{}   \and \Large Dianne Cook \(^{\ref{monash}}\)\vspace{0.05in} \newline\normalsize\emph{}   \and \Large Michael Lawrence
+\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author}\vspace{0.05in} \newline\normalsize\emph{}  }
 
@@
             bookmarks=true,
-            pdfauthor={Stuart Lee (Department of Econometrics and Business Statistics, Monash University\\
-\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
-Medicine Division, Walter and Eliza Hall Institute) and Di Cook (Department of Econometrics and Business Statistics, Monash University) and Michael Lawrence (Bioinformatics and Computational Biology, Genentech, Inc.)},
-             pdfkeywords = {bioconductor, grammar, genomes, data analysis},  
+            pdfauthor={Stuart Lee
+\footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia} () and Dianne Cook \(^{\ref{monash}}\) () and Michael Lawrence
+\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author} ()},
+             pdfkeywords = {Bioconductor, Grammar, Genomes, Data Analysis},  
             pdftitle={plyranges: A grammar of genomic data transformation},
@@
    \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
-\textbf{\authorfont Stuart Lee} \hskip 15pt \emph{\small Department of Econometrics and Business Statistics, Monash University\\
-\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
-Medicine Division, Walter and Eliza Hall Institute}   \par \textbf{\authorfont Di Cook} \hskip 15pt \emph{\small Department of Econometrics and Business Statistics, Monash University}   \par \textbf{\authorfont Michael Lawrence} \hskip 15pt \emph{\small Bioinformatics and Computational Biology, Genentech, Inc.}
+\textbf{\authorfont Stuart Lee
+\footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}} \hskip 15pt \emph{\small }   \par \textbf{\authorfont Dianne Cook \(^{\ref{monash}}\)} \hskip 15pt \emph{\small }   \par \textbf{\authorfont Michael Lawrence
+\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author}} \hskip 15pt \emph{\small }
 
@@
 
-\vskip 8.5pt \noindent \emph{Keywords}: bioconductor, grammar, genomes, data analysis \par
+\vskip 8.5pt \noindent \emph{Keywords}: Bioconductor, Grammar, Genomes, Data Analysis \par
 
@@
 
-\noindent  \hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\noindent  \hypertarget{background}{%
+\section{Background}\label{background}}
 
@@
 
-\hypertarget{genomic-relational-algebra}{%
-\section{Genomic relational algebra}\label{genomic-relational-algebra}}
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{genomic-relation-algebra}{%
+\subsection{Genomic Relation Algebra}\label{genomic-relation-algebra}}
 
 \hypertarget{data-model}{%
-\subsection{Data model}\label{data-model}}
+\subsubsection{Data Model}\label{data-model}}
 
@@
 operations on genomic ranges. As a core data structure, GRanges enables
-interoperability between plyranges and the rest of Bioconductor.
-Adhering to a single data structure simplifies the API and makes it
-easier to learn and understand, in part because operations become
-endomorphic, i.e., they return the same type as their input.
+interoperability between \texttt{plyranges} and the rest of
+Bioconductor. Adhering to a single data structure simplifies the API and
+makes it easier to learn and understand, in part because operations
+become endomorphic, i.e., they return the same type as their input.
 
@@
 remains compatible with the general relational operations defined by
-dplyr. Thus, we define our algebra as an extension of the dplyr algebra,
-and borrow its syntax conventions and design principles.
+dplyr. Thus, we define our algebra as an extension of the \texttt{dplyr}
+algebra, and borrow its syntax conventions and design principles.
 
@@
 \hypertarget{algebraic-operations}{%
-\subsection{Algebraic operations}\label{algebraic-operations}}
+\subsubsection{Algebraic operations}\label{algebraic-operations}}
 
@@
 \emph{difference}, and \emph{complement}. There are two sets of these:
-parallel and merging. The parallel operations map to infix operators,
-which we surround with \emph{\%} symbols, a convention borrowed from R
-syntax. For example, the parallel intersection (\emph{x \%intersect\%
-y}) finds the intersecting range between \emph{xi} and \emph{yi} for
-\emph{i} in \emph{1\ldots{}n}, where \emph{n} is the length of both
-\emph{x} and \emph{y}. In contrast, the merging intersection
-(\emph{intersect\_ranges(x, y)}) returns a new set of disjoint ranges
-representing wherever there was overlap between a range in \emph{x} and
-a range in \emph{y}. We use the infix syntax for the parallel
-operations, since it is the conventional syntax for parallel, binary
-operations in R. Finding the parallel union will fail when two ranges
-have a gap, so we introduce a \emph{span} operator that takes the union
-while filling any gap. The \emph{complement} operation is unique in that
-it is unary. It finds the regions not covered by any of the ranges in a
-single set. Closely related is the \emph{between} parallel operation,
-which finds the gap separating \emph{xi} and \emph{yi}. The binary
-operations are callable from within arithmetic, restriction and
-aggregation expressions.
+parallel and merging. For example, the parallel intersection (\emph{x
+\%intersect\% y}) finds the intersecting range between \emph{xi} and
+\emph{yi} for \emph{i} in \emph{1\ldots{}n}, where \emph{n} is the
+length of both \emph{x} and \emph{y}. In contrast, the merging
+intersection (\emph{intersect\_ranges(x, y)}) returns a new set of
+disjoint ranges representing wherever there was overlap between a range
+in \emph{x} and a range in \emph{y}. Finding the parallel union will
+fail when two ranges have a gap, so we introduce a \emph{span} operator
+that takes the union while filling any gap. The \emph{complement}
+operation is unique in that it is unary. It finds the regions not
+covered by any of the ranges in a single set. Closely related is the
+\emph{between} parallel operation, which finds the gap separating
+\emph{xi} and \emph{yi}. The binary operations are callable from within
+arithmetic, restriction and aggregation expressions.
 
@@
 
-\hypertarget{design-principles}{%
-\subsection{Design principles}\label{design-principles}}
-
-The design of \texttt{plyranges} adheres to well understood principles
-of language and API design: cognitive consistency, cohesion,
-expressiveness and endomorphism \cite{Green1996-qg}. To varying degrees,
-these principles also underlie the design of \texttt{dplyr} and the
-Bioconductor infrastructure.
-
-\hypertarget{cognitive-consistency-and-fluency}{%
-\subsubsection{Cognitive consistency and
-fluency}\label{cognitive-consistency-and-fluency}}
-
-We aim for our interfaces to have a simple and direct mapping to the
-user's cognitive model, i.e., how the user thinks about the data. This
-requires careful selection of the level of abstraction so that the user
-can express workflows in the language of genomics. This motivates the
-adoption of the tidy GRanges object as our central data structure. The
-basic data.frame and \texttt{dplyr} tibble lack any notion of genomic
-ranges and so could not easily support our genomic grammar, with its
-specific verbs for range-oriented data manipulation. Another example of
-cognitive consistency is how \texttt{plyranges} is insensitive to
-direction/strand by default when, e.g., detecting overlaps.
-\texttt{GenomicRanges} has the opposite behavior. We believe that
-defaulting to purely spatial overlap is most intuitive to most users.
-
-Like \texttt{dplyr}, \texttt{plyranges} verbs are functional: they are
-free of side effects and return their result. This enables chaining of
-verbs through syntax like the forward pipe operator
-(\texttt{\%\textgreater{}\%}, read as ``then'') of the \texttt{magrittr}
-package \cite{R-magrittr}. This has syntax has a direct cognitive
-mapping to natural language and the intuitive notion of pipelines. The
-low-level object-oriented APIs of Bioconductor tend to manipulate data
-via sub-replacement functions, like \texttt{start(gr)\ \textless{}-\ x}.
-These ultimately produce the side effect of replacing a symbol mapping
-in the current environment and thus are not amenable to so-called fluent
-syntax.
-
-\hypertarget{cohesion}{%
-\subsubsection{Cohesion}\label{cohesion}}
-
-A function is cohesive if it performs a singular task without producing
-any side-effects. Singular tasks are not necessarily atomic; they can
-always be broken down further at lower levels of abstraction. For
-example, to resize a range, the user needs to specify which position
-(start, end, midpoint) should be invariant over the transformation. The
-\texttt{resize()} function from the \texttt{GenomicRanges} package has a
-\texttt{fix} argument that sets the anchor, so calling \texttt{resize()}
-coalesces anchoring and width modification. The coupling at the function
-call level is justified since the effect of setting the width depends on
-the anchor. However, plyranges increases cohesion and decouples the
-anchoring into its own function call.
-
-Increasing cohesion simplifies the interface to each operation, makes
-the meaning of arguments more intuitive, and relies on function names as
-the primary means of expression, instead of a more complex mixture of
-function and argument names. Since functions are superordinate to their
-arguments, flattening the API at the function level enables the user to
-conceptualize the API as a catalog of functions, without having to
-descend further. A flat function catalog also enhances API
-discoverability, particularly through auto-completion in IDEs. One
-downside of pushing cohesion to this extreme is that function calls
-become coupled, and care is necessary to treat them as a group when
-modifying code.
-
-\hypertarget{expressiveness}{%
-\subsubsection{Expressiveness}\label{expressiveness}}
-
-Expressiveness relates to the information content in code: the
-programmer should be able to clarify intent without unnecessary
-verbosity. For example, our overlap-based join operations are more
-concise than the multiple steps necessary to achieve the same effect in
-the original \texttt{GenomicRanges} API. In other cases, the
-\texttt{plyranges} API increases verbosity for the sake of clarity and
-cohesion. Explicitly calling \texttt{anchor()} can require more typing,
-but the code is easier to comprehend. Another example is the set of
-routines for importing genomic annotations, including
-\texttt{read\textbackslash{}\_gff()},
-\texttt{read\textbackslash{}\_bed()}, and \texttt{read\_bam()}. Compared
-to the generic \texttt{import()} in \texttt{rtracklayer}, the explicit
-format-based naming in plyranges clarifies intent and the type of data
-being returned. Similarly, every plyranges function that computes with
-strand information indicates its intentions by including suffixes such
-as \texttt{directed}, \texttt{upstream} or \texttt{downstream} in its
-name, otherwise strand is ignored. The \texttt{GenomicRanges} API does
-not make this distinction explicit in its function naming, instead
-relying on a parameter that defaults to strand sensitivity, an arguably
-unintuitive behavior.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\hypertarget{developing-workflows-with-plyranges}{%
+\subsection{\texorpdfstring{Developing workflows with
+\texttt{plyranges}}{Developing workflows with plyranges}}\label{developing-workflows-with-plyranges}}
 
@@
 \hypertarget{peak-finding}{%
-\subsection{Peak Finding}\label{peak-finding}}
+\subsubsection{Peak Finding}\label{peak-finding}}
 
@@
 \hypertarget{computing-windowed-statistics}{%
-\subsection{Computing Windowed
+\subsubsection{Computing Windowed
 Statistics}\label{computing-windowed-statistics}}
@@
 \hypertarget{quality-control-metrics}{%
-\subsection{Quality Control Metrics}\label{quality-control-metrics}}
+\subsubsection{Quality Control Metrics}\label{quality-control-metrics}}
 
@@
 
-\hypertarget{discussion-and-conclusion}{%
-\section{Discussion and conclusion}\label{discussion-and-conclusion}}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The design of \texttt{plyranges} adheres to well understood principles
+of language and API design: cognitive consistency, cohesion,
+expressiveness and endomorphism \cite{Green1996-qg}. To varying degrees,
+these principles also underlie the design of \texttt{dplyr} and the
+Bioconductor infrastructure.
+
+We have aimed for \texttt{plyranges} to have a simple and direct mapping
+to the user's cognitive model, i.e., how the user thinks about the data.
+This requires careful selection of the level of abstraction so that the
+user can express workflows in the language of genomics. This motivates
+the adoption of the tidy GRanges object as our central data structure.
+The basic data.frame and \texttt{dplyr} tibble lack any notion of
+genomic ranges and so could not easily support our genomic grammar, with
+its specific verbs for range-oriented data manipulation. Another example
+of cognitive consistency is how \texttt{plyranges} is insensitive to
+direction/strand by default when, e.g., detecting overlaps.
+\texttt{GenomicRanges} has the opposite behavior. We believe that
+defaulting to purely spatial overlap is most intuitive to most users.
+
+Like \texttt{dplyr}, \texttt{plyranges} verbs are functional: they are
+free of side effects and return their result. This enables chaining of
+verbs through syntax like the forward pipe operator
+(\texttt{\%\textgreater{}\%}, read as ``then'') of the \texttt{magrittr}
+package \cite{R-magrittr}. This has syntax has a direct cognitive
+mapping to natural language and the intuitive notion of pipelines. The
+low-level object-oriented APIs of Bioconductor tend to manipulate data
+via sub-replacement functions, like \texttt{start(gr)\ \textless{}-\ x}.
+These ultimately produce the side effect of replacing a symbol mapping
+in the current environment and thus are not amenable to so-called fluent
+syntax.
+
+A function is cohesive if it performs a singular task without producing
+any side-effects. Singular tasks can always be broken down further at
+lower levels of abstraction. For example, to resize a range, the user
+needs to specify which position (start, end, midpoint) should be
+invariant over the transformation. The \texttt{resize()} function from
+the \texttt{GenomicRanges} package has a \texttt{fix} argument that sets
+the anchor, so calling \texttt{resize()} coalesces anchoring and width
+modification. The coupling at the function call level is justified since
+the effect of setting the width depends on the anchor. However,
+\texttt{plyranges} increases cohesion and decouples the anchoring into
+its own function call.
+
+Increasing cohesion simplifies the interface to each operation, makes
+the meaning of arguments more intuitive, and relies on function names as
+the primary means of expression, instead of a more complex mixture of
+function and argument names. This results in the user being able to
+conceptualize the \texttt{plyranges} DSL as a flat catalog of functions,
+without having to descend further into documentation to understand a
+function's arguments. A flat function catalog also enhances API
+discoverability, particularly through auto-completion in IDEs. One
+downside of pushing cohesion to this extreme is that function calls
+become coupled, and care is necessary to treat them as a group when
+modifying code.
+
+Expressiveness relates to the information content in code: the
+programmer should be able to clarify intent without unnecessary
+verbosity. For example, our overlap-based join operations are more
+concise than the multiple steps necessary to achieve the same effect in
+the original \texttt{GenomicRanges} API. In other cases, the
+\texttt{plyranges} API increases verbosity for the sake of clarity and
+cohesion. Explicitly calling \emph{anchor()} can require more typing,
+but the code is easier to comprehend. Another example is the set of
+routines for importing genomic annotations, including
+\emph{read\_gff()}, \emph{read\_bed()}, and \emph{read\_bam()}. Compared
+to the generic \emph{import()} in \texttt{rtracklayer}, the explicit
+format-based naming in \texttt{plyranges} clarifies intent and the type
+of data being returned. Similarly, every \texttt{plyranges} function
+that computes with strand information indicates its intentions by
+including suffixes such as \emph{directed}, \emph{upstream} or
+\emph{downstream} in its name, otherwise strand is ignored. The
+\texttt{GenomicRanges} API does not make this distinction explicit in
+its function naming, instead relying on a parameter that defaults to
+strand sensitivity, an arguably unintuitive behavior.
+
+A caveat to constructing a compatible interface with \texttt{dplyr} is
+that \texttt{plyranges} makes extensive use of non-standard evaluation
+in R via the \texttt{rlang} package \cite{R-rlang}. Simply, this means
+that computations are evaluated in the context of the GRanges objects.
+Both \texttt{dplyr} and \texttt{plyranges} are based on the
+\texttt{rlang} language, because it allows for more expressive code that
+is free of repeated references to the container. Implicitly referencing
+the container is particularly convenient when programming interactively.
+Consequently, when programming with \texttt{plyranges}, a user needs to
+generally understand the \texttt{rlang} language and how to adapt their
+code accordingly. Users familiar with the tidyverse should already have
+such knowledge.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
@@
 by the Bioconductor ecosystem and the R language.
-
-A caveat to constructing a compatible interface with \texttt{dplyr} is
-that \texttt{plyranges} makes extensive use of non-standard evaluation
-in R via the \texttt{rlang} package \cite{R-rlang}. Simply, this means
-that computations are evaluated in the context of the GRanges objects.
-Both dplyr and plyranges are based on the \texttt{rlang} language,
-because it allows for more expressive code that is free of repeated
-references to the container. Implicitly referencing the container is
-particularly convenient when programming interactively. Consequently,
-when programming with \texttt{plyranges}, a user needs to generally
-understand the \texttt{rlang} language and how to adapt their code
-accordingly. Users familiar with the tidyverse should already have such
-knowledge.
 
@@
 access and computing patterns, so the deferred mechanisms within
-plyranges should help support interactive genomics applications.
+\texttt{plyranges} should help support interactive genomics
+applications.
 
@@
 \url{https://github.com/sa-lee/plyranges}.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\hypertarget{data-availability}{%
+\subsection{Data Availability}\label{data-availability}}
+
+The BigWig file for the H3K27Me3 primary T CD8+ memory cells from
+peripheral blood ChIP-seq data was downloaded from the
+\texttt{AnnotationHub} package (2.12.0) under accession AH33458. The BAM
+file corresponding to the H1 cell line ChIP-seq data is available at GEO
+under accession GSM433167. The SNP array data for the H1 cell line data
+is available at GEO under accession GSM1463263.
+
+\hypertarget{software-versions}{%
+\subsection{Software Versions}\label{software-versions}}
+
+To produce the workflows as described in results section we used R
+version 3.5 with the development version of \texttt{plyranges} (1.1.4)
+and the \texttt{BioStrings} package (2.48.0) installed.
+
+All code required to reproduce this article is available at
+\url{https://github.com/sa-lee/plyranges-paper}.
 
@@
 \texttt{knitr} \cite{R-knitr} and the figures were made with
-\texttt{ggbio} \cite{R-ggbio}. All code required to reproduce this
-article is available at \url{https://github.com/sa-lee/plyranges-paper}.
+\texttt{ggbio} \cite{R-ggbio}.
 

### 2018-06-27 paper.Rmd
some tweaks to the abstract but probably messed up yaml

@@
     email: di.cook@monash.edu 
-  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author}
+  - name: Michael Lawrence \footnote{Bioinformatics and Computational
+    Biology, Genentech Research and Early Development, South San
+    Francisco, United States of America}\footnote{corresponding
+    author} 
 	email: michael.lawrence@gene.com
-abstract: "The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments.
-However, there is a cognitive load placed on a user in learning a data abstraction and understanding its appropriate use.
-Throughout a standard workflow, a user must navigate and know many of these abstractions to perform an genomic analysis task, when a single data abstraction, a GRanges object will suffice.
-The GRanges class naturally represent genomic intervals and their associated measurements.
-By recognising that the GRanges class follows 'tidy' data principles we have created a grammar of genomic data transformation.
-The grammar defines verbs for performing actions on and between genomic interval data.
-It provides a principled way of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows.
-We have implemented this grammar as a Bioconductor/R package called \\texttt{plyranges}."
+
+abstract: "The Bioconductor project provides a host of interoperable data abstractions for analysing high-throughput genomics experiments.
+Implementing a typical genomic workflow with Bioconductor requires learning the abstractions and understanding them at an integrative level.
+This places a large cognitive burden on the user.
+We have created a grammar of genomic data transformation that operates on single, central Bioconductor data structure, GRanges, which naturally represents genomic intervals and their associated measurements.
+The grammar defines verbs for performing actions on and between genomic interval data through a simplified, coherent interface to existing Bioconductor infrastructure, resulting in fluent analysis workflows.
+We have implemented this grammar as an R/Bioconductor package called \\texttt{plyranges}."
+
 keywords: "Bioconductor, Grammar, Genomes, Data Analysis" 
@@
 standard formats like BED. An example of the former is the Genome Query 
-Language (GQL) and its distributed implementation GenAp which use an 
+Language (GQL) and its distributed implementation GenAp which use a 
 SQL-like syntax for fast retrieval of information of unprocessed 

### 2018-07-17 paper.Rmd
tidying up abstract

@@
 
-abstract: "The Bioconductor project provides a host of interoperable data abstractions for analysing high-throughput genomics experiments.
-Implementing a typical genomic workflow with Bioconductor requires learning the abstractions and understanding them at an integrative level.
+abstract: "The Bioconductor project provides many interoperable data abstractions for analysing high-throughput genomics experiments, however implementing a typical genomic workflow with Bioconductor requires learning these abstractions and understanding them at an integrative level.
 This places a large cognitive burden on the user.

### 2018-07-17 paper.Rmd
tidying abstract; adding in para to conclusion

--- 2018-06-27 paper.Rmd
@@
     email: di.cook@monash.edu 
-  - name: Michael Lawrence \footnote{Bioinformatics and Computational
-    Biology, Genentech Research and Early Development, South San
-    Francisco, United States of America}\footnote{corresponding
-    author} 
-	email: michael.lawrence@gene.com
-
-abstract: "The Bioconductor project provides a host of interoperable data abstractions for analysing high-throughput genomics experiments.
-Implementing a typical genomic workflow with Bioconductor requires learning the abstractions and understanding them at an integrative level.
-This places a large cognitive burden on the user.
-We have created a grammar of genomic data transformation that operates on single, central Bioconductor data structure, GRanges, which naturally represents genomic intervals and their associated measurements.
+  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author} 
+    email: lawrence.michael@gene.com
+abstract: "The Bioconductor project provides many interoperable data abstractions for analysing high-throughput genomics experiments; however implementing a typical genomic workflow with Bioconductor requires learning these abstractions and understanding them at an integrative level.
+This places a large cognitive burden on the user, especially for non-programmers.
+To reduce this burden we have created a grammar of genomic data transformation that operates on a single, central Bioconductor data structure, GRanges, which naturally represents genomic intervals and their associated measurements.
 The grammar defines verbs for performing actions on and between genomic interval data through a simplified, coherent interface to existing Bioconductor infrastructure, resulting in fluent analysis workflows.
 We have implemented this grammar as an R/Bioconductor package called \\texttt{plyranges}."
-
 keywords: "Bioconductor, Grammar, Genomes, Data Analysis" 
@@
 
+We also note that the grammar elements and design principles we have described
+are programming language agnostic and could be easily be implemented in another
+language where genomic information could be represented as a tabular data
+structure. We chose R because it is what we are familiar with and because
+the aforementioned Bioconductor packages have implemented the GRanges data
+structure.
+
 We aim to continue developing the `plyranges` package and to extend it

### 2018-07-17 paper.Rmd
fix link; render paper

@@
   download.file(
-    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GPL18952&format=file&file=GPL18952%5FHumanOmni2%2D5%2D8%2Dv1%2D1%2DC%2Ecsv%2Egz",
+    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GPL18952&format=file&file=GPL18952%5FHumanOmni25M%2D8v1%2D1%5FB%2Eannotated%2Etxt%2Egz",
     destfile = array_info

### 2018-09-12 paper.Rmd
typo and spelling fixes

--- 2018-07-17 paper.Rmd
@@
 
-abstract: "The Bioconductor project provides many interoperable data abstractions for analysing high-throughput genomics experiments, however implementing a typical genomic workflow with Bioconductor requires learning these abstractions and understanding them at an integrative level.
+abstract: "The Bioconductor project provides a host of interoperable data abstractions for analysing high-throughput genomics experiments.
+Implementing a typical genomic workflow with Bioconductor requires learning the abstractions and understanding them at an integrative level.
 This places a large cognitive burden on the user.
@@
    & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
-   Aggregation & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
+   Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
    &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
@@
    & \textbf{\emph{select()}} & select columns \\
-  Arithmetic (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
+  Modify (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
    & \emph{stretch()} & extend range by fixed amount \\
@@
    & \emph{compute\_coverage} & coverage over all ranges \\
-  Arithmetic (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
+  Modify (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
    & \emph{between()} & row-wise gap range \\
@@
     & \emph{join\_follow} & merge by following ranges \\
-    Merging & \emph{join\_precedes} & merge by preceding ranges \\
+    Merge & \emph{join\_precedes} & merge by preceding ranges \\
     & \emph{union\_ranges} & range-wise union \\
@@
    & \emph{anchor\_direction()} & fix coordinates at direction \\
-  Modifier & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
+  Operate & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
    & \emph{group\_by\_overlaps()} & partition by overlaps \\
@@
    & \textbf{\emph{filter()}} & subset rows \\
-  Restriction & \emph{filter\_by\_overlaps()} & subset by overlap \\
+  Restrict & \emph{filter\_by\_overlaps()} & subset by overlap \\
     & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
@@
 \caption{Overview of the \texttt{plyranges} grammar. The core verbs are
-briefly described and categorised into one of: aggregation, unary or binary
-arithmetic, merging, modifier, or restriction. A verb is given bold text if
+briefly described and categorised into one of the following higher level 
+catgeories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
 its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
@@
 To support merging, our algebra recasts finding overlaps or 
-nearest neighbours between two
+nearest neighbors between two
 genomic regions as variants of the relational join operator. A join
@@
 The join semantics alter what is returned: for an \\textbf{inner} join the query range is returned for each match, for an \\textbf{intersect} the intersection is taken between overlapping ranges, and for a \\textbf{left} join all query ranges are returned even if the subject range does not overlap them.
-This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
+This principle is generally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
 knitr::include_graphics("diagrams/diagrams-002.png")
@@
 enables easy access to public datasets and methods for analysis and 
-visualisation.
+visualization.
 
@@
 
-Returning to the GRanges object containing normalised coverage scores, 
+Returning to the GRanges object containing normalized coverage scores, 
 we filter to find the coordinates of the peak containing the maximum coverage 
@@
 distinction explicit in its function naming, instead relying on a
-parameter that defaults to strand sensitivity, an arguably unintuitive
+parameter that defaults to strand sensitivity, an arguably confusing
 behavior.

### 2018-09-12 paper.tex

@@
 
-\noindent The Bioconductor project has created many useful data abstractions for
-analysing high-throughput genomics experiments. However, there is a
-cognitive load placed on a user in learning a data abstraction and
-understanding its appropriate use. Throughout a standard workflow, a
-user must navigate and know many of these abstractions to perform an
-genomic analysis task, when a single data abstraction, a GRanges object
-will suffice. The GRanges class naturally represent genomic intervals
-and their associated measurements. By recognising that the GRanges class
-follows `tidy' data principles we have created a grammar of genomic data
-transformation. The grammar defines verbs for performing actions on and
-between genomic interval data. It provides a principled way of
-performing common genomic data analysis tasks through a coherent
-interface to existing Bioconductor infrastructure, resulting in human
-readable analysis workflows. We have implemented this grammar as a
-Bioconductor/R package called \texttt{plyranges}.
+\noindent The Bioconductor project provides many interoperable data abstractions
+for analysing high-throughput genomics experiments; however implementing
+a typical genomic workflow with Bioconductor requires learning these
+abstractions and understanding them at an integrative level. This places
+a large cognitive burden on the user, especially for non-programmers. To
+reduce this burden we have created a grammar of genomic data
+transformation that operates on a single, central Bioconductor data
+structure, GRanges, which naturally represents genomic intervals and
+their associated measurements. The grammar defines verbs for performing
+actions on and between genomic interval data through a simplified,
+coherent interface to existing Bioconductor infrastructure, resulting in
+fluent analysis workflows. We have implemented this grammar as an
+R/Bioconductor package called \texttt{plyranges}.
 
@@
 standard formats like BED. An example of the former is the Genome Query
-Language (GQL) and its distributed implementation GenAp which use an
+Language (GQL) and its distributed implementation GenAp which use a
 SQL-like syntax for fast retrieval of information of unprocessed
@@
 
+We also note that the grammar elements and design principles we have
+described are programming language agnostic and could be easily be
+implemented in another language where genomic information could be
+represented as a tabular data structure. We chose R because it is what
+we are familiar with and because the aforementioned Bioconductor
+packages have implemented the GRanges data structure.
+
 We aim to continue developing the \texttt{plyranges} package and to

### 2018-09-18 paper.Rmd
avoid relational algebra before it is defined
it's Codd not Cobb

@@
 sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the Genometric 
-Query Language (GMQL) implements a relational algebra for combining genomic datasets 
+Query Language (GMQL) implements a DSL for combining genomic datasets 
 \cite{Kaitoua2017-pw}.  The command line application BEDtools develops an extensive 
@@
 the left range. Finally, the overlap _left_ join is akin to left outer
-join in Cobb's relational algebra: it performs an overlap inner join
+join in Codd's relational algebra: it performs an overlap inner join
 but also returns all query ranges that are not hit by the subject.

### 2018-09-18 paper.Rmd
avoid using relational algebra before it is defined.

@@
 and the variables measured. The `dplyr` package \cite{Wickham2017-dplyr} 
-defines an API that maps notions from the general relational algebra to 
-operations on tidy data. It expresses each operation as a cohesive, endomorphic
-verb. Taken together these features enable a user to write human
-readable analysis workflows.
+defines an application programming interface (API) that maps notions from 
+database languages such as SQL to verbs that act on tidy data. These verbs
+can be composed together on one or more tidy datasets with the pipe operator 
+from the `magrittr` package \cite{R-magrittr}. Taken together these features 
+enable a user to write human readable analysis workflows.
 

### 2018-09-18 paper.Rmd
go back to what we had :P

@@
 defines an application programming interface (API) that maps notions from 
-database languages such as SQL to verbs that act on tidy data. These verbs
+the general relational algebra to verbs that act on tidy data. These verbs
 can be composed together on one or more tidy datasets with the pipe operator 
@@
 The `plyranges` DSL is built on the core Bioconductor data structure
-GRanges, which is essentially a constrained table, with fixed columns
+GRanges, which is a constrained table, with fixed columns
 for the chromosome, start and end coordinates, and the strand, along

### 2018-09-18 paper.Rmd
simplify language

@@
 
-GRanges follows the intuitive tidy data pattern: it is a rectangular
+GRanges follow the intuitive tidy data pattern: it is a rectangular
 table corresponding to a single biological context. Each row contains
@@
 remains compatible with the general relational operations defined by
-dplyr. Thus, we define our algebra as an extension of the `dplyr`
+`dplyr`. Thus, we define our algebra as an extension of the `dplyr`
 algebra, and borrow its syntax conventions and design principles.
@@
 genomic regions as variants of the relational join operator. A join
-acts on two GRanges objects, a query and a subject. The join operator
-is relational in the sense that metadata from the query and subject
-ranges is retained in the joined range.  All join operators in the
+acts on two GRanges objects: _x_ and _y_. The join operator
+is relational in the sense that metadata from the _x_ and _y_
+ranges are retained in the joined range.  All join operators in the
 `plyranges` DSL generate a set of hits based on overlap or proximity
@@
 \ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further restrict the
-matching by whether the query is completely _within_ the subject, and
-adding the _directed_ suffix ensures that matching ranges have the
-same direction (strand).
+matching by whether _y_ is completely _within_ the _x_, and ensure that 
+matching ranges have the same direction (strand) using the _directed_ suffix.
 

### 2018-09-18 paper.Rmd
reorder of design, added paragraph about memory usage;  extended note about summarized experiment

@@
 
-## Genomic Relation Algebra
+## Genomic Relational Algebra
 
@@
 join in Codd's relational algebra: it performs an overlap inner join
-but also returns all query ranges that are not hit by the subject.
+but also returns all _x_ ranges that are not hit by the _y_ ranges.
 
@@
 
-Like `dplyr`, `plyranges` verbs are functional: they are free of side
-effects and return their result. This enables chaining of verbs
-through syntax like the forward pipe operator (`%>%`, read as "then")
-of the `magrittr` package \cite{R-magrittr}. This has syntax has a direct
-cognitive mapping to natural language and the intuitive notion of
-pipelines. The low-level object-oriented APIs of Bioconductor tend to
-manipulate data via sub-replacement functions, like `start(gr) <-
-x`. These ultimately produce the side effect of replacing a symbol
-mapping in the current environment and thus are not amenable to
-so-called fluent syntax.
-
-A function is cohesive if it performs a singular task without
+To further enable cognitive consistency, `plyranges` functions are cohesive. 
+A function is defined to be cohesive if it performs a singular task without
 producing any side-effects. Singular tasks can always be broken 
@@
 a function's arguments. A flat function catalog also enhances API
-discoverability, particularly through auto-completion in IDEs. One
-downside of pushing cohesion to this extreme is that function calls
-become coupled, and care is necessary to treat them as a group when
-modifying code.
+discoverability, particularly through auto-completion in 
+integrated developer environments (IDEs). One downside of pushing cohesion to 
+this extreme is that function calls become coupled, and care is 
+necessary to treat them as a group when modifying code.
+
+Like `dplyr`, `plyranges` verbs are functional: they are free of side
+effects and are generally endomorphic, meaning that when the input is a GRanges
+object they return a GRanges object. This enables chaining of verbs
+through syntax like the forward pipe operator from the `magrittr` package. 
+This syntax has a direct cognitive mapping to natural language and the intuitive notion of
+pipelines. The low-level object-oriented APIs of Bioconductor tend to
+manipulate data via sub-replacement functions, like `start(gr) <-
+x`. These ultimately produce the side effect of replacing a symbol
+mapping in the current environment and thus are not amenable to so-called 
+fluent syntax.
 
@@
 behavior.
+
+The implementation of `plyranges` is built on top of 
+Bioconductor infrastructure, meaning most functions are constructed
+by composing generic functions from core Bioconductor packages. As a result, any 
+Bioconductor packages that uses data structures that inherit from GRanges will
+be able to use `plyranges`  for free. Another consequence of building on top of
+Bioconductor generics is that the speed and memory usage of `plyranges` 
+functions are similar to the highly optimised methods implemented in Bioconductor
+for GRanges objects.
 
@@
 representing experimental results (e.g., counts) from multiple sample
-experiments in conjunction with feature and sample metadata. The
-grammar and design of the `plryanges` DSL are naturally extensible to
-SummarizedExperiment.
+experiments in conjunction with feature and sample metadata. Although,
+the SummarizedExperiment is not strictly tidy, it does consist of 
+three tidy data structures that are related by feature and sample identifiers. 
+Thererfore, the grammar and design of the `plryanges` DSL is naturally 
+extensible to the SummarizedExperiment.
 

### 2018-09-19 paper.Rmd
change of caption text

--- 2018-09-18 paper.Rmd
@@
 sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the Genometric 
-Query Language (GMQL) implements a DSL for combining genomic datasets 
+Query Language (GMQL) implements a relational algebra for combining genomic datasets 
 \cite{Kaitoua2017-pw}.  The command line application BEDtools develops an extensive 
@@
 and the variables measured. The `dplyr` package \cite{Wickham2017-dplyr} 
-defines an application programming interface (API) that maps notions from 
-the general relational algebra to verbs that act on tidy data. These verbs
-can be composed together on one or more tidy datasets with the pipe operator 
-from the `magrittr` package \cite{R-magrittr}. Taken together these features 
-enable a user to write human readable analysis workflows.
+defines an API that maps notions from the general relational algebra to 
+operations on tidy data. It expresses each operation as a cohesive, endomorphic
+verb. Taken together these features enable a user to write human
+readable analysis workflows.
 
@@
 The `plyranges` DSL is built on the core Bioconductor data structure
-GRanges, which is a constrained table, with fixed columns
+GRanges, which is essentially a constrained table, with fixed columns
 for the chromosome, start and end coordinates, and the strand, along
@@
 ways. There are four supported matching algorithms: _overlap_,
-_nearest_, _precede_, and _follow_ (figures 
-\ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further restrict the
+_nearest_, _precede_, and _follow_ (figure 
+\ref{fig:olaps-fig}). We can further restrict the
 matching by whether the query is completely _within_ the subject, and
@@
 ```{r olaps-fig, echo = FALSE, out.width="\\textwidth", fig.cap="Illustration of the three overlap join operators.
-Each join takes query and subject range as input (black and light gray rectangles, respectively).
-An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
-This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
-The join semantics alter what is returned: for an \\textbf{inner} join the query range is returned for each match, for an \\textbf{intersect} the intersection is taken between overlapping ranges, and for a \\textbf{left} join all query ranges are returned even if the subject range does not overlap them.
-This principle is generally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
-knitr::include_graphics("diagrams/diagrams-002.png")
+Each join takes two GRanges objects, \\textit{x} and \\textit{y} as input.
+An index object for the join is computed which consists of two components.
+The first component contains the indices of the ranges in \\textit{x} that have been overlapped (the rectangles of \\textit{x} that cross the orange lines).
+The second component consists of the indices of the ranges in \\textit{y} that overlap the ranges in \\textit{x}.
+In this case a range in \\textit{y} overlaps the ranges in \\textit{x} three times, so the index is repeated three times.
+This index is used to modify \\textit{x} by where it was 'hit' by \\textit{y}.
+This principle of is applied  The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \textit{x} ranges that are overlapped by \\textit{y} are returned.
+\\textbf{B}  For an \\textbf{intersect} join the intersection is taken between the overlapped \\textit{x} ranges and the \\textit{y} ranges.
+\\textbf{C} For the left join \\textbf{left} join all \\textit{x} ranges are returned regardless of whether they are overlapped by \\textit{y}."}
+knitr::include_graphics("diagrams/olap-fig.png")
 ```
@@
 the left range. Finally, the overlap _left_ join is akin to left outer
-join in Codd's relational algebra: it performs an overlap inner join
+join in Cobb's relational algebra: it performs an overlap inner join
 but also returns all query ranges that are not hit by the subject.
@@
 merging, these operations conform to the semantics and syntax of the 
-`dplyr` grammar.
+`dplyr` grammar. Consequently, this results
 

### 2018-09-19 paper.Rmd
fixing caption

@@
 Each join takes two GRanges objects, \\textit{x} and \\textit{y} as input.
-An index object for the join is computed which consists of two components.
+A 'Hits' object for the join is computed which consists of two components.
 The first component contains the indices of the ranges in \\textit{x} that have been overlapped (the rectangles of \\textit{x} that cross the orange lines).
@@
 In this case a range in \\textit{y} overlaps the ranges in \\textit{x} three times, so the index is repeated three times.
-This index is used to modify \\textit{x} by where it was 'hit' by \\textit{y}.
-This principle of is applied  The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \textit{x} ranges that are overlapped by \\textit{y} are returned.
+Te resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by \\textit{y}.
+This principle of is applied generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour operations.
+The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \textit{x} ranges that are overlapped by \\textit{y} are returned.
 \\textbf{B}  For an \\textbf{intersect} join the intersection is taken between the overlapped \\textit{x} ranges and the \\textit{y} ranges.

### 2018-09-20 paper.Rmd

--- 2018-09-18 paper.Rmd
@@
 ways. There are four supported matching algorithms: _overlap_,
-_nearest_, _precede_, and _follow_ (figures 
-\ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further restrict the
-matching by whether _y_ is completely _within_ the _x_, and ensure that 
-matching ranges have the same direction (strand) using the _directed_ suffix.
+_nearest_, _precede_, and _follow_ (figure 
+\ref{fig:olaps-fig}). We can further restrict the
+matching by whether the query is completely _within_ the subject, and
+adding the _directed_ suffix ensures that matching ranges have the
+same direction (strand).
 
 ```{r olaps-fig, echo = FALSE, out.width="\\textwidth", fig.cap="Illustration of the three overlap join operators.
-Each join takes query and subject range as input (black and light gray rectangles, respectively).
-An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
-This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
-The join semantics alter what is returned: for an \\textbf{inner} join the query range is returned for each match, for an \\textbf{intersect} the intersection is taken between overlapping ranges, and for a \\textbf{left} join all query ranges are returned even if the subject range does not overlap them.
-This principle is generally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
-knitr::include_graphics("diagrams/diagrams-002.png")
+Each join takes two GRanges objects, \\textit{x} and \\textit{y} as input.
+A 'Hits' object for the join is computed which consists of two components.
+The first component contains the indices of the ranges in \\textit{x} that have been overlapped (the rectangles of \\textit{x} that cross the orange lines).
+The second component consists of the indices of the ranges in \\textit{y} that overlap the ranges in \\textit{x}.
+In this case a range in \\textit{y} overlaps the ranges in \\textit{x} three times, so the index is repeated three times.
+Te resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by \\textit{y}.
+This principle of is applied generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour operations.
+The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \textit{x} ranges that are overlapped by \\textit{y} are returned.
+\\textbf{B}  For an \\textbf{intersect} join the intersection is taken between the overlapped \\textit{x} ranges and the \\textit{y} ranges.
+\\textbf{C} For the left join \\textbf{left} join all \\textit{x} ranges are returned regardless of whether they are overlapped by \\textit{y}."}
+knitr::include_graphics("diagrams/olap-fig.png")
 ```

### 2018-09-20 paper.Rmd
further clarification of joins

@@
 In this case a range in \\textit{y} overlaps the ranges in \\textit{x} three times, so the index is repeated three times.
-Te resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by \\textit{y}.
-This principle of is applied generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour operations.
-The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \textit{x} ranges that are overlapped by \\textit{y} are returned.
-\\textbf{B}  For an \\textbf{intersect} join the intersection is taken between the overlapped \\textit{x} ranges and the \\textit{y} ranges.
-\\textbf{C} For the left join \\textbf{left} join all \\textit{x} ranges are returned regardless of whether they are overlapped by \\textit{y}."}
+The resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by \\textit{y} and merge all metadata columns from \\textit{x} and \\textit{y} based on the indices contained in the 'Hits' object.
+This procedure is applied generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour operations.
+The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \\textit{x} ranges that are overlapped by \\textit{y} are returned.
+The returned ranges also include the metadata from the \\textit{y] range that overlapped the three \\textit{x} ranges.
+\\textbf{B} An \\textbf{intersect} join is identical to an inner join except that the intersection is taken between the overlapped \\textit{x} ranges and the \\textit{y} ranges.
+\\textbf{C} For the left join \\textbf{left} join all \\textit{x} ranges are returned regardless of whether they are overlapped by \\textit{y}.
+In this case the resulting third range  of the join would have missing values on metadata columns that came from \\textit{y}."}
 knitr::include_graphics("diagrams/olap-fig.png")

### 2018-09-20 paper.Rmd
use latex instead of rmd for figure, output data and code.

--- 2018-09-20 paper.Rmd
@@
 
+\begin{figure}
+\includegraphics[width=\textwidth]{diagrams/olap-fig}
+\caption{Illustration of the three overlap join operators. Each join takes two 
+GRanges objects, \\textit{x} and \\textit{y} as input. A 'Hits' object for the 
+join is computed which consists of two components. The first component contains 
+the indices of the ranges in \\textit{x} that have been overlapped 
+(the rectangles of \\textit{x} that cross the orange lines). The second 
+component consists of the indices of the ranges in \\textit{y} that overlap 
+the ranges in \\textit{x}. In this case a range in \\textit{y} overlaps the 
+ranges in \\textit{x} three times, so the index is repeated three times. The 
+resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by
+\\textit{y} and merge all metadata columns from \\textit{x} and \\textit{y} 
+based on the indices contained in the 'Hits' object. This procedure is applied 
+generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour
+operations. The join semantics alter what is returned: \\textbf{A}: for an 
+\\textbf{inner} join the \\textit{x} ranges that are overlapped by \\textit{y} 
+are returned. The returned ranges also include the metadata from the \\textit{y}
+range that overlapped the three \\textit{x} ranges. \\textbf{B} An 
+\\textbf{intersect} join is identical to an inner join except that the 
+intersection is taken between the overlapped \\textit{x} ranges and the 
+\\textit{y} ranges. \\textbf{C} For the left join \\textbf{left} join all 
+\\textit{x} ranges are returned regardless of whether they are overlapped by 
+\\textit{y}. In this case the resulting third range  of the join would have 
+missing values on metadata columns that came from \\textit{y}.}
+\label{fig:olaps-fig}
+\end{figure}
+
 To support merging, our algebra recasts finding overlaps or 
@@
 same direction (strand).
-
-```{r olaps-fig, echo = FALSE, out.width="\\textwidth", fig.cap="Illustration of the three overlap join operators.
-Each join takes two GRanges objects, \\textit{x} and \\textit{y} as input.
-A 'Hits' object for the join is computed which consists of two components.
-The first component contains the indices of the ranges in \\textit{x} that have been overlapped (the rectangles of \\textit{x} that cross the orange lines).
-The second component consists of the indices of the ranges in \\textit{y} that overlap the ranges in \\textit{x}.
-In this case a range in \\textit{y} overlaps the ranges in \\textit{x} three times, so the index is repeated three times.
-Te resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by \\textit{y}.
-This principle of is applied generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour operations.
-The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \textit{x} ranges that are overlapped by \\textit{y} are returned.
-\\textbf{B}  For an \\textbf{intersect} join the intersection is taken between the overlapped \\textit{x} ranges and the \\textit{y} ranges.
-\\textbf{C} For the left join \\textbf{left} join all \\textit{x} ranges are returned regardless of whether they are overlapped by \\textit{y}."}
-knitr::include_graphics("diagrams/olap-fig.png")
-```
 
@@
 merging, these operations conform to the semantics and syntax of the 
-`dplyr` grammar. Consequently, this results
+`dplyr` grammar. Consequently, `plyranges` code is generally more compact
+than the equivalent `GenomicRanges` code (figure \ref{code-fig}).
 
@@
 Here we provide illustrative examples of using the `plyranges` DSL to show
-how our grammar could be integrated into genomic data workflows. We also
+how our grammar could be integrated into genomic data workflows. As we
+construct the workflows we show the data output intermittantly to assist
+the reader in understanding the pipeline steps. The workflows
 highlight how interoperability with existing Bioconductor infrastructure, 
@@
   set_genome_info(genome = "hg19")
+chr10_scores
 ```
@@
   reduce_ranges(score = max(score))
+all_peaks
 ```
@@
   join_overlap_inner(chr10_max_score_region)
-
+peak_region
 ```
@@
   mutate(score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE)))
+alignments
 ```

### 2018-09-20 paper.Rmd
code comparison and latex fixes

@@
 \caption{Illustration of the three overlap join operators. Each join takes two 
-GRanges objects, \\textit{x} and \\textit{y} as input. A 'Hits' object for the 
+GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the 
 join is computed which consists of two components. The first component contains 
-the indices of the ranges in \\textit{x} that have been overlapped 
-(the rectangles of \\textit{x} that cross the orange lines). The second 
-component consists of the indices of the ranges in \\textit{y} that overlap 
-the ranges in \\textit{x}. In this case a range in \\textit{y} overlaps the 
-ranges in \\textit{x} three times, so the index is repeated three times. The 
-resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by
-\\textit{y} and merge all metadata columns from \\textit{x} and \\textit{y} 
+the indices of the ranges in \textit{x} that have been overlapped 
+(the rectangles of \textit{x} that cross the orange lines). The second 
+component consists of the indices of the ranges in \textit{y} that overlap 
+the ranges in \textit{x}. In this case a range in \textit{y} overlaps the 
+ranges in \textit{x} three times, so the index is repeated three times. The 
+resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by
+\textit{y} and merge all metadata columns from \textit{x} and \textit{y} 
 based on the indices contained in the 'Hits' object. This procedure is applied 
-generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour
-operations. The join semantics alter what is returned: \\textbf{A}: for an 
-\\textbf{inner} join the \\textit{x} ranges that are overlapped by \\textit{y} 
-are returned. The returned ranges also include the metadata from the \\textit{y}
-range that overlapped the three \\textit{x} ranges. \\textbf{B} An 
-\\textbf{intersect} join is identical to an inner join except that the 
-intersection is taken between the overlapped \\textit{x} ranges and the 
-\\textit{y} ranges. \\textbf{C} For the left join \\textbf{left} join all 
-\\textit{x} ranges are returned regardless of whether they are overlapped by 
-\\textit{y}. In this case the resulting third range  of the join would have 
-missing values on metadata columns that came from \\textit{y}.}
+generally in the \texttt{plyranges} DSL for both overlap and nearest neighbour
+operations. The join semantics alter what is returned: \textbf{A}: for an 
+\textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} 
+are returned. The returned ranges also include the metadata from the \textit{y}
+range that overlapped the three \textit{x} ranges. \textbf{B} An 
+\textbf{intersect} join is identical to an inner join except that the 
+intersection is taken between the overlapped \textit{x} ranges and the 
+\textit{y} ranges. \\textbf{C} For the left join \textbf{left} join all 
+\textit{x} ranges are returned regardless of whether they are overlapped by 
+\textit{y}. In this case the resulting third range  of the join would have 
+missing values on metadata columns that came from \textit{y}.}
 \label{fig:olaps-fig}
@@
 
+\begin{figure}
+\includegraphics[width=\textwidth]{diagrams/code-comparison.pdf}
+\caption{Idiomatic code examples for \texttt{plyranges} (\textbf(A)) and 
+\texttt{GenomicRanges} (\textbf(B)) illustrating an overlap and aggregate 
+operation that returns the same result. 
+In each example, we have two BED files consisting of SNPs that are
+genome-wide association study (GWAS) hits and reference exons. Each code block
+counts for each SNP the number of distinct exons it overlaps. The 
+\texttt(plyranges) code achieves this with an overlap join followed by partitioning
+and aggregation. Strand is ignored by default here. 
+The \texttt(GenomicRanges) code achieves this using the 'Hits' and 'List' 
+classes and their methods.}
+\label{fig:code}
+\end{figure}
+
 ```{r nn-fig, echo = FALSE, out.width = "\\textwidth", fig.cap="Illustration of neighbour finding joins.
@@
 `dplyr` grammar. Consequently, `plyranges` code is generally more compact
-than the equivalent `GenomicRanges` code (figure \ref{code-fig}).
+than the equivalent `GenomicRanges` code (figure \ref{fig:code}).
 

### 2018-09-20 paper.Rmd
example speed ups based on new plyranges

@@
 intersection is taken between the overlapped \textit{x} ranges and the 
-\textit{y} ranges. \\textbf{C} For the left join \textbf{left} join all 
+\textit{y} ranges. \textbf{C} For the left join \textbf{left} join all 
 \textit{x} ranges are returned regardless of whether they are overlapped by 
@@
 \includegraphics[width=\textwidth]{diagrams/code-comparison.pdf}
-\caption{Idiomatic code examples for \texttt{plyranges} (\textbf(A)) and 
-\texttt{GenomicRanges} (\textbf(B)) illustrating an overlap and aggregate 
+\caption{Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and 
+\texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate 
 operation that returns the same result. 
@@
 counts for each SNP the number of distinct exons it overlaps. The 
-\texttt(plyranges) code achieves this with an overlap join followed by partitioning
+\texttt{plyranges}code achieves this with an overlap join followed by partitioning
 and aggregation. Strand is ignored by default here. 
-The \texttt(GenomicRanges) code achieves this using the 'Hits' and 'List' 
+The \texttt{GenomicRanges} code achieves this using the 'Hits' and 'List' 
 classes and their methods.}
@@
 \end{figure}
-
-```{r nn-fig, echo = FALSE, out.width = "\\textwidth", fig.cap="Illustration of neighbour finding joins.
-Each join takes a query and subject range and computes a 'Hits' object.
-For the \\textbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range.
-For the \\textbf{follow} join there is only one query range that follows any subject range.
-Likewise for the \\textbf{precede} join, there is only one query range that precedes a subject range."}
-knitr::include_graphics("diagrams/diagrams-003.png")
-```
 
@@
 BAM file: in this case the reads were aligned to the hg19 genome build
-and there are no reads overlapping the mitochondrial genome. To generate 
-bins of fixed width 10000nt we apply the `tile_ranges()` function
-to the genomic coordinates in locations.
+and there are no reads overlapping the mitochondrial genome.
+
 ```{r bins}
@@
   get_genome_info() 
-
-bins <- locations %>% 
-  tile_ranges(width = 10000L)
 ```
@@
 for each alignment using the `letterFrequency()` function from the `Biostrings`
-package \cite{R-biostrings}.
+package \cite{R-biostrings}. After computing the GC proportion as the score
+column, we drop  all other columns in the GRanges object.
 
@@
   select(seq) %>%
-  mutate(score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE)))
+  mutate(score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE))) %>% 
+  select(score)
 alignments
@@
 
-Finally, we use `group_by_overlaps()` to see where the alignments overlap the 
-genomic windows and then apply `summarize()` to compute the
-total number of reads and average GC content within each window.
+Finally, we create 10000nt tiles over the genome and compute the number
+of reads and average GC content over all reads that fall within
+each tile using an overlap join and merging endpoints.
 
 ```{r, cache = TRUE}
+bins <- locations %>% 
+  tile_ranges(width = 10000L)
+
 alignments_summary <- bins %>%
-  group_by_overlaps(alignments %>% select(score)) %>% 
-  summarize(n_reads = n(), avg_gc = mean(score))
+  join_overlap_inner(alignments) %>% 
+  disjoin_ranges(n = n(), avg_gc = mean(score))
+alignments_summary
 ```
@@
 To produce the workflows as described in results section we used 
-R version 3.5 with the development version of `plyranges` (1.1.4) 
+R version 3.5 with the development version of `plyranges` (1.1.10) 
 and the  `BioStrings` package (2.48.0) installed.

### 2018-09-20 paper.Rmd
typos

@@
 
-After filtering for regions with a coverage score greater than 8, we can
-reduce individual runs to ranges representing the islands of coverage by
-using the `reduce_ranges()` function. This function allows a summary
-to be computer over each island: in this case we take the maximum of 
+We then filter for regions with a coverage score greater than 8, and following
+this reduce individual runs to ranges representing the islands of coverage.
+This is achieved with the `reduce_ranges()` function, which allows a summary
+to be computed over each island: in this case we take the maximum of 
 the scores to find the coverage peaks over chromosome 10.
@@
   select(seq) %>%
-  mutate(score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE))) %>% 
+  mutate(
+    score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE))
+  ) %>% 
   select(score)
@@
             ti_tv = sum(transition) / sum(!transition)) 
+ti_tv_results
 ```

### 2018-09-20 paper.Rmd
spelling and typo checks

@@
     email: lawrence.michael@gene.com
-abstract: "The Bioconductor project provides many interoperable data abstractions for analysing high-throughput genomics experiments; however implementing a typical genomic workflow with Bioconductor requires learning these abstractions and understanding them at an integrative level.
+abstract: "The Bioconductor project provides many interoperable data abstractions for analyzing high-throughput genomics experiments; however implementing a typical genomic workflow with Bioconductor requires learning these abstractions and understanding them at an integrative level.
 This places a large cognitive burden on the user, especially for non-programmers.
@@
 [@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
-analysing genomics data with R, an interactive data analysis environment that 
+analyzing genomics data with R, an interactive data analysis environment that 
 encourages reproducibility and provides high-level abstractions for manipulating,
@@
   \hline
-   & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
+   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
    Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
-   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
+   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
    \hline
@@
     & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
-    & \emph{join\_nearest} & merge by nearest neighbour ranges \\
+    & \emph{join\_nearest} & merge by nearest neighbor ranges \\
     & \emph{join\_follow} & merge by following ranges \\
@@
 \caption{Overview of the \texttt{plyranges} grammar. The core verbs are
-briefly described and categorised into one of the following higher level 
+briefly described and categorized into one of the following higher level 
 catgeories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
@@
 based on the indices contained in the 'Hits' object. This procedure is applied 
-generally in the \texttt{plyranges} DSL for both overlap and nearest neighbour
+generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor
 operations. The join semantics alter what is returned: \textbf{A}: for an 
@@
 intersection is taken between the overlapped \textit{x} ranges and the 
-\textit{y} ranges. \textbf{C} For the left join \textbf{left} join all 
+\textit{y} ranges. \textbf{C} For the \textbf{left} join all 
 \textit{x} ranges are returned regardless of whether they are overlapped by 
-\textit{y}. In this case the resulting third range  of the join would have 
-missing values on metadata columns that came from \textit{y}.}
+\textit{y}. In this case the third range (rectangle with the asterisk next to it)
+of the join would have missing values on metadata columns that came from \textit{y}.}
 \label{fig:olaps-fig}
@@
 how our grammar could be integrated into genomic data workflows. As we
-construct the workflows we show the data output intermittantly to assist
+construct the workflows we show the data output intermittently to assist
 the reader in understanding the pipeline steps. The workflows
@@
 Bioconductor generics is that the speed and memory usage of `plyranges` 
-functions are similar to the highly optimised methods implemented in Bioconductor
+functions are similar to the highly optimized methods implemented in Bioconductor
 for GRanges objects.
@@
 three tidy data structures that are related by feature and sample identifiers. 
-Thererfore, the grammar and design of the `plryanges` DSL is naturally 
+Therefore, the grammar and design of the `plryanges` DSL is naturally 
 extensible to the SummarizedExperiment.

### 2018-09-20 paper.tex
new pdf

@@
 \noindent The Bioconductor project provides many interoperable data abstractions
-for analysing high-throughput genomics experiments; however implementing
+for analyzing high-throughput genomics experiments; however implementing
 a typical genomic workflow with Bioconductor requires learning these
@@
 sequencing data \autocites{Kozanitis2014-va}{Kozanitis2016-bm}.
-Similarly, the Genometric Query Language (GMQL) implements a relational
-algebra for combining genomic datasets \cite{Kaitoua2017-pw}. The
-command line application BEDtools develops an extensive algebra for
-performing arithmetic between two or more sets of genomic regions
+Similarly, the Genometric Query Language (GMQL) implements a DSL for
+combining genomic datasets \cite{Kaitoua2017-pw}. The command line
+application BEDtools develops an extensive algebra for performing
+arithmetic between two or more sets of genomic regions
 \cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be
@@
 \autocites{r-core}{Lawrence2013-wg}{Huber2015-ei} define a DSL for
-analysing genomics data with R, an interactive data analysis environment
+analyzing genomics data with R, an interactive data analysis environment
 that encourages reproducibility and provides high-level abstractions for
@@
 design of an experiment and the variables measured. The \texttt{dplyr}
-package \cite{Wickham2017-dplyr} defines an API that maps notions from
-the general relational algebra to operations on tidy data. It expresses
-each operation as a cohesive, endomorphic verb. Taken together these
-features enable a user to write human readable analysis workflows.
+package \cite{Wickham2017-dplyr} defines an application programming
+interface (API) that maps notions from the general relational algebra to
+verbs that act on tidy data. These verbs can be composed together on one
+or more tidy datasets with the pipe operator from the \texttt{magrittr}
+package \cite{R-magrittr}. Taken together these features enable a user
+to write human readable analysis workflows.
 
@@
 
-\hypertarget{genomic-relation-algebra}{%
-\subsection{Genomic Relation Algebra}\label{genomic-relation-algebra}}
+\hypertarget{genomic-relational-algebra}{%
+\subsection{Genomic Relational
+Algebra}\label{genomic-relational-algebra}}
 
@@
 The \texttt{plyranges} DSL is built on the core Bioconductor data
-structure GRanges, which is essentially a constrained table, with fixed
-columns for the chromosome, start and end coordinates, and the strand,
-along with an arbitrary set of additional columns, consisting of
-measurements or metadata specific to the data type or experiment (figure
+structure GRanges, which is a constrained table, with fixed columns for
+the chromosome, start and end coordinates, and the strand, along with an
+arbitrary set of additional columns, consisting of measurements or
+metadata specific to the data type or experiment (figure
 \ref{fig:GRanges}). GRanges balances flexibility with formal
@@
 
-GRanges follows the intuitive tidy data pattern: it is a rectangular
+GRanges follow the intuitive tidy data pattern: it is a rectangular
 table corresponding to a single biological context. Each row contains a
@@
 remains compatible with the general relational operations defined by
-dplyr. Thus, we define our algebra as an extension of the \texttt{dplyr}
-algebra, and borrow its syntax conventions and design principles.
+\texttt{dplyr}. Thus, we define our algebra as an extension of the
+\texttt{dplyr} algebra, and borrow its syntax conventions and design
+principles.
 
@@
   \hline
-   & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
-   Aggregation & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
-   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
+   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
+   Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
+   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
    \hline
@@
     & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
-    & \emph{join\_nearest} & merge by nearest neighbour ranges \\
+    & \emph{join\_nearest} & merge by nearest neighbor ranges \\
     & \emph{join\_follow} & merge by following ranges \\
-    Merging & \emph{join\_precedes} & merge by preceding ranges \\
+    Merge & \emph{join\_precedes} & merge by preceding ranges \\
     & \emph{union\_ranges} & range-wise union \\
@@
 \caption{Overview of the \texttt{plyranges} grammar. The core verbs are
-briefly described and categorised into one of: aggregation, unary or binary
-arithmetic, merging, modifier, or restriction. A verb is given bold text if
+briefly described and categorized into one of the following higher level 
+catgeories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
 its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
@@
 
+\begin{figure}
+\includegraphics[width=\textwidth]{diagrams/olap-fig}
+\caption{Illustration of the three overlap join operators. Each join takes two 
+GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the 
+join is computed which consists of two components. The first component contains 
+the indices of the ranges in \textit{x} that have been overlapped 
+(the rectangles of \textit{x} that cross the orange lines). The second 
+component consists of the indices of the ranges in \textit{y} that overlap 
+the ranges in \textit{x}. In this case a range in \textit{y} overlaps the 
+ranges in \textit{x} three times, so the index is repeated three times. The 
+resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by
+\textit{y} and merge all metadata columns from \textit{x} and \textit{y} 
+based on the indices contained in the 'Hits' object. This procedure is applied 
+generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor
+operations. The join semantics alter what is returned: \textbf{A}: for an 
+\textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} 
+are returned. The returned ranges also include the metadata from the \textit{y}
+range that overlapped the three \textit{x} ranges. \textbf{B} An 
+\textbf{intersect} join is identical to an inner join except that the 
+intersection is taken between the overlapped \textit{x} ranges and the 
+\textit{y} ranges. \textbf{C} For the \textbf{left} join all 
+\textit{x} ranges are returned regardless of whether they are overlapped by 
+\textit{y}. In this case the third range (rectangle with the asterisk next to it)
+of the join would have missing values on metadata columns that came from \textit{y}.}
+\label{fig:olaps-fig}
+\end{figure}
+
 To support merging, our algebra recasts finding overlaps or nearest
-neighbours between two genomic regions as variants of the relational
-join operator. A join acts on two GRanges objects, a query and a
-subject. The join operator is relational in the sense that metadata from
-the query and subject ranges is retained in the joined range. All join
-operators in the \texttt{plyranges} DSL generate a set of hits based on
-overlap or proximity of ranges and use those hits to merge the two
-datasets in different ways. There are four supported matching
-algorithms: \emph{overlap}, \emph{nearest}, \emph{precede}, and
-\emph{follow} (figures \ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can
-further restrict the matching by whether the query is completely
-\emph{within} the subject, and adding the \emph{directed} suffix ensures
-that matching ranges have the same direction (strand).
-
-\begin{figure}
-
-{\centering \includegraphics[width=\textwidth]{diagrams/diagrams-002}
-
-}
-
-\caption{Illustration of the three overlap join operators.
-Each join takes query and subject range as input (black and light gray rectangles, respectively).
-An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range.
-This index is used to expand the query ranges by where it was 'hit' by the subject ranges.
-The join semantics alter what is returned: for an \textbf{inner} join the query range is returned for each match, for an \textbf{intersect} the intersection is taken between overlapping ranges, and for a \textbf{left} join all query ranges are returned even if the subject range does not overlap them.
-This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations.}\label{fig:olaps-fig}
-\end{figure}
+neighbors between two genomic regions as variants of the relational join
+operator. A join acts on two GRanges objects: \emph{x} and \emph{y}. The
+join operator is relational in the sense that metadata from the \emph{x}
+and \emph{y} ranges are retained in the joined range. All join operators
+in the \texttt{plyranges} DSL generate a set of hits based on overlap or
+proximity of ranges and use those hits to merge the two datasets in
+different ways. There are four supported matching algorithms:
+\emph{overlap}, \emph{nearest}, \emph{precede}, and \emph{follow}
+(figure \ref{fig:olaps-fig}). We can further restrict the matching by
+whether the query is completely \emph{within} the subject, and adding
+the \emph{directed} suffix ensures that matching ranges have the same
+direction (strand).
 
@@
 intersection instead of the left range. Finally, the overlap \emph{left}
-join is akin to left outer join in Cobb's relational algebra: it
-performs an overlap inner join but also returns all query ranges that
-are not hit by the subject.
+join is akin to left outer join in Codd's relational algebra: it
+performs an overlap inner join but also returns all \emph{x} ranges that
+are not hit by the \emph{y} ranges.
 
 \begin{figure}
-
-{\centering \includegraphics[width=\textwidth]{diagrams/diagrams-003}
-
-}
-
-\caption{Illustration of neighbour finding joins.
-Each join takes a query and subject range and computes a 'Hits' object.
-For the \textbf{nearest} join all query ranges are returned as they are all nearest neighbours of the second subject range.
-For the \textbf{follow} join there is only one query range that follows any subject range.
-Likewise for the \textbf{precede} join, there is only one query range that precedes a subject range.}\label{fig:nn-fig}
+\includegraphics[width=\textwidth]{diagrams/code-comparison.pdf}
+\caption{Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and 
+\texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate 
+operation that returns the same result. 
+In each example, we have two BED files consisting of SNPs that are
+genome-wide association study (GWAS) hits and reference exons. Each code block
+counts for each SNP the number of distinct exons it overlaps. The 
+\texttt{plyranges}code achieves this with an overlap join followed by partitioning
+and aggregation. Strand is ignored by default here. 
+The \texttt{GenomicRanges} code achieves this using the 'Hits' and 'List' 
+classes and their methods.}
+\label{fig:code}
 \end{figure}
@@
 and merging, these operations conform to the semantics and syntax of the
-\texttt{dplyr} grammar.
+\texttt{dplyr} grammar. Consequently, \texttt{plyranges} code is
+generally more compact than the equivalent \texttt{GenomicRanges} code
+(figure \ref{fig:code}).
 
@@
 DSL to show how our grammar could be integrated into genomic data
-workflows. We also highlight how interoperability with existing
-Bioconductor infrastructure, enables easy access to public datasets and
-methods for analysis and visualisation.
+workflows. As we construct the workflows we show the data output
+intermittently to assist the reader in understanding the pipeline steps.
+The workflows highlight how interoperability with existing Bioconductor
+infrastructure, enables easy access to public datasets and methods for
+analysis and visualization.
 
@@
 \StringTok{  }\KeywordTok{set_genome_info}\NormalTok{(}\DataTypeTok{genome =} \StringTok{"hg19"}\NormalTok{)}
+\NormalTok{chr10_scores}
 \end{Highlighting}
@@
 
-After filtering for regions with a coverage score greater than 8, we can
-reduce individual runs to ranges representing the islands of coverage by
-using the \texttt{reduce\_ranges()} function. This function allows a
-summary to be computer over each island: in this case we take the
-maximum of the scores to find the coverage peaks over chromosome 10.
+\begin{verbatim}
+#> GRanges object with 5789841 ranges and 1 metadata column:
+#>             seqnames              ranges strand |              score
+#>                <Rle>           <IRanges>  <Rle> |          <numeric>
+#>         [1]    chr10             1-60602      * | 0.0422799997031689
+#>         [2]    chr10         60603-60781      * |  0.163240000605583
+#>         [3]    chr10         60782-60816      * |  0.372139990329742
+#>         [4]    chr10         60817-60995      * |  0.163240000605583
+#>         [5]    chr10         60996-61625      * | 0.0422799997031689
+#>         ...      ...                 ...    ... .                ...
+#>   [5789837]    chr10 135524723-135524734      * |  0.144319996237755
+#>   [5789838]    chr10 135524735-135524775      * |  0.250230014324188
+#>   [5789839]    chr10 135524776-135524784      * |  0.427789986133575
+#>   [5789840]    chr10 135524785-135524806      * |  0.730019986629486
+#>   [5789841]    chr10 135524807-135524837      * |   1.03103005886078
+#>   -------
+#>   seqinfo: 25 sequences from hg19 genome
+\end{verbatim}
+
+We then filter for regions with a coverage score greater than 8, and
+following this reduce individual runs to ranges representing the islands
+of coverage. This is achieved with the \texttt{reduce\_ranges()}
+function, which allows a summary to be computed over each island: in
+this case we take the maximum of the scores to find the coverage peaks
+over chromosome 10.
 
@@
 \StringTok{  }\KeywordTok{reduce_ranges}\NormalTok{(}\DataTypeTok{score =} \KeywordTok{max}\NormalTok{(score))}
+\NormalTok{all_peaks}
 \end{Highlighting}
@@
 
-Returning to the GRanges object containing normalised coverage scores,
+\begin{verbatim}
+#> GRanges object with 1085 ranges and 1 metadata column:
+#>          seqnames              ranges strand |            score
+#>             <Rle>           <IRanges>  <Rle> |        <numeric>
+#>      [1]    chr10     1299144-1299370      * | 13.2264003753662
+#>      [2]    chr10     1778600-1778616      * | 8.20512008666992
+#>      [3]    chr10     4613068-4613078      * | 8.76027011871338
+#>      [4]    chr10     4613081-4613084      * | 8.43659973144531
+#>      [5]    chr10             4613086      * | 8.11507987976074
+#>      ...      ...                 ...    ... .              ...
+#>   [1081]    chr10 135344482-135344488      * | 9.23237991333008
+#>   [1082]    chr10 135344558-135344661      * |  11.843409538269
+#>   [1083]    chr10 135344663-135344665      * | 8.26965999603271
+#>   [1084]    chr10 135344670-135344674      * | 8.26965999603271
+#>   [1085]    chr10 135345440-135345441      * | 8.26965999603271
+#>   -------
+#>   seqinfo: 25 sequences from hg19 genome
+\end{verbatim}
+
+Returning to the GRanges object containing normalized coverage scores,
 we filter to find the coordinates of the peak containing the maximum
@@
 \StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(chr10_max_score_region)}
+\NormalTok{peak_region}
 \end{Highlighting}
 \end{Shaded}
+
+\begin{verbatim}
+#> GRanges object with 890 ranges and 2 metadata columns:
+#>         seqnames            ranges strand |            score.x
+#>            <Rle>         <IRanges>  <Rle> |          <numeric>
+#>     [1]    chr10 21805891-21805988      * | 0.0206599999219179
+#>     [2]    chr10 21805989-21806000      * | 0.0211200006306171
+#>     [3]    chr10 21806001-21806044      * |  0.022069999948144
+#>     [4]    chr10 21806045-21806049      * | 0.0215900000184774
+#>     [5]    chr10 21806050-21806081      * | 0.0211200006306171
+#>     ...      ...               ...    ... .                ...
+#>   [886]    chr10          21810878      * |   5.24951982498169
+#>   [887]    chr10          21810879      * |   5.83534002304077
+#>   [888]    chr10 21810880-21810884      * |   6.44267988204956
+#>   [889]    chr10 21810885-21810895      * |   7.07054996490479
+#>   [890]    chr10 21810896-21810911      * |   6.44267988204956
+#>                  score.y
+#>                <numeric>
+#>     [1] 29.9573001861572
+#>     [2] 29.9573001861572
+#>     [3] 29.9573001861572
+#>     [4] 29.9573001861572
+#>     [5] 29.9573001861572
+#>     ...              ...
+#>   [886] 29.9573001861572
+#>   [887] 29.9573001861572
+#>   [888] 29.9573001861572
+#>   [889] 29.9573001861572
+#>   [890] 29.9573001861572
+#>   -------
+#>   seqinfo: 25 sequences from hg19 genome
+\end{verbatim}
 
@@
 file: in this case the reads were aligned to the hg19 genome build and
-there are no reads overlapping the mitochondrial genome. To generate
-bins of fixed width 10000nt we apply the \texttt{tile\_ranges()}
-function to the genomic coordinates in locations.
+there are no reads overlapping the mitochondrial genome.
 
@@
 \StringTok{  }\KeywordTok{get_genome_info}\NormalTok{() }
-
-\NormalTok{bins <-}\StringTok{ }\NormalTok{locations }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{tile_ranges}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L)}
 \end{Highlighting}
@@
 proportion for each alignment using the \texttt{letterFrequency()}
-function from the \texttt{Biostrings} package \cite{R-biostrings}.
+function from the \texttt{Biostrings} package \cite{R-biostrings}. After
+computing the GC proportion as the score column, we drop all other
+columns in the GRanges object.
 
@@
 \StringTok{  }\KeywordTok{select}\NormalTok{(seq) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{score =} \KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{letterFrequency}\NormalTok{(seq, }\StringTok{"GC"}\NormalTok{, }\DataTypeTok{as.prob =} \OtherTok{TRUE}\NormalTok{)))}
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
+    \DataTypeTok{score =} \KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{letterFrequency}\NormalTok{(seq, }\StringTok{"GC"}\NormalTok{, }\DataTypeTok{as.prob =} \OtherTok{TRUE}\NormalTok{))}
+\NormalTok{  ) }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{select}\NormalTok{(score)}
+\NormalTok{alignments}
 \end{Highlighting}
@@
 
-Finally, we use \texttt{group\_by\_overlaps()} to see where the
-alignments overlap the genomic windows and then apply
-\texttt{summarize()} to compute the total number of reads and average GC
-content within each window.
+\begin{verbatim}
+#> GRanges object with 8275595 ranges and 1 metadata column:
+#>             seqnames            ranges strand |             score
+#>                <Rle>         <IRanges>  <Rle> |         <numeric>
+#>         [1]    chr10       50044-50119      - | 0.276315789473684
+#>         [2]    chr10       50050-50119      + |              0.25
+#>         [3]    chr10       50141-50213      - | 0.447368421052632
+#>         [4]    chr10       50203-50278      + | 0.263157894736842
+#>         [5]    chr10       50616-50690      + | 0.276315789473684
+#>         ...      ...               ...    ... .               ...
+#>   [8275591]     chrY 57772745-57772805      - | 0.513157894736842
+#>   [8275592]     chrY 57772751-57772800      + | 0.526315789473684
+#>   [8275593]     chrY 57772767-57772820      + | 0.565789473684211
+#>   [8275594]     chrY 57772812-57772845      + |              0.25
+#>   [8275595]     chrY 57772858-57772912      + | 0.592105263157895
+#>   -------
+#>   seqinfo: 24 sequences from an unspecified genome
+\end{verbatim}
+
+Finally, we create 10000nt tiles over the genome and compute the number
+of reads and average GC content over all reads that fall within each
+tile using an overlap join and merging endpoints.
 
@@
 \begin{Highlighting}[]
+\NormalTok{bins <-}\StringTok{ }\NormalTok{locations }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{tile_ranges}\NormalTok{(}\DataTypeTok{width =}\NormalTok{ 10000L)}
+
 \NormalTok{alignments_summary <-}\StringTok{ }\NormalTok{bins }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{group_by_overlaps}\NormalTok{(alignments }\OperatorTok{%>%}\StringTok{ }\KeywordTok{select}\NormalTok{(score)) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{n_reads =} \KeywordTok{n}\NormalTok{(), }\DataTypeTok{avg_gc =} \KeywordTok{mean}\NormalTok{(score))}
+\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(alignments) }\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{disjoin_ranges}\NormalTok{(}\DataTypeTok{n =} \KeywordTok{n}\NormalTok{(), }\DataTypeTok{avg_gc =} \KeywordTok{mean}\NormalTok{(score))}
+\NormalTok{alignments_summary}
 \end{Highlighting}
 \end{Shaded}
+
+\begin{verbatim}
+#> GRanges object with 286030 ranges and 2 metadata columns:
+#>            seqnames            ranges strand |         n            avg_gc
+#>               <Rle>         <IRanges>  <Rle> | <integer>         <numeric>
+#>        [1]    chr10       49999-59997      * |        88 0.369019138755981
+#>        [2]    chr10       59998-69997      * |        65 0.434210526315789
+#>        [3]    chr10       69998-79996      * |        56 0.386513157894737
+#>        [4]    chr10       79997-89996      * |        71  0.51297257227576
+#>        [5]    chr10       89997-99996      * |        64 0.387746710526316
+#>        ...      ...               ...    ... .       ...               ...
+#>   [286026]     chrY 57722961-57732958      * |        36 0.468201754385965
+#>   [286027]     chrY 57732959-57742957      * |        38 0.469529085872576
+#>   [286028]     chrY 57742958-57752956      * |        38 0.542936288088643
+#>   [286029]     chrY 57752957-57762955      * |        42 0.510651629072682
+#>   [286030]     chrY 57762956-57772954      * |       504 0.526942355889723
+#>   -------
+#>   seqinfo: 24 sequences from an unspecified genome; no seqlengths
+\end{verbatim}
 
@@
             \DataTypeTok{ti_tv =} \KeywordTok{sum}\NormalTok{(transition) }\OperatorTok{/}\StringTok{ }\KeywordTok{sum}\NormalTok{(}\OperatorTok{!}\NormalTok{transition)) }
+\NormalTok{ti_tv_results}
 \end{Highlighting}
 \end{Shaded}
+
+\begin{verbatim}
+#> DataFrame with 24 rows and 3 columns
+#>     seqnames    n_snps            ti_tv
+#>        <Rle> <integer>        <numeric>
+#> 1          Y      2226  1.4381161007667
+#> 2          6    154246 3.32013219807305
+#> 3         13     83736 3.40669403220714
+#> 4         10    120035 3.49400973418195
+#> 5          4    153243 3.29528828096533
+#> ...      ...       ...              ...
+#> 20        16     77538 3.19827819589583
+#> 21        12    113208 3.47851887016378
+#> 22        20     57073  3.7121036988111
+#> 23        21     32349 3.50480434479877
+#> 24         X     55495 3.58219800181653
+\end{verbatim}
 
@@
 
-Like \texttt{dplyr}, \texttt{plyranges} verbs are functional: they are
-free of side effects and return their result. This enables chaining of
-verbs through syntax like the forward pipe operator
-(\texttt{\%\textgreater{}\%}, read as ``then'') of the \texttt{magrittr}
-package \cite{R-magrittr}. This has syntax has a direct cognitive
-mapping to natural language and the intuitive notion of pipelines. The
-low-level object-oriented APIs of Bioconductor tend to manipulate data
-via sub-replacement functions, like \texttt{start(gr)\ \textless{}-\ x}.
-These ultimately produce the side effect of replacing a symbol mapping
-in the current environment and thus are not amenable to so-called fluent
-syntax.
-
-A function is cohesive if it performs a singular task without producing
-any side-effects. Singular tasks can always be broken down further at
-lower levels of abstraction. For example, to resize a range, the user
-needs to specify which position (start, end, midpoint) should be
-invariant over the transformation. The \texttt{resize()} function from
-the \texttt{GenomicRanges} package has a \texttt{fix} argument that sets
-the anchor, so calling \texttt{resize()} coalesces anchoring and width
-modification. The coupling at the function call level is justified since
-the effect of setting the width depends on the anchor. However,
-\texttt{plyranges} increases cohesion and decouples the anchoring into
-its own function call.
+To further enable cognitive consistency, \texttt{plyranges} functions
+are cohesive. A function is defined to be cohesive if it performs a
+singular task without producing any side-effects. Singular tasks can
+always be broken down further at lower levels of abstraction. For
+example, to resize a range, the user needs to specify which position
+(start, end, midpoint) should be invariant over the transformation. The
+\texttt{resize()} function from the \texttt{GenomicRanges} package has a
+\texttt{fix} argument that sets the anchor, so calling \texttt{resize()}
+coalesces anchoring and width modification. The coupling at the function
+call level is justified since the effect of setting the width depends on
+the anchor. However, \texttt{plyranges} increases cohesion and decouples
+the anchoring into its own function call.
 
@@
 function's arguments. A flat function catalog also enhances API
-discoverability, particularly through auto-completion in IDEs. One
-downside of pushing cohesion to this extreme is that function calls
-become coupled, and care is necessary to treat them as a group when
-modifying code.
+discoverability, particularly through auto-completion in integrated
+developer environments (IDEs). One downside of pushing cohesion to this
+extreme is that function calls become coupled, and care is necessary to
+treat them as a group when modifying code.
+
+Like \texttt{dplyr}, \texttt{plyranges} verbs are functional: they are
+free of side effects and are generally endomorphic, meaning that when
+the input is a GRanges object they return a GRanges object. This enables
+chaining of verbs through syntax like the forward pipe operator from the
+\texttt{magrittr} package. This syntax has a direct cognitive mapping to
+natural language and the intuitive notion of pipelines. The low-level
+object-oriented APIs of Bioconductor tend to manipulate data via
+sub-replacement functions, like \texttt{start(gr)\ \textless{}-\ x}.
+These ultimately produce the side effect of replacing a symbol mapping
+in the current environment and thus are not amenable to so-called fluent
+syntax.
 
@@
 its function naming, instead relying on a parameter that defaults to
-strand sensitivity, an arguably unintuitive behavior.
+strand sensitivity, an arguably confusing behavior.
+
+The implementation of \texttt{plyranges} is built on top of Bioconductor
+infrastructure, meaning most functions are constructed by composing
+generic functions from core Bioconductor packages. As a result, any
+Bioconductor packages that uses data structures that inherit from
+GRanges will be able to use \texttt{plyranges} for free. Another
+consequence of building on top of Bioconductor generics is that the
+speed and memory usage of \texttt{plyranges} functions are similar to
+the highly optimized methods implemented in Bioconductor for GRanges
+objects.
 
@@
 representing experimental results (e.g., counts) from multiple sample
-experiments in conjunction with feature and sample metadata. The grammar
-and design of the \texttt{plryanges} DSL are naturally extensible to
-SummarizedExperiment.
+experiments in conjunction with feature and sample metadata. Although,
+the SummarizedExperiment is not strictly tidy, it does consist of three
+tidy data structures that are related by feature and sample identifiers.
+Therefore, the grammar and design of the \texttt{plryanges} DSL is
+naturally extensible to the SummarizedExperiment.
 
@@
 To produce the workflows as described in results section we used R
-version 3.5 with the development version of \texttt{plyranges} (1.1.4)
+version 3.5 with the development version of \texttt{plyranges} (1.1.10)
 and the \texttt{BioStrings} package (2.48.0) installed.

### 2018-09-21 paper.Rmd
final checks

@@
 briefly described and categorized into one of the following higher level 
-catgeories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
+categories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
 its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
@@
 counts for each SNP the number of distinct exons it overlaps. The 
-\texttt{plyranges}code achieves this with an overlap join followed by partitioning
+\texttt{plyranges} code achieves this with an overlap join followed by partitioning
 and aggregation. Strand is ignored by default here. 

### 2018-09-21 paper.tex
final checks

@@
 
-\noindent  \hypertarget{background}{%
-\section{Background}\label{background}}
+\noindent  \section{Background}\label{background}
 
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{genomic-relational-algebra}{%
+\section{Results}\label{results}
+
 \subsection{Genomic Relational
-Algebra}\label{genomic-relational-algebra}}
-
-\hypertarget{data-model}{%
-\subsubsection{Data Model}\label{data-model}}
+Algebra}\label{genomic-relational-algebra}
+
+\subsubsection{Data Model}\label{data-model}
 
@@
 
-\hypertarget{algebraic-operations}{%
-\subsubsection{Algebraic operations}\label{algebraic-operations}}
+\subsubsection{Algebraic operations}\label{algebraic-operations}
 
@@
 
-\hypertarget{developing-workflows-with-plyranges}{%
 \subsection{\texorpdfstring{Developing workflows with
-\texttt{plyranges}}{Developing workflows with plyranges}}\label{developing-workflows-with-plyranges}}
+\texttt{plyranges}}{Developing workflows with plyranges}}\label{developing-workflows-with-plyranges}
 
@@
 
-\hypertarget{peak-finding}{%
-\subsubsection{Peak Finding}\label{peak-finding}}
+\subsubsection{Peak Finding}\label{peak-finding}
 
@@
 
-\hypertarget{computing-windowed-statistics}{%
 \subsubsection{Computing Windowed
-Statistics}\label{computing-windowed-statistics}}
+Statistics}\label{computing-windowed-statistics}
 
@@
 
-\hypertarget{quality-control-metrics}{%
-\subsubsection{Quality Control Metrics}\label{quality-control-metrics}}
+\subsubsection{Quality Control Metrics}\label{quality-control-metrics}
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section{Discussion}\label{discussion}
 
@@
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+\section{Conclusion}\label{conclusion}
 
@@
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-\hypertarget{data-availability}{%
-\subsection{Data Availability}\label{data-availability}}
+\section{Methods}\label{methods}
+
+\subsection{Data Availability}\label{data-availability}
 
@@
 
-\hypertarget{software-versions}{%
-\subsection{Software Versions}\label{software-versions}}
+\subsection{Software Versions}\label{software-versions}
 
@@
 
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\section{Acknowledgements}\label{acknowledgements}
 

### 2018-10-23 paper.Rmd
abstract and title change based on editor's comments

@@
 ---
-title: "plyranges: A grammar of genomic data transformation" 
+title: "plyranges: a grammar of genomic data transformation" 
 author:
@@
     email: lawrence.michael@gene.com
-abstract: "The Bioconductor project provides many interoperable data abstractions for analyzing high-throughput genomics experiments; however implementing a typical genomic workflow with Bioconductor requires learning these abstractions and understanding them at an integrative level.
-This places a large cognitive burden on the user, especially for non-programmers.
-To reduce this burden we have created a grammar of genomic data transformation that operates on a single, central Bioconductor data structure, GRanges, which naturally represents genomic intervals and their associated measurements.
-The grammar defines verbs for performing actions on and between genomic interval data through a simplified, coherent interface to existing Bioconductor infrastructure, resulting in fluent analysis workflows.
-We have implemented this grammar as an R/Bioconductor package called \\texttt{plyranges}."
+abstract: "There is a cognitive load placed on users in learning a data abstraction from the Bioconductor project and understanding its appropriate use.
+Users must navigate these abstractions to perform a genomic analysis task, when a single data abstraction, a GRanges object will suffice.
+By recognizing that the GRanges class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure.
+We implement this grammar as a Bioconductor/R package called \texttt{plyranges}."
 keywords: "Bioconductor, Grammar, Genomes, Data Analysis" 

### 2018-10-23 paper.tex
abstract and title change based on editor's comments

--- 2018-09-20 paper.tex
@@
 \renewcommand*{\thefootnote}{\fnsymbol{footnote}}
+
+% highlight changes
+\usepackage{soul}
 
@@
 package \cite{Wickham2017-dplyr} defines an application programming
-interface (API) that maps notions from the general relational algebra to
+interface (API) that maps notions from query languages such as SQL to
 verbs that act on tidy data. These verbs can be composed together on one

### 2018-10-24 paper.Rmd
declarations and software versioning

@@
     email: di.cook@monash.edu 
-  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author} 
+  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author \href{mailto:michael.lawerence@gene.com}{michael.lawerence@gene.com} }
     email: lawrence.michael@gene.com
@@
                fig.height = 3, 
-               fig.align = "center")
+               fig.align = "center",
+               fig.path = "diagrams")
 # for peak figures
@@
 ## Data Availability
+
 The BigWig file for the H3K27Me3 primary T CD8+ memory cells from peripheral 
-blood ChIP-seq data was downloaded from the `AnnotationHub` package (2.12.0)
-under accession AH33458. The BAM file corresponding to the
-H1 cell line ChIP-seq data is available at GEO under accession GSM433167. 
-The SNP array data for the H1 cell line data is available at GEO under accession
-GSM1463263.
+blood ChIP-seq data from the Human Roadmap Epigenomics project  
+was downloaded from the `AnnotationHub` package (2.13.1)
+under accession AH33458 \cite{R-ahub, Roadmap-Epigenomics-Consortium2015-pr}. 
+The BAM file corresponding to the H1 cell line ChIP-seq data is available at 
+NCBI GEO under accession [GSM433167](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM433167) \cite{ncbi-geo, Roadmap-Epigenomics-Consortium2015-pr}. 
+The SNP array data for the H1 cell line data is available at  NCBI GEO under accession
+[GPL18952](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952) 
+\cite{Roadmap-Epigenomics-Consortium2015-pr}.
 
@@
 R version 3.5 with the development version of `plyranges` (1.1.10)   
-and the  `BioStrings` package (2.48.0) installed.
+and the `BioStrings` package (2.49.0) installed \cite{r-core, plyranges-zenodo, R-biostrings}.
 
-All code required to reproduce this article is available at https://github.com/sa-lee/plyranges-paper.
+All code required to reproduce this article is available at 
+https://github.com/sa-lee/plyranges-paper.
 
-# Acknowledgements
+# Declarations
+
+## Acknowledgements
 
@@
 this work. We would also like to thank Lori Shepherd and Hèrve Pages
-for the code review they performed. This article was written with
-`knitr` \cite{R-knitr} and the figures were made with `ggbio` \cite{R-ggbio}.
+for the code review they performed and users who have submitted feedback
+and pull requests. This article was written with `knitr` \cite{R-knitr} and 
+the figures were made with `ggbio` \cite{R-ggbio}.
+
+## Funding
+
+SL is funded by an Australian Government Research Training Program (RTP) 
+scholarship and receives an additional top up scholarship from CSL. Funding 
+bodies had no role in data collection, analysis, and interpretation.
+
+## Authors Contributions
+
+SL, DC, ML contributed ideas towards the software. SL wrote 
+the manuscript and analysed data. SL and ML wrote the software. 
+DC and ML supervised the research and revised the manuscript. 
+All authors have read and approved the final version of the manuscript.
+
+## Ethics approval and consent to participate
+
+Not applicable to the current manuscript.
+
+## Competing Interests
+
+Not applicable to the current manuscript.
 

### 2018-10-24 paper.Rmd
final readable version

@@
 By recognizing that the GRanges class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure.
-We implement this grammar as a Bioconductor/R package called \texttt{plyranges}."
+We implement this grammar as a Bioconductor/R package called \\texttt{plyranges}."
 keywords: "Bioconductor, Grammar, Genomes, Data Analysis" 
@@
                fig.align = "center",
-               fig.path = "diagrams")
+               fig.path = "./diagrams/")
 # for peak figures
@@
     & \emph{setdiff\_ranges} & range-wise set difference \\
-    & \emph{complement\_ranges} & range-wise union \\
+    & \emph{complement\_ranges} & range-wise set complement \\
   \hline
@@
 language and API design: cognitive consistency, cohesion,
-expressiveness and endomorphism \cite{Green1996-qg}. 
+endomorphism and expressiveness \cite{Green1996-qg}. 
 To varying degrees, these principles also underlie the
@@
 
-The `plyranges` package can be obtained via the Bioconductor project
-website [https://bioconductor.org](https://bioconductor.org) or
-accessed via Github 
-[https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
+# Declarations
 
-# Methods
-## Data Availability
+## Acknowledgements
+
+We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
+Institute and Dr Paul Harrison for their feedback on earlier drafts of
+this work. We would also like to thank Lori Shepherd and Hèrve Pages
+for the code review they performed and users who have submitted feedback
+and pull requests.
+
+## Availabliity of data and Materials
 
@@
 
-## Software Versions
+The `plyranges` package is open source under an Artistic 2.0 license 
+\cite{R-plyranges-z}. 
+The  version (1.1.10) used in the paper is archived under DOI: [10.5281/zenodo.1469841](https://doi.org/10.5281/zenodo.1469841). The software 
+can be obtained via the Bioconductor project website 
+[https://bioconductor.org](https://bioconductor.org) or accessed via Github 
+[https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
+
 To produce the workflows as described in results section we used 
 R version 3.5 with the development version of `plyranges` (1.1.10)   
-and the `BioStrings` package (2.49.0) installed \cite{r-core, plyranges-zenodo, R-biostrings}.
+and the `BioStrings` package (2.49.0) installed \cite{r-core, R-plyranges-z, R-biostrings}.
 
-All code required to reproduce this article is available at 
-https://github.com/sa-lee/plyranges-paper.
-
-# Declarations
-
-## Acknowledgements
-
-We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
-Institute and Dr Paul Harrison for their feedback on earlier drafts of
-this work. We would also like to thank Lori Shepherd and Hèrve Pages
-for the code review they performed and users who have submitted feedback
-and pull requests. This article was written with `knitr` \cite{R-knitr} and 
-the figures were made with `ggbio` \cite{R-ggbio}.
+This article was written with `knitr` \cite{R-knitr} and 
+the figures were made with `ggbio` \cite{R-ggbio}. All code required to 
+reproduce this article is available at https://github.com/sa-lee/plyranges-paper.
 

### 2018-10-24 paper.tex
final readable version

@@
 \renewcommand*{\thefootnote}{\fnsymbol{footnote}}
-
-% highlight changes
-\usepackage{soul}
 
@@
 
-\title{plyranges: A grammar of genomic data transformation  }
+\title{plyranges: a grammar of genomic data transformation  }
 
@@
 \footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}\vspace{0.05in} \newline\normalsize\emph{}   \and \Large Dianne Cook \(^{\ref{monash}}\)\vspace{0.05in} \newline\normalsize\emph{}   \and \Large Michael Lawrence
-\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author}\vspace{0.05in} \newline\normalsize\emph{}  }
+\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author \href{mailto:michael.lawerence@gene.com}{michael.lawerence@gene.com} }\vspace{0.05in} \newline\normalsize\emph{}  }
 
@@
 \footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia} () and Dianne Cook \(^{\ref{monash}}\) () and Michael Lawrence
-\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author} ()},
+\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author \href{mailto:michael.lawerence@gene.com}{michael.lawerence@gene.com} } ()},
              pdfkeywords = {Bioconductor, Grammar, Genomes, Data Analysis},  
-            pdftitle={plyranges: A grammar of genomic data transformation},
+            pdftitle={plyranges: a grammar of genomic data transformation},
             colorlinks=true,
@@
 \footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}} \hskip 15pt \emph{\small }   \par \textbf{\authorfont Dianne Cook \(^{\ref{monash}}\)} \hskip 15pt \emph{\small }   \par \textbf{\authorfont Michael Lawrence
-\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author}} \hskip 15pt \emph{\small }
+\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author \href{mailto:michael.lawerence@gene.com}{michael.lawerence@gene.com} }} \hskip 15pt \emph{\small }
 
@@
 
-\noindent The Bioconductor project provides many interoperable data abstractions
-for analyzing high-throughput genomics experiments; however implementing
-a typical genomic workflow with Bioconductor requires learning these
-abstractions and understanding them at an integrative level. This places
-a large cognitive burden on the user, especially for non-programmers. To
-reduce this burden we have created a grammar of genomic data
-transformation that operates on a single, central Bioconductor data
-structure, GRanges, which naturally represents genomic intervals and
-their associated measurements. The grammar defines verbs for performing
-actions on and between genomic interval data through a simplified,
-coherent interface to existing Bioconductor infrastructure, resulting in
-fluent analysis workflows. We have implemented this grammar as an
-R/Bioconductor package called \texttt{plyranges}.
+\noindent There is a cognitive load placed on users in learning a data abstraction
+from the Bioconductor project and understanding its appropriate use.
+Users must navigate these abstractions to perform a genomic analysis
+task, when a single data abstraction, a GRanges object will suffice. By
+recognizing that the GRanges class follows `tidy' data principles, we
+create a grammar of genomic data transformation, defining verbs for
+performing actions on and between genomic interval data and providing a
+way of performing common data analysis tasks through a coherent
+interface to existing Bioconductor infrastructure. We implement this
+grammar as a Bioconductor/R package called \texttt{plyranges}.
 
@@
 package \cite{Wickham2017-dplyr} defines an application programming
-interface (API) that maps notions from query languages such as SQL to
+interface (API) that maps notions from the general relational algebra to
 verbs that act on tidy data. These verbs can be composed together on one
@@
 
-{\centering \includegraphics{paper_files/figure-latex/peak-viz-1}
+{\centering \includegraphics{./diagrams/peak-viz-1}
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/titv-viz-1}
+{\centering \includegraphics{./diagrams/titv-viz-1}
 
@@
 of language and API design: cognitive consistency, cohesion,
-expressiveness and endomorphism \cite{Green1996-qg}. To varying degrees,
+endomorphism and expressiveness \cite{Green1996-qg}. To varying degrees,
 these principles also underlie the design of \texttt{dplyr} and the
@@
 
-The \texttt{plyranges} package can be obtained via the Bioconductor
-project website \url{https://bioconductor.org} or accessed via Github
-\url{https://github.com/sa-lee/plyranges}.
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-\hypertarget{data-availability}{%
-\subsection{Data Availability}\label{data-availability}}
-
-The BigWig file for the H3K27Me3 primary T CD8+ memory cells from
-peripheral blood ChIP-seq data was downloaded from the
-\texttt{AnnotationHub} package (2.12.0) under accession AH33458. The BAM
-file corresponding to the H1 cell line ChIP-seq data is available at GEO
-under accession GSM433167. The SNP array data for the H1 cell line data
-is available at GEO under accession GSM1463263.
-
-\hypertarget{software-versions}{%
-\subsection{Software Versions}\label{software-versions}}
-
-To produce the workflows as described in results section we used R
-version 3.5 with the development version of \texttt{plyranges} (1.1.10)
-and the \texttt{BioStrings} package (2.48.0) installed.
-
-All code required to reproduce this article is available at
-\url{https://github.com/sa-lee/plyranges-paper}.
+\hypertarget{declarations}{%
+\section{Declarations}\label{declarations}}
 
 \hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\subsection{Acknowledgements}\label{acknowledgements}}
 
@@
 this work. We would also like to thank Lori Shepherd and Hèrve Pages for
-the code review they performed. This article was written with
-\texttt{knitr} \cite{R-knitr} and the figures were made with
-\texttt{ggbio} \cite{R-ggbio}.
+the code review they performed and users who have submitted feedback and
+pull requests.
+
+\hypertarget{availabliity-of-data-and-materials}{%
+\subsection{Availabliity of data and
+Materials}\label{availabliity-of-data-and-materials}}
+
+The BigWig file for the H3K27Me3 primary T CD8+ memory cells from
+peripheral blood ChIP-seq data from the Human Roadmap Epigenomics
+project\\
+was downloaded from the \texttt{AnnotationHub} package (2.13.1) under
+accession AH33458 \cite{R-ahub, Roadmap-Epigenomics-Consortium2015-pr}.
+The BAM file corresponding to the H1 cell line ChIP-seq data is
+available at NCBI GEO under accession
+\href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM433167}{GSM433167}
+\cite{ncbi-geo, Roadmap-Epigenomics-Consortium2015-pr}. The SNP array
+data for the H1 cell line data is available at NCBI GEO under accession
+\href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952}{GPL18952}
+\cite{Roadmap-Epigenomics-Consortium2015-pr}.
+
+The \texttt{plyranges} package is open source under an Artistic 2.0
+license \cite{R-plyranges-z}. The version (1.1.10) used in the paper is
+archived under DOI:
+\href{https://doi.org/10.5281/zenodo.1469841}{10.5281/zenodo.1469841}.
+The software can be obtained via the Bioconductor project website
+\url{https://bioconductor.org} or accessed via Github
+\url{https://github.com/sa-lee/plyranges}.
+
+To produce the workflows as described in results section we used R
+version 3.5 with the development version of \texttt{plyranges}
+(1.1.10)\\
+and the \texttt{BioStrings} package (2.49.0) installed
+\cite{r-core, R-plyranges-z, R-biostrings}.
+
+This article was written with \texttt{knitr} \cite{R-knitr} and the
+figures were made with \texttt{ggbio} \cite{R-ggbio}. All code required
+to reproduce this article is available at
+\url{https://github.com/sa-lee/plyranges-paper}.
+
+\hypertarget{funding}{%
+\subsection{Funding}\label{funding}}
+
+SL is funded by an Australian Government Research Training Program (RTP)
+scholarship and receives an additional top up scholarship from CSL.
+Funding bodies had no role in data collection, analysis, and
+interpretation.
+
+\hypertarget{authors-contributions}{%
+\subsection{Authors Contributions}\label{authors-contributions}}
+
+SL, DC, ML contributed ideas towards the software. SL wrote the
+manuscript and analysed data. SL and ML wrote the software. DC and ML
+supervised the research and revised the manuscript. All authors have
+read and approved the final version of the manuscript.
+
+\hypertarget{ethics-approval-and-consent-to-participate}{%
+\subsection{Ethics approval and consent to
+participate}\label{ethics-approval-and-consent-to-participate}}
+
+Not applicable to the current manuscript.
+
+\hypertarget{competing-interests}{%
+\subsection{Competing Interests}\label{competing-interests}}
+
+Not applicable to the current manuscript.
 
